### 2020-01-02 50sites.Rmd
added old code

---
title: "50sites"
author: "Yanlin Ma + Kevin Murgas"
date: "February 11, 2018"
output: word_document
---
Model 3: $$ y_{i,j,k} = \mu + b_j + I_{tumor} * ( \beta_T + c_j + d_k) + \epsilon_i $$
$$ b_j \sim N(0, \sigma_P), \ c_j \sim N(0,\sigma_{PT}),\ d_k \sim N(0, \sigma_T), \ \epsilon_i \sim N(0,\sigma_E)  $$
```{r echo=FALSE, include=FALSE}
library(rstan)
library(gtools)
library(ggplot2)
library(bayesplot)
```

```{r echo=FALSE, include=FALSE}
# Kevin's working directory
setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")

# Yanlin's working directory
# setwd("D:/DataPlus2017/Data")

# load data
load("myFA.Rdata")

# Extract single site data
site <- function (site_no) {
  # extract CpG site xx to start
  temp <- FullAnnotation[site_no,]

  # here we use all patient samples, excluding glands
  indices <- c(9:13, 15:39,46,47,57,58,72:75)
  patientLabel <- substr(colnames(temp[indices]),1,1)
  patientLabel[10:12] <- "K*"
  sideLabel <- substr(colnames(temp[indices]),2,2)
  tissueLabel <- sideLabel
  tissueLabel[tissueLabel %in% c("A","B")] <- "T"   #replace A and B with T
  tumorIndicator <- 1*(tissueLabel=="T")

  work <- data.frame(beta=logit(t(temp[indices])), patient = patientLabel, tissue=tissueLabel, side=sideLabel, tInd=tumorIndicator)
  colnames(work)[1] <- "beta"
  return(work)
}

# Using Model PPTT (previously model 3): add intra-tumoral variance
stanfitPPTT <- function (dataset) {
  stanDat <- list(pID = as.integer(factor(dataset$patient)),
                  tInd = dataset$tInd,
                  N = nrow(dataset),
                  P = nlevels(dataset$patient),
                  y = dataset[,1])
  stanFitPPTT <-
    stan(
      #file = "model_PPTT.stan",
      fit = emptyFit,
      warmup = 1000,
      iter = 4000,
      chains = 4,
      thin = 1,
      data = stanDat,
      control = list(adapt_delta = 0.999),
      refresh = 0
    )
  return(stanFitPPTT=stanFitPPTT)
}

# Plot raw data and fits
fitplot <- function(data,stanFit) {
  # plot beta and fits at this site for all patients, using mean estimates for b and c coefficients
  p1 <- ggplot() + geom_point(data = data, aes(x=tInd, y=beta, colour=patient))
  pats <- as.integer(factor(data$patient))
  nsamp <- length(pats)
  npats <- max(pats)
  sumTemp <- summary(stanFit)$summary
  mu <- sumTemp[dim(sumTemp)[1]-5,1]
  betaT <- sumTemp[dim(sumTemp)[1]-6,1]
  sampcoefs <- data.frame(pat = pats, b = sumTemp[pats], c = sumTemp[pats+npats], d = sumTemp[1:nsamp+2*npats])
  patcoefs <- data.frame(pat = 1:npats, b = sumTemp[1:npats], c = sumTemp[1:npats+npats])
  rm(sumTemp)
  # add individual sample estimates
  sampdf <- data.frame(pat = rep(factor(data$patient),2), patg = rep(factor(1:nsamp),2), tInd = c(rep(0,nsamp),rep(1,nsamp)), est = c(sampcoefs$b+mu, (sampcoefs$b+sampcoefs$c++sampcoefs$d+mu+betaT)))
  p1 <- p1 + geom_line(data=sampdf, aes(x=tInd, y=est, colour=pat, group=patg), linetype="dashed", alpha=0.5)
  # add patient mean estimates
  patdf <- data.frame(pat = rep(levels(data$patient),2), tInd = c(rep(0,npats),rep(1,npats)), est = c(patcoefs$b+mu, patcoefs$b+patcoefs$c+mu+betaT))
  p1 <- p1 + geom_line(data=patdf, aes(x=tInd, y=est, colour=pat))
  # add overall mean estimates
  meandf <- data.frame(tInd = c(0,1), est = c(mu,mu+betaT))
  p1 <- p1 + geom_point(data=meandf, aes(x=tInd, y=est), size=3, alpha=0.7) + geom_line(data=meandf, aes(x=tInd, y=est), size=2, alpha=0.5)

  return(p1)
}

# Plot posterior distribution
postplot <- function(stanFit) {
  p2 <- mcmc_areas(
    as.array(stanFit), 
    pars = c("sigma_p", "sigma_pt","sigma_t","sigma_e"),
    prob = 0.8, # 80% intervals
    prob_outer = 0.95, 
    point_est = "median"
  )
  return(p2)
}

## CODE
# Randomly choose 50 sites
set.seed(50)
siteInds <- sample(1:866836,50)
nsites <- length(siteInds)

# check for model_PPTT.rds, must add to prevent crash
checkFile <- "model_PPTT.rds"
if (file.exists(checkFile)) {
  file.remove(checkFile)
  print("caught one!") 
}

# create empty fit to use same model compilation throughout
data1<-site(1)
stanDat1 <- list(
  pID = as.integer(factor(data1$patient)),
  tInd = data1$tInd,
  N = nrow(data1),
  P = nlevels(data1$patient),
  y = data1[, 1]
)
emptyFit <- stan(file="model_PPTT.stan", data = stanDat1, chains = 0)

```

```{r echo=FALSE,include=TRUE,error=FALSE,message=FALSE,warning=FALSE,results='hide',fig.keep='all',fig.show='hold',fig.height=3.7,fig.width=5}
# fig.align='center'
ptm <- proc.time()
for (i in siteInds) {
  #print(paste("Site ",i,sep=''))
  data <- site(i)
  stanFit <- stanfitPPTT(data)
  p1 <- fitplot(data, stanFit) + ggtitle(paste("Site", i, ": Beta Fits by Patient"))
  p2 <- postplot(stanFit) + ggtitle(paste("Site", i, ": Sigma Posteriors"))
  print(p1)
  print(p2)
}
mtp <- proc.time() - ptm

```

```{r echo=FALSE,include=TRUE}

print(mtp)
print(paste("Completed run on sites:", paste(siteInds, collapse=" ")))

```

### 2020-01-02 QC_and_Annotation.Rmd
added old code

---
title: "Quality Control"
output: word_document
---

## Density Plots of Beta Values

```{r, echo=FALSE, warning=FALSE,message=FALSE,fig.width=13,fig.height=5}
library(minfi)
library(IlluminaHumanMethylationEPICmanifest)
library(IlluminaHumanMethylationEPICanno.ilm10b2.hg19)
library(gdata) #gdata package needs Perl language installed
library(RColorBrewer)

# here set the working directory that points to the data folder
# e.g. the folder with all datasets in it, should contain all the
# 1337-1387 folders
# please just comment out the directories already here, so each
# user can uncomment the setwd for them as we move code around

# Kevin's working directory
# setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")

# Yanlin's working directory
setwd("D:/DataPlus2017/Data")

####################### Function used for read in data #########################

# base_dir = directory containing basenames file
# targets = name of the dataset after reading in basenames file
# pat_dir = directory containing patient information file
# pat_file = WHOLE name (including directory) of the dataset after reading in patient information file
# work_name = name of the RGChannelSet

read.fun <- function(base_dir,pat_file) {
  # read in patient data from the .xlsx file
  pat_name <- read.xls(pat_file,header=TRUE)

  # Extract targets
  targets_name <- data.frame(pat_name[,"Complete.Barcode"])
  colnames(targets_name) <- "Basement"

  # Read in all .idat files
  targets_name$Basement <- file.path(base_dir, targets_name$Basement)
  work_name <- read.metharray(targets_name$Basement, verbose = TRUE)

  # Push the RGChannelSet to the global environment
  return(work_name)
}

############################### Read in data  #################################

## 1337 ##

base_dir_1337 <- "1337_Shibata EPIC DNA methylation data package/IDAT FILES"
pat_file_1337 <- "1337_Shibata EPIC DNA methylation data package/SAMPLE-ARRAY MAPPING/1337 (Shibata-8).xls"

work_1337 <- read.fun(base_dir_1337,pat_file_1337)

## 1345 ##

base_dir_1345 <- "1345_Shibata EPIC DNA methylation data package/IDAT FILES"
pat_file_1345 <- "1345_Shibata EPIC DNA methylation data package/SAMPLE-ARRAY MAPPING/1345 (Shibata-16).xlsx"

work_1345 <- read.fun(base_dir_1345,pat_file_1345)

# Two beadchips in plate 1345, read in seperately

pat_1345 <- read.xls(pat_file_1345)
targets_1345 <- data.frame(pat_1345[,"Complete.Barcode"])
colnames(targets_1345) <- 'Basement'

targets_1345$Basement <- file.path(base_dir_1345,targets_1345$Basement)
# Beadchip 1
work_1345a <- read.metharray(targets_1345$Basement[1:8],verbose=TRUE)
# Beadchip 2
work_1345b <- read.metharray(targets_1345$Basement[9:16],verbose = TRUE)

## 1350 ##

base_dir_1350 <- "1350_SHIBATA EPIC DNA METHYLATION DATA PACKAGE/IDAT files"
pat_file_1350 <- "1350_SHIBATA EPIC DNA METHYLATION DATA PACKAGE/SAMPLE-ARRAY MAPPING/1350 (Shibata-8).xlsx"

work_1350 <- read.fun(base_dir_1350,pat_file_1350)

## 1357 ##

base_dir_1357 <- "1357_Shibata EPIC DNA methylation data package/IDAT FILES"
pat_file_1357 <- "1357_Shibata EPIC DNA Methylation Data Package/SAMPLE-ARRAY MAPPING/1357 (Shibata-8).xlsx"

work_1357 <- read.fun(base_dir_1357,pat_file_1357)

## 1360 ##

base_dir_1360 <- "1360_Shibata EPIC Data Package/IDAT FILES"
pat_file_1360 <- "1360_Shibata EPIC Data Package/SAMPLE-ARRAY MAPPING/1360 (Shibata-8).xlsx"

work_1360 <- read.fun(base_dir_1360,pat_file_1360)

## 1378 ##

base_dir_1378 <- "1378_Shibata EPIC Data Package/IDAT FILES"
pat_file_1378 <- "1378_Shibata EPIC Data Package/SAMPLE-ARRAY MAPPING/1378 (Shbata-8).xls"

work_1378 <- read.fun(base_dir_1378,pat_file_1378)

## 1385 ##

base_dir_1385 <- "1385_Shibata EPIC Data Package/IDAT FILES"
pat_file_1385 <- "1385_Shibata EPIC Data Package/SAMPLE-ARRAY MAPPING/1385 (Shibata-8).xlsx"

work_1385 <- read.fun(base_dir_1385,pat_file_1385)

## 1387 ##

base_dir_1387 <- "1387_Shibata EPIC DNA methylation data package/IDAT FILES"
pat_file_1387 <- "1387_Shibata EPIC DNA methylation data package/SAMPLE-ARRAY MAPPING/1387 (Shibata-8).xls"

work_1387 <- read.fun(base_dir_1387,pat_file_1387)

############################# Bad Sample in 1337 ###############################

# Convert to MethylSet
mset_1337 <- preprocessRaw(work_1337)

# Check the methylation and unmethylation signals
# head(getMeth(mset_1337))
# head(getUnmeth((mset_1337)))

# Convert to RatioSet containing beta values and Mvalues(log(M/U))
rset_1337 <- ratioConvert(mset_1337,what = "both")

# Get beta
beta_1337 <- getBeta(rset_1337)

# Check beta values
# head(beta_1337)

#200360140022_R07C01 (JB) has bad quality
# dim(beta_1337)[1] == sum(is.na(beta_1337[,7]))

### read in new data without JB ###
base_dir_1337_new <- "1337_Shibata EPIC DNA methylation data package/IDAT FILES"

targets_1337_new <- read.csv("1337_Shibata EPIC DNA methylation data package/SAMPLE-ARRAY MAPPING/1337_Shibata-targets.csv", as.is = TRUE)

targets_1337_new$Basename <- file.path(base_dir_1337_new, targets_1337_new$Basename)
work_1337_new <- read.metharray(targets_1337_new$Basename, verbose = TRUE)

######################## Preprocessing/Normalization ##########################

### 1337 ###

## noob ##
noob_1337 <- preprocessNoob(work_1337_new)

rset_1337_noob <- ratioConvert(noob_1337,what="both")
beta_1337_noob <- getBeta(rset_1337_noob)

# check
# head(beta_1337_noob)
# Compare with raw beta values
# head(beta_1337)

## Genome Studio ##
lumi_1337 <- preprocessIllumina(work_1337_new,bg.correct = TRUE,
                                normalize = "controls",reference = 2)
rset_1337_lumi <- ratioConvert(lumi_1337,what="both")
beta_1337_lumi <- getBeta(rset_1337_lumi)

# check
# head(beta_1337_lumi)

# Compare with noob and raw
# head(beta_1337)
# head(beta_1337_noob)

### 1345 ###

## noob ##
noob_1345 <- preprocessNoob(work_1345)
noob_1345a <- preprocessNoob(work_1345a)
noob_1345b <- preprocessNoob(work_1345b)

## Genome Studio ##
lumi_1345 <- preprocessIllumina(work_1345,bg.correct = TRUE,
                                normalize = "controls",reference = 2)
lumi_1345a <- preprocessIllumina(work_1345a,bg.correct = TRUE,
                                normalize = "controls",reference = 2)
lumi_1345b <- preprocessIllumina(work_1345b,bg.correct = TRUE,
                                normalize = "controls",reference = 2)

### 1350 ###

## noob ##
noob_1350 <- preprocessNoob(work_1350)

## Genome Studio ##
lumi_1350 <- preprocessIllumina(work_1350,bg.correct = TRUE,
                                normalize = "controls",reference = 2)

### 1357 ###

## noob ##
noob_1357 <- preprocessNoob(work_1357)

## Genome Studio ##
lumi_1357 <- preprocessIllumina(work_1357,bg.correct = TRUE,
                                normalize = "controls",reference = 2)

### 1360 ###

## noob ##
noob_1360 <- preprocessNoob(work_1360)

## Genome Studio ##
lumi_1360 <- preprocessIllumina(work_1360,bg.correct = TRUE,
                                normalize = "controls",reference = 2)

### 1378 ###

## noob ##
noob_1378 <- preprocessNoob(work_1378)

## Genome Studio ##
lumi_1378 <- preprocessIllumina(work_1378,bg.correct = TRUE,
                                normalize = "controls",reference = 2)

### 1385 ###

## noob ##
noob_1385 <- preprocessNoob(work_1385)

## Genome Studio ##
lumi_1385 <- preprocessIllumina(work_1385,bg.correct = TRUE,
                                normalize = "controls",reference = 2)

### 1387 ###

## noob ##
noob_1387 <- preprocessNoob(work_1387)

## Genome Studio ##
lumi_1387 <- preprocessIllumina(work_1387,bg.correct = TRUE,
                                normalize = "controls",reference = 2)

############################### Quality Control ################################

####### Density Plots #########

## 1337 ##

pat_1337 <- read.xls("1337_Shibata EPIC DNA methylation data package/SAMPLE-ARRAY MAPPING/1337 (Shibata-8).xls")

par(mfrow=c(1,2),mar=c(4,2,4,1))
densityPlot(noob_1337,sampGroups = pat_1337$Sample_No,
            main = "Corrected Beta Values Using 'noob'",legend = FALSE)
legend("topright",legend = levels(pat_1337$Sample_No),
       text.col = brewer.pal(8,"Dark2"),cex=0.5)
densityPlot(lumi_1337,sampGroups = pat_1337$Sample_No,
            main = "Corrected Beta Values Using 'Illumina'",legend = FALSE)
legend("topright",legend = levels(pat_1337$Sample_No),
       text.col = brewer.pal(8,"Dark2"),cex=0.5)

## 1345 ##
pat_1345 <- read.xls("1345_Shibata EPIC DNA methylation data package/SAMPLE-ARRAY MAPPING/1345 (Shibata-16).xlsx")

par(mfrow=c(1,2),mar=c(4,2,4,1))
densityPlot(noob_1345,sampGroups = pat_1345$SAMPLE.ID,
            main = "Corrected Beta Values Using 'noob'",legend = FALSE)
legend("top",legend = levels(pat_1345$SAMPLE.ID),
       text.col = brewer.pal(8,"Dark2"),cex=0.5)
densityPlot(lumi_1345,sampGroups = pat_1345$SAMPLE.ID,
            main = "Corrected Beta Values Using 'Illumina'",legend = FALSE)
legend("top",legend = levels(pat_1345$SAMPLE.ID),
       text.col = brewer.pal(8,"Dark2"),cex=0.5)

## 1350 ##
pat_1350 <- read.xls("1350_SHIBATA EPIC DNA METHYLATION DATA PACKAGE/SAMPLE-ARRAY MAPPING/1350 (Shibata-8).xlsx")

par(mfrow=c(1,2),mar=c(4,2,4,1))
densityPlot(noob_1350,sampGroups = pat_1350$SAMPLE.ID,
            main = "Corrected Beta Values Using 'noob'",legend = FALSE)
legend("topright",legend = levels(pat_1350$SAMPLE.ID),
       text.col = brewer.pal(8,"Dark2"),cex=0.7)
densityPlot(lumi_1350,sampGroups = pat_1350$SAMPLE.ID,
            main = "Corrected Beta Values Using 'Illumina'",legend = FALSE)
legend("topright",legend = levels(pat_1350$SAMPLE.ID),
       text.col = brewer.pal(8,"Dark2"),cex=0.7)

## 1357 ## 
pat_1357 <- read.xls("1357_Shibata EPIC DNA Methylation Data Package/SAMPLE-ARRAY MAPPING/1357 (Shibata-8).xlsx")

par(mfrow=c(1,2),mar=c(4,2,4,1))
densityPlot(noob_1357,sampGroups = pat_1357$Sample.ID,
            main = "Corrected Beta Values Using 'noob'",legend = FALSE)
legend("top",legend = levels(pat_1357$Sample.ID),
       text.col = brewer.pal(8,"Dark2"),cex=0.7)
densityPlot(lumi_1357,sampGroups = pat_1357$Sample.ID,
            main = "Corrected Beta Values Using 'Illumina'",legend = FALSE)
legend("top",legend = levels(pat_1357$Sample.ID),
       text.col = brewer.pal(8,"Dark2"),cex=0.7)

## 1360 ##
pat_1360 <- read.xls("1360_Shibata EPIC Data Package/SAMPLE-ARRAY MAPPING/1360 (Shibata-8).xlsx")

par(mfrow=c(1,2),mar=c(4,2,4,1))
densityPlot(noob_1360,sampGroups = pat_1360$Sample.ID,
            main = "Corrected Beta Values Using 'noob'",legend = FALSE)
legend("top",legend = levels(pat_1360$Sample.ID),
       text.col = brewer.pal(8,"Dark2"),cex=0.7)
densityPlot(lumi_1360,sampGroups = pat_1360$Sample.ID,
            main = "Corrected Beta Values Using 'Illumina'",legend = FALSE)
legend("top",legend = levels(pat_1360$Sample.ID),
       text.col = brewer.pal(8,"Dark2"),cex=0.7)

## 1378 ##

pat_1378 <- read.xls("1378_Shibata EPIC Data Package/SAMPLE-ARRAY MAPPING/1378 (Shbata-8).xls")

par(mfrow=c(1,2),mar=c(4,2,4,1))
densityPlot(noob_1378,sampGroups = pat_1378$Tube.Label,
            main = "Corrected Beta Values Using 'noob'",legend = FALSE)
legend("top",legend = levels(pat_1378$Tube.Label),
       text.col = brewer.pal(8,"Dark2"),cex=0.7)
densityPlot(lumi_1378,sampGroups = pat_1378$Tube.Label,
            main = "Corrected Beta Values Using 'Illumina'",legend = FALSE)
legend("top",legend = levels(pat_1378$Tube.Label),
       text.col = brewer.pal(8,"Dark2"),cex=0.7)

## 1385 ##

pat_1385 <- read.xls("1385_Shibata EPIC Data Package/SAMPLE-ARRAY MAPPING/1385 (Shibata-8).xlsx")

par(mfrow=c(1,2),mar=c(4,2,4,1))
densityPlot(noob_1385,sampGroups = pat_1385$Tube.Label,
            main = "Corrected Beta Values Using 'noob'",legend = FALSE)
legend("top",legend = levels(pat_1385$Tube.Label),
       text.col = brewer.pal(8,"Dark2"),cex=0.7)
densityPlot(lumi_1385,sampGroups = pat_1385$Tube.Label,
            main = "Corrected Beta Values Using 'Illumina'",legend = FALSE)
legend("top",legend = levels(pat_1385$Tube.Label),
       text.col = brewer.pal(8,"Dark2"),cex=0.7)

## 1387 ##

pat_1387 <- read.xls("1387_Shibata EPIC DNA methylation data package/SAMPLE-ARRAY MAPPING/1387 (Shibata-8).xls")

par(mfrow=c(1,2),mar=c(4,2,4,1))
densityPlot(noob_1387,sampGroups = pat_1387$Sample.ID,
            main = "Corrected Beta Values Using 'noob'",legend = FALSE)
legend("top",legend = levels(pat_1387$Sample.ID),
       text.col = brewer.pal(8,"Dark2"),cex=0.7)
densityPlot(lumi_1387,sampGroups = pat_1387$Sample.ID,
            main = "Corrected Beta Values Using 'Illumina'",legend = FALSE)
legend("top",legend = levels(pat_1387$Sample.ID),
       text.col = brewer.pal(8,"Dark2"),cex=0.7)

```

``` {r,echo=FALSE, warning=FALSE,message=FALSE,eval=FALSE}

######### Median Plot (noob) ############

qc_1337_noob <- getQC(noob_1337)
qc_1345_noob <- getQC(noob_1345)
qc_1350_noob <- getQC(noob_1350)
qc_1357_noob <- getQC(noob_1357)
qc_1360_noob <- getQC(noob_1360)
qc_1378_noob <- getQC(noob_1378)
qc_1385_noob <- getQC(noob_1385)
qc_1387_noob <- getQC(noob_1387)

qc_noob <- rbind(qc_1337_noob,qc_1345_noob,qc_1350_noob,qc_1357_noob,qc_1360_noob, qc_1378_noob,qc_1385_noob,qc_1387_noob)

par(mfrow=c(1,1),mar=c(5,5,5,5))
plotQC(qc_noob,badSampleCutoff = 9)

######### Median Plot (Illumina) ############

qc_1337_lumi <- getQC(lumi_1337)
qc_1345_lumi <- getQC(lumi_1345)
qc_1350_lumi <- getQC(lumi_1350)
qc_1357_lumi <- getQC(lumi_1357)
qc_1360_lumi <- getQC(lumi_1360)
qc_1378_lumi <- getQC(lumi_1378)
qc_1385_lumi <- getQC(lumi_1385)
qc_1387_lumi <- getQC(lumi_1387)

qc_lumi <- rbind(qc_1337_lumi,qc_1345_lumi,qc_1350_lumi,qc_1357_lumi,qc_1360_lumi,
                 qc_1378_lumi,qc_1385_lumi,qc_1387_lumi)

par(mfrow=c(1,1),mar=c(5,5,5,5))
plotQC(qc_lumi,badSampleCutoff = 9)

```

## Multidimensional Scaling/Principle Component Analysis Plot

``` {r,echo=FALSE, eval=FALSE,warning=FALSE,message=FALSE,fig.width=17,fig.height=10}

############################### With JA ########################################

# Make sure var names are consistent
colnames(pat_1337)[8] <- "Sample.ID"
colnames(pat_1345)[8] <- "Sample.ID"
colnames(pat_1350)[8] <- "Sample.ID"
colnames(pat_1357)[8] <- "Sample.ID"
colnames(pat_1360)[8] <- "Sample.ID"
colnames(pat_1378)[8] <- "Sample.ID"
colnames(pat_1385)[8] <- "Sample.ID"
colnames(pat_1387)[8] <- "Sample.ID"

# Create a data frame that contains all patient information
pat <- rbind(pat_1337[,1:8],pat_1345[,1:8],pat_1350[,1:8],pat_1357[,1:8],
             pat_1360[,1:8],pat_1378[,1:8],pat_1385[,1:8],pat_1387[,1:8])

# Create a var indicating adenomas/carcinomas in pat
for(i in (1:dim(pat)[1])) {

     if ((pat$Sample.ID[i] == "KA") | (pat$Sample.ID[i] == "KB") |
      pat$Sample.ID[i] == "XA" | pat$Sample.ID[i] == "XB" |
      pat$Sample.ID[i] == "SA" | pat$Sample.ID[i] == "SB" |
      pat$Sample.ID[i] == "PA" | pat$Sample.ID[i] == "PB" |
      pat$Sample.ID[i] == "KA LL" | pat$Sample.ID[i] == "KB LL") {

       pat$type[i] <- "Adenomas"

      } else if (pat$Sample.ID[i] == "HA" | pat$Sample.ID[i] == "HB" |
                 pat$Sample.ID[i] == "HN" | pat$Sample.ID[i] == "JA" |
                 pat$Sample.ID[i] == "JN" | pat$Sample.ID[i] == "KA" |
                 pat$Sample.ID[i] == "EA" | pat$Sample.ID[i] == "EB" |
                 pat$Sample.ID[i] == "EN" | pat$Sample.ID[i] == "KA" |
                 pat$Sample.ID[i] == "KA (new)" | 
                 pat$Sample.ID[i] == "KB (new)" | 
                 pat$Sample.ID[i] == "KN (new)" | 
                 pat$Sample.ID[i] == "WA" | pat$Sample.ID[i] == "WB" |
                 pat$Sample.ID[i] == "WN" | pat$Sample.ID[i] == "TA" |
                 pat$Sample.ID[i] == "TB" | pat$Sample.ID[i] == "JB" |
                 pat$Sample.ID[i] == "KA" | pat$Sample.ID[i] == "CA" |
                 pat$Sample.ID[i] == "CB" | pat$Sample.ID[i] == "CN" |
                 pat$Sample.ID[i] == "MA" | pat$Sample.ID[i] == "MB" |
                 pat$Sample.ID[i] == "JA" | pat$Sample.ID[i] == "OA" |
                 pat$Sample.ID[i] == "OB" | pat$Sample.ID[i] == "FA" |
                 pat$Sample.ID[i] == "FB" | pat$Sample.ID[i] == "KA" |
                 pat$Sample.ID[i] == "CA LL" | pat$Sample.ID[i] == "CB LL" |
                 pat$Sample.ID[i] == "FA LL" | pat$Sample.ID[i] == "FB LL" |
                 pat$Sample.ID[i] == "DA" | pat$Sample.ID[i] == "DB" |
                 pat$Sample.ID[i] == "UA" | pat$Sample.ID[i] == "UB" |
                 pat$Sample.ID[i] == "HALL-1" | pat$Sample.ID[i] == "HALL-2"|
                 pat$Sample.ID[i] == "HBLL-3" | pat$Sample.ID[i] == "HBLL-4") {

         pat$type[i] <- "Carcinomas"

      } else {
         pat$type[i] <- "Other"
        }
}

## PCA plot ##
memory.limit(size=60000)
all_noob <- combineArrays(noob_1337,noob_1345)
all_noob <- combineArrays(all_noob,noob_1350)
all_noob <- combineArrays(all_noob,noob_1357)
all_noob <- combineArrays(all_noob,noob_1360)
all_noob <- combineArrays(all_noob,noob_1378)
all_noob <- combineArrays(all_noob,noob_1385)
all_noob <- combineArrays(all_noob,noob_1387)

all_lumi <- combineArrays(lumi_1337,lumi_1345)
all_lumi <- combineArrays(all_lumi,lumi_1350)
all_lumi <- combineArrays(all_lumi,lumi_1357)
all_lumi <- combineArrays(all_lumi,lumi_1360)
all_lumi <- combineArrays(all_lumi,lumi_1378)
all_lumi <- combineArrays(all_lumi,lumi_1385)
all_lumi <- combineArrays(all_lumi,lumi_1387)

par(mfrow=c(1,2),mar=c(5,3,5,3))
mdsPlot(all_noob,numPositions = 1000000,sampGroups = pat$type,
        main = "PCA Plot Using Noob (including JA)", pch=16)

mdsPlot(all_lumi,numPositions = 1000000,sampGroups = pat$type,
        main = "PCA Plot Using Illumina (including JA)", pch = 16)

```

``` {r,echo=FALSE,warning=FALSE,message=FALSE,fig.width=17,fig.height=10}
##################################### Without JA ########################################

### read in new data without JA ###
base_dir_1337_new2 <- "1337_Shibata EPIC DNA methylation data package/IDAT FILES"

targets_1337_new2 <- read.csv("1337_Shibata EPIC DNA methylation data package/SAMPLE-ARRAY MAPPING/1337_Shibata-targets.csv", as.is = TRUE)

targets_1337_new2 <- targets_1337_new2[c(1:5,7),]
targets_1337_new2 <- file.path(base_dir_1337_new, targets_1337_new2)
work_1337_new2 <- read.metharray(targets_1337_new2, verbose = TRUE)

# normalization

## noob ##
noob_1337_new <- preprocessNoob(work_1337_new2)

## Genome Studio ##
lumi_1337_new <- preprocessIllumina(work_1337_new2,bg.correct = TRUE,
                                normalize = "controls",reference = 2)

# Make sure var names are consistent
colnames(pat_1337)[8] <- "Sample.ID"
colnames(pat_1345)[8] <- "Sample.ID"
colnames(pat_1350)[8] <- "Sample.ID"
colnames(pat_1357)[8] <- "Sample.ID"
colnames(pat_1360)[8] <- "Sample.ID"
colnames(pat_1378)[8] <- "Sample.ID"
colnames(pat_1385)[8] <- "Sample.ID"
colnames(pat_1387)[8] <- "Sample.ID"

# Create a data frame that contains all patient information
pat <- rbind(pat_1337[,1:8],pat_1345[,1:8],pat_1350[,1:8],pat_1357[,1:8],
             pat_1360[,1:8],pat_1378[,1:8],pat_1385[,1:8],pat_1387[,1:8])

# Create a var indicating adenomas/carcinomas in pat
for(i in (1:dim(pat)[1])) {

     if ((pat$Sample.ID[i] == "KA") | (pat$Sample.ID[i] == "KB") |
      pat$Sample.ID[i] == "XA" | pat$Sample.ID[i] == "XB" |
      pat$Sample.ID[i] == "SA" | pat$Sample.ID[i] == "SB" |
      pat$Sample.ID[i] == "PA" | pat$Sample.ID[i] == "PB" |
      pat$Sample.ID[i] == "KA LL" | pat$Sample.ID[i] == "KB LL") {

       pat$type[i] <- "Adenomas"

      } else if (pat$Sample.ID[i] == "HA" | pat$Sample.ID[i] == "HB" |
                 pat$Sample.ID[i] == "HN" | pat$Sample.ID[i] == "JA" |
                 pat$Sample.ID[i] == "JN" | pat$Sample.ID[i] == "KA" |
                 pat$Sample.ID[i] == "EA" | pat$Sample.ID[i] == "EB" |
                 pat$Sample.ID[i] == "EN" | pat$Sample.ID[i] == "KA" |
                 pat$Sample.ID[i] == "KA (new)" | 
                 pat$Sample.ID[i] == "KB (new)" | 
                 pat$Sample.ID[i] == "KN (new)" | 
                 pat$Sample.ID[i] == "WA" | pat$Sample.ID[i] == "WB" |
                 pat$Sample.ID[i] == "WN" | pat$Sample.ID[i] == "TA" |
                 pat$Sample.ID[i] == "TB" | pat$Sample.ID[i] == "JB" |
                 pat$Sample.ID[i] == "KA" | pat$Sample.ID[i] == "CA" |
                 pat$Sample.ID[i] == "CB" | pat$Sample.ID[i] == "CN" |
                 pat$Sample.ID[i] == "MA" | pat$Sample.ID[i] == "MB" |
                 pat$Sample.ID[i] == "JA" | pat$Sample.ID[i] == "OA" |
                 pat$Sample.ID[i] == "OB" | pat$Sample.ID[i] == "FA" |
                 pat$Sample.ID[i] == "FB" | pat$Sample.ID[i] == "KA" |
                 pat$Sample.ID[i] == "CA LL" | pat$Sample.ID[i] == "CB LL" |
                 pat$Sample.ID[i] == "FA LL" | pat$Sample.ID[i] == "FB LL" |
                 pat$Sample.ID[i] == "DA" | pat$Sample.ID[i] == "DB" |
                 pat$Sample.ID[i] == "UA" | pat$Sample.ID[i] == "UB" |
                 pat$Sample.ID[i] == "HALL-1" | pat$Sample.ID[i] == "HALL-2"|
                 pat$Sample.ID[i] == "HBLL-3" | pat$Sample.ID[i] == "HBLL-4") {

         pat$type[i] <- "Carcinomas"

      } else {
         pat$type[i] <- "Other"
        }
}

## PCA plot ##
memory.limit(size=30000)
all_noob <- combineArrays(noob_1337_new,noob_1345)
all_noob <- combineArrays(all_noob,noob_1350)
all_noob <- combineArrays(all_noob,noob_1357)
all_noob <- combineArrays(all_noob,noob_1360)
all_noob <- combineArrays(all_noob,noob_1378)
all_noob <- combineArrays(all_noob,noob_1385)
all_noob <- combineArrays(all_noob,noob_1387)

all_lumi <- combineArrays(lumi_1337_new,lumi_1345)
all_lumi <- combineArrays(all_lumi,lumi_1350)
all_lumi <- combineArrays(all_lumi,lumi_1357)
all_lumi <- combineArrays(all_lumi,lumi_1360)
all_lumi <- combineArrays(all_lumi,lumi_1378)
all_lumi <- combineArrays(all_lumi,lumi_1385)
all_lumi <- combineArrays(all_lumi,lumi_1387)

par(mfrow=c(1,2),mar=c(5,3,5,3))
mdsPlot(all_noob,numPositions = 1000000,sampGroups = pat$type,
        main = "PCA Plot Using Noob (without JA)", pch=16)

mdsPlot(all_lumi,numPositions = 1000000,sampGroups = pat$type,
        main = "PCA Plot Using Illumina (without JA)", pch=16)

```

``` {r, echo=FALSE,warning=FALSE,message=FALSE,fig.width=17,fig.height=10}
####################################### Without Cell Lines and Xenografts #############################

####### 1360: all 
####### 1378: two

### read in 1378 without cell lines and xenografts###
base_dir_1378_new <- "1378_Shibata EPIC Data Package/IDAT FILES"

targets_1378_new <- read.xls("1378_Shibata EPIC Data Package/SAMPLE-ARRAY MAPPING/1378 (Shbata-8).xls")

targets_1378_new <- targets_1378_new[c(-1,-4),7]
targets_1378_new <- file.path(base_dir_1378_new, targets_1378_new)
work_1378_new <- read.metharray(targets_1378_new, verbose = TRUE)

# normalization

## noob ##
noob_1378_new <- preprocessNoob(work_1378_new)

## Genome Studio ##
lumi_1378_new <- preprocessIllumina(work_1378_new,bg.correct = TRUE,
                                normalize = "controls",reference = 2)

# Make sure var names are consistent
colnames(pat_1337)[8] <- "Sample.ID"
colnames(pat_1345)[8] <- "Sample.ID"
colnames(pat_1350)[8] <- "Sample.ID"
colnames(pat_1357)[8] <- "Sample.ID"
colnames(pat_1360)[8] <- "Sample.ID"
colnames(pat_1378)[8] <- "Sample.ID"
colnames(pat_1385)[8] <- "Sample.ID"
colnames(pat_1387)[8] <- "Sample.ID"

# Create a data frame that contains all patient information
pat <- rbind(pat_1337[,1:8],pat_1345[,1:8],pat_1350[,1:8],pat_1357[,1:8],
             pat_1360[,1:8],pat_1378[,1:8],pat_1385[,1:8],pat_1387[,1:8])

pat <- pat[-c(33:38,41:49,52),]

# Create a var indicating adenomas/carcinomas in pat
for(i in (1:dim(pat)[1])) {

     if ((pat$Sample.ID[i] == "KA") | (pat$Sample.ID[i] == "KB") |
      pat$Sample.ID[i] == "XA" | pat$Sample.ID[i] == "XB" |
      pat$Sample.ID[i] == "SA" | pat$Sample.ID[i] == "SB" |
      pat$Sample.ID[i] == "PA" | pat$Sample.ID[i] == "PB" |
      pat$Sample.ID[i] == "KA LL" | pat$Sample.ID[i] == "KB LL") {

       pat$type[i] <- "Adenomas"

      } else if (pat$Sample.ID[i] == "HA" | pat$Sample.ID[i] == "HB" |
                 pat$Sample.ID[i] == "HN" | pat$Sample.ID[i] == "JA" |
                 pat$Sample.ID[i] == "JN" | pat$Sample.ID[i] == "KA" |
                 pat$Sample.ID[i] == "EA" | pat$Sample.ID[i] == "EB" |
                 pat$Sample.ID[i] == "EN" | pat$Sample.ID[i] == "KA" |
                 pat$Sample.ID[i] == "KA (new)" | 
                 pat$Sample.ID[i] == "KB (new)" | 
                 pat$Sample.ID[i] == "KN (new)" | 
                 pat$Sample.ID[i] == "WA" | pat$Sample.ID[i] == "WB" |
                 pat$Sample.ID[i] == "WN" | pat$Sample.ID[i] == "TA" |
                 pat$Sample.ID[i] == "TB" | pat$Sample.ID[i] == "JB" |
                 pat$Sample.ID[i] == "KA" | pat$Sample.ID[i] == "CA" |
                 pat$Sample.ID[i] == "CB" | pat$Sample.ID[i] == "CN" |
                 pat$Sample.ID[i] == "MA" | pat$Sample.ID[i] == "MB" |
                 pat$Sample.ID[i] == "JA" | pat$Sample.ID[i] == "OA" |
                 pat$Sample.ID[i] == "OB" | pat$Sample.ID[i] == "FA" |
                 pat$Sample.ID[i] == "FB" | pat$Sample.ID[i] == "KA" |
                 pat$Sample.ID[i] == "CA LL" | pat$Sample.ID[i] == "CB LL" |
                 pat$Sample.ID[i] == "FA LL" | pat$Sample.ID[i] == "FB LL" |
                 pat$Sample.ID[i] == "DA" | pat$Sample.ID[i] == "DB" |
                 pat$Sample.ID[i] == "UA" | pat$Sample.ID[i] == "UB" |
                 pat$Sample.ID[i] == "HALL-1" | pat$Sample.ID[i] == "HALL-2"|
                 pat$Sample.ID[i] == "HBLL-3" | pat$Sample.ID[i] == "HBLL-4") {

         pat$type[i] <- "Carcinomas"

      } 
}

## PCA plot ##
memory.limit(size=30000)
all_noob <- combineArrays(noob_1337_new,noob_1345)
all_noob <- combineArrays(all_noob,noob_1350)
all_noob <- combineArrays(all_noob,noob_1357)
all_noob <- combineArrays(all_noob,noob_1378_new)
all_noob <- combineArrays(all_noob,noob_1385)
all_noob <- combineArrays(all_noob,noob_1387)

all_lumi <- combineArrays(lumi_1337_new,lumi_1345)
all_lumi <- combineArrays(all_lumi,lumi_1350)
all_lumi <- combineArrays(all_lumi,lumi_1357)
all_lumi <- combineArrays(all_lumi,lumi_1378_new)
all_lumi <- combineArrays(all_lumi,lumi_1385)
all_lumi <- combineArrays(all_lumi,lumi_1387)

par(mfrow=c(1,2),mar=c(5,3,5,3))
mdsPlot(all_noob,numPositions = 1000000,sampGroups = pat$type,
        main = "PCA Plot Using Noob (without JA)", pch=16)

mdsPlot(all_lumi,numPositions = 1000000,sampGroups = pat$type,
        main = "PCA Plot Using Illumina (without JA)", pch=16)

```

## Predicting Sex

``` {r,echo=FALSE, warning=FALSE,message=FALSE}

all_gmset <- mapToGenome(all_noob)
all_sex <- getSex(all_gmset)
plotSex(all_sex)

```

## Percent of Probes

```{r,echo=FALSE, warning=FALSE}

########################### Percent of probes ################################

per_fun <- function (work) {

  mset <- preprocessRaw(work)

  rset <- ratioConvert(mset,what = "both")

  beta <- getBeta(rset)

  per <- 1-apply(is.na(beta),2,sum)/dim(beta)[1]

  return(per)
}

per_1337 <- per_fun(work_1337)
per_1345 <- per_fun(work_1345)
per_1350 <- per_fun(work_1350)
per_1357 <- per_fun(work_1357)
per_1360 <- per_fun(work_1360)
per_1378 <- per_fun(work_1378)
per_1385 <- per_fun(work_1385)
per_1387 <- per_fun(work_1387)

```

* Plate 1337

``` {r,echo=FALSE}
per_1337
```

* Plate 1345

``` {r,echo=FALSE}
per_1345
```

* Plate 1350

``` {r,echo=FALSE}
per_1350
```

* Plate 1357

``` {r,echo=FALSE}
per_1357
```

* Plate 1360

``` {r,echo=FALSE}
per_1360
```

* Plate 1378

``` {r,echo=FALSE}
per_1378
```

* Plate 1385

``` {r,echo=FALSE}
per_1385
```

* Plate 1387

``` {r,echo=FALSE}
per_1387
```

```{r,echo=FALSE,eval=FALSE}
################################ Annotation ##################################

# here we select, from the EPIC characterizing data, the columns for:
# ILmnID, CHR, MAPINFO(geneposition), UCSC_RefGene_Name, USCS_RefGene_Group, CpG Island name, Phantom 4, Phantom5
selectedCols <- c("character", rep("NULL", 10), "character", "integer", rep("NULL", 2), "character", "NULL", rep("character", 2), "NULL", rep("character", 2), rep("NULL", 25))
EPICchar <- read.csv("EPIC MANIFEST AND SUPPORTING INFORMATION/MethylationEPIC_v-1-0_B1.csv", as.is=TRUE, colClasses = selectedCols)

# only choose rows with cpg sites in preprocessed data
DataChar <- EPICchar[EPICchar$IlmnID %in% row.names(noob_1337),]

## then add each sample
# data from noob preprocessed data
SampleData <- data.frame(getBeta(noob_1337), getBeta(noob_1345), getBeta(noob_1350), getBeta(noob_1357), getBeta(noob_1360), getBeta(noob_1378), getBeta(noob_1385), getBeta(noob_1387))
# rewrite sample names from patient file
sampleNames <- unlist(list(pat_1337$Sample_No, pat_1345$SAMPLE.ID, pat_1350$SAMPLE.ID, pat_1357$Sample.ID, pat_1360$Sample.ID, pat_1378$Tube.Label, pat_1385$Tube.Label, pat_1387$Sample.ID))
colnames(SampleData) <- sampleNames[-7]
# reorder sample data by cpg site (row name)
SampleData <- SampleData[order(row.names(SampleData)), ]

# save/load to save time
save(DataChar,file="myDC.Rdata")
save(SampleData,file="mySD.Rdata")
load("myDC.Rdata")
load("mySD.Rdata")

# merge two, then sort by IlmnID
FullAnnotation <- cbind(DataChar, SampleData)

save(FullAnnotation,file="myFA.Rdata")
load("myFA.Rdata")

```

### 2020-01-02 stan_Model.Rmd
added old code

---
title: "Stan Model"
output: word_document
---

This is the code for modeling our data using lmer function. The model we are using is:

$$A_i = b_i + \beta_T + b_T_i + \mu + \epsilon_i$$
$$B_i = b_i + \beta_T + b_T_i + \mu + \epsilon_i$$
$$N_i = b_i ~~~~~~~~~~~~~~~~~~+ \mu + \epsilon_i$$

* $A_i$, $B_i$, $N_i$ are the mythelation for side A, B, and N respectively
* $\beta_T$ = effect of tumor tissue (fixed)
* $\mu$ = grand mean
* $b_i \sim N(0,\sigma^2_P)$
* $b_T_i \sim N(0,\sigma^2_T)$
* $\epsilon_i \sim N(0,\sigma^2_e)$

Please download the "model.stan" file from Git and save it in the folder where all the other data are stored. Please also make sure the annotation file is saved in that folder.

This version of codes assigns uniform distribution as a prior for $\beta_T$, $\mu$, $\sigma^2_P$, $\sigma^2_T$, and $\sigma^2_y$.

``` {r, echo=FALSE, warning=FALSE, message=FALSE}
### install the following package
# install.package("rstan")

library(rstan)

# here set the working directory that points to the data folder
# e.g. the folder with annotated data saved as "myFA.Rdata"
# please just comment out the directories already here, so each
# user can uncomment the setwd for them as we move code around

# Kevin's working directory
# setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")

# Yanlin's working directory
setwd("D:/DataPlus2017/Data")

# Set your directory here

### LOAD DATA

# load fully annotated data (saved from LoadDataAndQC.R)
load("myFA.Rdata")

# extract CpG site 1 to start
firstSite <- head(FullAnnotation, 1)

# convert first site to table, using only H E K* W C J (tumor+normal)
# 18 rows, with columns for patient, tumor/normal
# index in order A,B,N
indices <- c(11, 12, 13, 17, 18, 16, 20, 21, 19, 25, 26, 24, 33, 34, 32, 
             39, 31, 15)
patientLabel <- c(rep("H", 3),rep("E", 3),rep("K*", 3),rep("W", 3),rep("C", 3),
                  rep("J", 3))
tissueLabel <- rep(c("T", "T", "N"), 6)
sideLabel <- rep(c("A", "B", ""), 6)

# here we use all samples
indices <- c(9:13, 15:39,46,47,57,58,72:75)
patientLabel <- substr(colnames(firstSite[indices]),1,1)
patientLabel[10:12] <- "K*"
sideLabel <- substr(colnames(firstSite[indices]),2,2)
tissueLabel <- sideLabel
tissueLabel[tissueLabel %in% c("A","B")] <- "T"   #replace A and B with T
tissueIndicator <- 1*(tissueLabel=="T")

firstData <- data.frame(beta=t(firstSite[indices]), patient = patientLabel, tissue=tissueLabel, side=sideLabel, tInd=tissueIndicator)

### FIT WITH STAN

# Logit transform on beta values

firstData$y <- log(firstData$X1/(1-firstData$X1))

# Stan data

stanDat <- list(pID = as.integer(factor(firstData$patient)),
                tInd = firstData$tInd,
                N = nrow(firstData),
                P = nlevels(firstData$patient),
                y = firstData$y)

stanFit <- stan(file="model.stan",data=stanDat)
```

#### Summary of the Stan Model

``` {r, echo=FALSE,warning=FALSE}

print(stanFit, probs = c(0.025, 0.5, 0.975))

```

#### Estimatesfor the first three samples

``` {r, echo=FALSE,warning=FALSE}

est <- as.matrix(stanFit,pars=c("mu","betaT","b_pat","bT_pat"))

# Estimates for first three samples
head(est[,c(1:5,19:21)])
```

#### Graphs (still working on finding how useful these graphs are)

``` {r, echo=FALSE, warning=FALSE}
# Diagnostic Graphs

plot(stanFit, pars=c("b_pat"), main="Random Effect of Patients")
plot(stanFit, pars=c("bT_pat"), main="Random Effect of Tumor Tissue Varying Between Patients")

```

### 2020-01-02 GetPosteriorsSingleSites.R

@@ -1,0 +1,225 @@
+# script for extracting variance from Stan complex model (model3)
+# input taskID corresponding to 10000 site chunks starting at site (N-1)*10000+1
+# Now incorporating parallel processing by RStan methods
+
+library(doParallel)
+library(rstan)
+library(gtools)
+library(ggplot2)
+library(bayesplot)
+
+
+## INITIALIZE
+rstan_options(auto_write = FALSE)
+print(paste("Cores: ", detectCores()))
+
+# Kevin's working directory
+setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
+
+# Yanlin's working directory
+#setwd("D:/DataPlus2017/Data")
+
+# load data
+load("myFA.Rdata")
+
+#args <- commandArgs(trailingOnly = TRUE)
+#print(args)
+#N <- as.numeric(args[1])
+N <- 1
+print(paste("Task #: ", N))
+#out.file <- args[2]
+log.file <- paste("logTask",N,".txt",sep="")
+writeLines(c(""), log.file)
+
+# randomly select 5 sites in 1-866836
+siteInds <- sample(1:866836,5)
+nsites <- length(siteInds)
+
+# select chunk within FullAnnotation, then remove FA
+print(paste("Making chunk of nsites =", nsites))
+chunk <- FullAnnotation[siteInds,]
+rm(FullAnnotation)
+gc()
+
+# must add to prevent crash
+checkFile <- "model3.rds"
+
+
+### FXNS ###
+# Function used to read in data from each site
+# changed to now use a "chunk" instead of FullAnnotation, for speed and memory
+# requires index 1-10000 instead of siteIndex
+site <- function (ind) {
+  # extract CpG site xx to start
+  temp <- chunk[ind,]
+  
+  # here we use all patient samples, excluding glands
+  indices <- c(9:13, 15:39, 46, 47, 57, 58, 72:75)
+  patientLabel <- substr(colnames(temp[indices]), 1, 1)
+  patientLabel[10:12] <- "K*"
+  sideLabel <- substr(colnames(temp[indices]), 2, 2)
+  tissueLabel <- sideLabel
+  tissueLabel[tissueLabel %in% c("A", "B")] <-
+    "T"   #replace A and B with T
+  tumorIndicator <- 1 * (tissueLabel == "T")
+  
+  work <-
[…]

### 2020-01-02 LoadDataAndQC.R

@@ -1,0 +1,457 @@
+#################################################################################
+######## This program is used for reading in dataset and qualify control ########
+######### Refer to the RMarkdown file (coming soon) for final results ###########
+#################################################################################
+
+
+library(minfi)
+library(minfiData)
+library(IlluminaHumanMethylationEPICmanifest)
+library(IlluminaHumanMethylationEPICanno.ilm10b2.hg19)
+library(gdata) #gdata package needs Perl language installed
+library(RColorBrewer)
+
+# here set the working directory that points to the data folder
+# e.g. the folder with all datasets in it, should contain all the
+# 1337-1387 folders
+# please just comment out the directories already here, so each
+# user can uncomment the setwd for them as we move code around
+
+# Kevin's working directory
+setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
+
+# Yanlin's working directory
+setwd("D:/DataPlus2017/Data")
+
+############### Function used for read in data ####################
+
+# base_dir = directory containing basenames file
+# targets = name of the dataset after reading in basenames file
+# pat_dir = directory containing patient information file
+# pat_file = WHOLE name (including directory) of the dataset after reading in patient information file
+# work_name = name of the RGChannelSet
+
+read.fun <- function(base_dir,pat_file) {
+  # read in patient data from the .xlsx file
+  pat_name <- read.xls(pat_file)
+  
+  # Extract targets
+  targets_name <- data.frame(pat_name[,"Complete.Barcode"])
+  colnames(targets_name) <- "Basement"
+  
+  # Read in all .idat files
+  targets_name$Basement <- file.path(base_dir, targets_name$Basement)
+  work_name <- read.metharray(targets_name$Basement, verbose = TRUE)
+  
[…]

### 2020-01-02 NormalSimulation.R

@@ -1,0 +1,123 @@
+#################################################################################
+######################### Normal Tissue Data Simulation #########################
+#################################################################################
+
+
+library(rstan)
+library(gtools)
+library(ggplot2)
+library(bayesplot)
+library(MASS)
+
+# Kevin's working directory
+setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
+
+# Yanlin's working directory
+#setwd("D:/DataPlus2017/Data")
+
+# load data
+load("myFA.Rdata")
+
+
+# Extract single site data
+site <- function (site_no) {
+  # extract CpG site xx to start
+  temp <- FullAnnotation[site_no,]
+  
+  # here we use all patient samples, excluding glands
+  indices <- c(9:13, 15:39,46,47,57,58,72:75)
+  patientLabel <- substr(colnames(temp[indices]),1,1)
+  patientLabel[10:12] <- "K*"
+  sideLabel <- substr(colnames(temp[indices]),2,2)
+  tissueLabel <- sideLabel
+  tissueLabel[tissueLabel %in% c("A","B")] <- "T"   #replace A and B with T
+  tumorIndicator <- 1*(tissueLabel=="T")
+  
+  work <- data.frame(beta=logit(t(temp[indices])), patient = patientLabel, tissue=tissueLabel, side=sideLabel, tInd=tumorIndicator)
+  colnames(work)[1] <- "beta"
+  return(work)
+}
+
+# Using Model 3: add intra-tumoral variance
+
+stanfit3 <- function (dataset) {
+  
+  stanDat <- list(pID = as.integer(factor(dataset$patient)),
+                  tInd = dataset$tInd,
+                  N = nrow(dataset),
+                  P = nlevels(dataset$patient),
+                  y = dataset[,1])
+  
+  
+  
+  stanFit3 <- stan(file="model3.stan", data=stanDat, control = list(adapt_delta = 0.999))
+  
+  return(stanFit3=stanFit3)
+}
+
+# Plot posterior distribution
+posterior <- function(dataset) {
+  stan <- stanfit3(dataset)
+  mcmc_areas(
+    as.array(stan), 
+    pars = c("sigma_p", "sigma_pt","sigma_t","sigma_e"),
+    prob = 0.8, # 80% intervals
+    prob_outer = 0.95, 
+    point_est = "median"
+  )
+}
+
+# Randomly choose 5 sites
[…]

### 2020-01-02 PlotBetas.R

@@ -1,0 +1,390 @@
+### plot beta values between specific samples
+
+### INITIALIZE
+
+library(ggplot2)
+library(minfi)
+library(MASS)
+
+# here set the working directory that points to the data folder
+# e.g. the folder with annotated data saved as "myFA.Rdata"
+# please just comment out the directories already here, so each
+# user can uncomment the setwd for them as we move code around
+
+# Kevin's working directory
+setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
+
+# Yanlin's working directory
+setwd("D:/DataPlus2017/Data")
+
+
+##### LOAD DATA
+
+# load fully annotated data (saved from LoadDataAndQC.R)
+load("myFA.Rdata")
+
+
+### mds plot
+
+# mds plot
+cutData <- FullAnnotation[-(1:8)]
+
+pat <- data.frame(Sample.ID = colnames(cutData))
+
+pat$Sample.ID = colnames(cutData)
+
+aInd = c(1,2,14,15,21,22,29,30,52:55)
+cInd = c(3:13,16:20,23:28,31,38,39,49,50,56:71)
+oInd = c(32:37,40:48)
+
+pat$type[aInd] <- "Adenomas"
+pat$type[cInd] <- "Carcinomas"
+pat$type[oInd] <- "Other"
+
+patLabel <- pat$Sample.ID
+
+cutData2 <- cutData[pat$type=="Adenomas"|pat$type=="Carcinomas"]
+
+mdsPlot(as.matrix(cutData),numPositions = 1000000,sampNames=patLabel,sampGroups = pat$type,
+        main = "PCA Plot Using Noob (with JA)", pch=16,cex=0.5)
+
+
+
+
+##### COMPARE BETA
+
+### tumor vs normal for patients H E K* W C J (have both)
+# average tumor side A and B for tumor data
+
+# H
+betaHNor <- FullAnnotation$HN
+betaHTum <- (FullAnnotation$HA + FullAnnotation$HB)/2
[…]

### 2020-01-02 Plots_FullData.R

@@ -1,0 +1,373 @@
+################### Plots ############################
+
+library(ggplot2)
+library(dunn.test)
+
+# here set the working directory that points to the data folder
+# e.g. the folder with annotated data saved as "myFA.Rdata"
+# please just comment out the directories already here, so each
+# user can uncomment the setwd for them as we move code around
+
+# Kevin's working directory
+setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
+
+# Yanlin's working directory
+setwd("D:/DataPlus2017/Data")
+
+
+### LOAD DATA
+load("StanCFullResults.RData")
+
+sigma <- data.frame(sigmaP_Cfull$p50,sigmaPT_Cfull$p50,sigmaT_Cfull$p50,sigmaE_Cfull$p50)
+
+
+### Estimates Distribution
+ggplot(mu_Cfull,aes(x=mu_Cfull$p50)) + geom_histogram(binwidth = 0.1) + labs(title="Distribution of mu",x="mu (median)")
+
+ggplot(betaT_Cfull,aes(x=betaT_Cfull$p50)) + geom_histogram(binwidth = 0.1) + labs(title="Distribution of betaT",x="betaT (median)")
+
+ggplot(sigmaP_Cfull,aes(x=sigmaP_Cfull$p50)) + geom_histogram(binwidth = 0.05) + labs(title="Distribution of sigmaP",x="sigmaP (median)") + coord_cartesian(ylim=c(0,175000),xlim=c(0,3))
+
+ggplot(sigmaPT_Cfull,aes(x=sigmaPT_Cfull$p50)) + geom_histogram(binwidth = 0.05) + labs(title="Distribution of sigmaPT",x="sigmaPT (median)") + coord_cartesian(ylim=c(0,175000),xlim=c(0,3))
+
+ggplot(sigmaT_Cfull,aes(x=sigmaT_Cfull$p50)) + geom_histogram(binwidth = 0.05) + labs(title="Distribution of sigmaT",x="sigmaT (median)") + coord_cartesian(ylim=c(0,175000),xlim=c(0,3))
+
[…]

### 2020-01-02 Score.R

@@ -1,0 +1,135 @@
+########################## Scoring ################################
+
+library(doParallel)
+
+### Initialize
+
+print(paste("Cores: ", detectCores()))
+
+# here set the working directory that points to the data folder
+# e.g. the folder with annotated data saved as "myFA.Rdata"
+# please just comment out the directories already here, so each
+# user can uncomment the setwd for them as we move code around
+
+# Kevin's working directory
+setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
+
+# Yanlin's working directory
+setwd("D:/DataPlus2017/Data")
+
+
+### LOAD DATA
+
+# load fully annotated data (saved from LoadDataAndQC.R)
+load("myFA.Rdata")
+load("GeneInfo.Rdata")
+load("StanCfullResults3.Rdata")
+
+N<-1
+
+# make vector of unique genes and regions
+uniqueGenes <- unique(unlist(geneNames))
+regionTypes <- unique(unlist(geneRegions))
+
+# make dataframes for median values of each parameter
+dfMedians <- cbind.data.frame(mu_Cfull3$p50, betaT_Cfull3$p50, sigmaE_Cfull3$p50, sigmaP_Cfull3$p50, sigmaPT_Cfull3$p50, sigmaT_Cfull3$p50)
+colnames(dfMedians) <- c("mu","betaT","sigmaE","sigmaP","sigmaPT","sigmaT")
+
+# calculate log ratios: P/T, P/PT, PT/T
+logRatios <- cbind.data.frame(log(dfMedians$sigmaP/dfMedians$sigmaT), log(dfMedians$sigmaP/dfMedians$sigmaPT), log(dfMedians$sigmaPT/dfMedians$sigmaT))
+colnames(logRatios) <- c("logP/T","logP/PT","logPT/T")
+logRatioMeans <- colMeans(logRatios)
+# insert means by region
+
+# Choose 1000 genes from uniqueGenes by N, and select geneChunk within gene list
+geneIDs <- (1:1000) + (N - 1) * 1000
+if (N > 27) {
+  geneIDs <- geneIDs[1:383]
+}
+geneIDs <- 1:20
+numGenes <- length(geneIDs)
+
+print(paste("Making chunk of numGenes =", numGenes))
+geneChunk <- uniqueGenes[geneIDs]
+
+
+### Functions
+
+# find indices corresponding to geneString in the list geneNames
[…]

### 2020-01-02 ScoreGenesParallel.R

@@ -1,0 +1,121 @@
+# script for calculating scores on individual genes (27383 total genes)
+# using full-scale results from Stan complex model (model3), including PTprob (run 3)
+# input taskID (1-28) corresponding to 1000 gene chunks starting at site (N-1)*1000+1
+# incorporating parallel processing to run through all genes faster
+# just need "StanCfullResults3.Rdata" and "GeneInfo" in working directory
+
+library(doParallel)
+
+### Initialize
+
+print(paste("Cores: ", detectCores()))
+
+args <- commandArgs(trailingOnly = TRUE)
+print(args)
+N <- as.numeric(args[1])
+print(paste("Task #:", N))
+out.file <- args[2]
+log.file <- paste0("logTask",N,".txt")
+writeLines(c(""), log.file)
+
+# load data: Stan parameters, geneNames and geneRegions lists
+load("StanCfullResults3.Rdata")
+load("GeneInfo.Rdata")
+
+# make vector of unique genes and regions
+uniqueGenes <- unique(unlist(geneNames))
+regionTypes <- unique(unlist(geneRegions))
+
+# count sites at each geneName
+geneCount <- data.frame(table(unlist(geneNames)))
+
+# make dataframes for median values of each parameter
+dfMedians <- cbind.data.frame(mu_Cfull$p50, betaT_Cfull$p50, sigmaE_Cfull$p50, sigmaP_Cfull$p50, sigmaPT_Cfull$p50, sigmaT_Cfull$p50)
+colnames(dfMedians) <- c("mu","betaT","sigmaE","sigmaP","sigmaPT","sigmaT")
+
+# calculate log ratios: P/T, P/PT, PT/T
+logRatios <- cbind.data.frame(log(dfMedians$sigmaP/dfMedians$sigmaT), log(dfMedians$sigmaP/dfMedians$sigmaPT), log(dfMedians$sigmaPT/dfMedians$sigmaT))
+colnames(logRatios) <- c("logP/T","logP/PT","logPT/T")
+logRatioMeans <- colMeans(logRatios)
+
+# Choose 1000 genes from uniqueGenes by N, and select geneChunk within gene list
+geneIDs <- (1:1000) + (N - 1) * 1000
+if (N > 27) {
+  geneIDs <- geneIDs[1:383]
+}
+numGenes <- length(geneIDs)
+
+print(paste("Making chunk of numGenes =", numGenes))
[…]

### 2020-01-02 StanCParallel.R

@@ -1,0 +1,239 @@
+# script for extracting variance from Stan complex model (model3)
+# input taskID corresponding to 10000 site chunks starting at site (N-1)*10000+1
+# Now incorporating parallel processing by RStan methods
+
+library(doParallel)
+library(rstan)
+library(gtools)
+
+rstan_options(auto_write = FALSE)
+print(paste("Cores: ", detectCores()))
+
+# load data
+load("myFA.Rdata")
+
+args <- commandArgs(trailingOnly = TRUE)
+print(args)
+N <- as.numeric(args[1])
+print(paste("Task #: ", N))
+out.file <- args[2]
+log.file <- paste("logTask",N,".txt",sep="")
+writeLines(c(""), log.file)
+
+# Choose 10000 sites by N, and select chunk within FullAnnotation, then remove FA
+siteInds <- (1:10000) + (N - 1) * 10000
+if (N > 86) {
+  siteInds <- siteInds[1:6836]
+}
+nsites <- length(siteInds)
+
+print(paste("Making chunk of nsites =", nsites))
+chunk <- FullAnnotation[siteInds,]
+rm(FullAnnotation)
+gc()
+
+# must add to prevent crash
+checkFile <- "model3.rds"
+
+
+### FXNS ###
+# Function used to read in data from each site
+# changed to now use a "chunk" instead of FullAnnotation, for speed and memory
+# requires index 1-10000 instead of siteIndex
+site <- function (ind) {
+  # extract CpG site xx to start
+  temp <- chunk[ind,]
+  
+  # here we use all patient samples, excluding glands
+  indices <- c(9:13, 15:39, 46, 47, 57, 58, 72:75)
+  patientLabel <- substr(colnames(temp[indices]), 1, 1)
+  patientLabel[10:12] <- "K*"
+  sideLabel <- substr(colnames(temp[indices]), 2, 2)
+  tissueLabel <- sideLabel
+  tissueLabel[tissueLabel %in% c("A", "B")] <-
+    "T"   #replace A and B with T
+  tumorIndicator <- 1 * (tissueLabel == "T")
+  
+  work <-
+    data.frame(
+      beta = logit(t(temp[indices])),
+      patient = patientLabel,
+      tissue = tissueLabel,
[…]

### 2020-01-02 StanCParallel2.R

@@ -1,0 +1,263 @@
+# script for extracting variance from Stan complex model (model3)
+# input taskID corresponding to site chunks, here set as 1
+# testing on first 100 sites
+# Now incorporating parallel processing with parallel package
+
+library(doParallel)
+library(rstan)
+library(gtools)
+library(pryr)
+
+rstan_options(auto_write = FALSE)
+#options(mc.cores = detectCores())
+print(paste("Cores: ", detectCores()))
+
+# Kevin's working directory
+setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
+# load data
+load("myFA.Rdata")
+
+N <- 1
+#args <- commandArgs(trailingOnly = TRUE)
+#print(args)
+#N <- as.numeric(args[1])
+print(paste("Task #: ", N))
+#out.file <- args[2]
+log.file <- paste("logTask",N,".txt",sep="")
+writeLines(c(""), log.file)
+
+# Choose 1000  sites by N, and select chunk within FullAnnotation
+# siteInds <- (1:1000) + (N - 1) * 1000
+# if (N > 866) {
+#   siteInds <- siteInds[1:836]
+# }
+# nsites <- length(siteInds)
+
+siteInds <- (1:3) + (N - 1) * 3
+nsites <- length(siteInds)
+
+print(paste("Making chunk of nsites =", nsites))
+chunk <- FullAnnotation[siteInds,]
+rm(FullAnnotation)
+gc()
+
+# must add to prevent crash, by checking for presence of rogue decoy Stan (.rds) file
+checkFile <- "model3.rds"
+
+
+### FXNS ###
+# Function used to read in data from each site
+# changed to now use a "chunk" instead of FullAnnotation, for speed and memory
+# requires index 1-1000 instead of siteIndex
+site <- function (ind) {
+  # extract CpG site xx to start
+  temp <- chunk[ind,]
+  
+  # here we use all patient samples, excluding glands
+  indices <- c(9:13, 15:39, 46, 47, 57, 58, 72:75)
+  patientLabel <- substr(colnames(temp[indices]), 1, 1)
+  patientLabel[10:12] <- "K*"
+  sideLabel <- substr(colnames(temp[indices]), 2, 2)
+  tissueLabel <- sideLabel
[…]

### 2020-01-02 StanCVarsScript1.R

@@ -1,0 +1,137 @@
+# script for extracting variance from Stan model on random 5K sites
+# part 1
+
+
+library(rstan)
+library(gtools)
+
+# load data
+load("myFA.Rdata")
+
+args <- commandArgs(trailingOnly = TRUE)
+N <- as.numeric(args[1])
+out.file <- paste("Results_", args[1], ".Rdata", sep="")
+
+# Function used to read in data from each site
+
+site <- function (site_no) {
+  # extract CpG site xx to start
+  temp <- FullAnnotation[site_no,]
+  
+  # here we use all patient samples, excluding glands
+  indices <- c(9:13, 15:39,46,47,57,58,72:75)
+  patientLabel <- substr(colnames(temp[indices]),1,1)
+  patientLabel[10:12] <- "K*"
+  sideLabel <- substr(colnames(temp[indices]),2,2)
+  tissueLabel <- sideLabel
+  tissueLabel[tissueLabel %in% c("A","B")] <- "T"   #replace A and B with T
+  tumorIndicator <- 1*(tissueLabel=="T")
+  
+  work <- data.frame(beta=logit(t(temp[indices])), patient = patientLabel, tissue=tissueLabel, side=sideLabel, tInd=tumorIndicator)
+  
+  return(work)
+}
+
+
+
+# Function to build stan model for each site
+
+stanfit3 <- function (dataset) {
+  
+  stanDat <- list(pID = as.integer(factor(dataset$patient)),
+                  tInd = dataset$tInd,
+                  N = nrow(dataset),
+                  P = nlevels(dataset$patient),
+                  y = dataset[,1])
+  
+  
+  # Using Model 3: add intra-tumoral variances
+  stanFit3 <- stan(file="model3.stan", data=stanDat, control = list(adapt_delta = 0.999))
+  
+  return(stanFit3=stanFit3)
+}
+
+
+
+# Function to get mode
+
+getmode <- function(v) {
+  uniqv <- unique(v)
+  uniqv[which.max(tabulate(match(v, uniqv)))]
+}
+
+
+# Choose 1k sites by N
+
+siteInds <- ((N*1000-999):(N*1000))
+if (N>866) { siteInds <- ((N*1000-999):866836) }
+nsites <- length(siteInds)
+
+
+
[…]

### 2020-01-02 estimateTCGApriors.R

@@ -1,0 +1,283 @@
+# script for estimating sigma priors from TCGA patient data
+# loads and combines patient data (with annotation)
+# then algebraically calculates prior distribution estimates for each site
+# to ultimately produce a full prior distribution then fit with a parameterized gamma
+# currently test version
+
+library(gtools)
+library(ggplot2)
+
+### INITIALIZE ###
+
+log.file <- "logTCGA.txt"
+writeLines(c(""), log.file)
+
+# Kevin's working directory
+setwd("/Users/kevinmurgas/Documents/Data+ project/TCGA data")
+
+# Yanlin's working directory
+#setwd("D:/DataPlus2017/Data")
+
+# Load and concatenate all available data
+# need to: search all folders for data, load each one just the beta and label
+fileList <- list.files(pattern = "_hg38.txt$", recursive = TRUE)
+fileList <- fileList[-grep("tion27", fileList)] # filters out 27K files
+# also check that A6-2672 is not utilized (only tumor B in 450K)
+
+# intialize data frame
+ptmTot <- proc.time()
+print("Initializing data frame with annotations")
+temp <- read.table(fileList[1],header=TRUE,sep="\t",fill=TRUE)
+concatData <- as.data.frame.matrix(temp[,-2])
+
+print("Now adding patient data")
+for(f in fileList) {
+  ptm <- proc.time()
+  print(paste("File: ",which(f==fileList),"/",length(fileList)))
+  dataLabel <- substr(f,regexpr("TCGA",f),regexpr("TCGA",f)+15)
+  print(paste("Label: ", dataLabel))
+  temp <- read.table(f,header=TRUE,sep="\t",fill=TRUE)
+  concatData[dataLabel] <- temp[,2]
+  print(proc.time() - ptm)
+}
+rm(temp)
+gc()
+
+print("Total time to load data:")
+print(proc.time()-ptmTot)
+
+save(concatData, file="testdata_allpats.Rdata")
+load("testdata_allpats.Rdata") # start here to load concatenated patient data
+
+# create labels
[…]

### 2020-01-02 individualGenes.R

@@ -1,0 +1,296 @@
+# mark individual genes on histogram of Stan results
+
+library(lme4)
+library(arm)
+library(rstan)
+library(coda)
+library(gtools)
+library(ggplot2)
+library(bayesplot)
+library(reshape2)
+
+# Kevin's working directory
+setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
+
+# Yanlin's working directory
+setwd("D:/DataPlus2017/Data")
+
+# load annotation, Stan results, and gene scores
+load("myFA.Rdata")
+load("StanCfullResults.Rdata")
+load("StanCfullResults2.Rdata")
+load("StanCfullResults3.Rdata")
+load("ScoresFull.Rdata")
+
+
+### INITIALIZE
+
+# # create list of unique gene names and gene regions at each site
+# geneNames <- list(866836)
+# geneRegions <- list(866836)
+# for (i in 1:866836) {
+#   if (!(i %% 86684)) {
+#     print(paste(i/8668.4,"% complete"))
+#   }
+#   geneNames[[i]] <- unique(unlist(strsplit(FullAnnotation$UCSC_RefGene_Name[i],split=";")))
+#   worklist <- unique(unlist(strsplit(FullAnnotation$UCSC_RefGene_Group[i],split=";")))
+#   worklist[worklist=="3'UTR"] <- "3UTR"
+#   worklist[worklist=="5'UTR"] <- "5UTR"
+#   worklist[worklist=="5URT"] <- "5UTR"
+#   if (length(worklist)==0) { worklist <- "blank"}
+#   geneRegions[[i]] <- worklist
+# }
+# save(geneNames, geneRegions, file = "GeneInfo.Rdata")
+load("GeneInfo.Rdata")
+
+# make vector of unique genes and regions
+uniqueGenes <- unique(unlist(geneNames))
+regionTypes <- unique(unlist(geneRegions))
+
+# make dataframes for median and mean values of each parameter
+dfMedians <- cbind.data.frame(mu_Cfull$p50, betaT_Cfull$p50, sigmaE_Cfull$p50, sigmaP_Cfull$p50, sigmaPT_Cfull$p50, sigmaT_Cfull$p50)
+colnames(dfMedians) <- c("mu","betaT","sigmaE","sigmaP","sigmaPT","sigmaT")
[…]

### 2020-01-02 loadSCRuns.R

@@ -1,0 +1,167 @@
+# load StanCRun data
+# produce full size data frames for mu,betaT,sigmas (P,PT,T,E)
+# read in individual save files containing 10000 site runs
+# store 10000 site chunks in corresponding positions in full data frame
+# repeat for each completed run
+
+library(ggplot2)
+library(reshape2)
+
+### LOADING FOR STAN MODEL RESULTS
+
+# Set working directory to normal data wd then folder ResultsC
+setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data/NewResultsC")
+
+betaT_Cfull3 <-
+  data.frame(
+    mean = numeric(866836),
+    p2.5 = numeric(866836),
+    p25 = numeric(866836),
+    p50 = numeric(866836),
+    p75 = numeric(866836),
+    p97.5 = numeric(866836)
+  )
+
+mu_Cfull3 <- betaT_Cfull3
+sigmaE_Cfull3 <- betaT_Cfull3
+sigmaP_Cfull3 <- betaT_Cfull3
+sigmaPT_Cfull3 <- betaT_Cfull3
+sigmaT_Cfull3 <- betaT_Cfull3
+PTprob3 <- numeric(866836)
+
+nsites <- 10000
+filesPresent <- (1:87)
+fileNames <- dir(pattern = "StanCParResults_*")
+allInds <- numeric()
+for (i in filesPresent) {
+  print(paste("Chunk:",i))
+  load(fileNames[fileNames == paste("StanCParResults_",i,".Rdata", sep="")])
+  inds <- (1:nsites) + (i - 1) * nsites
+  if (i==87) {
+    inds <- inds[1:6836]
+  }
+  allInds <- append(allInds, inds)
+  betaT_Cfull3[inds, ] <- betaT_C
+  mu_Cfull3[inds, ] <- mu_C
+  sigmaE_Cfull3[inds, ] <- sigmaE_C
+  sigmaP_Cfull3[inds, ] <- sigmaP_C
+  sigmaPT_Cfull3[inds, ] <- sigmaPT_C
+  sigmaT_Cfull3[inds, ] <- sigmaT_C
+  PTprob3[inds] <- PTprob
+}
+save(mu_Cfull3, betaT_Cfull3, sigmaP_Cfull3, sigmaT_Cfull3,
[…]

### 2020-01-02 varyChainParams.R

@@ -1,0 +1,384 @@
+# script for extracting variance from Stan complex model (model3)
+# input taskID corresponding to 10000 site chunks starting at site (N-1)*10000+1
+# Now incorporating parallel processing by RStan methods
+
+library(doParallel)
+library(rstan)
+library(gtools)
+library(ggplot2)
+library(bayesplot)
+
+## INITIALIZE
+rstan_options(auto_write = FALSE)
+print(paste("Cores: ", detectCores()))
+
+# Kevin's working directory
+setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
+
+# Yanlin's working directory
+#setwd("D:/DataPlus2017/Data")
+
+# load data
+load("myFA.Rdata")
+
+#args <- commandArgs(trailingOnly = TRUE)
+#print(args)
+#N <- as.numeric(args[1])
+N <- 1
+print(paste("Task #: ", N))
+#out.file <- args[2]
+log.file <- paste("logTask",N,".txt",sep="")
+writeLines(c(""), log.file)
+
+# randomly select 5 sites in 1-866836
+siteInds <- sample(1:866836,1)
+nsites <- length(siteInds)
+
+# select chunk within FullAnnotation, then remove FA
+print(paste("Making chunk of nsites =", nsites))
+chunk <- FullAnnotation[siteInds,]
+rm(FullAnnotation)
+gc()
+
+# must add to prevent crash
+checkFile <- "model3.rds"
+
+
+### FXNS ###
+# Function used to read in data from each site
+# changed to now use a "chunk" instead of FullAnnotation, for speed and memory
+# requires index 1-10000 instead of siteIndex
+site <- function (ind) {
+  # extract CpG site xx to start
+  temp <- chunk[ind,]
+  
+  # here we use all patient samples, excluding glands
+  indices <- c(9:13, 15:39, 46, 47, 57, 58, 72:75)
+  patientLabel <- substr(colnames(temp[indices]), 1, 1)
+  patientLabel[10:12] <- "K*"
+  sideLabel <- substr(colnames(temp[indices]), 2, 2)
+  tissueLabel <- sideLabel
+  tissueLabel[tissueLabel %in% c("A", "B")] <-
+    "T"   #replace A and B with T
+  tumorIndicator <- 1 * (tissueLabel == "T")
+  
[…]

### 2020-09-25 50sites files/50sites.Rmd
updated code

---
title: "50sites"
author: "Yanlin Ma + Kevin Murgas"
date: "April 13, 2020"
output: word_document
---
Model 3: $$ y_{i,j,k} = \mu + b_j + I_{tumor} * ( \beta_T + c_j + d_k) + \epsilon_i $$
$$ b_j \sim N(0, \sigma_P), \ c_j \sim N(0,\sigma_{PT}),\ d_k \sim N(0, \sigma_T), \ \epsilon_i \sim N(0,\sigma_E)  $$
Stan Chain parameters: \n
warmup=200, iter=2000, chains=4, thin=1 (no thin), adapt_delta=0.999, seed=123 \n
```{r echo=FALSE, include=FALSE}
library(rstan)
library(gtools)
library(ggplot2)
library(bayesplot)
library(knitr)
library(ggpubr)
```
\n Using Relaxed Gamma Priors (x5 variance) \n
```{r echo=FALSE, include=FALSE}
# Kevin's working directory
setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")

# Yanlin's working directory
# setwd("D:/DataPlus2017/Data")

# load data
load("myFA_bulkonly.Rdata")

# Extract single site data
site <- function (site_no) {
  # extract CpG site xx to start
  temp <- FullAnnotation[site_no,]

  # here we use all patient samples, excluding glands
  #indices <- c(10:14, 16:40,47,48,58,59,73:76,87:96) # if using full data file
  indices <- c(10:57) # if using bulk only data file
  patientLabel <- substr(colnames(temp[indices]),1,1)
  patientLabel[10:12] <- "K*"
  sideLabel <- substr(colnames(temp[indices]),2,2)
  tissueLabel <- sideLabel
  tissueLabel[!(sideLabel %in% c("N"))] <- "T"   #replace A/B/D/M with T
  tumorIndicator <- 1*(tissueLabel=="T")

  work <- data.frame(beta=logit(t(temp[indices])), patient = patientLabel, tissue=tissueLabel, side=sideLabel, tInd=tumorIndicator)
  colnames(work)[1] <- "beta"
  return(work)
}

# Using Model with TCGA priors (previously model 3)
stanfit_model <- function (dataset) {
  stanDat <- list(
    N = nrow(dataset),
    P = nlevels(dataset$patient),
    pID = as.integer(factor(dataset$patient)),
    tInd = dataset$tInd,
    y = dataset[, 1]
    )
  stanFit <-
    stan(
      fit = emptyFit,
      warmup = 200,
      iter = 2000,
      chains = 4,
      data = stanDat,
      control = list(adapt_delta = 0.999),
      refresh = 0,
      seed=123
    )
  return(stanFit=stanFit)
}

# Plot raw data and fits
fitplot <- function(data,stanFit) {
  # plot beta and fits at this site for all patients, using mean estimates for b and c coefficients
  # first plot data points
  p1 <- ggplot() + geom_point(data = data, aes(x=tInd, y=beta, colour=patient))

  # extract coefficients from stan fit
  pats <- as.integer(factor(data$patient))
  nsamp <- length(pats)
  npats <- max(pats)
  sumTemp <- summary(stanFit)$summary
  mu <- sumTemp[1,1] # mean estimate
  betaT <- sumTemp[2,1] # mean
  sampcoefs <- data.frame(pat = pats, b = sumTemp[6+pats], c = sumTemp[6+npats+pats], d = sumTemp[6+2*npats+c(1:nsamp)])
  patcoefs <- data.frame(pat = 1:npats, b = sumTemp[6+1:npats], c = sumTemp[6+npats+1:npats])
  rm(sumTemp)
  # add individual sample estimates
  sampdf <- data.frame(pat = rep(factor(data$patient),2), patg = rep(factor(1:nsamp),2), tInd = c(rep(0,nsamp),rep(1,nsamp)), est = c(sampcoefs$b+mu, (sampcoefs$b+sampcoefs$c++sampcoefs$d+mu+betaT)))
  p1 <- p1 + geom_line(data=sampdf, aes(x=tInd, y=est, colour=pat, group=patg), linetype="dashed", alpha=0.5)
  # add patient mean estimates
  patdf <- data.frame(pat = rep(levels(data$patient),2), tInd = c(rep(0,npats),rep(1,npats)), est = c(patcoefs$b+mu, patcoefs$b+patcoefs$c+mu+betaT))
  p1 <- p1 + geom_line(data=patdf, aes(x=tInd, y=est, colour=pat))
  # add overall mean estimates as black dots
  meandf <- data.frame(tInd = c(0,1), est = c(mu,mu+betaT))
  p1 <- p1 + geom_point(data=meandf, aes(x=tInd, y=est), size=3, alpha=0.7) + geom_line(data=meandf, aes(x=tInd, y=est), size=2, alpha=0.5)

  return(p1)
}

# Plot posterior distribution
postplot <- function(stanFit) {
  p2 <- mcmc_areas(
    as.array(stanFit), 
    pars = c("sigma_p", "sigma_pt","sigma_t","sigma_e"),
    prob = 0.8, # 80% intervals
    prob_outer = 0.95, 
    point_est = "median"
  )
  return(p2)
}

## CODE
# Randomly choose 50 sites
set.seed(123)
siteInds <- sample(1:866091,50)
nsites <- length(siteInds)

# check for model_TCGApriors.rds, must add to prevent crash
checkFile <- "model_relaxedgamma5.rds"
if (file.exists(checkFile)) {
  file.remove(checkFile)
  print("caught one!") 
}

# create empty fit to use same model compilation throughout
data1<-site(1)
stanDat1 <- list(
  N = nrow(data1),
  P = nlevels(data1$patient),
  pID = as.integer(factor(data1$patient)),
  tInd = data1$tInd,
  y = data1[, 1]
)
emptyFit <- stan(file="model_relaxedgamma5.stan", data = stanDat1, chains = 0)

```

```{r echo=FALSE,include=TRUE,error=FALSE,message=FALSE,warning=FALSE,results='asis',fig.keep='all',fig.show='asis',fig.height=2.9,fig.width=5}
# fig.align='center'
ptm <- proc.time()
n_eff_sum<-numeric(6)
for (i in siteInds) {
  # fit model at given site
  data <- site(i)
  stanFit <- stanfit_model(data)

  # plot the fit plot (data and fit overlays), and posteriors of sigmas
  p1 <- fitplot(data, stanFit) + ggtitle(paste("Site", i, ": Beta Fits by Patient"))
  p2 <- postplot(stanFit) + ggtitle("Sigma Posteriors")
  p3 <- traceplot(stanFit,pars=c("sigma_p","sigma_pt","sigma_t","sigma_e"),inc_warmup=TRUE,ncol=1,window=c(0,1000)) + theme(axis.title.x=element_blank(),axis.text.x=element_blank(),axis.ticks.x=element_blank(),axis.text.y=element_text(angle=45),legend.position="none")
  p4 <- ggarrange(p2,p3,nrow=1,ncol=2)
  print(p1)
  print(p4)

  # text output:
  cat(paste("  \nSite ",i,": Parameter posterior summary\n",sep=''))
  # print (or table) mean, median, R_hat, n_eff for each of the 6 params
  datasumm <-summary(stanFit)$summary[1:6,c(1,6,9,10)]
  print(kable(datasumm))
  cat("  \n")
  n_eff_sum <- n_eff_sum + datasumm[,3]
}
mtp <- proc.time() - ptm

```

```{r echo=FALSE,include=TRUE}

print(mtp)
print(paste("Completed run on sites:", paste(siteInds, collapse=" ")))
print(paste("n_eff averages:",paste(round(n_eff_sum/nsites), collapse=" ")))
```

### 2020-09-25 50sites files/50sites_modelPT.Rmd
updated code

---
title: "50sites_modelPT"
author: "Yanlin Ma + Kevin Murgas"
date: "April 4, 2018"
output: word_document
---
Model 5: $$ y_{i,j,k} = \mu + b_j + I_{tumor} * ( \beta_T + d_k) + \epsilon_i$$
$$ b_j \sim N(0, \sigma_P), \ d_k \sim N(0, \sigma_T), \ \epsilon_i \sim N(0,\sigma_E)  $$
```{r echo=FALSE, include=FALSE}
library(rstan)
library(gtools)
library(ggplot2)
library(bayesplot)
```

```{r echo=FALSE, include=FALSE}
# Kevin's working directory
setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")

# Yanlin's working directory
# setwd("D:/DataPlus2017/Data")

# load data
load("myFA.Rdata")

# Extract single site data
site <- function (site_no) {
  # extract CpG site xx to start
  temp <- FullAnnotation[site_no,]

  # here we use all patient samples, excluding glands
  indices <- c(9:13, 15:39,46,47,57,58,72:75)
  patientLabel <- substr(colnames(temp[indices]),1,1)
  patientLabel[10:12] <- "K*"
  sideLabel <- substr(colnames(temp[indices]),2,2)
  tissueLabel <- sideLabel
  tissueLabel[tissueLabel %in% c("A","B")] <- "T"   #replace A and B with T
  tumorIndicator <- 1*(tissueLabel=="T")

  work <- data.frame(beta=logit(t(temp[indices])), patient = patientLabel, tissue=tissueLabel, side=sideLabel, tInd=tumorIndicator)
  colnames(work)[1] <- "beta"
  return(work)
}

# Using Model PT (previously model5): add intra-tumoral variance
stanfitPT <- function (dataset) {
  stanDat <- list(pID = as.integer(factor(dataset$patient)),
                  tInd = dataset$tInd,
                  N = nrow(dataset),
                  P = nlevels(dataset$patient),
                  y = dataset[,1])
  stanFitPT <-
    stan(
      #file = "model_PT.stan",
      fit = emptyFit,
      warmup = 1000,
      iter = 4000,
      chains = 4,
      thin = 1,
      data = stanDat,
      control = list(adapt_delta = 0.999),
      refresh = 0
    )
  return(stanFitPT=stanFitPT)
}

# Plot raw data and fits
fitplot <- function(data,stanFit) {
  # plot beta and fits at this site for all patients, using mean estimates for b coefficients
  p1 <- ggplot() + geom_point(data = data, aes(x=tInd, y=beta, colour=patient))
  pats <- as.integer(factor(data$patient))
  nsamp <- length(pats)
  npats <- max(pats)
  sumTemp <- summary(stanFit)$summary
  mu <- sumTemp[dim(sumTemp)[1]-4,1]
  betaT <- sumTemp[dim(sumTemp)[1]-5,1]
  sampcoefs <- data.frame(pat = pats, b = sumTemp[pats], d = sumTemp[1:nsamp+npats])
  rm(sumTemp)
  # add individual sample estimates
  sampdf <- data.frame(pat = rep(factor(data$patient),2), patg = rep(factor(1:nsamp),2), tInd = c(rep(0,nsamp),rep(1,nsamp)), est = c(sampcoefs$b+mu, (sampcoefs$b+mu+sampcoefs$d+betaT)))
  p1 <- p1 + geom_line(data=sampdf, aes(x=tInd, y=est, colour=pat, group=patg))
  # add overall mean estimate
  meandf <- data.frame(tInd = c(0,1), est = c(mu,mu+betaT))
  p1 <- p1 + geom_point(data=meandf, aes(x=tInd, y=est), size=3, alpha=0.7) + geom_line(data=meandf, aes(x=tInd, y=est), size=2, alpha=0.5)

  return(p1)
}

# Plot posterior distribution
postplot <- function(stanFit) {
  p2 <- mcmc_areas(
    as.array(stanFit), 
    pars = c("sigma_p","sigma_t","sigma_e"),
    prob = 0.8, # 80% intervals
    prob_outer = 0.95, 
    point_est = "median"
  )
  return(p2)
}

## CODE
# Randomly choose 50 sites
set.seed(50)
siteInds <- sample(1:866836,50)
nsites <- length(siteInds)

# check for model_PT.rds, must add to prevent crash
checkFile <- "model_PT.rds"
if (file.exists(checkFile)) {
  file.remove(checkFile)
  print("caught one!") 
}

# create empty fit to use same model compilation throughout
data1<-site(1)
stanDat1 <- list(
  pID = as.integer(factor(data1$patient)),
  tInd = data1$tInd,
  N = nrow(data1),
  P = nlevels(data1$patient),
  y = data1[, 1]
)
emptyFit <- stan(file="model_PT.stan", data = stanDat1, chains = 0)

```

```{r echo=FALSE,include=TRUE,error=FALSE,message=FALSE,warning=FALSE,results='hide',fig.keep='all',fig.show='hold',fig.height=3.7,fig.width=5}
# fig.align='center'
ptm <- proc.time()
for (i in siteInds) {
  #print(paste("Site ",i,sep=''))
  data <- site(i)
  stanFit <- stanfitPT(data)
  p1 <- fitplot(data, stanFit) + ggtitle(paste("Site", i, ": Beta Fits by Patient"))
  p2 <- postplot(stanFit) + ggtitle(paste("Site", i, ": Sigma Posteriors"))
  print(p1)
  print(p2)
}
mtp <- proc.time() - ptm

```

```{r echo=FALSE,include=TRUE}

print(mtp)
print(paste("Completed run on sites:", paste(siteInds, collapse=" ")))

```

### 2020-09-25 Paper_Figures.Rmd
updated code

---
title: "DNA Methylation Hierarchical Variance Model\nPaper Figures"
author: "Kevin Murgas"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: true
    number_section: true
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gt)
library(rlang)
library(ggpubr)
library(knitr)

# set working directory
setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
library(here)

select <- dplyr::select
rename <- dplyr::rename
filter <- dplyr::filter
slice <- dplyr::slice

set.seed(1234)
```

```{r Load_data, echo=FALSE, include=TRUE, message=FALSE}
# Load raw methylation data
load(here("myFA_bulkonly.Rdata"))

stanFile = "FullResultsTCGA_relaxgamma3_postsum3.Rdata"
load(here(stanFile))

prepdir <- "data-stan_rg3ps3"
cat(sprintf("Load prepped data dir: %s (needs to be pre-processed)...",prepdir))
gened_data <- read_csv(here(prepdir, "gened_data.csv"), col_types = list("CHR" = col_character()))
prepped_data <- read_csv(here(prepdir, "prepped_data.csv"))
permed_data <- read_csv(here(prepdir, "permed_data.csv"))

```

\newpage
\section{Figure 1: Model Overview, Priors, Example Site Fit}
\subsection{Data}
Logit-transformed beta values, from EPIC methylation array pre-processed with minfi Noob algorithm. \newline
\newline Sample counts (total): 48 samples from 21 patients. \newline
By tissue: 6 normal samples, 42 paired bulk tumor samples.
\subsection{Model equation}
Let i=sample, j=patient
$$ y_i = \mu + a_j +  \text{tInd}(i)*(\beta_T + b_j + c_i) $$
$$ a_j \sim N(0,\sigma_P) , b_j \sim N(0,\sigma_{PT}) , c_i \sim N(0,\sigma_T) $$
\newline Model is fit at each individual CpG site using RStan's MCMC sampling algorithm. \newline
Stan Parameters: 4 chains x 2000 iterations (200 warmup) = 7200 total draws. adapt_delta = 0.999

\subsection{Prior Probabilities}
Priors for each parameter (mu, betaT, sigmaP/PT/T) were defined using TCGA 450k array data from colorectal tumor patients with paired multiple tumor and/or normal+tumor samples. \newline Empirical values of mean and standard deviations were assesed for all 450k CpG sites. \newline The resulting distributions were fit with: bimodal gaussian (mu), Cauchy (betaT), or gamma distributions (sigmas). \newline
SigmaE (error variance) prior was defined using a non-informative gamma distribution
\newline Additionally: Prior relaxation was performed for the Cauchy and gamma distribtions by algebraically increasing the variance of the distributions while maintaining the same mode (peak)

```{r Priors, echo=FALSE, fig.fullwidth=TRUE, warning=FALSE, comment=NA}
# load TCGA estimates for mu, betaT, sigmaP, sigmaPT, sigmaT
load(file="/Users/kevinmurgas/Documents/Data+ project/TCGA data/estimates_allpats.Rdata")

# remove all zeros, and take square root of variance (stored variable) to get sigma/stdev
sigmaP_ests <- sqrt(sigmaP_ests[which(!!sigmaP_ests)])
sigmaPT_ests <- sqrt(sigmaPT_ests[which(!!sigmaPT_ests)])
sigmaT_ests <- sqrt(sigmaT_ests[which(!!sigmaT_ests)])
mu_ests <- mu_ests[which(!!mu_ests)]
betaT_ests <- betaT_ests[which(!!betaT_ests)]

# for each parameter, plot the distribution and overlay the fits
fBG <- function(x, a, m1, s1, m2, s2) a*dnorm(x,m1,s1) + (1-a)*dnorm(x,m2,s2) # bimodal gaussian
post_prior_histogram <- function(vals,priorfun1,priorfun2,bw,xl,titlestr,legFlag=FALSE) {
  # stacked histogram colored by gene-associated sites
  # and prior curve overlaid
  plot_df <- data.frame(val=vals)
  p1 <- ggplot(plot_df, aes(x=val)) +
    geom_histogram(aes(y=..count../(length(vals)*bw), fill="Empirical"), binwidth=bw) +
    stat_function(fun=function(x) priorfun1(x),geom="line",aes(color="Fit")) +
    stat_function(fun=function(x) priorfun2(x),geom="line",aes(color="3x Relax")) +
    ggtitle(titlestr) + xlim(xl) + theme_classic() + xlab(titlestr) + ylab("Freq.") +
    scale_colour_manual(NULL, values = c("Empirical" = "gray","Fit" ="red","3x Relax" = "blue")) +
    scale_fill_manual(NULL,values = c("Empirical"="gray")) +
    theme(legend.position = c(0.75, 0.65))
  if (!legFlag) {p1 <- p1 + theme(legend.position = "none")}
  p1
}

# plot mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT with priors overlaid
pM <- post_prior_histogram(mu_ests, function(x) fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), function(x) fBG(x,0.3557,-3.0675, sqrt(3)*0.75214,1.1290, sqrt(3)*1.39197), 0.1, c(-5,5), "Mu")
pB <- post_prior_histogram(betaT_ests, function(x) dcauchy(x,-0.04051, 0.19330), function(x) dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "BetaT")
pP <- post_prior_histogram(sigmaP_ests, function(x) dgamma(x, 5.1989, 16.6192), function(x) dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
pPT <- post_prior_histogram(sigmaPT_ests, function(x) dgamma(x, 3.5539, 5.5324), function(x) dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
pT <- post_prior_histogram(sigmaT_ests, function(x) dgamma(x, 6.8383, 25.7560), function(x) dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT",legFlag=TRUE)
pE <- ggplot() + stat_function(fun=function(x) dgamma(x, 2, 2),geom="line",color="blue") + ggtitle("SigmaE") + xlim(c(0,2)) + theme_classic() + theme(legend.position = "none") + xlab("SigmaE") + ylab("Freq.")

pBlank <- ggplot(data=data.frame(len=c(1,2)),aes(x=len)) + theme_void() + geom_bar(aes(fill="Empirical")) +
    stat_function(fun=function(x) 3,geom="line",aes(color="1x variance")) +
    stat_function(fun=function(x) 4,geom="line",aes(color="3x variance")) + xlim(0,1) + ylim(0,1) + scale_colour_manual(NULL, values = c("Empirical" = "gray","1x variance" ="red","3x variance" = "blue")) + scale_fill_manual(NULL,values = "gray") + theme(legend.position = c(2, 0.75))

p_all <- ggarrange(pM,pB,pE,pP,pPT,pT,hjust=0,nrow=2,ncol=3)
print(p_all)
rm(pM,pB,pBlank,pP,pPT,pT,pE)

```

\subsection{CpG Conservation score}
We want to understand which sites of DNA methylation are fundamentally conserved within tumors, i.e. lower variation in methlyation within tumor (sigmaT) relative to the variation within normal healthy tissue (sigmaP). Therefore we choose the log-ratio of sigmaP/sigmaT to represent a conservation score, which is positive when tumor variation is lower than normal variation. We take the median of the posterior distribution for this parameter.

$$\text{score}_\text{med} = \text{Median} (\text{log}\frac{\sigma_{P}}{\sigma_T}) $$

\newpage
\subsection{Example site fit}
![Fit](/Users/kevinmurgas/Documents/Data+ project/Paper Figures/example_fit.png)
\newline \newline
![Posteriors](/Users/kevinmurgas/Documents/Data+ project/Paper Figures/example_posteriors.png)
\newpage
\section{Figure 2: Model Fit Results}
\subsection{Parameter Distributions}
Histograms are median values of posterior distribution
```{r Params_site, echo=FALSE, fig.fullwidth=TRUE, warning=FALSE, comment=NA}
genemask = !(FullAnnotation$UCSC_RefGene_Name=="")
cat(sprintf("N sites: total=%i, gene-associated=%i, non-gene=%i  \n",length(genemask),sum(genemask),length(genemask)-sum(genemask)))

fBG <- function(x, a, m1, s1, m2, s2) a*dnorm(x,m1,s1) + (1-a)*dnorm(x,m2,s2) # bimodal gaussian
post_prior_histogram <- function(postvals,mask,priorfun,bw,xl,titlestr) {
  # stacked histogram colored by gene-associated sites
  # and prior curve overlaid
  plot_df <- data.frame(val=postvals, gene_assoc=mask)
  p1 <- ggplot(plot_df, aes(x=val)) +
    geom_histogram(aes(y=..count../length(mask),fill=gene_assoc), binwidth=bw) +
    stat_function(fun=function(x) bw*priorfun(x),geom="line",color="black") +
    ggtitle(titlestr) + xlim(xl) + theme_classic() + theme(legend.position = "none") + xlab(titlestr) + ylab("Freq.")
  p1
}

# plot mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT with priors overlaid
pM <- post_prior_histogram(mu_full$p50, genemask, function(x) 1.5*fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), 0.1, c(-5,5), "Mu")
pB <- post_prior_histogram(betaT_full$p50, genemask, function(x) 1.7*dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "BetaT")
pE <- post_prior_histogram(sigmaE_full$p50, genemask, function(x) 4*dgamma(x,2,2), 0.05, c(0,2), "SigmaE")
pP <- post_prior_histogram(sigmaP_full$p50, genemask, function(x) 2*dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
pPT <- post_prior_histogram(sigmaPT_full$p50, genemask, function(x) 1.8*dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
pT <- post_prior_histogram(sigmaT_full$p50, genemask, function(x) 1.2*dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT")
# log sigmas
pLP <- post_prior_histogram(log(sigmaP_full$p50), genemask, function(x) 0.52*dgamma(exp(x), 2.7693, 7.0029), 0.1, c(-5,5), "log(SigmaP)")
pLPT <- post_prior_histogram(log(sigmaPT_full$p50), genemask, function(x) 0.7*dgamma(exp(x), 2.1457, 2.4819), 0.1, c(-5,5), "log(SigmaPT)")
pLT <- post_prior_histogram(log(sigmaT_full$p50), genemask, function(x) 0.4*dgamma(exp(x), 3.3643, 10.430), 0.1, c(-5,5), "log(SigmaT)") + theme(legend.position = c(0.8, 0.6), legend.text=element_text(size=6), legend.title=element_text(size=6))

# arrange each parameter as a subplot # ,labels=c("M","B","E","P","PT","T","lP","lPT","lT")
p_all <- ggarrange(pM,pB,pE,pP,pPT,pT,pLP,pLPT,pLT,hjust=0,nrow=3,ncol=3)
print(p_all)
rm(pM,pB,pE,pP,pPT,pT,pLP,pLPT,pLT)
```

\subsection{Conservation score}
```{r Fig1B, echo=FALSE, fig.height=2, fig.width=6.5, fig.fullwidth=TRUE, warning=FALSE}

post_histogram <- function(postvals,mask,bw,xl,titlestr) {
  # stacked histogram colored by gene-associated sites
  # and prior curve overlaid
  plot_df <- data.frame(val=postvals, gene_assoc=mask)
  p1 <- ggplot(plot_df, aes(x=val)) +
    geom_histogram(aes(y=..count../length(mask),fill=gene_assoc), binwidth=bw) +
    ggtitle(titlestr) + xlim(xl) + theme_classic() + theme(legend.position = "none") + xlab(titlestr) + ylab("Freq.")
}

# plot the two conservations score prob sums
pP1 <- post_histogram(prob1_full$median,genemask,0.05,c(-3,3),expression("Median log("~sigma[P]/sigma[T]~")"))
#prob1_proxy = log(sigmaP_full$p50/sigmaT_full$p50)
#pP2 <- post_histogram(prob1_proxy,genemask,0.05,c(-3,3),expression(" log( median" ~sigma[P]/~"median"~sigma[T]~")"))
p_scores <- pP1
print(p_scores)
```

\newpage
\subsection{Summary of neff and Rhat for each parameter}
```{r neffRhat, echo=FALSE, warning=FALSE}
# for each parameter (mu, betaT, sigmaP, sigmaPT, sigmaT, sigmaE, lp__)
# reporting: mean n_eff, min n_eff, # n_eff<500, %, # n_eff<100, %
n_eff_mat <-
  data.frame(
  mu = mu_full$n_eff,
  betaT = betaT_full$n_eff,
  sigmaP = sigmaP_full$n_eff,
  sigmaPT = sigmaPT_full$n_eff,
  sigmaT = sigmaT_full$n_eff,
  sigmaE = sigmaE_full$n_eff,
  lp = lp_full$n_eff
  ) %>% gather(param,value,mu:lp) %>% group_by(param) %>% summarize(
  mean = mean(value),
  min = min(value),
  s500 = sum(value < 500),
  p500 = 100 * mean(value < 500),
  s100 = sum(value < 100),
  p100 = 100 * mean(value < 100),
  .groups = "drop"
  ) %>% arrange(match(.$param,c("mu","betaT","sigmaP","sigmaPT","sigmaT","sigmaE","lp")))

kable(n_eff_mat, col.names=c("Parameter", "Mean n_eff", "Min n_eff", "# <500", "% <500", "# <100", "% <100"), digits = 2)

# again for Rhat
Rhat_mat <-
  data.frame(
  mu = mu_full$Rhat,
  betaT = betaT_full$Rhat,
  sigmaP = sigmaP_full$Rhat,
  sigmaPT = sigmaPT_full$Rhat,
  sigmaT = sigmaT_full$Rhat,
  sigmaE = sigmaE_full$Rhat,
  lp = lp_full$Rhat
  ) %>% gather(param,value,mu:lp) %>% group_by(param) %>% summarize(
  mean = mean(value),
  max = max(value),
  s11 = sum(value > 1.1),
  p11 = 100 * mean(value > 1.1),
  s101 = sum(value > 1.01),
  p101 = 100 * mean(value > 1.01),
  .groups = "drop"
  ) %>% arrange(match(.$param,c("mu","betaT","sigmaP","sigmaPT","sigmaT","sigmaE","lp")))

kable(Rhat_mat, col.names=c("Parameter", "Mean Rhat", "Max Rhat", "# >1.1", "% >1.1", "# >1.01", "% >1.01"), digits = 2)
cat("  \n")

# indices to filter out sites Rhat>1,1
QCinds <- lp_full$Rhat > 1.1
```

For the remainder of these results, a filter is applied to remove all sites with Rhat > 1.1 in lp (log-posterior likelihood of entire site fit). Another filter is later applied to remove all genes with <5 CpG sites.

\newpage
\subsection{Intra-gene conservation}
We examine intra gene conservation 2 ways: a table looking at gene regulatory regions, and a CpG-distance analysis

\subsection{Regulatory regions}
Table comparing different groups of CpG sites:
```{r Table1, echo=FALSE, warning=FALSE}

#, include=TRUE, results='asis', fig.keep='all', fig.show='asis', fig.height=3, fig.width=5

prepped_data %>%
  group_by(`Island relation` = Relation_to_UCSC_CpG_Island) %>%
  summarise_at(vars(sigmaP:sigmaT,score_median), mean, na.rm = TRUE) %>%
  dplyr::rename(ir = `Island relation`) %>%
  mutate(ir = case_when(is.na(ir) ~ "Sea",
                        str_detect(ir, "^S") ~ str_replace(ir, "^S_", "South "),
                        str_detect(ir, "^N") ~ str_replace(ir, "^N_", "North "),
                        TRUE ~ ir)) %>%
  dplyr::rename(`Island relation` = ir) %>%
  arrange(c(1,4,2,5,3,6)) %>% # for island, N shore, S shore, N shelf, S shelf, sea
  rename(`CpG Conservation Score`=score_median) %>%
  gt() %>%
  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, `CpG Conservation Score`), decimals = 3) %>%
  tab_spanner(label = "Hierarchical Variances", columns = vars(sigmaP, sigmaPT, sigmaT)) %>% cols_align("center")

summarize_conservation <- function(data,str) {
  indsin <- str_detect(data$UCSC_RefGene_Group, str) & !is.na(data$UCSC_RefGene_Group)
  indsout <- !str_detect(data$UCSC_RefGene_Group, str) | is.na(data$UCSC_RefGene_Group)
  if (str=="any") {
    indsin <- !is.na(data$UCSC_RefGene_Group)
    indsout <- is.na(data$UCSC_RefGene_Group)
  }
  data.frame(cons_in=mean(data$score_median[indsin], na.rm=TRUE),
         cons_out=mean(data$score_median[indsout],na.rm=TRUE),
         nsites=sum(indsin),
         pval=wilcox.test(data$score_median[indsin],data$score_median[indsout])$p.value)
}

bind_rows({prepped_data %>% summarize_conservation("any") %>% mutate(region="Gene-Associated")},
          {prepped_data %>% summarize_conservation("TSS1500") %>% mutate(region="TSS1500")},
          {prepped_data %>% summarize_conservation("TSS200") %>% mutate(region="TSS200")},
          {prepped_data %>% summarize_conservation("5'UTR|5UTR|5URT") %>% mutate(region="5'UTR")},
          {prepped_data %>% summarize_conservation("1stExon") %>% mutate(region="1stExon")},
          {prepped_data %>% summarize_conservation("Body") %>% mutate(region="Body")},
          {prepped_data %>% summarize_conservation("3'UTR|5UTR") %>% mutate(region="3'UTR")}) %>%
  select(c(region,cons_in,cons_out,nsites,pval)) %>%
  gt() %>%
  fmt_number(columns = vars(cons_in,cons_out,pval), decimals = 3)
```

\newpage
\subsection{Figure 1: CpG Distance Analysis}
Corrected for gene direction, based on Ensembl gene database \newline
```{r Fig1, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=3, comment=NA}
# load ensembl db
library(EnsDb.Hsapiens.v75) # ensembl annotation with gene position (hg19 version)
edb <- EnsDb.Hsapiens.v75
geneKeys <- keys(edb, keytype="GENENAME") %>% intersect(unique(gened_data$UCSC_RefGene_Name))
edbData<- AnnotationDbi::select(edb, keys=geneKeys, keytype="GENENAME",
                                columns=c("GENEBIOTYPE", "GENESEQSTART", "GENESEQEND", "SEQNAME", "SEQSTRAND")) %>%
  filter(SEQNAME %in% c(as.character(1:22),"X","Y")) %>%
  mutate(GENESTART = GENESEQSTART*(SEQSTRAND==1) + GENESEQEND*(SEQSTRAND==-1)) %>%
  select(-c(GENESEQSTART,GENESEQEND))

# calculate CpG distance based on Ensembl gene start position and direction
distanced_data <- gened_data %>%
  filter(!is.na(UCSC_RefGene_Name)) %>%
  filter(UCSC_RefGene_Name %in% geneKeys) %>%
  mutate(EDBCHR=edbData$SEQNAME[match(UCSC_RefGene_Name,edbData$GENENAME)],
         EDBSTART=edbData$GENESTART[match(UCSC_RefGene_Name,edbData$GENENAME)],
         EDBDIR=edbData$SEQSTRAND[match(UCSC_RefGene_Name,edbData$GENENAME)],
         EDBTYPE=edbData$GENEBIOTYPE[match(UCSC_RefGene_Name,edbData$GENENAME)]
         ) %>%
  filter(CHR==EDBCHR) %>%
  mutate(cpgdist = if_else(EDBDIR==1, MAPINFO-EDBSTART, EDBSTART-MAPINFO))

distance_length <- distanced_data %>%
  filter(!is.na(UCSC_RefGene_Name)) %>%
  select(CHR, UCSC_RefGene_Name, cpgdist, c(sigmaP:sigmaT,score_median:score_reg)) %>%
  mutate(location_bin = cut(cpgdist, c(seq(-1500, 20000, by = 100)), include.lowest = TRUE)) %>%
  group_by(location_bin) %>%
  summarise(mean_sigmaP = mean(sigmaP), mean_sigmaPT = mean(sigmaPT), mean_sigmaT = mean(sigmaT),  mean_score_median = mean(score_median), mean_score_median2 = mean(score_median2), n_obs = n(), .groups='drop') %>%
  mutate(location_bin = as.numeric(str_extract(location_bin, "(?<=,)(.*)(?=])")))

distance_length %>%
  drop_na() %>%
  pivot_longer(mean_sigmaP:mean_sigmaT) %>% 
  mutate(name = factor(name, c("mean_sigmaP","mean_sigmaPT","mean_sigmaT"), c("SigmaP","SigmaPT","SigmaT"))) %>%
  sample_frac() %>%
  ggplot(aes(location_bin, value, color = name, fill = name)) +
  geom_point(key_glyph = draw_key_rect) +
  ggforce::facet_zoom(location_bin < 5000) +
  theme_light() +
  theme(legend.position = "bottom") +
  labs(title = NULL,
       fill = NULL,
       color = NULL,
       x = "Distance from gene start site (binned to 100 bp)",
       y = "Average Sigma Value", 
       caption = "Figure 1: Average Sigma values of single CpGs as a function of position relative to gene.") +
  scale_color_brewer(palette = "Set2") +
  scale_fill_brewer(palette = "Set2")
cat("\n")

# again for conservation scores
distance_length %>%
  drop_na() %>%
  pivot_longer(mean_score_median) %>% 
  mutate(name = factor(name, c("mean_score_median"), c("Score_median"))) %>%
  sample_frac() %>%
  ggplot(aes(location_bin, value, color = name, fill = name)) +
  geom_point(key_glyph = draw_key_rect) +
  ggforce::facet_zoom(location_bin < 5000) +
  theme_light() +
  theme(legend.position = "bottom") +
  labs(title = NULL,
       fill = NULL,
       color = NULL,
       x = "Distance from gene start site (binned to 100 bp)",
       y = "Average Conservation Score", 
       caption = "Figure 1: Average conservation scores of single CpGs as a function of position relative to gene.") +
  scale_color_brewer(palette = "Set2") +
  scale_fill_brewer(palette = "Set2")
cat("\n")

# and look at number of CpG in each bin
distance_length %>%
  drop_na() %>%
  pivot_longer(n_obs) %>%
  mutate(name = factor(name, c("n_obs"), c("n_obs"))) %>%
  sample_frac() %>%
  ggplot(aes(location_bin, value, color = name, fill = name)) +
  geom_point(key_glyph = draw_key_rect) +
  ggforce::facet_zoom(location_bin < 5000) +
  theme_light() +
  theme(legend.position = "bottom") +
  labs(title = NULL,
       fill = NULL,
       color = NULL,
       x = "Distance from gene start site (binned to 100 bp)",
       y = "Number of CpGs", 
       caption = "Figure 1: Frequency of single CpGs as a function of position relative to gene start site.") +
  scale_color_brewer(palette = "Set2") +
  scale_fill_brewer(palette = "Set2")
```

\newpage
\section{Gene Conservation Scores}
Genes are scored by average of all CpG sites within gene
\subsection{Example Genes: APC, TP53, TTN, HLA-A, GAPDH}
```{r examplegenes, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=2, comment=NA}
# for these genes, show the CpG site distribution, and add vertical lines for the CpG sites in the gene (gened_data?), then a bold yellow vertical line for the overall mean

plotGeneScore <- function(geneStr) {
  CpGscores <- gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^",geneStr,"$"))) %>% pull(score_median)

  ggplot(prepped_data) + geom_histogram(aes(x=score_median, y=..count../nrow(prepped_data)), binwidth=0.05) + xlim(-3,3) + ggtitle(sprintf("%s: CpG scores (%i sites)",geneStr, length(CpGscores))) + xlab("CpG conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") + geom_vline(xintercept=CpGscores, color="red", alpha=0.5, size=0.2) + geom_vline(xintercept=mean(CpGscores), color="blue", alpha=0.8, size=2)
}

cat("\n")
plotGeneScore("TP53")
cat("\n")
plotGeneScore("TTN")
cat("\n")
plotGeneScore("HLA-A")
```

\subsection{COSMIC Cancer Genes}
```{r cancergenes, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=2, comment=NA}
# make gene-wise averages
gene_summarize <- function(data, id, value, fun = mean, ...) {
  grouped_data <- data %>% group_by({{id}})

  bind_cols(
    summarise_at(grouped_data, dplyr::vars({{value}}), fun, ...),
    count(grouped_data) %>% dplyr::ungroup() %>% dplyr::select(.data$n)
  )
}

gene_summary <- gened_data %>%
  filter(!is.na(UCSC_RefGene_Name)) %>%
  gene_summarize(UCSC_RefGene_Name, c(score_median))

cancergenes <- read_csv("/Users/kevinmurgas/Documents/Data+ project/Cancer gene lists/SangerCensus_COSMIC_CRC.csv")
genescores <- gene_summary %>% filter(UCSC_RefGene_Name %in% (cancergenes %>% pull(`Gene Symbol`))) %>% pull(score_median)
ggplot(gene_summary) + geom_histogram(aes(x=score_median, y=..count../nrow(gene_summary)), binwidth=0.05) + xlim(-3,3) + ggtitle(sprintf("Gene Scores of %i CRC Genes", length(genescores))) + xlab("Gene conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") + geom_vline(xintercept=genescores, color="blue", alpha=0.5, size=0.2)
cat("\n")

# source 2
cancergenes <- read_csv("/Users/kevinmurgas/Documents/Data+ project/Cancer gene lists/CRCgenes_atlasgeneticsoncology.csv")
genescores <- gene_summary %>% filter(UCSC_RefGene_Name %in% (cancergenes %>% pull(CancerGenes))) %>% pull(score_median)
ggplot(gene_summary) + geom_histogram(aes(x=score_median, y=..count../nrow(gene_summary)), binwidth=0.05) + xlim(-3,3) + ggtitle(sprintf("Gene Scores of %i CRC Genes", length(genescores))) + xlab("Gene conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") + geom_vline(xintercept=genescores, color="blue", alpha=0.1, size=0.2)
```

\newpage
\section{Gene Set Conservation Analysis}
```{r load_gene_sets, include=FALSE}
# load in the C5 gene sets from MSigDB
library(GSA)
gmtfile <- file.path("/Users/kevinmurgas/Documents/Data+ project/EPIC data","c2.cp.v7.1.symbols.gmt")
genesets <- GSA.read.gmt(gmtfile)
```

```{r GSCA, echo=FALSE, comment=NA, fig.width=8}
# make a data.frame for gene sets and scores
nsets <- length(genesets[[1]])
GSEA_df <- data.frame(GeneSet=genesets[[2]][1:nsets], score=numeric(nsets),
                      num_gene = unlist(lapply(genesets[[1]][1:nsets],length)),
                      num_gene_cut = numeric(nsets))

# have to order all genes first by rank of conservation (conserved genes come first)
genenames = permed_data$UCSC_RefGene_Name
ngene = length(genenames)

# reorder by evidence rank
ind = order(-permed_data$score_median) # use ranked conservation score as evidence
evidence = seq(ngene,1,-1)
genenames = genenames[ind]

# plot evidence ranked: permed_data$score_median
ranks <- permed_data$score_median
names(ranks) <- permed_data$UCSC_RefGene_Name
barplot(sort(ranks, decreasing = T))

ptm <- proc.time()
for (iset in 1:nsets) {
  # take only genes that we have in full gene list
  set <- unlist(genesets[[1]][iset]) %>% intersect(genenames)
  GSEA_df$num_gene_cut[iset] = length(set)

  # skip if num_gene <10 or >500
  if (length(set)<10 | length(set)>500) {next}

  # compute running sum (as PDF first)
  runsum_in = numeric(ngene)
  runsum_out = numeric(ngene)
  for (i in 1:ngene) {
    if (genenames[i] %in% set) {
      # evidence = permed_data$median, or ranks
      runsum_in[i] = evidence[i]
      runsum_out[i] = 0
    } else {
      runsum_in[i] = 0
      runsum_out[i] = 1
    }
  }
  # convert to CDFs
  runsum_in <- cumsum(runsum_in)/sum(runsum_in)
  runsum_out <- cumsum(runsum_out)/sum(runsum_out)

  # compute K-S-like statistic
  v <- runsum_in - runsum_out
  score <- max(c(0,v)) + min(c(0,v))
  GSEA_df$score[iset] <- score
}
cat(proc.time()-ptm)
cat(sprintf("Processed %i genes, %i pathways. Top 50 paths:",ngene,nsets))

# insert significance bootstrapping

# make table of top scores
GSEA_df %>% arrange(-score) %>%
  select(score,num_gene,GeneSet) %>%
  slice(1:50) %>%
  mutate(GeneSet = gsub('(.{1,90})(\\s|$)', '\\1\n', GeneSet)) %>%
  gt()
```

```{r Exit, include=FALSE}
knit_exit()
```

``` {r enrich, include=FALSE}
# expression data
# downloaded file: E-MTAB-2836-query-results.tpms.tsv
# source: https://www.ebi.ac.uk/gxa/experiments/E-MTAB-2836/Results
expression_atlas <- read_tsv(here("data-raw", "E-MTAB-2836-query-results.tpms.tsv"), skip = 4, 
                             col_types = cols(
                               .default = col_double(),
                               `Gene ID` = col_character(),
                               `Gene Name` = col_character()
                             ))

# Gene Set Enrichment
tidy_gsea <- function(genes) {
  AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, genes, 'ENTREZID', 'SYMBOL') %>%
    ReactomePA::enrichPathway()
}

enrich_sigmaP <- permed_data %>%
  #filter(sigmaP_v3 < quantile(sigmaP_v3, 0.05, na.rm = TRUE)) %>%
  filter(sigmaP_v3 < 0.05) %>%
  pull(UCSC_RefGene_Name) %>% tidy_gsea()
enrich_sigmaPT <- permed_data %>%
  #filter(sigmaPT_v3 < quantile(sigmaPT_v3, 0.05, na.rm = TRUE)) %>%
  filter(sigmaPT_v3 < 0.05) %>%
  pull(UCSC_RefGene_Name) %>% tidy_gsea()
enrich_sigmaT <- permed_data %>%
  #filter(sigmaT_v3 < quantile(sigmaT_v3, 0.05, na.rm = TRUE)) %>%
  filter(sigmaT_v3 < 0.05) %>%
  pull(UCSC_RefGene_Name) %>% tidy_gsea()
#enrich_score_prob <- permed_data %>%
#  filter(score_prob_v3 > quantile(score_prob_v3, 0.95, na.rm = TRUE)) %>%
#  pull(UCSC_RefGene_Name) %>% tidy_gsea()
#enrich_score_mean <- permed_data %>%
#  filter(score_mean_v3 > quantile(score_mean_v3, 0.95, na.rm = TRUE)) %>%
#  pull(UCSC_RefGene_Name) %>% tidy_gsea()
enrich_score_median <- permed_data %>%
  #filter(score_median_v3 > quantile(score_median_v3, 0.95, na.rm = TRUE)) %>%
  filter(score_median_v3 > 0.95) %>%
  pull(UCSC_RefGene_Name) %>% tidy_gsea()
enrich_score_median2 <- permed_data %>%
  #filter(score_median2_v3 > quantile(score_median2_v3, 0.95, na.rm = TRUE)) %>%
  filter(score_median2_v3 > 0.95) %>%
  pull(UCSC_RefGene_Name) %>% tidy_gsea()
enrich_score_reg <- permed_data %>%
  #filter(score_reg_v3 > quantile(score_reg_v3, 0.95, na.rm = TRUE)) %>%
  filter(score_reg_v3 > 0.95) %>%
  pull(UCSC_RefGene_Name) %>% tidy_gsea()
```

\newpage
\subsection{Gene-average Parameter Distributions}
```{r Params_gene, echo=FALSE, fig.fullwidth=TRUE, warning=FALSE, comment=NA}
# make gene-wise averages
gene_summarize <- function(data, id, value, fun = mean, ...) {
  grouped_data <- data %>% group_by({{id}})

  bind_cols(
    summarise_at(grouped_data, dplyr::vars({{value}}), fun, ...),
    count(grouped_data) %>% dplyr::ungroup() %>% dplyr::select(.data$n)
  )
}

cat("Summarize by gene, taking mean value of all sites in gene")
ptm<-proc.time()
gene_summary <- gened_data %>%
  filter(!is.na(UCSC_RefGene_Name)) %>%
  gene_summarize(UCSC_RefGene_Name, c(mu:sigmaT,score_median:score_reg))
cat(proc.time()-ptm)

post_histogram <- function(postvals,bw,xl,titlestr) {
  # stacked histogram colored by gene averages
  # and prior curve overlaid
  plot_df <- data.frame(val=postvals)
  p1 <- ggplot(plot_df, aes(x=val)) +
    geom_histogram(aes(y=..density..), binwidth=bw) +
    ggtitle(titlestr) + xlim(xl) + theme_classic() + xlab(titlestr) + ylab("Freq.")
  p1
}

# plot: n_sites, mu, betaT, sigmaP, sigmaPT, sigmaT
n_cutlo = 5
n_cuthi = quantile(gene_summary$n, 0.99)
pN <- post_histogram(gene_summary$n,1,c(0,200),paste0("N sites (max = ",max(gene_summary$n),")")) + geom_vline(xintercept=n_cutlo-0.5,linetype = "dashed",color="red") + annotate(geom="text",x=80,y=0.03,size=2,label=paste0("# genes <",n_cutlo," sites:\n", sum(gene_summary$n<n_cutlo)," / ",nrow(gene_summary))) +
geom_vline(xintercept=n_cuthi,linetype = "dashed",color="red") +annotate(geom="text",x=120,y=0.013,size=2,label=sprintf("# genes >%0.1f sites:\n %i / %i (99th pctile)", n_cuthi,sum(gene_summary$n>n_cuthi),nrow(gene_summary)))
gene_summary <- gene_summary %>% filter(n>=n_cutlo) %>% filter(n<n_cuthi)
pM <- post_histogram(gene_summary$mu,0.1,c(-5,5),"Mu")
pB <- post_histogram(gene_summary$betaT,0.1,c(-5,5),"BetaT")
pP <- post_histogram(gene_summary$sigmaP,0.05,c(0,2),"SigmaP")
pPT <- post_histogram(gene_summary$sigmaPT,0.05,c(0,2),"SigmaPT")
pT <- post_histogram(gene_summary$sigmaT,0.05,c(0,2),"SigmaT")
pP1 <- post_histogram(gene_summary$score_median,0.05,c(-3,3),"Score_median")
pP2 <- post_histogram(gene_summary$score_median2,0.05,c(-3,3),"Score_median2")
pP3 <- post_histogram(gene_summary$score_reg,0.05,c(-3,3),"Score_reg")

#pP1 <- post_histogram(gene_summary$score_prob,0.01,c(0,1),"Score_prob")
#pP2 <- post_histogram(gene_summary$score_mean,0.05,c(-3,3),"Score_mean")

# arrange each parameter as a subplot
p_all <- ggarrange(pN,pM,pB,pP,pPT,pT,pP1,pP2,pP3,hjust=0,nrow=3,ncol=3)
print(p_all)
rm(pN,pM,pB,pP,pPT,pT,pP1,pP2,pP3)

```

\newpage
\subsection{Gene-average parameters vs N sites}
```{r Params_vsN, echo=FALSE, fig.fullwidth=TRUE, warning=FALSE, comment=NA, dev = "png"}
pM <- ggplot(gene_summary) + geom_point(aes(x=n,y=mu), alpha=0.01) + ggtitle("Mu vs N-sites")
pB <- ggplot(gene_summary) + geom_point(aes(x=n,y=betaT), alpha=0.01) + ggtitle("BetaT vs N-sites")
pE <- ggplot(gene_summary) + geom_point(aes(x=n,y=sigmaE), alpha=0.01) + ggtitle("SigmaE vs N-sites")
pP <- ggplot(gene_summary) + geom_point(aes(x=n,y=sigmaP), alpha=0.01) + ggtitle("SigmaP vs N-sites")
pPT <- ggplot(gene_summary) + geom_point(aes(x=n,y=sigmaPT), alpha=0.01) + ggtitle("SigmaPT vs N-sites")
pT <- ggplot(gene_summary) + geom_point(aes(x=n,y=sigmaT), alpha=0.01) + ggtitle("SigmaT vs N-sites")

#pP1 <- ggplot(gene_summary) + geom_point(aes(x=n,y=score_prob), alpha=0.01) + ggtitle("Score_prob vs N-sites")
#pP2 <- ggplot(gene_summary) + geom_point(aes(x=n,y=score_mean), alpha=0.01) + ggtitle("Score_mean vs N-sites")
pP1 <- ggplot(gene_summary) + geom_point(aes(x=n,y=score_median), alpha=0.01) + ggtitle("Score_median vs N-sites")
pP2 <- ggplot(gene_summary) + geom_point(aes(x=n,y=score_median2), alpha=0.01) + ggtitle("Score_median2 vs N-sites")
pP3 <- ggplot(gene_summary) + geom_point(aes(x=n,y=score_reg), alpha=0.01) + ggtitle("Score_reg vs N-sites")

# arrange each parameter as a subplot
p_all <- ggarrange(pM,pB,pE,pP,pPT,pT,pP1,pP2,pP3,hjust=0,nrow=3,ncol=3)
print(p_all)
rm(pM,pB,pE,pP,pPT,pT,pP1,pP2,pP3)
```

\subsection{Figure 2: Bootstrapping}
```{r Fig2, echo=FALSE, warning=FALSE, fig.height=4, comment=NA}
# Figure 2
permed_data %>%
  pivot_longer(dplyr::matches("1|3")) %>%
  separate(name, into = c("type", "method"), sep = "_v") %>%
  mutate(method = factor(method, c("1", "3"), c("Naive bootstrap", "Adjusted bootstrap")),
         type = factor(type, c("sigmaP","sigmaPT","sigmaT"), c("SigmaP","SigmaPT","SigmaT"))) %>%
  drop_na() %>%
  ggplot(aes(value, fill = type)) +
  geom_histogram(bins = 50, color = "grey30") +
  facet_grid(method~type, scales = "free") +
  theme_minimal() +
  labs(x = "p-value",
       y = "Number of genes",
       title = NULL,
       caption = "Figure 2A: Distribution of boot-strapping p-values for genes as a function of sigma level and bootstrap method.")  +
  theme(legend.position = "none") +
  scale_color_brewer(palette = "Set2") +
  scale_fill_brewer(palette = "Set2")

cat("***")
permed_data %>%
  pivot_longer(dplyr::matches("1|3")) %>%
  separate(name, into = c("type", "method"), sep = "_v") %>%
  mutate(method = factor(method, c("1", "3"), c("Naive bootstrap", "Adjusted bootstrap")),
         type = factor(type, c("score_median","score_median2","score_reg"), c("Score_median","Score_median2","Score_reg"))) %>%
  drop_na() %>%
  ggplot(aes(value, fill = type)) +
  geom_histogram(bins = 50, color = "grey30") +
  facet_grid(method~type, scales = "free") +
  theme_minimal() +
  labs(x = "p-value",
       y = "Number of genes",
       title = NULL,
       caption = "Figure 2B: Distribution of boot-strapping p-values for genes as a function of conservation score and bootstrap method.")  +
  theme(legend.position = "none") +
  scale_color_brewer(palette = "Set2") +
  scale_fill_brewer(palette = "Set2")
```

\newpage
\subsection{Figure 3: Reactome Pathway Enrichment Analysis}
Enrichment defined from the group of genes <5th percentile of each sigma, or >95th percentile of conservation score.
```{r Fig3, echo=FALSE, fig.height=3.5, comment=NA}
cat("Number of enrichment pathways (sigmas):")
c(sigmaP = length(enrich_sigmaP$Description),
  sigmaPT = length(enrich_sigmaPT$Description),
  sigmaT = length(enrich_sigmaT$Description)) %>%
  sort()

list(
  sigmaP = enrich_sigmaP$Description,
  sigmaPT = enrich_sigmaPT$Description,
  sigmaT = enrich_sigmaT$Description
) %>%
  UpSetR::fromList() %>%
  UpSetR::upset(sets.bar.color = RColorBrewer::brewer.pal(3, "Set2")[c(3, 2, 1)])
grid::grid.text("Figure 3A: Frequency of overlap between pathways that are called \nas most conserved for each hierarchical variance.", x=0.65, y=0.05, gp=grid::gpar(fontsize=10))

cat("Number of enrichment pathways (probs):")
c(score_median = length(enrich_score_median$Description),
  score_median2 = length(enrich_score_median2$Description),
  score_reg = length(enrich_score_reg$Description)) %>%
  sort()

list(
  score_median = enrich_score_median$Description,
  score_median2 = enrich_score_median2$Description,
  score_reg = enrich_score_reg$Description
) %>%
  UpSetR::fromList() %>%
  UpSetR::upset(sets.bar.color = RColorBrewer::brewer.pal(3, "Set2")[c( 2, 1)])
grid::grid.text("Figure 3B: Frequency of overlap between pathways that are called \nas most conserved for each conservation score.", x=0.65, y=0.05, gp=grid::gpar(fontsize=10))
```

\newpage
\subsection{Figure 4: Pathway Enrichment Network}
```{r Fig4, echo=FALSE, message=FALSE, fig.height=2.5}
# Figure 4
cat("sigmaP:")
ggg <- enrich_sigmaP %>%
  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
  labs(caption = "Figure 4A: Relationship between pathways that are called as conserved based on sigmaP.", title = NULL)

ggg$layers <- ggg$layers[1:2]

ggg + coord_flip() + 
  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
  scale_color_gradientn(colors = "black") +
  guides(color = "none", size = "none")

#ggsave(filename = here("figures", "figure-4.png"), width = 7, dpi = 320, height = 4.326)

# sigmaPT
cat("sigmaPT:")
ggg <- enrich_sigmaPT %>%
  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
  labs(caption = "Figure 4B: Relationship between pathways that are called as conserved based on sigmaPT.", title = NULL)

ggg$layers <- ggg$layers[1:2]
ggg + coord_flip() + 
  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
  scale_color_gradientn(colors = "black") +
  guides(color = "none", size = "none")

# sigmaT
cat("sigmaT:")
ggg <- enrich_sigmaT %>%
  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
  labs(caption = "Figure 4C: Relationship between pathways that are called as conserved based on sigmaT.", title = NULL)

ggg$layers <- ggg$layers[1:2]
ggg + coord_flip() + 
  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
  scale_color_gradientn(colors = "black") +
  guides(color = "none", size = "none")
```

\newpage
```{r Fig4DEF, echo=FALSE, message=FALSE, fig.height=2.5}
# score_median
cat("score_median")
ggg <- enrich_score_median %>%
  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
  labs(caption = "Figure 4D: Relationship between pathways that are called as conserved based on score_median.", title = NULL)
ggg$layers <- ggg$layers[1:2]
ggg + coord_flip() + 
  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
  scale_color_gradientn(colors = "black") +
  guides(color = "none", size = "none")

# score_median2
cat("score_median2")
ggg <- enrich_score_median2 %>%
  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
  labs(caption = "Figure 4E: Relationship between pathways that are called as conserved based on score_median2.", title = NULL)
ggg$layers <- ggg$layers[1:2]
ggg + coord_flip() + 
  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
  scale_color_gradientn(colors = "black") +
  guides(color = "none", size = "none")

# score_reg
cat("score_reg")
#ggg <- enrich_score_reg %>%
#  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
#  labs(caption = "Figure 4G: Relationship between pathways that are called as conserved based on score_reg.", title = NULL)
#ggg$layers <- ggg$layers[1:2]
#ggg + coord_flip() +
#  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
#  scale_color_gradientn(colors = "black") +
#  guides(color = "none", size = "none")
```

\subsection{Figure 5: Gene Conservation vs Gene Expression}
```{r Fig5, echo=FALSE, message=FALSE}
# Figure 5A
figure5_data <- permed_data %>% 
  left_join(by = c("UCSC_RefGene_Name" = "gene"), promoted_data) %>%
  left_join(expression_atlas %>% select(`Gene Name`, colon),
            by = c("UCSC_RefGene_Name" = "Gene Name"))

bind_rows(
  figure5_data %>%
    select(sigmaP_v3, sigmaP_pro, colon) %>%
    pivot_longer(-colon) %>%
    drop_na() %>%
    rename(expression = colon),
  figure5_data %>%
    select(sigmaPT_v3, sigmaPT_pro, colon) %>%
    pivot_longer(-colon) %>%
    drop_na() %>%
    rename(expression = colon),
  figure5_data %>%
    select(sigmaT_v3, sigmaT_pro, colon) %>%
    pivot_longer(-colon) %>%
    drop_na() %>%
    rename(expression = colon)
) %>%
  group_by(name) %>%
  mutate(value = factor(11-as.numeric(cut_number(value, 10)))) %>%
  separate(name, c("type", "method"), sep = "_", extra = "merge") %>%
  mutate(type = factor(type, c("sigmaP","sigmaPT","sigmaT"), c("SigmaP-Colon","SigmaPT-Colon","SigmaT-colon")),
         method = factor(method, c("v3", "pro"), c("Boot-strapped values", "Promoter region"))) %>%
  ggplot(aes(value, expression, fill = type)) +
  geom_boxplot() +
  scale_y_log10() +
  guides(color = "none") +
  facet_grid(method ~ type) +
  labs(x = "quantile of conservation by gene", y = "Expression", title = NULL,
       caption = "Figure 5A: Relationship between hierarchical variance (Sigmas) and gene expression.") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +
  theme(legend.position = "none")

# Figure 5B
bind_rows(
  figure5_data %>%
    select(score_median_v3, score_median_pro, colon) %>%
    rename(scoremedian_v3 = score_median_v3, scoremedian_pro = score_median_pro) %>%
    pivot_longer(-colon) %>%
    drop_na() %>%
    rename(expression = colon),
  figure5_data %>%
    select(score_median2_v3, score_median2_pro, colon) %>%
    rename(scoremedian2_v3 = score_median2_v3, scoremedian2_pro = score_median2_pro) %>%
    pivot_longer(-colon) %>%
    drop_na() %>%
    rename(expression = colon),
  figure5_data %>%
    select(score_reg_v3, score_reg_pro, colon) %>%
    rename(scorereg_v3 = score_reg_v3, scorereg_pro = score_reg_pro) %>%
    pivot_longer(-colon) %>%
    drop_na() %>%
    rename(expression = colon)
) %>%
  group_by(name) %>%
  mutate(value = factor(as.numeric(cut_number(value, 10)))) %>%
  separate(name, c("type", "method"), sep = "_", extra = "merge") %>%
  mutate(type = factor(type, c("scoremedian","scoremedian2","scorereg"), c("ScoreMedian-colon","ScoreMedian2-colon","Scorereg-colon")),
         method = factor(method, c("v3", "pro"), c("Boot-strapped values", "Promoter region"))) %>%
  ggplot(aes(value, expression, fill = type)) +
  geom_boxplot() +
  scale_y_log10() +
  guides(color = "none") +
  facet_grid(method ~ type) +
  labs(x = "quantile of conservation by gene", y = "Expression", title = NULL,
       caption = "Figure 5B: Relationship between conservation scores and gene expression.") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +
  theme(legend.position = "none")
```

<!-- \newpage -->
<!-- \section{Patient-specific analysis} -->
<!-- Basic overview: using a specified number of genes (here 100), take top genes (here using score_prob bootstrapping percentiles) to select a "gene signature" of sorts. Then for each patient, calculate the mean pair-wise distance within the individual patients' tumor samples. Method 1 is these raw values, method 2 normalizes by the mean PWD of all normal samples. Method 3 just examines the overall mean PWD for all tumor samples, normalized by mean PWD of all normal. -->

<!-- ```{r PatientPWD, echo=FALSE, message=FALSE, warning=FALSE, comment=NA, fig.height=9, fig.width=7} -->
<!-- library(gtools) -->

<!-- # pull out all samples beta values -->
<!-- all_beta <- FullAnnotation[,10:57] -->

<!-- # extract patient and tissue labels -->
<!-- patientLabel <- substr(colnames(all_beta), 1, 1) -->
<!-- patientLabel[10:12] <- "K*" -->
<!-- tissueLabel <- substr(colnames(all_beta), 2, 2) -->
<!-- tissueLabel[!(tissueLabel %in% "N")] <- "T"   #replace A/B/D/M with T -->

<!-- pats = patientLabel %>% sort %>% unique -->
<!-- npat = length(pats) -->

<!-- enrichedGenes <- permed_data %>% filter(score_prob_v3 < quantile(score_prob_v3, 0.05, na.rm = TRUE)) %>% -->
<!--   select(UCSC_RefGene_Name, score_prob, score_prob_v3, n) %>% -->
<!--   rename(Gene=UCSC_RefGene_Name, Pctile=score_prob_v3) -->

<!-- #ngene = 70 # using top N genes -->
<!-- #genes = enrichedGenes %>% top_n(-ngene,Pctile) %>% pull(Gene) -->
<!-- #genes = genes[1:ngene] -->

<!-- genes = enrichedGenes$Gene -->
<!-- ngene = length(genes) -->

<!-- pwd <- function(x) { -->
<!--   mean(dist(x)) -->
<!-- } -->
<!-- pwd2 <- function(x) { -->
<!--   abs(diff(x)) -->
<!-- } -->

<!-- signature_mat1 = matrix(,ngene,npat) # raw tumor pwd -->
<!-- signature_mat2 = matrix(,ngene,npat) # tumor pwd normalized by all normal pwd -->
<!-- for (igene in 1:ngene) { -->
<!--   # each gene, get gene inds, calculate PWD for the tumor samples -->
<!--   geneind = grep(genes[igene], FullAnnotation$UCSC_RefGene_Name) -->

<!--   # calculate PWD of each tumor -->
<!--   tum_pwd <- numeric(npat) -->
<!--   for (ipat in 1:npat) { -->
<!--     # each patient, get tumor sample inds -->
<!--     patind <- which(patientLabel == pats[ipat] & tissueLabel == "T") -->
<!--     tum_pwd[ipat] <- all_beta[geneind,patind] %>% logit %>% apply(1,pwd2) %>% mean -->
<!--   } -->

<!--   # calculate PWD of all normal -->
<!--   norm_pwd <- all_beta[geneind,tissueLabel == "N"] %>% logit %>% apply(1, pwd) %>% mean -->

<!--   signature_mat1[igene,] = tum_pwd -->
<!--   signature_mat2[igene,] = tum_pwd/norm_pwd -->
<!-- } -->

<!-- #cat("  \n\\newline  \nMethod 1 - raw intratumor PWD within each patient") -->
<!-- #signature_mat1 %>% as_tibble %>% `colnames<-`(pats) %>% gather %>% -->
<!-- #  mutate(gene=rep(genes,npat)) %>% rename(Patient=key,PWD=value) %>% -->
<!-- #  ggplot + geom_tile(aes(y=gene,x=Patient,fill=PWD)) + ylab(paste0(ngene," genes")) + -->
<!-- #  theme(axis.text.y = element_blank()) -->
<!-- #cat("  \n\\newline  \nMethod 2 - intratumor PWD normalized by all normal PWD") -->
<!-- #signature_mat2 %>% as_tibble %>% `colnames<-`(pats) %>% gather %>% -->
<!-- #  mutate(gene=rep(genes,npat)) %>% rename(Patient=key,PWD=value) %>% -->
<!-- #  ggplot + geom_tile(aes(y=gene,x=Patient,fill=PWD)) + ylab(paste0(ngene," genes")) + -->
<!-- #  theme(axis.text.y = element_blank()) -->
<!-- #cat("  \n\\newline  \nMethod 3 - all tumor PWD normalized by all normal PWD") -->
<!-- #signature_mat3 %>% as_tibble %>% mutate(gene=genes) %>% rename(PWD=V1) %>% -->
<!-- #  ggplot + geom_tile(aes(y=gene,x=1,fill=PWD)) + ylab(paste0(ngene," genes")) + -->
<!-- #  theme(axis.text.y = element_blank()) -->

<!-- library(pheatmap) -->
<!-- library(dendsort) -->

<!-- cat("  \n\\newline  \nMethod 1 - raw intratumor PWD within each patient") -->
<!-- df1 = as.data.frame(signature_mat1) -->
<!-- rownames(df1) <- genes -->
<!-- colnames(df1) <- pats -->
<!-- sort_hclust <- function(...) as.hclust(dendsort(as.dendrogram(...))) -->
<!-- mat_cluster_cols <- sort_hclust(hclust(dist(t(df1)))) -->
<!-- mat_cluster_rows <- sort_hclust(hclust(dist(df1))) -->
<!-- pheatmap( -->
<!--   mat=df1, -->
<!--   cluster_cols = mat_cluster_cols, -->
<!--   cluster_rows = mat_cluster_rows, -->
<!--   show_rownames = FALSE, -->
<!--   show_colnames = TRUE, -->
<!--   main = "Top 1118 Genes: Tumor PWD" -->
<!-- ) -->

<!-- cat("  \n\\newline  \nMethod 2 - intratumor PWD normalized by all normal PWD") -->
<!-- df1 = as.data.frame(signature_mat2) -->
<!-- rownames(df1) <- genes -->
<!-- colnames(df1) <- pats -->
<!-- sort_hclust <- function(...) as.hclust(dendsort(as.dendrogram(...))) -->
<!-- mat_cluster_cols <- sort_hclust(hclust(dist(t(df1)))) -->
<!-- mat_cluster_rows <- sort_hclust(hclust(dist(df1))) -->
<!-- pheatmap( -->
<!--   mat=df1, -->
<!--   cluster_cols = mat_cluster_cols, -->
<!--   cluster_rows = mat_cluster_rows, -->
<!--   show_rownames = FALSE, -->
<!--   show_colnames = TRUE, -->
<!--   main = "Top 1118 Genes: Tumor PWD" -->
<!-- ) -->
<!-- ``` -->

\newpage
\section{Appendix: Enrichment Gene/Pathway Lists}
\subsection{A1: SigmaP}
Genes are selected with cutoff <5th percentile (adjusted bootstrap), showing top 30 genes/pathways
```{r App_sigmaP, echo=FALSE, fig.fullwidth=TRUE, comment=NA, message=FALSE}
enrichedGenes <- permed_data %>%
  select(UCSC_RefGene_Name, sigmaP, sigmaP_v3, n) %>%
  rename(Gene=UCSC_RefGene_Name, Pctile=sigmaP_v3) %>%
  filter(Pctile < quantile(Pctile, 0.05, na.rm = TRUE)) %>%
  arrange(Pctile)

enrichedPaths <- tibble(Pathway = enrich_sigmaP$Description, ngenes = enrich_sigmaP$Count, padj = enrich_sigmaP$p.adjust) %>% arrange(padj)

ngene <- dim(enrichedGenes)[1]
npath <- length(enrich_sigmaP$Description)

cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath))

nshow = 30
if (ngene > nshow) {
  enrichedGenes <- enrichedGenes[1:nshow,]
}
if (npath > nshow) {
  enrichedPaths <- enrichedPaths[1:nshow,]
}

library(pander)
enrichedGenes %>% pander()

cat("SigmaP top 30 enriched paths:  \n")
enrichedPaths %>% pander()
```

\newpage
\subsection{A2: SigmaPT}
Genes are selected with cutoff <5th percentile (adjusted bootstrap), showing top 30 genes/pathways
```{r App_sigmaPT, echo=FALSE, comment=NA, message=FALSE}
enrichedGenes <- permed_data %>%
  select(UCSC_RefGene_Name, sigmaPT, sigmaPT_v3, n) %>%
  rename(Gene=UCSC_RefGene_Name, Pctile=sigmaPT_v3) %>%
  filter(Pctile < quantile(Pctile, 0.05, na.rm = TRUE)) %>%
  arrange(Pctile)

enrichedPaths <- tibble(Pathway = enrich_sigmaPT$Description, ngenes = enrich_sigmaPT$Count, padj = enrich_sigmaPT$p.adjust) %>% arrange(padj)

ngene <- dim(enrichedGenes)[1]
npath <- length(enrich_sigmaPT$Description)

cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath))

nshow = 30
if (ngene > nshow) {
  enrichedGenes <- enrichedGenes[1:nshow,]
}
if (npath > nshow) {
  enrichedPaths <- enrichedPaths[1:nshow,]
}

enrichedGenes %>% pander()

cat("\nSigmaPT top 30 enriched paths:  \n")
enrichedPaths %>% pander()
```

\newpage
\subsection{A3: SigmaT}
Genes are selected with cutoff <5th percentile (adjusted bootstrap), showing top 30 genes/pathways
```{r App_sigmaT, echo=FALSE, comment=NA, message=FALSE}
enrichedGenes <- permed_data %>%
  select(UCSC_RefGene_Name, sigmaT, sigmaT_v3, n) %>%
  rename(Gene=UCSC_RefGene_Name, Pctile=sigmaT_v3) %>%
  filter(Pctile < quantile(Pctile, 0.05, na.rm = TRUE)) %>%
  arrange(Pctile)

enrichedPaths <- tibble(Pathway = enrich_sigmaT$Description, ngenes = enrich_sigmaT$Count, padj = enrich_sigmaT$p.adjust) %>% arrange(padj)

ngene <- dim(enrichedGenes)[1]
npath <- length(enrich_sigmaT$Description)

cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath))

nshow = 30
if (ngene > nshow) {
  enrichedGenes <- enrichedGenes[1:nshow,]
}
if (npath > nshow) {
  enrichedPaths <- enrichedPaths[1:nshow,]
}

enrichedGenes %>% pander()

cat("SigmaT top 30 enriched paths:  \n")
enrichedPaths %>% pander()
```

<!-- \newpage -->
<!-- \subsection{A4: Score Prob} -->
<!-- Genes are selected with cutoff >95th percentile (adjusted bootstrap), showing top 50 genes/pathways -->
<!-- ```{r App_score_prob, echo=FALSE, comment=NA, message=FALSE} -->
<!-- enrichedGenes <- permed_data %>%  -->
<!--   select(UCSC_RefGene_Name, score_prob, score_prob_v3, n) %>% -->
<!--   rename(Gene=UCSC_RefGene_Name, Score_median=score_prob, Pctile=score_prob_v3) %>% -->
<!--   filter(Pctile > quantile(Pctile, 0.95, na.rm = TRUE)) %>% -->
<!--   arrange(Pctile) -->

<!-- enrichedPaths <- tibble(Pathway = enrich_score_prob$Description, ngenes = enrich_score_prob$Count, padj = enrich_score_prob$p.adjust) %>% arrange(padj) -->

<!-- ngene <- dim(enrichedGenes)[1] -->
<!-- npath <- length(enrich_score_prob$Description) -->

<!-- cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath)) -->

<!-- nshow = 50 -->
<!-- if (ngene > nshow) { -->
<!--   enrichedGenes <- enrichedGenes[1:nshow,] -->
<!-- } -->
<!-- if (npath > nshow) { -->
<!--   enrichedPaths <- enrichedPaths[1:nshow,] -->
<!-- } -->

<!-- enrichedGenes %>% pander() -->

<!-- cat("Score_prob top 50 enriched paths:  \n") -->
<!-- enrichedPaths %>% pander() -->
<!-- ``` -->

<!-- \newpage -->
<!-- \subsection{A5: Score Mean} -->
<!-- Genes are selected with cutoff >95th percentile (adjusted bootstrap), showing top 50 genes/pathways -->
<!-- ```{r App_score_mean, echo=FALSE, comment=NA, message=FALSE} -->
<!-- enrichedGenes <- permed_data %>%  -->
<!--   select(UCSC_RefGene_Name, score_mean, score_mean_v3, n) %>% -->
<!--   rename(Gene=UCSC_RefGene_Name, Score_median=score_mean, Pctile=score_mean_v3) %>% -->
<!--   filter(Pctile > quantile(Pctile, 0.95, na.rm = TRUE)) %>% -->
<!--   arrange(Pctile) -->

<!-- enrichedPaths <- tibble(Pathway = enrich_score_mean$Description, ngenes = enrich_score_mean$Count, padj = enrich_score_mean$p.adjust) %>% arrange(padj) -->

<!-- ngene <- dim(enrichedGenes)[1] -->
<!-- npath <- length(enrich_score_mean$Description) -->

<!-- cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath)) -->

<!-- nshow = 50 -->
<!-- if (ngene > nshow) { -->
<!--   enrichedGenes <- enrichedGenes[1:nshow,] -->
<!-- } -->
<!-- if (npath > nshow) { -->
<!--   enrichedPaths <- enrichedPaths[1:nshow,] -->
<!-- } -->

<!-- enrichedGenes %>% pander() -->

<!-- cat("Score_mean top 50 enriched paths:  \n") -->
<!-- enrichedPaths %>% pander() -->
<!-- ``` -->

\newpage
\subsection{A6: Score Median}
Genes are selected with cutoff >95th percentile (adjusted bootstrap), showing top 50 genes/pathways
```{r App_score_median, echo=FALSE, comment=NA, message=FALSE}
enrichedGenes <- permed_data %>% 
  select(UCSC_RefGene_Name, score_median, score_median_v3, n) %>%
  rename(Gene=UCSC_RefGene_Name, Score_median=score_median, Pctile=score_median_v3) %>%
  filter(Pctile > quantile(Pctile, 0.95, na.rm = TRUE)) %>%
  arrange(Pctile)

enrichedPaths <- tibble(Pathway = enrich_score_median$Description, ngenes = enrich_score_median$Count, padj = enrich_score_median$p.adjust) %>% arrange(padj)

ngene <- dim(enrichedGenes)[1]
npath <- length(enrich_score_median$Description)

cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath))

nshow = 50
if (ngene > nshow) {
  enrichedGenes <- enrichedGenes[1:nshow,]
}
if (npath > nshow) {
  enrichedPaths <- enrichedPaths[1:nshow,]
}

enrichedGenes %>% pander()

cat("Score_median top 50 enriched paths:  \n")
enrichedPaths %>% pander()
```

\newpage
\subsection{A7: Score Median2}
Genes are selected with cutoff >95th percentile (adjusted bootstrap), showing top 50 genes/pathways
```{r App_score_median2, echo=FALSE, comment=NA, message=FALSE}
enrichedGenes <- permed_data %>% 
  select(UCSC_RefGene_Name, score_median2, score_median2_v3, n) %>%
  rename(Gene=UCSC_RefGene_Name, Score_median2=score_median2, Pctile=score_median2_v3) %>%
  filter(Pctile > quantile(Pctile, 0.95, na.rm = TRUE)) %>%
  arrange(Pctile)

enrichedPaths <- tibble(Pathway = enrich_score_median2$Description, ngenes = enrich_score_median2$Count, padj = enrich_score_median2$p.adjust) %>% arrange(padj)

ngene <- dim(enrichedGenes)[1]
npath <- length(enrich_score_median2$Description)

cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath))

nshow = 50
if (ngene > nshow) {
  enrichedGenes <- enrichedGenes[1:nshow,]
}
if (npath > nshow) {
  enrichedPaths <- enrichedPaths[1:nshow,]
}

enrichedGenes %>% pander()

cat("Score_median2 top 50 enriched paths:  \n")
enrichedPaths %>% pander()
```

\newpage
\subsection{A9: Score Regularized}
Genes are selected with cutoff >95th percentile (adjusted bootstrap), showing top 50 genes/pathways
```{r App_score_reg, echo=FALSE, comment=NA, message=FALSE}
enrichedGenes <- permed_data %>% 
  select(UCSC_RefGene_Name, score_reg, score_reg_v3, n) %>%
  rename(Gene=UCSC_RefGene_Name, Score_reg=score_reg, Pctile=score_reg_v3) %>%
  filter(Pctile > quantile(Pctile, 0.95, na.rm = TRUE)) %>%
  arrange(Pctile)

enrichedPaths <- tibble(Pathway = enrich_score_reg$Description, ngenes = enrich_score_reg$Count, padj = enrich_score_reg$p.adjust) %>% arrange(padj)

ngene <- dim(enrichedGenes)[1]
npath <- length(enrich_score_reg$Description)

cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath))

nshow = 50
if (ngene > nshow) {
  enrichedGenes <- enrichedGenes[1:nshow,]
}
if (npath > nshow) {
  enrichedPaths <- enrichedPaths[1:nshow,]
}

enrichedGenes %>% pander()

cat("Score_reg top 50 enriched paths:  \n")
enrichedPaths %>% pander()
```

### 2020-09-25 Results Archive/StanResultsSummary.Rmd
updated code

---
title: "StanResultsSummary"
author: "Kevin Murgas"
date: "4/24/2020"
output: pdf_document
---
Summary of Stan Model Fit Results: Full-Scale (850K sites)
\newline Gamma priors (using exact gamma fits)
```{r Setup, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(ggpubr)
library(knitr)
```

```{r Load, echo=FALSE, include=FALSE}
setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
load("myFA_bulkonly.Rdata")
load("FullResultsTCGA_gamma.Rdata")

# make a dataframe of the 3 sigma vectors (take median of posterior)
# and do some log transforms
sigma_df <- data.frame(sigmaP=sigmaP_full$p50,sigmaPT=sigmaPT_full$p50,sigmaT=sigmaT_full$p50)
logsig_df <- log(sigma_df)
logRatio_df <- data.frame(PPTr=(logsig_df$sigmaP-logsig_df$sigmaPT),PTr=(logsig_df$sigmaP-logsig_df$sigmaT),PTTr=(logsig_df$sigmaPT-logsig_df$sigmaT))
sigmaMeans <- colMeans(sigma_df)
logsigMeans <- colMeans(logsig_df)
logRatioMeans <- colMeans(logRatio_df)
```

n_eff + Rhat Summary: \newline
```{r Diags, echo=FALSE,include=TRUE,error=FALSE,message=FALSE,warning=FALSE,results='asis',fig.keep='all',fig.show='asis',fig.height=3.5,fig.width=6.5}
library(knitr)
# making matrix of data for each parameter (mu, betaT, sigmaP, sigmaPT, sigmaT, sigmaE, lp__)
# reporting: average n_eff, min n_eff, site of min n_eff, # n_eff<500, (%), # n_eff<100, (%)
n_eff_mat <- matrix(nrow=7, ncol=7)
n_eff_mat[1,] <- c(mean(mu_full$n_eff), min(mu_full$n_eff), which(mu_full$n_eff==min(mu_full$n_eff)), sum(mu_full$n_eff<500), sum(mu_full$n_eff<500)*100/length(mu_full$n_eff), sum(mu_full$n_eff<100), sum(mu_full$n_eff<100)*100/length(mu_full$n_eff))
n_eff_mat[2,] <- c(mean(betaT_full$n_eff), min(betaT_full$n_eff), which(betaT_full$n_eff==min(betaT_full$n_eff)), sum(betaT_full$n_eff<500), sum(betaT_full$n_eff<500)*100/length(betaT_full$n_eff), sum(betaT_full$n_eff<100), sum(betaT_full$n_eff<100)*100/length(betaT_full$n_eff))
n_eff_mat[3,] <- c(mean(sigmaP_full$n_eff), min(sigmaP_full$n_eff), which(sigmaP_full$n_eff==min(sigmaP_full$n_eff)), sum(sigmaP_full$n_eff<500), sum(sigmaP_full$n_eff<500)*100/length(sigmaP_full$n_eff), sum(sigmaP_full$n_eff<100), sum(sigmaP_full$n_eff<100)*100/length(sigmaP_full$n_eff))
n_eff_mat[4,] <- c(mean(sigmaPT_full$n_eff), min(sigmaPT_full$n_eff), which(sigmaPT_full$n_eff==min(sigmaPT_full$n_eff)), sum(sigmaPT_full$n_eff<500), sum(sigmaPT_full$n_eff<500)*100/length(sigmaPT_full$n_eff), sum(sigmaPT_full$n_eff<100), sum(sigmaPT_full$n_eff<100)*100/length(sigmaPT_full$n_eff))
n_eff_mat[5,] <- c(mean(sigmaT_full$n_eff), min(sigmaT_full$n_eff), which(sigmaT_full$n_eff==min(sigmaT_full$n_eff)), sum(sigmaT_full$n_eff<500), sum(sigmaT_full$n_eff<500)*100/length(sigmaT_full$n_eff), sum(sigmaT_full$n_eff<100), sum(sigmaT_full$n_eff<100)*100/length(sigmaT_full$n_eff))
n_eff_mat[6,] <- c(mean(sigmaE_full$n_eff), min(sigmaE_full$n_eff), which(sigmaE_full$n_eff==min(sigmaE_full$n_eff)), sum(sigmaE_full$n_eff<500), sum(sigmaE_full$n_eff<500)*100/length(sigmaE_full$n_eff), sum(sigmaE_full$n_eff<100), sum(sigmaE_full$n_eff<100)*100/length(sigmaE_full$n_eff))
n_eff_mat[7,] <- c(mean(lp_full$n_eff), min(lp_full$n_eff), which(lp_full$n_eff==min(lp_full$n_eff)), sum(lp_full$n_eff<500), sum(lp_full$n_eff<500)*100/length(lp_full$n_eff), sum(lp_full$n_eff<100), sum(lp_full$n_eff<100)*100/length(lp_full$n_eff))
rownames(n_eff_mat)<-c("mu","betaT","sigmaP","sigmaPT","sigmaT","sigmaE","lp__")

kable(n_eff_mat, col.names=c("Avg n_eff", "Min n_eff", "< Site", "# <500","% <500","# <100", "% <100"), digits = 2)

# again for Rhat
Rhat_mat <- matrix(nrow=7, ncol=7)
Rhat_mat[1,] <- c(mean(mu_full$Rhat), max(mu_full$Rhat), which(mu_full$Rhat==max(mu_full$Rhat)), sum(mu_full$Rhat>1.1), sum(mu_full$Rhat>1.1)*100/length(mu_full$Rhat), sum(mu_full$Rhat>1.01), sum(mu_full$Rhat>1.01)*100/length(mu_full$Rhat))
Rhat_mat[2,] <- c(mean(betaT_full$Rhat), max(betaT_full$Rhat), which(betaT_full$Rhat==max(betaT_full$Rhat)), sum(betaT_full$Rhat>1.1), sum(betaT_full$Rhat>1.1)*100/length(betaT_full$Rhat), sum(betaT_full$Rhat>1.01), sum(betaT_full$Rhat>1.01)*100/length(betaT_full$Rhat))
Rhat_mat[3,] <- c(mean(sigmaP_full$Rhat), max(sigmaP_full$Rhat), which(sigmaP_full$Rhat==max(sigmaP_full$Rhat)), sum(sigmaP_full$Rhat>1.1), sum(sigmaP_full$Rhat>1.1)*100/length(sigmaP_full$Rhat), sum(sigmaP_full$Rhat>1.01), sum(sigmaP_full$Rhat>1.01)*100/length(sigmaP_full$Rhat))
Rhat_mat[4,] <- c(mean(sigmaPT_full$Rhat), max(sigmaPT_full$Rhat), which(sigmaPT_full$Rhat==max(sigmaPT_full$Rhat)), sum(sigmaPT_full$Rhat>1.1), sum(sigmaPT_full$Rhat>1.1)*100/length(sigmaPT_full$Rhat), sum(sigmaPT_full$Rhat>1.01), sum(sigmaPT_full$Rhat>1.01)*100/length(sigmaPT_full$Rhat))
Rhat_mat[5,] <- c(mean(sigmaT_full$Rhat), max(sigmaT_full$Rhat), which(sigmaT_full$Rhat==max(sigmaT_full$Rhat)), sum(sigmaT_full$Rhat>1.1), sum(sigmaT_full$Rhat>1.1)*100/length(sigmaT_full$Rhat), sum(sigmaT_full$Rhat>1.01), sum(sigmaT_full$Rhat>1.01)*100/length(sigmaT_full$Rhat))
Rhat_mat[6,] <- c(mean(sigmaE_full$Rhat), max(sigmaE_full$Rhat), which(sigmaE_full$Rhat==max(sigmaE_full$Rhat)), sum(sigmaE_full$Rhat>1.1), sum(sigmaE_full$Rhat>1.1)*100/length(sigmaE_full$Rhat), sum(sigmaE_full$Rhat>1.01), sum(sigmaE_full$Rhat>1.01)*100/length(sigmaE_full$Rhat))
Rhat_mat[7,] <- c(mean(lp_full$Rhat), max(lp_full$Rhat), which(lp_full$Rhat==max(lp_full$Rhat)), sum(lp_full$Rhat>1.1), sum(lp_full$Rhat>1.1)*100/length(lp_full$Rhat), sum(lp_full$Rhat>1.01), sum(lp_full$Rhat>1.01)*100/length(lp_full$Rhat))
rownames(Rhat_mat)<-c("mu","betaT","sigmaP","sigmaPT","sigmaT","sigmaE","lp__")

kable(Rhat_mat, col.names=c("Avg Rhat", "Max Rhat", "< Site", "# >1.1","% >1.1","# >1.01", "% >1.01"), digits = 2)
probsites <- unique(c(n_eff_mat[,3],Rhat_mat[,3]))
print(paste("Problem sites:",paste(probsites,collapse=" ")))
```
\newpage
Parameter Distributions (based on medians of posteriors): \newline
```{r Params, echo=FALSE,include=TRUE,error=FALSE,message=FALSE,warning=FALSE,results='asis',fig.keep='all',fig.show='asis',fig.height=3.2,fig.width=6.5}

# plot mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT
fBG <- function(x, a, m1, s1, m2, s2) a*dnorm(x,m1,s1) + (1-a)*dnorm(x,m2,s2)
pM <- ggplot(data.frame(mu=mu_full$p50), aes(x=mu)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("Mu") + xlim(c(-5,5)) + geom_vline(xintercept=mean(mu_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197)/0.1906566,geom="line",color="red")
pB <- ggplot(data.frame(betaT=betaT_full$p50), aes(x=betaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("BetaT") + xlim(c(-5,5)) + geom_vline(xintercept=mean(betaT_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) dcauchy(x,-0.04051, 0.19330)/1.646714,geom="line",color="red")
pE <- ggplot(data.frame(sigmaE=sigmaE_full$p50), aes(x=sigmaE)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaE") + xlim(c(0,2)) + geom_vline(xintercept=mean(sigmaE_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x,2,2)/0.7357589, geom="line",color="red")
pP <- ggplot(sigma_df, aes(x=sigmaP)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaP") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[1], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x,5.1989,16.6192)/3.1721, geom="line",color="red")
pPT <- ggplot(sigma_df, aes(x=sigmaPT)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaPT") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[2], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x,3.5539, 5.5324)/1.33697, geom="line",color="red")
pT <- ggplot(sigma_df, aes(x=sigmaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaT") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[3], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x,6.8383, 25.7560)/4.1923, geom="line",color="red")

p6 <- ggarrange(pM,pB,pE,pP,pPT,pT,labels=c("M","B","E","P","PT","T"),hjust=0,nrow=2,ncol=3)
print(p6)
rm(pM,pB,pE,pP,pPT,pT)

# plot the log sigmas
pP <- ggplot(logsig_df, aes(x=sigmaP)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaP)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[1], linetype="dotted") + theme_bw()  + stat_function(fun=function(x) dgamma(exp(x), 5.1989,16.6192)/3.1721, geom="line",color="red")
pPT <- ggplot(logsig_df, aes(x=sigmaPT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaPT)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[2], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(exp(x), 3.5539, 5.5324)/1.33697, geom="line",color="red")
pT <- ggplot(logsig_df, aes(x=sigmaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaT)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[3], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(exp(x), 6.8383, 25.7560)/4.1923, geom="line",color="red")
# and log ratios
pPPTr <- ggplot(logRatio_df, aes(x=PPTr)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle(expression("log("*sigma*"P/"*sigma*"PT)")) + xlim(c(-5,5)) + geom_vline(xintercept=logRatioMeans[1], linetype="dotted") + theme_bw()
pPTr <- ggplot(logRatio_df, aes(x=PTr)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle(expression("log("*sigma*"P/"*sigma*"T)")) + xlim(c(-5,5)) + geom_vline(xintercept=logRatioMeans[2], linetype="dotted") + theme_bw()
pPTTr <- ggplot(logRatio_df, aes(x=PTTr)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle(expression("log("*sigma*"PT/"*sigma*"T)")) + xlim(c(-5,5)) + geom_vline(xintercept=logRatioMeans[3], linetype="dotted") + theme_bw()

p6 <- ggarrange(pP,pPT,pT,pPPTr,pPTr,pPTTr,labels=c("logP","logPT","logT","PPTr","PTr","PTTr"),hjust=0,nrow=2,ncol=3)
print(p6)
rm(pPPTr,pPTr,pPTTr,p6)

# density plots of sigmas x sigmas
# too long to load/print with full data! using ~1:10 thinning
logsig_df_thin <- logsig_df
pPxPT <- ggplot(logsig_df_thin, aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1), show.legend=FALSE) + ggtitle("log-P vs log-PT") + xlim(-5,2) + ylim(-5,2)
pPxT <- ggplot(logsig_df_thin, aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1), show.legend=FALSE) + ggtitle("log-P vs log-T") + xlim(-5,2) + ylim(-5,2)
pPTxT <- ggplot(logsig_df_thin, aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1), show.legend=FALSE) + ggtitle("log-PT vs log-T") + xlim(-5,2) + ylim(-5,2)
pSxS1 <- ggarrange(pPxT,pPTxT,pPxPT,nrow=2,ncol=2)
logsig_df_thin <- logsig_df[sample(1:dim(logsig_df)[1], 60000, replace=F),]
pPxPT <- ggplot(logsig_df_thin, aes(x=sigmaP,y=sigmaPT)) + geom_density2d() + ggtitle("log-P vs log-PT") + xlim(-3,1) + ylim(-3,1)
pPxT <- ggplot(logsig_df_thin, aes(x=sigmaP,y=sigmaT)) + geom_density2d() + ggtitle("log-P vs log-T") + xlim(-3,1) + ylim(-3,1)
pPTxT <- ggplot(logsig_df_thin, aes(x=sigmaPT,y=sigmaT)) + geom_density2d() + ggtitle("log-PT vs log-T") + xlim(-3,1) + ylim(-3,1)
pSxS2 <- ggarrange(pPxT,pPTxT,pPxPT,nrow=2,ncol=2)
pSxS <- ggarrange(pSxS1,pSxS2,nrow=1,ncol=2)
print(pSxS)
rm(pSxS1,pSxS2)
```
\newpage
Principal Component Analysis (on all sites log-sigma values) \newline
```{r PCA, echo=FALSE,include=TRUE,error=FALSE,message=FALSE,warning=FALSE,results='asis',fig.keep='all',fig.show='asis',fig.height=2.5,fig.width=6.5}
pca <- princomp(logsig_df)
pve <- pca$sdev^2/sum(pca$sdev^2)
summ <- cbind(t(pca$loadings[,]),pve)
colnames(summ)[4] <- "Frac. var explained"
kable(summ,digits=4)

# plot scores of sites along PCA axes
# first just axes 1+2 #geom_bin2d(binwidth=c(0.1,0.1), show.legend=FALSE)
pca_df <- data.frame(pca$scores)
pca_df_thin <- pca_df[sample(1:dim(logsig_df)[1], 60000, replace=F),]
p1<-ggplot(data=pca_df_thin, aes(x=Comp.1, y=Comp.2)) + geom_density2d() + ggtitle("Comp 1 vs 2")
p2<-ggplot(data=pca_df_thin, aes(x=Comp.1, y=Comp.3)) + geom_density2d() + ggtitle("Comp 1 vs 3")
p3<-ggplot(data=pca_df_thin, aes(x=Comp.2, y=Comp.3)) + geom_density2d() + ggtitle("Comp 2 vs 3")
p4<-ggarrange(p1,p2,p3,nrow=1,ncol=3)
print(p4)
```
\newpage
Clustering Analysis: \newline 
```{r clust, echo=FALSE,include=TRUE,error=FALSE,message=FALSE,warning=FALSE,results='asis',fig.keep='all',fig.show='asis',fig.height=4,fig.width=6.5}
cat("\\newline")
print("K-means clustering, 3 clusters:")
ptm<-proc.time()
kclust <- kmeans(logsig_df,3,nstart=5)
print(proc.time()-ptm)
clust <- kclust$cluster

# sigmaP vs sigmaPT
p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
pPxPT_kclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + scale_fill_identity() + theme_bw() + ggtitle("log-P x log-PT") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaP") + ylab("sigmaPT")

# sigmaP vs sigmaT
p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
pPxT_kclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + scale_fill_identity() + theme_bw() + ggtitle("log-P x log-T") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaP") + ylab("sigmaT")

# sigmaPT vs sigmaT
p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
pPTxT_kclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + scale_fill_identity() + theme_bw() + ggtitle("log-PT x log-T") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaPT") + ylab("sigmaT")
pSxS_kclust3 <- ggarrange(pPxT_kclust,pPTxT_kclust,pPxPT_kclust,nrow=2,ncol=2)

cat("\\newline")
print("K-means clustering, 4 clusters:")
ptm<-proc.time()
kclust <- kmeans(logsig_df,4,nstart=5)
print(proc.time()-ptm)
clust <- kclust$cluster

# sigmaP vs sigmaPT
p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
p4 <- ggplot(data=logsig_df[which(clust==4),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="orange")
pPxPT_kclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + geom_tile(data = layer_data(p4)) + scale_fill_identity() + theme_bw() + ggtitle("log-P x log-PT") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaP") + ylab("sigmaPT")

# sigmaP vs sigmaT
p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
p4 <- ggplot(data=logsig_df[which(clust==4),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="orange")
pPxT_kclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + geom_tile(data = layer_data(p4)) + scale_fill_identity() + theme_bw() + ggtitle("log-P x log-T") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaP") + ylab("sigmaT")

# sigmaPT vs sigmaT
p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
p4 <- ggplot(data=logsig_df[which(clust==4),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="orange")
pPTxT_kclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + geom_tile(data = layer_data(p4)) + scale_fill_identity() + theme_bw() + ggtitle("log-PT x log-T") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaPT") + ylab("sigmaT")
pSxS_kclust4 <- ggarrange(pPxT_kclust,pPTxT_kclust,pPxPT_kclust,nrow=2,ncol=2)
pSxS_kclust <- ggarrange(pSxS_kclust3,pSxS_kclust4,nrow=1,ncol=2)
cat("\\newline")
print(pSxS_kclust)
rm(p1,p2,p3,p4,pPxPT_kclust,pPxT_kclust,pPTxT_kclust,pSxS_kclust3,pSxS_kclust4,pSxS_kclust)

cat("\\newline")
print("Fuzzy C-means clustering, 3 clusters:")
library(e1071)
ptm<-proc.time()
cclust <- cmeans(logsig_df,3)
print(proc.time()-ptm)
clust <- cclust$cluster

# sigmaP vs sigmaPT
p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
pPxPT_cclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + scale_fill_identity() + theme_bw() + ggtitle("log-P x log-PT") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaP") + ylab("sigmaPT")

# sigmaP vs sigmaT
p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
pPxT_cclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + scale_fill_identity() + theme_bw() + ggtitle("log-P x log-T") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaP") + ylab("sigmaT")

# sigmaPT vs sigmaT
p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
pPTxT_cclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + scale_fill_identity() + theme_bw() + ggtitle("log-PT x log-T") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaPT") + ylab("sigmaT")
pSxS_cclust3 <- ggarrange(pPxT_cclust,pPTxT_cclust,pPxPT_cclust,nrow=2,ncol=2)

cat("\\newline")
print("Fuzzy C-means clustering, 4 clusters:")
library(e1071)
ptm<-proc.time()
cclust <- cmeans(logsig_df,4)
print(proc.time()-ptm)
clust <- cclust$cluster

# sigmaP vs sigmaPT
p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
p4 <- ggplot(data=logsig_df[which(clust==4),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="orange")
pPxPT_cclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + geom_tile(data = layer_data(p4)) + scale_fill_identity() + theme_bw() + ggtitle("log-P x log-PT") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaP") + ylab("sigmaPT")

# sigmaP vs sigmaT
p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
p4 <- ggplot(data=logsig_df[which(clust==4),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="orange")
pPxT_cclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + geom_tile(data = layer_data(p4)) + scale_fill_identity() + theme_bw() + ggtitle("log-P x log-T") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaP") + ylab("sigmaT")

# sigmaPT vs sigmaT
p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
p4 <- ggplot(data=logsig_df[which(clust==4),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="orange")
pPTxT_cclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + geom_tile(data = layer_data(p4)) + scale_fill_identity() + theme_bw() + ggtitle("log-PT x log-T") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaPT") + ylab("sigmaT")
pSxS_cclust4 <- ggarrange(pPxT_cclust,pPTxT_cclust,pPxPT_cclust,nrow=2,ncol=2)
pSxS_cclust <- ggarrange(pSxS_cclust3,pSxS_cclust4,nrow=1,ncol=2)
cat("\\newline")
print(pSxS_cclust)
rm(p1,p2,p3,p4,pPxPT_cclust,pPxT_cclust,pPTxT_cclust,pSxS_cclust3,pSxS_cclust4,pSxS_cclust)
```

### 2020-09-25 Results Archive/StanResultsSummary_gamma.Rmd
updated code

--- 2020-09-25 Results Archive/StanResultsSummary.Rmd
@@
 Parameter Distributions (based on medians of posteriors): \newline
-```{r Params, echo=FALSE,include=TRUE,error=FALSE,message=FALSE,warning=FALSE,results='asis',fig.keep='all',fig.show='asis',fig.height=3.2,fig.width=6.5}
+```{r Params, echo=FALSE,include=TRUE,error=FALSE,message=FALSE,warning=FALSE,results='asis',fig.keep='all',fig.show='asis',fig.height=3,fig.width=6.5}
 

### 2020-09-25 Results Archive/StanResultsSummary_invgamma.Rmd
updated code

--- 2020-09-25 Results Archive/StanResultsSummary_gamma.Rmd
@@
 Summary of Stan Model Fit Results: Full-Scale (850K sites)
-\newline Gamma priors (using exact gamma fits)
+\newline Inverse-Gamma priors (using exact inv-gamma fits)
 ```{r Setup, echo=FALSE, include=FALSE}
@@
 load("myFA_bulkonly.Rdata")
-load("FullResultsTCGA_gamma.Rdata")
+load("FullResultsTCGA_invgamma.Rdata")
 
@@
 fBG <- function(x, a, m1, s1, m2, s2) a*dnorm(x,m1,s1) + (1-a)*dnorm(x,m2,s2)
+dIG <- function(x, a, b) 1/(b*gamma(a))*(b/x)^(a+1)*exp(-b/x)
 pM <- ggplot(data.frame(mu=mu_full$p50), aes(x=mu)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("Mu") + xlim(c(-5,5)) + geom_vline(xintercept=mean(mu_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197)/0.1906566,geom="line",color="red")
 pB <- ggplot(data.frame(betaT=betaT_full$p50), aes(x=betaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("BetaT") + xlim(c(-5,5)) + geom_vline(xintercept=mean(betaT_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) dcauchy(x,-0.04051, 0.19330)/1.646714,geom="line",color="red")
-pE <- ggplot(data.frame(sigmaE=sigmaE_full$p50), aes(x=sigmaE)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaE") + xlim(c(0,2)) + geom_vline(xintercept=mean(sigmaE_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x,2,2)/0.7357589, geom="line",color="red")
-pP <- ggplot(sigma_df, aes(x=sigmaP)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaP") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[1], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x,5.1989,16.6192)/3.1721, geom="line",color="red")
-pPT <- ggplot(sigma_df, aes(x=sigmaPT)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaPT") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[2], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x,3.5539, 5.5324)/1.33697, geom="line",color="red")
-pT <- ggplot(sigma_df, aes(x=sigmaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaT") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[3], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x,6.8383, 25.7560)/4.1923, geom="line",color="red")
+pE <- ggplot(data.frame(sigmaE=sigmaE_full$p50), aes(x=sigmaE)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaE") + xlim(c(0,2)) + geom_vline(xintercept=mean(sigmaE_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) dIG(x,1,1)/0.54134, geom="line",color="red")
+pP <- ggplot(sigma_df, aes(x=sigmaP)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaP") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[1], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dIG(x,6.3641, 1.6631)/4.09616,geom="line",color="red")
+pPT <- ggplot(sigma_df, aes(x=sigmaPT)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaPT") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[2], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dIG(x,3.2624, 1.5397)/1.7114,geom="line",color="red")
+pT <- ggplot(sigma_df, aes(x=sigmaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaT") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[3], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dIG(x,6.5392, 1.4894)/4.7565,geom="line",color="red")
 
@@
 # plot the log sigmas
-pP <- ggplot(logsig_df, aes(x=sigmaP)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaP)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[1], linetype="dotted") + theme_bw()  + stat_function(fun=function(x) dgamma(exp(x), 5.1989,16.6192)/3.1721, geom="line",color="red")
-pPT <- ggplot(logsig_df, aes(x=sigmaPT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaPT)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[2], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(exp(x), 3.5539, 5.5324)/1.33697, geom="line",color="red")
-pT <- ggplot(logsig_df, aes(x=sigmaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaT)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[3], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(exp(x), 6.8383, 25.7560)/4.1923, geom="line",color="red")
+pP <- ggplot(logsig_df, aes(x=sigmaP)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaP)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[1], linetype="dotted") + theme_bw()  + stat_function(fun=function(x) dIG(exp(x), 6.3641, 1.6631)/4.09616, geom="line",color="red")
+pPT <- ggplot(logsig_df, aes(x=sigmaPT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaPT)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[2], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dIG(exp(x), 3.2624, 1.5397)/1.7114, geom="line",color="red")
+pT <- ggplot(logsig_df, aes(x=sigmaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaT)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[3], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dIG(exp(x), 6.5392, 1.4894)/4.7565, geom="line",color="red")
 # and log ratios

### 2020-09-25 Results Archive/StanResultsSummary_relaxedgamma2.Rmd
updated code

--- 2020-09-25 Results Archive/StanResultsSummary_gamma.Rmd
@@
 Summary of Stan Model Fit Results: Full-Scale (850K sites)
-\newline Gamma priors (using exact gamma fits)
+\newline Relaxed Gamma priors (using relaxed gamma fits with x2 variance)
 ```{r Setup, echo=FALSE, include=FALSE}
@@
 load("myFA_bulkonly.Rdata")
-load("FullResultsTCGA_gamma.Rdata")
+load("FullResultsTCGA_relaxedgamma.Rdata")
 
@@
 pM <- ggplot(data.frame(mu=mu_full$p50), aes(x=mu)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("Mu") + xlim(c(-5,5)) + geom_vline(xintercept=mean(mu_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197)/0.1906566,geom="line",color="red")
-pB <- ggplot(data.frame(betaT=betaT_full$p50), aes(x=betaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("BetaT") + xlim(c(-5,5)) + geom_vline(xintercept=mean(betaT_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) dcauchy(x,-0.04051, 0.19330)/1.646714,geom="line",color="red")
+pB <- ggplot(data.frame(betaT=betaT_full$p50), aes(x=betaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("BetaT") + xlim(c(-5,5)) + geom_vline(xintercept=mean(betaT_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) dcauchy(x,-0.04051, 0.38660)/0.82336,geom="line",color="red")
 pE <- ggplot(data.frame(sigmaE=sigmaE_full$p50), aes(x=sigmaE)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaE") + xlim(c(0,2)) + geom_vline(xintercept=mean(sigmaE_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x,2,2)/0.7357589, geom="line",color="red")
-pP <- ggplot(sigma_df, aes(x=sigmaP)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaP") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[1], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x,5.1989,16.6192)/3.1721, geom="line",color="red")
-pPT <- ggplot(sigma_df, aes(x=sigmaPT)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaPT") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[2], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x,3.5539, 5.5324)/1.33697, geom="line",color="red")
-pT <- ggplot(sigma_df, aes(x=sigmaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaT") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[3], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x,6.8383, 25.7560)/4.1923, geom="line",color="red")
+pP <- ggplot(sigma_df, aes(x=sigmaP)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaP") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[1], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x, 3.4016, 9.5057)/2.3641, geom="line",color="red")
+pPT <- ggplot(sigma_df, aes(x=sigmaPT)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaPT") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[2], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x, 2.5210, 3.2948)/1.0097, geom="line",color="red")
+pT <- ggplot(sigma_df, aes(x=sigmaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaT") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[3], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x, 4.2574, 14.370)/3.0964, geom="line",color="red")
 
@@
 # plot the log sigmas
-pP <- ggplot(logsig_df, aes(x=sigmaP)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaP)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[1], linetype="dotted") + theme_bw()  + stat_function(fun=function(x) dgamma(exp(x), 5.1989,16.6192)/3.1721, geom="line",color="red")
-pPT <- ggplot(logsig_df, aes(x=sigmaPT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaPT)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[2], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(exp(x), 3.5539, 5.5324)/1.33697, geom="line",color="red")
-pT <- ggplot(logsig_df, aes(x=sigmaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaT)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[3], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(exp(x), 6.8383, 25.7560)/4.1923, geom="line",color="red")
+pP <- ggplot(logsig_df, aes(x=sigmaP)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaP)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[1], linetype="dotted") + theme_bw()  + stat_function(fun=function(x) dgamma(exp(x), 3.4016, 9.5057)/2.3641, geom="line",color="red")
+pPT <- ggplot(logsig_df, aes(x=sigmaPT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaPT)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[2], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(exp(x), 2.5210, 3.2948)/1.0097, geom="line",color="red")
+pT <- ggplot(logsig_df, aes(x=sigmaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaT)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[3], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(exp(x), 4.2574, 14.370)/3.0964, geom="line",color="red")
 # and log ratios

### 2020-09-25 Results Archive/StanResultsSummary_relaxedgamma3.Rmd
updated code

--- 2020-09-25 Results Archive/StanResultsSummary_gamma.Rmd
@@
 author: "Kevin Murgas"
-date: "4/24/2020"
+date: "4/30/2020"
 output: pdf_document
@@
 Summary of Stan Model Fit Results: Full-Scale (850K sites)
-\newline Gamma priors (using exact gamma fits)
+\newline Relaxed Gamma priors (using relaxed gamma fits with x3 variance)
 ```{r Setup, echo=FALSE, include=FALSE}
@@
 load("myFA_bulkonly.Rdata")
-load("FullResultsTCGA_gamma.Rdata")
+load("FullResultsTCGA_relaxedgamma3.Rdata")
 
@@
 pM <- ggplot(data.frame(mu=mu_full$p50), aes(x=mu)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("Mu") + xlim(c(-5,5)) + geom_vline(xintercept=mean(mu_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197)/0.1906566,geom="line",color="red")
-pB <- ggplot(data.frame(betaT=betaT_full$p50), aes(x=betaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("BetaT") + xlim(c(-5,5)) + geom_vline(xintercept=mean(betaT_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) dcauchy(x,-0.04051, 0.19330)/1.646714,geom="line",color="red")
+pB <- ggplot(data.frame(betaT=betaT_full$p50), aes(x=betaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("BetaT") + xlim(c(-5,5)) + geom_vline(xintercept=mean(betaT_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) dcauchy(x,-0.04051, 0.5799)/0.54890,geom="line",color="red")
 pE <- ggplot(data.frame(sigmaE=sigmaE_full$p50), aes(x=sigmaE)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaE") + xlim(c(0,2)) + geom_vline(xintercept=mean(sigmaE_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x,2,2)/0.7357589, geom="line",color="red")
-pP <- ggplot(sigma_df, aes(x=sigmaP)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaP") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[1], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x,5.1989,16.6192)/3.1721, geom="line",color="red")
-pPT <- ggplot(sigma_df, aes(x=sigmaPT)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaPT") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[2], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x,3.5539, 5.5324)/1.33697, geom="line",color="red")
-pT <- ggplot(sigma_df, aes(x=sigmaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaT") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[3], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x,6.8383, 25.7560)/4.1923, geom="line",color="red")
+pP <- ggplot(sigma_df, aes(x=sigmaP)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaP") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[1], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x, 2.7693, 7.0029)/2.0046, geom="line",color="red")
+pPT <- ggplot(sigma_df, aes(x=sigmaPT)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaPT") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[2], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x, 2.1457, 2.4819)/0.8615, geom="line",color="red")
+pT <- ggplot(sigma_df, aes(x=sigmaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaT") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[3], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x, 3.3643, 10.430)/2.6129, geom="line",color="red")
 
@@
 # plot the log sigmas
-pP <- ggplot(logsig_df, aes(x=sigmaP)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaP)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[1], linetype="dotted") + theme_bw()  + stat_function(fun=function(x) dgamma(exp(x), 5.1989,16.6192)/3.1721, geom="line",color="red")
-pPT <- ggplot(logsig_df, aes(x=sigmaPT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaPT)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[2], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(exp(x), 3.5539, 5.5324)/1.33697, geom="line",color="red")
-pT <- ggplot(logsig_df, aes(x=sigmaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaT)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[3], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(exp(x), 6.8383, 25.7560)/4.1923, geom="line",color="red")
+pP <- ggplot(logsig_df, aes(x=sigmaP)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaP)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[1], linetype="dotted") + theme_bw()  + stat_function(fun=function(x) dgamma(exp(x), 2.7693, 7.0029)/2.0046, geom="line",color="red")
+pPT <- ggplot(logsig_df, aes(x=sigmaPT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaPT)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[2], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(exp(x), 2.1457, 2.4819)/0.8615, geom="line",color="red")
+pT <- ggplot(logsig_df, aes(x=sigmaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaT)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[3], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(exp(x), 3.3643, 10.430)/2.6129, geom="line",color="red")
 # and log ratios
@@
 ```
-\newpage
-Clustering Analysis: \newline 
-```{r clust, echo=FALSE,include=TRUE,error=FALSE,message=FALSE,warning=FALSE,results='asis',fig.keep='all',fig.show='asis',fig.height=4,fig.width=6.5}
-cat("\\newline")
-print("K-means clustering, 3 clusters:")
-ptm<-proc.time()
-kclust <- kmeans(logsig_df,3,nstart=5)
-print(proc.time()-ptm)
-clust <- kclust$cluster
-
-# sigmaP vs sigmaPT
-p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
-p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
-p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
-pPxPT_kclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + scale_fill_identity() + theme_bw() + ggtitle("log-P x log-PT") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaP") + ylab("sigmaPT")
-
-# sigmaP vs sigmaT
-p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
-p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
-p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
-pPxT_kclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + scale_fill_identity() + theme_bw() + ggtitle("log-P x log-T") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaP") + ylab("sigmaT")
-
-# sigmaPT vs sigmaT
-p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
-p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
-p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
-pPTxT_kclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + scale_fill_identity() + theme_bw() + ggtitle("log-PT x log-T") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaPT") + ylab("sigmaT")
-pSxS_kclust3 <- ggarrange(pPxT_kclust,pPTxT_kclust,pPxPT_kclust,nrow=2,ncol=2)
-
-cat("\\newline")
-print("K-means clustering, 4 clusters:")
-ptm<-proc.time()
-kclust <- kmeans(logsig_df,4,nstart=5)
-print(proc.time()-ptm)
-clust <- kclust$cluster
-
-# sigmaP vs sigmaPT
-p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
-p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
-p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
-p4 <- ggplot(data=logsig_df[which(clust==4),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="orange")
-pPxPT_kclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + geom_tile(data = layer_data(p4)) + scale_fill_identity() + theme_bw() + ggtitle("log-P x log-PT") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaP") + ylab("sigmaPT")
-
-# sigmaP vs sigmaT
-p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
-p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
-p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
-p4 <- ggplot(data=logsig_df[which(clust==4),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="orange")
-pPxT_kclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + geom_tile(data = layer_data(p4)) + scale_fill_identity() + theme_bw() + ggtitle("log-P x log-T") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaP") + ylab("sigmaT")
-
-# sigmaPT vs sigmaT
-p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
-p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
-p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
-p4 <- ggplot(data=logsig_df[which(clust==4),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="orange")
-pPTxT_kclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + geom_tile(data = layer_data(p4)) + scale_fill_identity() + theme_bw() + ggtitle("log-PT x log-T") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaPT") + ylab("sigmaT")
-pSxS_kclust4 <- ggarrange(pPxT_kclust,pPTxT_kclust,pPxPT_kclust,nrow=2,ncol=2)
-pSxS_kclust <- ggarrange(pSxS_kclust3,pSxS_kclust4,nrow=1,ncol=2)
-cat("\\newline")
-print(pSxS_kclust)
-rm(p1,p2,p3,p4,pPxPT_kclust,pPxT_kclust,pPTxT_kclust,pSxS_kclust3,pSxS_kclust4,pSxS_kclust)
-
-cat("\\newline")
-print("Fuzzy C-means clustering, 3 clusters:")
-library(e1071)
-ptm<-proc.time()
-cclust <- cmeans(logsig_df,3)
-print(proc.time()-ptm)
-clust <- cclust$cluster
-
-# sigmaP vs sigmaPT
-p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
-p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
-p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
-pPxPT_cclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + scale_fill_identity() + theme_bw() + ggtitle("log-P x log-PT") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaP") + ylab("sigmaPT")
-
-# sigmaP vs sigmaT
-p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
-p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
-p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
-pPxT_cclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + scale_fill_identity() + theme_bw() + ggtitle("log-P x log-T") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaP") + ylab("sigmaT")
-
-# sigmaPT vs sigmaT
-p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
-p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
-p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
-pPTxT_cclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + scale_fill_identity() + theme_bw() + ggtitle("log-PT x log-T") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaPT") + ylab("sigmaT")
-pSxS_cclust3 <- ggarrange(pPxT_cclust,pPTxT_cclust,pPxPT_cclust,nrow=2,ncol=2)
-
-cat("\\newline")
-print("Fuzzy C-means clustering, 4 clusters:")
-library(e1071)
-ptm<-proc.time()
-cclust <- cmeans(logsig_df,4)
-print(proc.time()-ptm)
-clust <- cclust$cluster
-
-# sigmaP vs sigmaPT
-p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
-p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
-p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
-p4 <- ggplot(data=logsig_df[which(clust==4),],aes(x=sigmaP,y=sigmaPT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="orange")
-pPxPT_cclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + geom_tile(data = layer_data(p4)) + scale_fill_identity() + theme_bw() + ggtitle("log-P x log-PT") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaP") + ylab("sigmaPT")
-
-# sigmaP vs sigmaT
-p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
-p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
-p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
-p4 <- ggplot(data=logsig_df[which(clust==4),],aes(x=sigmaP,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="orange")
-pPxT_cclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + geom_tile(data = layer_data(p4)) + scale_fill_identity() + theme_bw() + ggtitle("log-P x log-T") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaP") + ylab("sigmaT")
-
-# sigmaPT vs sigmaT
-p1 <- ggplot(data=logsig_df[which(clust==1),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="red")
-p2 <- ggplot(data=logsig_df[which(clust==2),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="green")
-p3 <- ggplot(data=logsig_df[which(clust==3),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="blue")
-p4 <- ggplot(data=logsig_df[which(clust==4),],aes(x=sigmaPT,y=sigmaT)) + geom_bin2d(binwidth=c(0.1,0.1)) + scale_fill_gradient(low="transparent", high="orange")
-pPTxT_cclust <- ggplot(data = layer_data(p1), aes(x = x, y = y, fill = fill)) + geom_tile() + geom_tile(data = layer_data(p2)) + geom_tile(data = layer_data(p3)) + geom_tile(data = layer_data(p4)) + scale_fill_identity() + theme_bw() + ggtitle("log-PT x log-T") + xlim(-3,1) + ylim(-3,1) + xlab("sigmaPT") + ylab("sigmaT")
-pSxS_cclust4 <- ggarrange(pPxT_cclust,pPTxT_cclust,pPxPT_cclust,nrow=2,ncol=2)
-pSxS_cclust <- ggarrange(pSxS_cclust3,pSxS_cclust4,nrow=1,ncol=2)
-cat("\\newline")
-print(pSxS_cclust)
-rm(p1,p2,p3,p4,pPxPT_cclust,pPxT_cclust,pPTxT_cclust,pSxS_cclust3,pSxS_cclust4,pSxS_cclust)
-```

### 2020-09-25 Results Archive/StanResultsSummary_relaxedgamma5.Rmd
updated code

--- 2020-09-25 Results Archive/StanResultsSummary_gamma.Rmd
@@
 author: "Kevin Murgas"
-date: "4/24/2020"
+date: "4/25/2020"
 output: pdf_document
@@
 Summary of Stan Model Fit Results: Full-Scale (850K sites)
-\newline Gamma priors (using exact gamma fits)
+\newline Relaxed Gamma priors (using relaxed gamma fits with x5 variance)
 ```{r Setup, echo=FALSE, include=FALSE}
@@
 load("myFA_bulkonly.Rdata")
-load("FullResultsTCGA_gamma.Rdata")
+load("FullResultsTCGA_relaxedgamma5.Rdata")
 
@@
 pM <- ggplot(data.frame(mu=mu_full$p50), aes(x=mu)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("Mu") + xlim(c(-5,5)) + geom_vline(xintercept=mean(mu_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197)/0.1906566,geom="line",color="red")
-pB <- ggplot(data.frame(betaT=betaT_full$p50), aes(x=betaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("BetaT") + xlim(c(-5,5)) + geom_vline(xintercept=mean(betaT_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) dcauchy(x,-0.04051, 0.19330)/1.646714,geom="line",color="red")
+pB <- ggplot(data.frame(betaT=betaT_full$p50), aes(x=betaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("BetaT") + xlim(c(-5,5)) + geom_vline(xintercept=mean(betaT_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) dcauchy(x,-0.04051, 0.96650)/0.32934,geom="line",color="red")
 pE <- ggplot(data.frame(sigmaE=sigmaE_full$p50), aes(x=sigmaE)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaE") + xlim(c(0,2)) + geom_vline(xintercept=mean(sigmaE_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x,2,2)/0.7357589, geom="line",color="red")
-pP <- ggplot(sigma_df, aes(x=sigmaP)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaP") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[1], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x,5.1989,16.6192)/3.1721, geom="line",color="red")
-pPT <- ggplot(sigma_df, aes(x=sigmaPT)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaPT") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[2], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x,3.5539, 5.5324)/1.33697, geom="line",color="red")
-pT <- ggplot(sigma_df, aes(x=sigmaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaT") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[3], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x,6.8383, 25.7560)/4.1923, geom="line",color="red")
+pP <- ggplot(sigma_df, aes(x=sigmaP)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaP") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[1], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x, 2.2298, 4.8674)/1.6384, geom="line",color="red")
+pPT <- ggplot(sigma_df, aes(x=sigmaPT)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaPT") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[2], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x, 1.8166, 1.7689)/0.70791, geom="line",color="red")
+pT <- ggplot(sigma_df, aes(x=sigmaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaT") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[3], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x, 2.6144, 7.1221)/2.1250, geom="line",color="red")
 
@@
 # plot the log sigmas
-pP <- ggplot(logsig_df, aes(x=sigmaP)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaP)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[1], linetype="dotted") + theme_bw()  + stat_function(fun=function(x) dgamma(exp(x), 5.1989,16.6192)/3.1721, geom="line",color="red")
-pPT <- ggplot(logsig_df, aes(x=sigmaPT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaPT)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[2], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(exp(x), 3.5539, 5.5324)/1.33697, geom="line",color="red")
-pT <- ggplot(logsig_df, aes(x=sigmaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaT)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[3], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(exp(x), 6.8383, 25.7560)/4.1923, geom="line",color="red")
+pP <- ggplot(logsig_df, aes(x=sigmaP)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaP)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[1], linetype="dotted") + theme_bw()  + stat_function(fun=function(x) dgamma(exp(x), 2.2298, 4.8674)/1.6384, geom="line",color="red")
+pPT <- ggplot(logsig_df, aes(x=sigmaPT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaPT)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[2], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(exp(x), 1.8166, 1.7689)/0.70791, geom="line",color="red")
+pT <- ggplot(logsig_df, aes(x=sigmaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaT)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[3], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(exp(x), 2.6144, 7.1221)/2.1250, geom="line",color="red")
 # and log ratios

### 2020-09-25 Results Archive/StanResultsSummary_relaxedinvgamma.Rmd
updated code

--- 2020-09-25 Results Archive/StanResultsSummary_relaxedgamma5.Rmd
@@
 Summary of Stan Model Fit Results: Full-Scale (850K sites)
-\newline Relaxed Gamma priors (using relaxed gamma fits with x5 variance)
+\newline Inverse-Gamma priors (using relaxed inv-gamma fits with x2 variance)
 ```{r Setup, echo=FALSE, include=FALSE}
@@
 load("myFA_bulkonly.Rdata")
-load("FullResultsTCGA_relaxedgamma5.Rdata")
+load("FullResultsTCGA_relaxedinvgamma.Rdata")
 
@@
 fBG <- function(x, a, m1, s1, m2, s2) a*dnorm(x,m1,s1) + (1-a)*dnorm(x,m2,s2)
+dIG <- function(x, a, b) 1/(b*gamma(a))*(b/x)^(a+1)*exp(-b/x)
 pM <- ggplot(data.frame(mu=mu_full$p50), aes(x=mu)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("Mu") + xlim(c(-5,5)) + geom_vline(xintercept=mean(mu_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197)/0.1906566,geom="line",color="red")
-pB <- ggplot(data.frame(betaT=betaT_full$p50), aes(x=betaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("BetaT") + xlim(c(-5,5)) + geom_vline(xintercept=mean(betaT_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) dcauchy(x,-0.04051, 0.96650)/0.32934,geom="line",color="red")
-pE <- ggplot(data.frame(sigmaE=sigmaE_full$p50), aes(x=sigmaE)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaE") + xlim(c(0,2)) + geom_vline(xintercept=mean(sigmaE_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x,2,2)/0.7357589, geom="line",color="red")
-pP <- ggplot(sigma_df, aes(x=sigmaP)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaP") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[1], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x, 2.2298, 4.8674)/1.6384, geom="line",color="red")
-pPT <- ggplot(sigma_df, aes(x=sigmaPT)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaPT") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[2], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x, 1.8166, 1.7689)/0.70791, geom="line",color="red")
-pT <- ggplot(sigma_df, aes(x=sigmaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaT") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[3], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(x, 2.6144, 7.1221)/2.1250, geom="line",color="red")
+pB <- ggplot(data.frame(betaT=betaT_full$p50), aes(x=betaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("BetaT") + xlim(c(-5,5)) + geom_vline(xintercept=mean(betaT_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) dcauchy(x,-0.04051, 0.38660)/0.823357,geom="line",color="red")
+pE <- ggplot(data.frame(sigmaE=sigmaE_full$p50), aes(x=sigmaE)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaE") + xlim(c(0,2)) + geom_vline(xintercept=mean(sigmaE_full$p50), linetype="dotted") + theme_bw() + stat_function(fun=function(x) dIG(x,1,1)/0.54134, geom="line",color="red")
+pP <- ggplot(sigma_df, aes(x=sigmaP)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaP") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[1], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dIG(x, 3.7350, 1.0694)/2.9791,geom="line",color="red")
+pPT <- ggplot(sigma_df, aes(x=sigmaPT)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaPT") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[2], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dIG(x, 2.4884, 1.2601)/1.4368,geom="line",color="red")
+pT <- ggplot(sigma_df, aes(x=sigmaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.05) + ggtitle("sigmaT") + xlim(c(0,2)) + geom_vline(xintercept=sigmaMeans[3], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dIG(x, 4.8358, 1.1529)/3.9851,geom="line",color="red")
 
@@
 # plot the log sigmas
-pP <- ggplot(logsig_df, aes(x=sigmaP)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaP)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[1], linetype="dotted") + theme_bw()  + stat_function(fun=function(x) dgamma(exp(x), 2.2298, 4.8674)/1.6384, geom="line",color="red")
-pPT <- ggplot(logsig_df, aes(x=sigmaPT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaPT)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[2], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(exp(x), 1.8166, 1.7689)/0.70791, geom="line",color="red")
-pT <- ggplot(logsig_df, aes(x=sigmaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaT)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[3], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dgamma(exp(x), 2.6144, 7.1221)/2.1250, geom="line",color="red")
+pP <- ggplot(logsig_df, aes(x=sigmaP)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaP)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[1], linetype="dotted") + theme_bw()  + stat_function(fun=function(x) dIG(exp(x), 3.7350, 1.0694)/2.9791, geom="line",color="red")
+pPT <- ggplot(logsig_df, aes(x=sigmaPT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaPT)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[2], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dIG(exp(x), 2.4884, 1.2601)/1.4368, geom="line",color="red")
+pT <- ggplot(logsig_df, aes(x=sigmaT)) + geom_histogram(aes(y=..ncount..), binwidth=0.1) + ggtitle("log(sigmaT)") + xlim(c(-5,5)) + geom_vline(xintercept=logsigMeans[3], linetype="dotted") + theme_bw() + stat_function(fun=function(x) dIG(exp(x), 4.8358, 1.1529)/3.9851, geom="line",color="red")
 # and log ratios

### 2020-09-25 StanAnalysis_updated.Rmd
updated code

--- 2020-09-25 Paper_Figures.Rmd
@@
 ---
-title: "DNA Methylation Hierarchical Variance Model\nPaper Figures"
+title: "DNA Methylation Hierarchical Variance Model\nStan Results"
 author: "Kevin Murgas"
@@
     toc: true
+params:
+  prep: true
 ---
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-library(tidyverse)
-library(gt)
-library(rlang)
-library(ggpubr)
-library(knitr)
-
-# set working directory
-setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
-library(here)
-
-select <- dplyr::select
-rename <- dplyr::rename
-filter <- dplyr::filter
-slice <- dplyr::slice
-
-set.seed(1234)
-```
-
-```{r Load_data, echo=FALSE, include=TRUE, message=FALSE}
-# Load raw methylation data
-load(here("myFA_bulkonly.Rdata"))
-
-stanFile = "FullResultsTCGA_relaxgamma3_postsum3.Rdata"
-load(here(stanFile))
-
-prepdir <- "data-stan_rg3ps3"
-cat(sprintf("Load prepped data dir: %s (needs to be pre-processed)...",prepdir))
-gened_data <- read_csv(here(prepdir, "gened_data.csv"), col_types = list("CHR" = col_character()))
-prepped_data <- read_csv(here(prepdir, "prepped_data.csv"))
-permed_data <- read_csv(here(prepdir, "permed_data.csv"))
-
-```
-
-\newpage
-\section{Figure 1: Model Overview, Priors, Example Site Fit}
+\newpage
+\section{Model Overview}
 \subsection{Data}
@@
 
-```{r Priors, echo=FALSE, fig.fullwidth=TRUE, warning=FALSE, comment=NA}
-# load TCGA estimates for mu, betaT, sigmaP, sigmaPT, sigmaT
-load(file="/Users/kevinmurgas/Documents/Data+ project/TCGA data/estimates_allpats.Rdata")
-
-# remove all zeros, and take square root of variance (stored variable) to get sigma/stdev
-sigmaP_ests <- sqrt(sigmaP_ests[which(!!sigmaP_ests)])
-sigmaPT_ests <- sqrt(sigmaPT_ests[which(!!sigmaPT_ests)])
-sigmaT_ests <- sqrt(sigmaT_ests[which(!!sigmaT_ests)])
-mu_ests <- mu_ests[which(!!mu_ests)]
-betaT_ests <- betaT_ests[which(!!betaT_ests)]
-
-# for each parameter, plot the distribution and overlay the fits
-fBG <- function(x, a, m1, s1, m2, s2) a*dnorm(x,m1,s1) + (1-a)*dnorm(x,m2,s2) # bimodal gaussian
-post_prior_histogram <- function(vals,priorfun1,priorfun2,bw,xl,titlestr,legFlag=FALSE) {
-  # stacked histogram colored by gene-associated sites
-  # and prior curve overlaid
-  plot_df <- data.frame(val=vals)
-  p1 <- ggplot(plot_df, aes(x=val)) +
-    geom_histogram(aes(y=..count../(length(vals)*bw), fill="Empirical"), binwidth=bw) +
-    stat_function(fun=function(x) priorfun1(x),geom="line",aes(color="Fit")) +
-    stat_function(fun=function(x) priorfun2(x),geom="line",aes(color="3x Relax")) +
-    ggtitle(titlestr) + xlim(xl) + theme_classic() + xlab(titlestr) + ylab("Freq.") +
-    scale_colour_manual(NULL, values = c("Empirical" = "gray","Fit" ="red","3x Relax" = "blue")) +
-    scale_fill_manual(NULL,values = c("Empirical"="gray")) +
-    theme(legend.position = c(0.75, 0.65))
-  if (!legFlag) {p1 <- p1 + theme(legend.position = "none")}
-  p1
-}
-
-# plot mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT with priors overlaid
-pM <- post_prior_histogram(mu_ests, function(x) fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), function(x) fBG(x,0.3557,-3.0675, sqrt(3)*0.75214,1.1290, sqrt(3)*1.39197), 0.1, c(-5,5), "Mu")
-pB <- post_prior_histogram(betaT_ests, function(x) dcauchy(x,-0.04051, 0.19330), function(x) dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "BetaT")
-pP <- post_prior_histogram(sigmaP_ests, function(x) dgamma(x, 5.1989, 16.6192), function(x) dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
-pPT <- post_prior_histogram(sigmaPT_ests, function(x) dgamma(x, 3.5539, 5.5324), function(x) dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
-pT <- post_prior_histogram(sigmaT_ests, function(x) dgamma(x, 6.8383, 25.7560), function(x) dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT",legFlag=TRUE)
-pE <- ggplot() + stat_function(fun=function(x) dgamma(x, 2, 2),geom="line",color="blue") + ggtitle("SigmaE") + xlim(c(0,2)) + theme_classic() + theme(legend.position = "none") + xlab("SigmaE") + ylab("Freq.")
-
-pBlank <- ggplot(data=data.frame(len=c(1,2)),aes(x=len)) + theme_void() + geom_bar(aes(fill="Empirical")) +
-    stat_function(fun=function(x) 3,geom="line",aes(color="1x variance")) +
-    stat_function(fun=function(x) 4,geom="line",aes(color="3x variance")) + xlim(0,1) + ylim(0,1) + scale_colour_manual(NULL, values = c("Empirical" = "gray","1x variance" ="red","3x variance" = "blue")) + scale_fill_manual(NULL,values = "gray") + theme(legend.position = c(2, 0.75))
-
-p_all <- ggarrange(pM,pB,pE,pP,pPT,pT,hjust=0,nrow=2,ncol=3)
-print(p_all)
-rm(pM,pB,pBlank,pP,pPT,pT,pE)
-
-```
-
-\subsection{CpG Conservation score}
+\subsection{Conservation score}
 We want to understand which sites of DNA methylation are fundamentally conserved within tumors, i.e.
@@
 Therefore we choose the log-ratio of sigmaP/sigmaT to represent a conservation score, which is positive when tumor variation is lower than normal variation.
-We take the median of the posterior distribution for this parameter.
-
+We take the median of the posterior distribution for this parameter. Additionally, we calculate a regularization term based on the inverse of the L2 norm of sigmaP and sigmaT.
+
+<!--  $$ \text{score}_\text{prob} = \Pr (\text{log}\frac{\sigma_{P}}{\sigma_T} > 0) = \int \text{log}\frac{\sigma_{P}}{\sigma_T} > 0 ~ dp $$
+$$ \text{score}_\text{mean} = \text{Mean} (\text{log}\frac{\sigma_{P}}{\sigma_T}) ,    ~ -->
 $$\text{score}_\text{med} = \text{Median} (\text{log}\frac{\sigma_{P}}{\sigma_T}) $$
 
-\newpage
-\subsection{Example site fit}
-![Fit](/Users/kevinmurgas/Documents/Data+ project/Paper Figures/example_fit.png)
-\newline \newline
-![Posteriors](/Users/kevinmurgas/Documents/Data+ project/Paper Figures/example_posteriors.png)
-\newpage
-\section{Figure 2: Model Fit Results}
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+library(tidyverse)
+library(gt)
+library(rlang)
+library(ggpubr)
+library(knitr)
+
+# set working directory
+setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
+library(here)
+
+select <- dplyr::select
+rename <- dplyr::rename
+filter <- dplyr::filter
+
+set.seed(1234)
+```
+
+```{r Load, include=FALSE}
+# Load raw methylation data
+load(here("myFA_bulkonly.Rdata"))
+
+stanFile = "FullResultsTCGA_relaxgamma3_postsum3.Rdata"
+load(here(stanFile))
+```
+
+\newpage
+\section{Fit Results}
 \subsection{Parameter Distributions}
@@
 pLPT <- post_prior_histogram(log(sigmaPT_full$p50), genemask, function(x) 0.7*dgamma(exp(x), 2.1457, 2.4819), 0.1, c(-5,5), "log(SigmaPT)")
-pLT <- post_prior_histogram(log(sigmaT_full$p50), genemask, function(x) 0.4*dgamma(exp(x), 3.3643, 10.430), 0.1, c(-5,5), "log(SigmaT)") + theme(legend.position = c(0.8, 0.6), legend.text=element_text(size=6), legend.title=element_text(size=6))
+pLT <- post_prior_histogram(log(sigmaT_full$p50), genemask, function(x) 0.4*dgamma(exp(x), 3.3643, 10.430), 0.1, c(-5,5), "log(SigmaT)") + theme(legend.position = c(0.8, 0.6))
 
@@
 
-\subsection{Conservation score}
+\subsection{Conservation scores}
 ```{r Fig1B, echo=FALSE, fig.height=2, fig.width=6.5, fig.fullwidth=TRUE, warning=FALSE}
@@
 pP1 <- post_histogram(prob1_full$median,genemask,0.05,c(-3,3),expression("Median log("~sigma[P]/sigma[T]~")"))
-#prob1_proxy = log(sigmaP_full$p50/sigmaT_full$p50)
-#pP2 <- post_histogram(prob1_proxy,genemask,0.05,c(-3,3),expression(" log( median" ~sigma[P]/~"median"~sigma[T]~")"))
-p_scores <- pP1
-print(p_scores)
+prob1_proxy = log(sigmaP_full$p50/sigmaT_full$p50)
+pP2 <- post_histogram(prob1_proxy,genemask,0.05,c(-3,3),expression(" log( median" ~sigma[P]/~"median"~sigma[T]~")"))
+# regularized version
+L2norm = sqrt(sigmaP_full$p50^2 + sigmaT_full$p50^2)
+prob1_reg = log(sigmaP_full$p50/sigmaT_full$p50) - 1/L2norm
+pP3 <- post_histogram(prob1_reg,genemask,0.1,c(-10,5),"Regularized")
+p_probs <- ggarrange(pP1,pP2,pP3,nrow=1,ncol=3)
+print(p_probs)
 ```
@@
 
-\newpage
-\subsection{Intra-gene conservation}
-We examine intra gene conservation 2 ways: a table looking at gene regulatory regions, and a CpG-distance analysis
-
-\subsection{Regulatory regions}
-Table comparing different groups of CpG sites:
-```{r Table1, echo=FALSE, warning=FALSE}
-
-#, include=TRUE, results='asis', fig.keep='all', fig.show='asis', fig.height=3, fig.width=5
-
-prepped_data %>%
-  group_by(`Island relation` = Relation_to_UCSC_CpG_Island) %>%
-  summarise_at(vars(sigmaP:sigmaT,score_median), mean, na.rm = TRUE) %>%
-  dplyr::rename(ir = `Island relation`) %>%
-  mutate(ir = case_when(is.na(ir) ~ "Sea",
-                        str_detect(ir, "^S") ~ str_replace(ir, "^S_", "South "),
-                        str_detect(ir, "^N") ~ str_replace(ir, "^N_", "North "),
-                        TRUE ~ ir)) %>%
-  dplyr::rename(`Island relation` = ir) %>%
-  arrange(c(1,4,2,5,3,6)) %>% # for island, N shore, S shore, N shelf, S shelf, sea
-  rename(`CpG Conservation Score`=score_median) %>%
-  gt() %>%
-  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, `CpG Conservation Score`), decimals = 3) %>%
-  tab_spanner(label = "Hierarchical Variances", columns = vars(sigmaP, sigmaPT, sigmaT)) %>% cols_align("center")
-
-summarize_conservation <- function(data,str) {
-  indsin <- str_detect(data$UCSC_RefGene_Group, str) & !is.na(data$UCSC_RefGene_Group)
-  indsout <- !str_detect(data$UCSC_RefGene_Group, str) | is.na(data$UCSC_RefGene_Group)
-  if (str=="any") {
-    indsin <- !is.na(data$UCSC_RefGene_Group)
-    indsout <- is.na(data$UCSC_RefGene_Group)
-  }
-  data.frame(cons_in=mean(data$score_median[indsin], na.rm=TRUE),
-         cons_out=mean(data$score_median[indsout],na.rm=TRUE),
-         nsites=sum(indsin),
-         pval=wilcox.test(data$score_median[indsin],data$score_median[indsout])$p.value)
-}
-
-bind_rows({prepped_data %>% summarize_conservation("any") %>% mutate(region="Gene-Associated")},
-          {prepped_data %>% summarize_conservation("TSS1500") %>% mutate(region="TSS1500")},
-          {prepped_data %>% summarize_conservation("TSS200") %>% mutate(region="TSS200")},
-          {prepped_data %>% summarize_conservation("5'UTR|5UTR|5URT") %>% mutate(region="5'UTR")},
-          {prepped_data %>% summarize_conservation("1stExon") %>% mutate(region="1stExon")},
-          {prepped_data %>% summarize_conservation("Body") %>% mutate(region="Body")},
-          {prepped_data %>% summarize_conservation("3'UTR|5UTR") %>% mutate(region="3'UTR")}) %>%
-  select(c(region,cons_in,cons_out,nsites,pval)) %>%
-  gt() %>%
-  fmt_number(columns = vars(cons_in,cons_out,pval), decimals = 3)
-```
-
-\newpage
-\subsection{Figure 1: CpG Distance Analysis}
-Corrected for gene direction, based on Ensembl gene database \newline
-```{r Fig1, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=3, comment=NA}
-# load ensembl db
-library(EnsDb.Hsapiens.v75) # ensembl annotation with gene position (hg19 version)
-edb <- EnsDb.Hsapiens.v75
-geneKeys <- keys(edb, keytype="GENENAME") %>% intersect(unique(gened_data$UCSC_RefGene_Name))
-edbData<- AnnotationDbi::select(edb, keys=geneKeys, keytype="GENENAME",
-                                columns=c("GENEBIOTYPE", "GENESEQSTART", "GENESEQEND", "SEQNAME", "SEQSTRAND")) %>%
-  filter(SEQNAME %in% c(as.character(1:22),"X","Y")) %>%
-  mutate(GENESTART = GENESEQSTART*(SEQSTRAND==1) + GENESEQEND*(SEQSTRAND==-1)) %>%
-  select(-c(GENESEQSTART,GENESEQEND))
-
-# calculate CpG distance based on Ensembl gene start position and direction
-distanced_data <- gened_data %>%
-  filter(!is.na(UCSC_RefGene_Name)) %>%
-  filter(UCSC_RefGene_Name %in% geneKeys) %>%
-  mutate(EDBCHR=edbData$SEQNAME[match(UCSC_RefGene_Name,edbData$GENENAME)],
-         EDBSTART=edbData$GENESTART[match(UCSC_RefGene_Name,edbData$GENENAME)],
-         EDBDIR=edbData$SEQSTRAND[match(UCSC_RefGene_Name,edbData$GENENAME)],
-         EDBTYPE=edbData$GENEBIOTYPE[match(UCSC_RefGene_Name,edbData$GENENAME)]
-         ) %>%
-  filter(CHR==EDBCHR) %>%
-  mutate(cpgdist = if_else(EDBDIR==1, MAPINFO-EDBSTART, EDBSTART-MAPINFO))
-
-distance_length <- distanced_data %>%
-  filter(!is.na(UCSC_RefGene_Name)) %>%
-  select(CHR, UCSC_RefGene_Name, cpgdist, c(sigmaP:sigmaT,score_median:score_reg)) %>%
-  mutate(location_bin = cut(cpgdist, c(seq(-1500, 20000, by = 100)), include.lowest = TRUE)) %>%
-  group_by(location_bin) %>%
-  summarise(mean_sigmaP = mean(sigmaP), mean_sigmaPT = mean(sigmaPT), mean_sigmaT = mean(sigmaT),  mean_score_median = mean(score_median), mean_score_median2 = mean(score_median2), n_obs = n(), .groups='drop') %>%
-  mutate(location_bin = as.numeric(str_extract(location_bin, "(?<=,)(.*)(?=])")))
-
-distance_length %>%
-  drop_na() %>%
-  pivot_longer(mean_sigmaP:mean_sigmaT) %>% 
-  mutate(name = factor(name, c("mean_sigmaP","mean_sigmaPT","mean_sigmaT"), c("SigmaP","SigmaPT","SigmaT"))) %>%
-  sample_frac() %>%
-  ggplot(aes(location_bin, value, color = name, fill = name)) +
-  geom_point(key_glyph = draw_key_rect) +
-  ggforce::facet_zoom(location_bin < 5000) +
-  theme_light() +
-  theme(legend.position = "bottom") +
-  labs(title = NULL,
-       fill = NULL,
-       color = NULL,
-       x = "Distance from gene start site (binned to 100 bp)",
-       y = "Average Sigma Value", 
-       caption = "Figure 1: Average Sigma values of single CpGs as a function of position relative to gene.") +
-  scale_color_brewer(palette = "Set2") +
-  scale_fill_brewer(palette = "Set2")
-cat("\n")
-
-# again for conservation scores
-distance_length %>%
-  drop_na() %>%
-  pivot_longer(mean_score_median) %>% 
-  mutate(name = factor(name, c("mean_score_median"), c("Score_median"))) %>%
-  sample_frac() %>%
-  ggplot(aes(location_bin, value, color = name, fill = name)) +
-  geom_point(key_glyph = draw_key_rect) +
-  ggforce::facet_zoom(location_bin < 5000) +
-  theme_light() +
-  theme(legend.position = "bottom") +
-  labs(title = NULL,
-       fill = NULL,
-       color = NULL,
-       x = "Distance from gene start site (binned to 100 bp)",
-       y = "Average Conservation Score", 
-       caption = "Figure 1: Average conservation scores of single CpGs as a function of position relative to gene.") +
-  scale_color_brewer(palette = "Set2") +
-  scale_fill_brewer(palette = "Set2")
-cat("\n")
-
-# and look at number of CpG in each bin
-distance_length %>%
-  drop_na() %>%
-  pivot_longer(n_obs) %>%
-  mutate(name = factor(name, c("n_obs"), c("n_obs"))) %>%
-  sample_frac() %>%
-  ggplot(aes(location_bin, value, color = name, fill = name)) +
-  geom_point(key_glyph = draw_key_rect) +
-  ggforce::facet_zoom(location_bin < 5000) +
-  theme_light() +
-  theme(legend.position = "bottom") +
-  labs(title = NULL,
-       fill = NULL,
-       color = NULL,
-       x = "Distance from gene start site (binned to 100 bp)",
-       y = "Number of CpGs", 
-       caption = "Figure 1: Frequency of single CpGs as a function of position relative to gene start site.") +
-  scale_color_brewer(palette = "Set2") +
-  scale_fill_brewer(palette = "Set2")
-```
-
-\newpage
-\section{Gene Conservation Scores}
-Genes are scored by average of all CpG sites within gene
-\subsection{Example Genes: APC, TP53, TTN, HLA-A, GAPDH}
-```{r examplegenes, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=2, comment=NA}
-# for these genes, show the CpG site distribution, and add vertical lines for the CpG sites in the gene (gened_data?), then a bold yellow vertical line for the overall mean
-
-plotGeneScore <- function(geneStr) {
-  CpGscores <- gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^",geneStr,"$"))) %>% pull(score_median)
-
-  ggplot(prepped_data) + geom_histogram(aes(x=score_median, y=..count../nrow(prepped_data)), binwidth=0.05) + xlim(-3,3) + ggtitle(sprintf("%s: CpG scores (%i sites)",geneStr, length(CpGscores))) + xlab("CpG conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") + geom_vline(xintercept=CpGscores, color="red", alpha=0.5, size=0.2) + geom_vline(xintercept=mean(CpGscores), color="blue", alpha=0.8, size=2)
-}
-
-cat("\n")
-plotGeneScore("TP53")
-cat("\n")
-plotGeneScore("TTN")
-cat("\n")
-plotGeneScore("HLA-A")
-```
-
-\subsection{COSMIC Cancer Genes}
-```{r cancergenes, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=2, comment=NA}
-# make gene-wise averages
+\newpage Prep data
+```{r prep, echo=FALSE, include=TRUE, eval=params$prep, comment=NA}
+cat("Param:prep=TRUE -> prepare data...")
+
+med_vals <- tibble(Name=FullAnnotation[,1], mu=mu_full$p50, betaT=betaT_full$p50, sigmaE=sigmaE_full$p50, sigmaP=sigmaP_full$p50, sigmaPT=sigmaPT_full$p50, sigmaT=sigmaT_full$p50, Rhat=lp_full$Rhat) %>% mutate(logP=log(sigmaP), logPT=log(sigmaPT), logT=log(sigmaT)) %>%
+  mutate(L2norm = sqrt(sigmaP^2 + sigmaT^2)) %>%
+  mutate(score_median=prob1_full$median, score_median2=log(sigmaP/sigmaT), score_reg=log(sigmaP/sigmaT * L2norm))
+# score_prob=prob1_full$prob, score_mean=prob1_full$mean,
+
+# define filter cutoffs
+Rhat_cutoff=1.1 # maximum Rhat cutoff
+n_cutoff=5 # cutoff for gene minimum number of CpG sites
+
+# prep data by joining with FullAnnotation (columns: IlmnID, CHR, MAPINFO, RefGene_Name+Group, CpG_Island)
+cat("Prep data with annotation, filter out sites Rhat>1.1 (and save)")
+ptm<-proc.time()
+prepped_data <- tibble(FullAnnotation[,c(1,2,3,5,6,8)]) %>% rename(Name = IlmnID) %>%
+  left_join(by="Name", med_vals) %>% filter(Rhat<Rhat_cutoff)
+write_csv(prepped_data, here("data-stan_rg3ps3", "prepped_data.csv"))
+cat(proc.time()-ptm)
+
+# separate by rows by genes
+cat("Separate by gene")
+ptm<-proc.time()
+gened_data <- prepped_data %>%
+  arrange(CHR, MAPINFO) %>%
+  separate_rows(UCSC_RefGene_Name, sep=";") %>% 
+  distinct()
+write_csv(gened_data, here("data-stan_rg3ps3", "gened_data.csv"))
+cat(proc.time()-ptm)
+
 gene_summarize <- function(data, id, value, fun = mean, ...) {
@@
 
+cat(paste0("Summarize by gene, taking mean value of all sites in gene, filter out sites n<",n_cutoff))
+ptm<-proc.time()
 gene_summary <- gened_data %>%
   filter(!is.na(UCSC_RefGene_Name)) %>%
-  gene_summarize(UCSC_RefGene_Name, c(score_median))
-
-cancergenes <- read_csv("/Users/kevinmurgas/Documents/Data+ project/Cancer gene lists/SangerCensus_COSMIC_CRC.csv")
-genescores <- gene_summary %>% filter(UCSC_RefGene_Name %in% (cancergenes %>% pull(`Gene Symbol`))) %>% pull(score_median)
-ggplot(gene_summary) + geom_histogram(aes(x=score_median, y=..count../nrow(gene_summary)), binwidth=0.05) + xlim(-3,3) + ggtitle(sprintf("Gene Scores of %i CRC Genes", length(genescores))) + xlab("Gene conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") + geom_vline(xintercept=genescores, color="blue", alpha=0.5, size=0.2)
-cat("\n")
-
-# source 2
-cancergenes <- read_csv("/Users/kevinmurgas/Documents/Data+ project/Cancer gene lists/CRCgenes_atlasgeneticsoncology.csv")
-genescores <- gene_summary %>% filter(UCSC_RefGene_Name %in% (cancergenes %>% pull(CancerGenes))) %>% pull(score_median)
-ggplot(gene_summary) + geom_histogram(aes(x=score_median, y=..count../nrow(gene_summary)), binwidth=0.05) + xlim(-3,3) + ggtitle(sprintf("Gene Scores of %i CRC Genes", length(genescores))) + xlab("Gene conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") + geom_vline(xintercept=genescores, color="blue", alpha=0.1, size=0.2)
-```
-
-\newpage
-\section{Gene Set Conservation Analysis}
-```{r load_gene_sets, include=FALSE}
-# load in the C5 gene sets from MSigDB
-library(GSA)
-gmtfile <- file.path("/Users/kevinmurgas/Documents/Data+ project/EPIC data","c2.cp.v7.1.symbols.gmt")
-genesets <- GSA.read.gmt(gmtfile)
-```
-
-```{r GSCA, echo=FALSE, comment=NA, fig.width=8}
-# make a data.frame for gene sets and scores
-nsets <- length(genesets[[1]])
-GSEA_df <- data.frame(GeneSet=genesets[[2]][1:nsets], score=numeric(nsets),
-                      num_gene = unlist(lapply(genesets[[1]][1:nsets],length)),
-                      num_gene_cut = numeric(nsets))
-
-# have to order all genes first by rank of conservation (conserved genes come first)
-genenames = permed_data$UCSC_RefGene_Name
-ngene = length(genenames)
-
-# reorder by evidence rank
-ind = order(-permed_data$score_median) # use ranked conservation score as evidence
-evidence = seq(ngene,1,-1)
-genenames = genenames[ind]
-
-# plot evidence ranked: permed_data$score_median
-ranks <- permed_data$score_median
-names(ranks) <- permed_data$UCSC_RefGene_Name
-barplot(sort(ranks, decreasing = T))
-
-ptm <- proc.time()
-for (iset in 1:nsets) {
-  # take only genes that we have in full gene list
-  set <- unlist(genesets[[1]][iset]) %>% intersect(genenames)
-  GSEA_df$num_gene_cut[iset] = length(set)
-
-  # skip if num_gene <10 or >500
-  if (length(set)<10 | length(set)>500) {next}
-
-  # compute running sum (as PDF first)
-  runsum_in = numeric(ngene)
-  runsum_out = numeric(ngene)
-  for (i in 1:ngene) {
-    if (genenames[i] %in% set) {
-      # evidence = permed_data$median, or ranks
-      runsum_in[i] = evidence[i]
-      runsum_out[i] = 0
-    } else {
-      runsum_in[i] = 0
-      runsum_out[i] = 1
-    }
-  }
-  # convert to CDFs
-  runsum_in <- cumsum(runsum_in)/sum(runsum_in)
-  runsum_out <- cumsum(runsum_out)/sum(runsum_out)
-
-  # compute K-S-like statistic
-  v <- runsum_in - runsum_out
-  score <- max(c(0,v)) + min(c(0,v))
-  GSEA_df$score[iset] <- score
-}
+  gene_summarize(UCSC_RefGene_Name, c(mu:sigmaT,score_median:score_reg)) %>% filter(n>=n_cutoff)
 cat(proc.time()-ptm)
-cat(sprintf("Processed %i genes, %i pathways. Top 50 paths:",ngene,nsets))
-
-# insert significance bootstrapping
-
-# make table of top scores
-GSEA_df %>% arrange(-score) %>%
-  select(score,num_gene,GeneSet) %>%
-  slice(1:50) %>%
-  mutate(GeneSet = gsub('(.{1,90})(\\s|$)', '\\1\n', GeneSet)) %>%
-  gt()
+
+boot_naive <- function(data, data_full, value, id, reps) {
+  lengths <- sort(unique(data$n))
+  varname <- paste0(as_name(ensym(value)), "_v1")
+
+  # perm_v1_inner fxn: draw n_rep samples of n_sites CpG sites, randomly out of all sites
+  # take mean of each draw
+  sample_naive <- function(n_sites, n_rep, data) {
+    replicate(n = n_rep, expr = mean(sample(data, as.numeric(n_sites), TRUE)))
+  }
+
+  # run naive_sample on all lengths to get res (list of null distributions for each length)
+  # then map a percentile to each gene based on null distrib. of that length
+  names(lengths) <- lengths
+  res <- lapply(lengths, sample_naive, n_rep = reps, data = pull(data_full, {{value}}))
+  values <- map2_dbl(pull(data, {{value}}), data$n, ~ mean(res[[as.character(.y)]] < .x))
+
+  data %>% mutate(!!varname := values)
+}
+
+boot_adj <- function(data, data_full, value, id, reps) {
+  lengths <- sort(unique(data$n))
+  varname <- paste0(as_name(ensym(value)), "_v3")
+
+  sample_adj <- function(n_sites, n_rep, data, id, value) {
+    perm3_data <- data %>% group_by({{id}}) %>% mutate(left = n() - row_number() + 1)
+    starting_index <- sample(which(n_sites <= perm3_data$left), n_rep, replace = TRUE)
+    map_dbl(starting_index, ~ mean(pull(perm3_data, {{value}})[.x + seq_len(n_sites) - 1]))
+  }
+
+  names(lengths) <- lengths
+  res <- lapply(lengths, sample_adj, n_rep = reps, data = data_full, id = {{id}}, value = {{value}})
+  values <- map2_dbl(pull(data, {{value}}), data$n, ~ mean(res[[as.character(.y)]] < .x))
+
+  data %>% mutate(!!varname := values)
+}
+
+cat("Permutate data with bootstrapping")
+ptm<-proc.time()
+# 2nd function of methcon5
+permed_data <- gene_summary %>%
+  filter(n < quantile(n, 0.99)) %>%
+  boot_naive(gened_data, sigmaP, UCSC_RefGene_Name, 1000) %>%
+  boot_naive(gened_data, sigmaPT, UCSC_RefGene_Name, 1000) %>%
+  boot_naive(gened_data, sigmaT, UCSC_RefGene_Name, 1000) %>%
+  boot_naive(gened_data, score_median, UCSC_RefGene_Name, 1000) %>%
+  boot_naive(gened_data, score_median2, UCSC_RefGene_Name, 1000) %>%
+  boot_naive(gened_data, score_reg, UCSC_RefGene_Name, 1000) %>%
+  boot_adj(gened_data, sigmaP, UCSC_RefGene_Name, 1000) %>%
+  boot_adj(gened_data, sigmaPT, UCSC_RefGene_Name, 1000) %>%
+  boot_adj(gened_data, sigmaT, UCSC_RefGene_Name, 1000) %>%
+  boot_adj(gened_data, score_median, UCSC_RefGene_Name, 1000) %>%
+  boot_adj(gened_data, score_median2, UCSC_RefGene_Name, 1000) %>%
+  boot_adj(gened_data, score_reg, UCSC_RefGene_Name, 1000)
+write_csv(permed_data, here("data-stan_rg3ps3", "permed_data.csv"))
+cat(proc.time()-ptm)
+#  boot_naive(gened_data, score_prob, UCSC_RefGene_Name, 1000) %>%
+#  boot_naive(gened_data, score_mean, UCSC_RefGene_Name, 1000) %>%
+
+# look at mean methylation conservation in promoters
+cat("Calculate mean methylation conservation in promoters")
+ptm<-proc.time()
+promoted_data <- prepped_data %>%
+  mutate(gene = str_extract(UCSC_RefGene_Name, "^[^;]+")) %>%
+  filter(!is.na(gene)) %>%
+  left_join(read_csv(here("data-raw","MethylationEPIC_v-1-0_B4.csv"), skip = 7) %>%  
+              select(Name, Regulatory_Feature_Group), 
+            by = "Name") %>%
+  filter(Regulatory_Feature_Group == "Promoter_Associated") %>%
+  group_by(gene) %>%
+  summarise(sigmaP_pro = mean(sigmaP), sigmaPT_pro = mean(sigmaPT), sigmaT_pro = mean(sigmaT), 
+            score_median_pro = mean(score_median), score_median2_pro = mean(score_median2), score_reg_pro = mean(score_reg))
+write_csv(promoted_data, here("data-stan_rg3ps3", "promoted_data.csv"))
+cat(proc.time()-ptm)
+cat("Done with data preparation (BOTH)")
 ```
@@
 knit_exit()
+```
+
+```{r load_prep, echo=FALSE, include=TRUE, message=FALSE, eval=!params$prep}
+cat("Param:prep=FALSE -> load saved data (BOTH)...")
+gened_data <- read_csv(here("data-stan_rg3ps3", "gened_data.csv"), col_types = list("CHR" = col_character()))
+prepped_data <- read_csv(here("data-stan_rg3ps3", "prepped_data.csv"))
+permed_data <- read_csv(here("data-stan_rg3ps3", "permed_data.csv"))
+promoted_data <- read_csv(here("data-stan_rg3ps3", "promoted_data.csv"))
+cat("done")
 ```
@@
 \newpage
+\subsection{Tables comparing different groups of CpG sites:}
+```{r Table1, echo=FALSE, warning=FALSE}
+#, include=TRUE, results='asis', fig.keep='all', fig.show='asis', fig.height=3, fig.width=5
+# Table 1
+prepped_data %>%
+  group_by(`Gene Associated` = !is.na(UCSC_RefGene_Name)) %>%
+  summarise_at(vars(sigmaP:sigmaT,score_median:score_reg), mean, na.rm = TRUE) %>%
+  mutate(`Gene Associated` = factor(`Gene Associated`, c(TRUE, FALSE), c("Yes", "No"))) %>%
+  arrange(rev(rownames(.))) %>%
+  gt() %>%
+  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, score_median, score_median2, score_reg), decimals = 3) %>%
+  tab_spanner(label = "Hierarchical Variance", columns = vars(sigmaP, sigmaPT, sigmaT)) %>%
+  tab_spanner(label = "Conservation Scores", columns = vars(score_median, score_median2, score_reg))
+#score_prob, score_mean,
+
+prepped_data %>%
+  group_by(`Island relation` = Relation_to_UCSC_CpG_Island) %>%
+  summarise_at(vars(sigmaP:sigmaT,score_median:score_reg), mean, na.rm = TRUE) %>%
+  dplyr::rename(ir = `Island relation`) %>%
+  mutate(ir = case_when(is.na(ir) ~ "Sea",
+                        str_detect(ir, "^S") ~ str_replace(ir, "^S_", "South "),
+                        str_detect(ir, "^N") ~ str_replace(ir, "^N_", "North "),
+                        TRUE ~ ir)) %>%
+  dplyr::rename(`Island relation` = ir) %>%
+  arrange(rev(rownames(.))) %>%
+  gt() %>%
+  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, score_median, score_median2, score_reg), decimals = 3) %>%
+  tab_spanner(label = "Hierarchical Variance", columns = vars(sigmaP, sigmaPT, sigmaT)) %>%
+  tab_spanner(label = "Conservation Scores", columns = vars(score_median, score_median2, score_reg))
+
+prepped_data %>%
+  group_by(`5'UTR` = str_detect(UCSC_RefGene_Group, "5'UTR") & !is.na(UCSC_RefGene_Group)) %>%
+  summarise_at(vars(sigmaP:sigmaT,score_median:score_reg), mean, na.rm = TRUE) %>%
+  mutate(`5'UTR` = factor(`5'UTR`, c(TRUE, FALSE), c("Yes", "No"))) %>%
+  arrange(rev(rownames(.))) %>%
+  gt() %>%
+  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, score_median, score_median2, score_reg), decimals = 3) %>%
+  tab_spanner(label = "Hierarchical Variance", columns = vars(sigmaP, sigmaPT, sigmaT)) %>%
+  tab_spanner(label = "Conservation Scores", columns = vars(score_median, score_median2, score_reg))
+
+prepped_data %>%
+  group_by(`TSS1500` = str_detect(UCSC_RefGene_Group, "TSS1500") & !is.na(UCSC_RefGene_Group)) %>%
+  summarise_at(vars(sigmaP:sigmaT,score_median:score_reg), mean, na.rm = TRUE) %>%
+  mutate(`TSS1500` = factor(`TSS1500`, c(TRUE, FALSE), c("Yes", "No"))) %>%
+  arrange(rev(rownames(.))) %>%
+  gt() %>%
+  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, score_median, score_median2, score_reg), decimals = 3) %>%
+  tab_spanner(label = "Hierarchical Variance", columns = vars(sigmaP, sigmaPT, sigmaT)) %>%
+  tab_spanner(label = "Conservation Scores", columns = vars(score_median, score_median2, score_reg))
+
+prepped_data %>%
+  group_by(`TSS200` = str_detect(UCSC_RefGene_Group, "TSS200") & !is.na(UCSC_RefGene_Group)) %>%
+  summarise_at(vars(sigmaP:sigmaT,score_median:score_reg), mean, na.rm = TRUE) %>%
+  mutate(`TSS200` = factor(`TSS200`, c(TRUE, FALSE), c("Yes", "No"))) %>%
+  arrange(rev(rownames(.))) %>%
+  gt() %>%
+  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, score_median, score_median2, score_reg), decimals = 3) %>%
+  tab_spanner(label = "Hierarchical Variance", columns = vars(sigmaP, sigmaPT, sigmaT)) %>%
+  tab_spanner(label = "Conservation Scores", columns = vars(score_median, score_median2, score_reg))
+```
+
+\newpage
 \subsection{Gene-average Parameter Distributions}
@@
 rm(pM,pB,pE,pP,pPT,pT,pP1,pP2,pP3)
+```
+
+\newpage
+\subsection{Figure 1: CpG Distance Analysis}
+Corrected for gene direction, based on Ensembl gene database
+```{r Fig1, echo=FALSE, warning=FALSE, fig.width=7, fig.height=2.5, comment=NA}
+# load ensembl db
+library(EnsDb.Hsapiens.v75) # ensembl annotation with gene position (hg19 version)
+edb <- EnsDb.Hsapiens.v75
+geneKeys <- keys(edb, keytype="GENENAME") %>% intersect(unique(gened_data$UCSC_RefGene_Name))
+edbData<- AnnotationDbi::select(edb, keys=geneKeys, keytype="GENENAME",
+                                columns=c("GENEBIOTYPE", "GENESEQSTART", "GENESEQEND", "SEQNAME", "SEQSTRAND")) %>%
+  filter(SEQNAME %in% c(as.character(1:22),"X","Y")) %>%
+  mutate(GENESTART = GENESEQSTART*(SEQSTRAND==1) + GENESEQEND*(SEQSTRAND==-1)) %>%
+  select(-c(GENESEQSTART,GENESEQEND))
+
+# calculate CpG distance based on Ensembl gene start position and direction
+distanced_data <- gened_data %>%
+  filter(!is.na(UCSC_RefGene_Name)) %>%
+  filter(UCSC_RefGene_Name %in% geneKeys) %>%
+  mutate(EDBCHR=edbData$SEQNAME[match(UCSC_RefGene_Name,edbData$GENENAME)],
+         EDBSTART=edbData$GENESTART[match(UCSC_RefGene_Name,edbData$GENENAME)],
+         EDBDIR=edbData$SEQSTRAND[match(UCSC_RefGene_Name,edbData$GENENAME)],
+         EDBTYPE=edbData$GENEBIOTYPE[match(UCSC_RefGene_Name,edbData$GENENAME)]
+         ) %>%
+  filter(CHR==EDBCHR) %>%
+  mutate(cpgdist = if_else(EDBDIR==1, MAPINFO-EDBSTART, EDBSTART-MAPINFO))
+
+distance_length <- distanced_data %>%
+  filter(!is.na(UCSC_RefGene_Name)) %>%
+  select(CHR, UCSC_RefGene_Name, cpgdist, c(sigmaP:sigmaT,score_median:score_reg)) %>%
+  mutate(location_bin = cut(cpgdist, c(seq(-1500, 20000, by = 100)), include.lowest = TRUE)) %>%
+  group_by(location_bin) %>%
+  summarise(mean_sigmaP = mean(sigmaP), mean_sigmaPT = mean(sigmaPT), mean_sigmaT = mean(sigmaT),  mean_score_median = mean(score_median), mean_score_median2 = mean(score_median2), n_obs = n(), .groups='drop') %>%
+  mutate(location_bin = as.numeric(str_extract(location_bin, "(?<=,)(.*)(?=])")))
+
+distance_length %>%
+  drop_na() %>%
+  pivot_longer(mean_sigmaP:mean_sigmaT) %>% 
+  mutate(name = factor(name, c("mean_sigmaP","mean_sigmaPT","mean_sigmaT"), c("SigmaP","SigmaPT","SigmaT"))) %>%
+  sample_frac() %>%
+  ggplot(aes(location_bin, value, color = name, fill = name)) +
+  geom_point(key_glyph = draw_key_rect) +
+  ggforce::facet_zoom(location_bin < 5000) +
+  theme_light() +
+  theme(legend.position = "bottom") +
+  labs(title = NULL,
+       fill = NULL,
+       color = NULL,
+       x = "Distance from gene start site (binned to 100 bp)",
+       y = "Average Sigma Value", 
+       caption = "Figure 1: Average Sigma values of single CpGs as a function of position relative to gene.") +
+  scale_color_brewer(palette = "Set2") +
+  scale_fill_brewer(palette = "Set2")
+cat("\n")
+
+# again for conservation scores
+distance_length %>%
+  drop_na() %>%
+  pivot_longer(mean_score_median:mean_score_median2) %>% 
+  mutate(name = factor(name, c("mean_score_median","mean_score_median2"), c("Score_median","Score_median2"))) %>%
+  sample_frac() %>%
+  ggplot(aes(location_bin, value, color = name, fill = name)) +
+  geom_point(key_glyph = draw_key_rect) +
+  ggforce::facet_zoom(location_bin < 5000) +
+  theme_light() +
+  theme(legend.position = "bottom") +
+  labs(title = NULL,
+       fill = NULL,
+       color = NULL,
+       x = "Distance from gene start site (binned to 100 bp)",
+       y = "Average Conservation Score", 
+       caption = "Figure 1: Average conservation scores of single CpGs as a function of position relative to gene.") +
+  scale_color_brewer(palette = "Set2") +
+  scale_fill_brewer(palette = "Set2")
+cat("\n")
+
+# and look at number of CpG in each bin
+distance_length %>%
+  drop_na() %>%
+  pivot_longer(n_obs) %>%
+  mutate(name = factor(name, c("n_obs"), c("n_obs"))) %>%
+  sample_frac() %>%
+  ggplot(aes(location_bin, value, color = name, fill = name)) +
+  geom_point(key_glyph = draw_key_rect) +
+  ggforce::facet_zoom(location_bin < 5000) +
+  theme_light() +
+  theme(legend.position = "bottom") +
+  labs(title = NULL,
+       fill = NULL,
+       color = NULL,
+       x = "Distance from gene start site (binned to 100 bp)",
+       y = "Number of CpGs", 
+       caption = "Figure 1: Frequency of single CpGs as a function of position relative to gene start site.") +
+  scale_color_brewer(palette = "Set2") +
+  scale_fill_brewer(palette = "Set2")
 ```

### 2020-09-25 conserved_genes_and_paths.Rmd
updated code

---
title: "Significant Genes and Enriched Pathways"
author: "Kevin Murgas"
date: "`r Sys.Date()`"
output: pdf_document
---

The genes included here are significantly conserved based on the conservation probability score derived from the posterior distribution of log sigmaP/sigmaT ratio (which will be high when tumor variability is lower than normal variability):
$$ \text{score}_\text{median} = \text{Median} (\text{log}\frac{\sigma_{P}}{\sigma_T}) $$
We select significantly conserved genes based on the top 95th percentile of adjusted bootstrapped values.

```{r load, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(pander)
library(gt)
library(dplyr)
# set working directory
setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
library(here)

rename <- dplyr::rename
select <- dplyr::select
slice <- dplyr::slice

permed_data <- read_csv(here("data-stan_rg3ps3", "permed_data.csv"))
#enrichedGenes <- read_csv(here("prob1genes.csv"))

```

\section{Reactome Enriched Pathways}
```{r paths, echo=FALSE, message=FALSE, comment=NA}

# select enriched genes from permed data
# filter out lowest quartile of sigmaP genes
enrichedGenes <- permed_data %>%
  filter(sigmaP > quantile(sigmaP, 0.25, na.rm=TRUE)) %>%
  select(UCSC_RefGene_Name, score_median, score_median_v3, n) %>%
  rename(Gene=UCSC_RefGene_Name, bootstrap=score_median_v3) %>%
  #filter(score_median > quantile(score_median, 0.95, na.rm = TRUE)) %>%
  filter(bootstrap > 0.95) %>%
  arrange(-bootstrap,-score_median)

# enrichedGenes <- permed_data %>%
#   #filter(sigmaP > quantile(sigmaP, 0.25, na.rm=TRUE)) %>%
#   select(UCSC_RefGene_Name, score_reg, score_reg_v3, n) %>%
#   rename(Gene=UCSC_RefGene_Name, bootstrap=score_reg_v3) %>%
#   #filter(score_reg > quantile(score_reg, 0.95, na.rm = TRUE)) %>%
#   filter(bootstrap > 0.95) %>%
#   arrange(bootstrap,score_reg)

#write.csv(enrichedGenes %>% select(Gene), here("prob1genes.csv"), col.names=FALSE, row.names = FALSE, quote = FALSE)

# Gene Set Enrichment
tidy_gsea <- function(genes) {
  AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, genes, 'ENTREZID', 'SYMBOL') %>%
    ReactomePA::enrichPathway()
}

cat("Begin reactome GSEA...")
ptm<-proc.time()
enrich_reactome <-enrichedGenes %>%
  pull(Gene) %>% tidy_gsea()
proc.time()-ptm

enrichedPaths <- tibble(Pathway = enrich_reactome$Description, ngenes = enrich_reactome$Count, padj = enrich_reactome$p.adjust) %>% arrange(padj)
geneList <- enrich_reactome$geneID %>% strsplit(split='/') %>%
  lapply(function(x) AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, x, 'SYMBOL', 'ENTREZID')) %>%
  lapply(function(x) paste(x,collapse=";")) %>% unlist
enrichedPaths$Genes = geneList

#write.csv(enrichedPaths, here("ReactomeRPACKAGE.csv"), col.names=FALSE, row.names = FALSE, quote = TRUE)

ngene <- dim(enrichedGenes)[1]
npath <- length(enrich_reactome$Description)

cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath))

cat("score_median top enriched paths:  \n")
enrichedPaths %>% pander()

knit_exit()

```

<!-- \newpage -->
<!-- \section{Top Conserved Genes} -->
<!-- ```{r genes, echo=FALSE, comment=NA} -->
<!-- cat(paste0("score_median top enriched genes:  \n Total number of genes: ", ngene)) -->
<!-- #enrichedGenes %>% arrange(-bootstrap) %>% rename(nsites=n) %>% pander() -->

<!-- ``` -->

\newpage
\section{GSVA-type analysis}
```{r load_geneset, include=FALSE}
# load in the C5 gene sets from MSigDB
library(GSA)
gmtfile <- file.path("/Users/kevinmurgas/Documents/Data+ project/EPIC data","c2.cp.v7.1.symbols.gmt")
genesets <- GSA.read.gmt(gmtfile)
nsets <- length(genesets[[1]])

# make a data.frame for gene sets and scores
nsets <- length(genesets[[1]])
GSEA_df <- data.frame(GeneSet=genesets[[2]][1:nsets], score=numeric(nsets),
                      num_gene = unlist(lapply(genesets[[1]][1:nsets],length)),
                      num_gene_cut = numeric(nsets))

```

\subsection{Evidence: conservation score rank}
```{r GSVA_rank, echo=FALSE, comment=NA, fig.width=8}
# Procedure:
# 1. Order and rank all genes by evidence (currently using log-P/T-ratio integral > 0)
# 2. For each gene set:
#     Create two CDFs/running sums:
#     1. In-set genes: increment by the evidence magnitude or RANK
#     2. Out-of-set genes: increment by 1
#     add 0 to other sum if gene in/not in
#     Normalize by max at the end to get CDFs
#     Take modified K-S statistic, by measuring the difference in the max and min differences of the two CDFs
# 3. Each gene set now has a value to rank enrichment
#  KS statistic
# 4. For significance (FDR), need to compare to KS statistic
# if in set: X = sqrt((N-G)/G) ... with G = # genes in set, N = # genes total
# if not in set: XN=−sqrt(G/(N-G))

# have to order all genes first by rank of conservation (conserved genes come first)
genenames = permed_data$UCSC_RefGene_Name
ngene = length(genenames)

# reorder by evidence rank
genenames = genenames[order(-permed_data$score_median)] # use ranked conservation score as evidence
evidence = seq(ngene,1,-1)

# plot evidence ranked: permed_data$score_median
ranks <- permed_data$score_median
names(ranks) <- permed_data$UCSC_RefGene_Name
barplot(sort(ranks, decreasing = T), main = "Ranked conservation scores")

cat("Begin calculating random-walk statistics...")
ptm <- proc.time()
for (iset in 1:nsets) {
  # take only genes that we have in full gene list
  set <- unlist(genesets[[1]][iset]) %>% intersect(genenames)
  GSEA_df$num_gene_cut[iset] = length(set)

  # skip if num_gene <10 or >500
  if (length(set)<10 | length(set)>500) {next}

  # compute running sum (as PDF first)
  # create the random walk, with correct scaling for in/out genes
  inmask = genenames %in% set
  runsum = evidence/sum(evidence[inmask]) * inmask - 1/(ngene-length(set)) * !inmask

  # convert to CDF, compute K-S-like statistic
  v <- cumsum(runsum)
  score <- max(c(0,v)) + min(c(0,v))
  GSEA_df$score[iset] <- score
}
cat(proc.time()-ptm)
cat(sprintf("Processed %i genes, %i pathways/genesets. Begin bootstrapping...",ngene,nsets))

# calculate bootstraps
setlengths <- unique(GSEA_df$num_gene_cut)
setlengths <- sort(setlengths[which(!(setlengths < 10 | setlengths > 500))])
# run thru each set length
# store 1000 bootstrapped scores using random gene sets

shuffle_genesets <- function(n_geneset, n_rep, data) {
  bootstrap = numeric(n_rep)
  for (ishuf in 1:n_rep) {
    # for random shuffles just make a random logical mask
    inds = sample(1:ngene, n_geneset, replace = F)
    inmask = logical(ngene)
    inmask[inds] = TRUE
    runsum = evidence/sum(evidence[inmask]) * inmask - 1/(ngene - length(set)) * !inmask

    # convert to CDF, compute K-S-like statistic
    v <- cumsum(runsum)
    score <- max(c(0, v)) + min(c(0, v))
    bootstrap[ishuf] <- score
  }
  bootstrap
}

names(setlengths) <- setlengths
cat(sprintf("Shuffling %i set-lengths: 1000x random shuffles (w/o replacement)",length(setlengths)))
ptm<-proc.time()
res <- lapply(setlengths, shuffle_genesets, n_rep = 1000, data = evidence)
proc.time()-ptm

# map enrichment scores to get percentiles on bootstrap null distribution using upper tail probability
pvals <- map2_dbl(GSEA_df$score, GSEA_df$num_gene_cut, ~ mean(res[[as.character(.y)]] > .x))

# Benjamini-Hochberg step-up correction procedure
# this matches r p.adjust("BH")
ptemp <- pvals[!is.na(pvals)]
m <- length(ptemp)
BHp <- sort(ptemp,decreasing=T,na.last=TRUE)*m/seq(m,1)
BHp[order(ptemp,decreasing=T)] <- cummin(BHp)
q <- pvals*NA
q[!is.na(pvals)] <- BHp

GSEA_df$pval <- pvals
GSEA_df$FDR <- q

nsig <- sum(GSEA_df$FDR<0.05, na.rm=TRUE)
ntot <- sum(!is.na(GSEA_df$FDR))
cat(sprintf("%i significant pathways out of %i tested (%i excluded by size <10 | >500)", nsig, ntot, nsets-ntot))
# make table of top scores
GSEA_df %>% arrange(FDR) %>%
  filter(FDR<0.05) %>%
  select(score,pval,FDR,num_gene_cut,GeneSet) %>%
  rename(Ngene=num_gene_cut) %>%
  #mutate(GeneSet = gsub('(.{1,90})(\\s|$)', '\\1\n', GeneSet)) %>%
  gt() %>% cols_align(align="left") %>% fmt_number(vars(score,pval,FDR),decimals=3)

# save as .csv
enrichedGenes <- permed_data %>%
  filter(sigmaP > quantile(sigmaP, 0.25, na.rm=TRUE)) %>%
  select(UCSC_RefGene_Name, score_median, score_median_v3, n) %>%
  rename(Gene=UCSC_RefGene_Name, bootstrap=score_median_v3) %>%
  filter(bootstrap > 0.95) %>%
  arrange(-bootstrap,-score_median)

write.csv(enrichedGenes, here("significantgenes_filtered.csv"), col.names=FALSE, row.names = FALSE, quote = FALSE)
write.csv(GSEA_df %>% arrange(FDR) %>%
  filter(FDR<0.05) %>%
  select(score,pval,FDR,num_gene_cut,GeneSet) %>%
  rename(Ngene=num_gene_cut), here("GSVApathways_significantFDR.csv"), col.names=FALSE, row.names = FALSE, quote = FALSE)
```

<!-- \newpage -->
<!-- \section{Patient-specific analysis} -->
<!-- Basic overview: using a specified number of genes (here 100), take top genes (here using score_med bootstrapping percentiles) to select a "gene signature" of sorts. Then for each patient, calculate the mean pair-wise distance within the individual patients' tumor samples. Method 1 is these raw values, method 2 normalizes by the mean PWD of all normal samples. Method 3 just examines the overall mean PWD for all tumor samples, normalized by mean PWD of all normal. -->

<!-- ```{r PatientPWD, echo=FALSE, message=FALSE, warning=FALSE, comment=NA, fig.height=9, fig.width=7} -->
<!-- library(gtools) -->

<!-- # Load raw methylation data -->
<!-- load(here("myFA_bulkonly.Rdata")) -->

<!-- # pull out all samples beta values -->
<!-- all_beta <- FullAnnotation[,10:57] -->

<!-- # extract patient and tissue labels -->
<!-- patientLabel <- substr(colnames(all_beta), 1, 1) -->
<!-- patientLabel[10:12] <- "K*" -->
<!-- tissueLabel <- substr(colnames(all_beta), 2, 2) -->
<!-- tissueLabel[!(tissueLabel %in% "N")] <- "T"   #replace A/B/D/M with T -->

<!-- pats = patientLabel %>% sort %>% unique -->
<!-- npat = length(pats) -->

<!-- enrichedGenes <- permed_data %>% filter(score_median_v3 > 0.95) %>% -->
<!--   select(UCSC_RefGene_Name, score_median, score_median_v3, n) %>% -->
<!--   rename(Gene=UCSC_RefGene_Name, Pctile=score_median_v3) -->

<!-- #ngene = 70 # using top N genes -->
<!-- #genes = enrichedGenes %>% top_n(-ngene,Pctile) %>% pull(Gene) -->
<!-- #genes = genes[1:ngene] -->

<!-- genes = enrichedGenes$Gene -->
<!-- ngene = length(genes) -->

<!-- pwd <- function(x) { -->
<!--   mean(dist(x)) -->
<!-- } -->
<!-- pwd2 <- function(x) { -->
<!--   abs(diff(x)) -->
<!-- } -->

<!-- signature_mat1 = matrix(,ngene,npat) # raw tumor pwd -->
<!-- for (igene in 1:ngene) { -->
<!--   # each gene, get gene inds, calculate PWD for the tumor samples -->
<!--   geneind = grep(genes[igene], FullAnnotation$UCSC_RefGene_Name) -->

<!--   # calculate PWD of each tumor -->
<!--   tum_pwd <- numeric(npat) -->
<!--   for (ipat in 1:npat) { -->
<!--     # each patient, get tumor sample inds -->
<!--     patind <- which(patientLabel == pats[ipat] & tissueLabel == "T") -->
<!--     tum_pwd[ipat] <- all_beta[geneind,patind] %>% logit %>% apply(1,pwd2) %>% mean -->
<!--   } -->

<!--   signature_mat1[igene,] = tum_pwd -->
<!-- } -->

<!-- library(pheatmap) -->
<!-- library(dendsort) -->

<!-- cat("  \n\\newline  \nMethod 1 - raw intratumor PWD within each patient") -->
<!-- df1 = as.data.frame(signature_mat1) -->
<!-- rownames(df1) <- genes -->
<!-- colnames(df1) <- pats -->
<!-- sort_hclust <- function(...) as.hclust(dendsort(as.dendrogram(...))) -->
<!-- mat_cluster_cols <- sort_hclust(hclust(dist(t(df1)))) -->
<!-- mat_cluster_rows <- sort_hclust(hclust(dist(df1))) -->
<!-- pheatmap( -->
<!--   mat=df1, -->
<!--   cluster_cols = mat_cluster_cols, -->
<!--   cluster_rows = mat_cluster_rows, -->
<!--   show_rownames = FALSE, -->
<!--   show_colnames = TRUE, -->
<!--   main = sprintf("Top %i Genes: Tumor PWD", ngene) -->
<!-- ) -->

<!-- df2 = df1 -->
<!-- for (ipat in 1:npat) { -->
<!--   df2[,ipat] = df1[,ipat] > quantile(df1[,ipat],0.5) -->
<!-- } -->
<!-- cat("  \n\\newline  \nMethod 1 - Lowest 50% PWD Genes within each patient") -->
<!-- pheatmap( -->
<!--   mat=df2, -->
<!--   cluster_cols = mat_cluster_cols, -->
<!--   cluster_rows = mat_cluster_rows, -->
<!--   show_rownames = FALSE, -->
<!--   show_colnames = TRUE, -->
<!--   main = sprintf("Top %i Genes: Lowest 50 percentile Tumor PWD per patient", ngene) -->
<!-- ) -->

<!-- ``` -->

### 2020-09-25 GetPosteriorsSingleSites.R

@@ -20,47 +20,28 @@ setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
 #setwd("D:/DataPlus2017/Data")
 
 # load data
-load("myFA.Rdata")
+load("myFA_bulkonly.Rdata")
 
-#args <- commandArgs(trailingOnly = TRUE)
-#print(args)
-#N <- as.numeric(args[1])
-N <- 1
-print(paste("Task #: ", N))
-#out.file <- args[2]
-log.file <- paste("logTask",N,".txt",sep="")
-writeLines(c(""), log.file)
-
-# randomly select 5 sites in 1-866836
-siteInds <- sample(1:866836,5)
+# randomly select 5 sites in 1-866091
+#siteInds <- sample(1:866091,5)
+siteInds <- c(525490, 570532, 827025, 287316, 854143)
 nsites <- length(siteInds)
-
-# select chunk within FullAnnotation, then remove FA
-print(paste("Making chunk of nsites =", nsites))
-chunk <- FullAnnotation[siteInds,]
-rm(FullAnnotation)
-gc()
-
-# must add to prevent crash
-checkFile <- "model3.rds"
 
 
 ### FXNS ###
 # Function used to read in data from each site
-# changed to now use a "chunk" instead of FullAnnotation, for speed and memory
-# requires index 1-10000 instead of siteIndex
 site <- function (ind) {
   # extract CpG site xx to start
-  temp <- chunk[ind,]
+  temp <- FullAnnotation[ind,]
   
   # here we use all patient samples, excluding glands
-  indices <- c(9:13, 15:39, 46, 47, 57, 58, 72:75)
+  #indices <- c(10:14, 16:40,47,48,58,59,73:76,87:96) # if using full data file
+  indices <- c(10:57) # if using bulk only data file
   patientLabel <- substr(colnames(temp[indices]), 1, 1)
   patientLabel[10:12] <- "K*"
   sideLabel <- substr(colnames(temp[indices]), 2, 2)
   tissueLabel <- sideLabel
-  tissueLabel[tissueLabel %in% c("A", "B")] <-
-    "T"   #replace A and B with T
+  tissueLabel[!(sideLabel %in% c("N"))] <- "T"   #replace A/B/D/M with T
   tumorIndicator <- 1 * (tissueLabel == "T")
   
   work <-
[…]

### 2020-09-25 LoadDataAndQC.R

@@ -51,24 +51,16 @@ read.fun <- function(base_dir,pat_file) {
 ############################### Read in data  #################################
 
 ## 1337 ##
-
 base_dir_1337 <- "1337_Shibata EPIC DNA methylation data package/IDAT FILES"
 pat_file_1337 <- "1337_Shibata EPIC DNA methylation data package/SAMPLE-ARRAY MAPPING/1337 (Shibata-8).xls"
-
 work_1337 <- read.fun(base_dir_1337,pat_file_1337)
 
-
-
 ## 1345 ##
-
 base_dir_1345 <- "1345_Shibata EPIC DNA methylation data package/IDAT FILES"
 pat_file_1345 <- "1345_Shibata EPIC DNA methylation data package/SAMPLE-ARRAY MAPPING/1345 (Shibata-16).xlsx"
-
 work_1345 <- read.fun(base_dir_1345,pat_file_1345)
 
-
 # Two beadchips in plate 1345, read in seperately
-
 pat_1345 <- read.xls(pat_file_1345)
 targets_1345 <- data.frame(pat_1345[,"Complete.Barcode"])
 colnames(targets_1345) <- 'Basement'
@@ -79,54 +71,55 @@ work_1345a <- read.metharray(targets_1345$Basement[1:8],verbose=TRUE)
 # Beadchip 2
 work_1345b <- read.metharray(targets_1345$Basement[9:16],verbose = TRUE)
 
-
-
 ## 1350 ##
-
 base_dir_1350 <- "1350_SHIBATA EPIC DNA METHYLATION DATA PACKAGE/IDAT files"
 pat_file_1350 <- "1350_SHIBATA EPIC DNA METHYLATION DATA PACKAGE/SAMPLE-ARRAY MAPPING/1350 (Shibata-8).xlsx"
-
 work_1350 <- read.fun(base_dir_1350,pat_file_1350)
  
-
 ## 1357 ##
-
 base_dir_1357 <- "1357_Shibata EPIC DNA methylation data package/IDAT FILES"
 pat_file_1357 <- "1357_Shibata EPIC DNA Methylation Data Package/SAMPLE-ARRAY MAPPING/1357 (Shibata-8).xlsx"
-
 work_1357 <- read.fun(base_dir_1357,pat_file_1357)
 
-
 ## 1360 ##
-
 base_dir_1360 <- "1360_Shibata EPIC Data Package/IDAT FILES"
[…]

### 2020-09-25 NormalSimulation.R

@@ -1,123 +1,0 @@
-#################################################################################
-######################### Normal Tissue Data Simulation #########################
-#################################################################################
-
-
-library(rstan)
-library(gtools)
-library(ggplot2)
-library(bayesplot)
-library(MASS)
-
-# Kevin's working directory
-setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
-
-# Yanlin's working directory
-#setwd("D:/DataPlus2017/Data")
-
-# load data
-load("myFA.Rdata")
-
-
-# Extract single site data
-site <- function (site_no) {
-  # extract CpG site xx to start
-  temp <- FullAnnotation[site_no,]
-  
-  # here we use all patient samples, excluding glands
-  indices <- c(9:13, 15:39,46,47,57,58,72:75)
-  patientLabel <- substr(colnames(temp[indices]),1,1)
-  patientLabel[10:12] <- "K*"
-  sideLabel <- substr(colnames(temp[indices]),2,2)
-  tissueLabel <- sideLabel
-  tissueLabel[tissueLabel %in% c("A","B")] <- "T"   #replace A and B with T
-  tumorIndicator <- 1*(tissueLabel=="T")
-  
-  work <- data.frame(beta=logit(t(temp[indices])), patient = patientLabel, tissue=tissueLabel, side=sideLabel, tInd=tumorIndicator)
-  colnames(work)[1] <- "beta"
-  return(work)
-}
-
-# Using Model 3: add intra-tumoral variance
-
-stanfit3 <- function (dataset) {
-  
-  stanDat <- list(pID = as.integer(factor(dataset$patient)),
-                  tInd = dataset$tInd,
-                  N = nrow(dataset),
-                  P = nlevels(dataset$patient),
-                  y = dataset[,1])
-  
-  
-  
-  stanFit3 <- stan(file="model3.stan", data=stanDat, control = list(adapt_delta = 0.999))
-  
-  return(stanFit3=stanFit3)
-}
-
-# Plot posterior distribution
-posterior <- function(dataset) {
-  stan <- stanfit3(dataset)
-  mcmc_areas(
-    as.array(stan), 
-    pars = c("sigma_p", "sigma_pt","sigma_t","sigma_e"),
-    prob = 0.8, # 80% intervals
-    prob_outer = 0.95, 
-    point_est = "median"
-  )
-}
-
-# Randomly choose 5 sites
[…]

### 2020-09-25 Plots_FullData.R

@@ -1,373 +1,0 @@
-################### Plots ############################
-
-library(ggplot2)
-library(dunn.test)
-
-# here set the working directory that points to the data folder
-# e.g. the folder with annotated data saved as "myFA.Rdata"
-# please just comment out the directories already here, so each
-# user can uncomment the setwd for them as we move code around
-
-# Kevin's working directory
-setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
-
-# Yanlin's working directory
-setwd("D:/DataPlus2017/Data")
-
-
-### LOAD DATA
-load("StanCFullResults.RData")
-
-sigma <- data.frame(sigmaP_Cfull$p50,sigmaPT_Cfull$p50,sigmaT_Cfull$p50,sigmaE_Cfull$p50)
-
-
-### Estimates Distribution
-ggplot(mu_Cfull,aes(x=mu_Cfull$p50)) + geom_histogram(binwidth = 0.1) + labs(title="Distribution of mu",x="mu (median)")
-
-ggplot(betaT_Cfull,aes(x=betaT_Cfull$p50)) + geom_histogram(binwidth = 0.1) + labs(title="Distribution of betaT",x="betaT (median)")
-
-ggplot(sigmaP_Cfull,aes(x=sigmaP_Cfull$p50)) + geom_histogram(binwidth = 0.05) + labs(title="Distribution of sigmaP",x="sigmaP (median)") + coord_cartesian(ylim=c(0,175000),xlim=c(0,3))
-
-ggplot(sigmaPT_Cfull,aes(x=sigmaPT_Cfull$p50)) + geom_histogram(binwidth = 0.05) + labs(title="Distribution of sigmaPT",x="sigmaPT (median)") + coord_cartesian(ylim=c(0,175000),xlim=c(0,3))
-
-ggplot(sigmaT_Cfull,aes(x=sigmaT_Cfull$p50)) + geom_histogram(binwidth = 0.05) + labs(title="Distribution of sigmaT",x="sigmaT (median)") + coord_cartesian(ylim=c(0,175000),xlim=c(0,3))
-
[…]

### 2020-09-25 Score.R

@@ -1,135 +1,0 @@
-########################## Scoring ################################
-
-library(doParallel)
-
-### Initialize
-
-print(paste("Cores: ", detectCores()))
-
-# here set the working directory that points to the data folder
-# e.g. the folder with annotated data saved as "myFA.Rdata"
-# please just comment out the directories already here, so each
-# user can uncomment the setwd for them as we move code around
-
-# Kevin's working directory
-setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
-
-# Yanlin's working directory
-setwd("D:/DataPlus2017/Data")
-
-
-### LOAD DATA
-
-# load fully annotated data (saved from LoadDataAndQC.R)
-load("myFA.Rdata")
-load("GeneInfo.Rdata")
-load("StanCfullResults3.Rdata")
-
-N<-1
-
-# make vector of unique genes and regions
-uniqueGenes <- unique(unlist(geneNames))
-regionTypes <- unique(unlist(geneRegions))
-
-# make dataframes for median values of each parameter
-dfMedians <- cbind.data.frame(mu_Cfull3$p50, betaT_Cfull3$p50, sigmaE_Cfull3$p50, sigmaP_Cfull3$p50, sigmaPT_Cfull3$p50, sigmaT_Cfull3$p50)
-colnames(dfMedians) <- c("mu","betaT","sigmaE","sigmaP","sigmaPT","sigmaT")
-
-# calculate log ratios: P/T, P/PT, PT/T
-logRatios <- cbind.data.frame(log(dfMedians$sigmaP/dfMedians$sigmaT), log(dfMedians$sigmaP/dfMedians$sigmaPT), log(dfMedians$sigmaPT/dfMedians$sigmaT))
-colnames(logRatios) <- c("logP/T","logP/PT","logPT/T")
-logRatioMeans <- colMeans(logRatios)
-# insert means by region
-
-# Choose 1000 genes from uniqueGenes by N, and select geneChunk within gene list
-geneIDs <- (1:1000) + (N - 1) * 1000
-if (N > 27) {
-  geneIDs <- geneIDs[1:383]
-}
-geneIDs <- 1:20
-numGenes <- length(geneIDs)
-
-print(paste("Making chunk of numGenes =", numGenes))
-geneChunk <- uniqueGenes[geneIDs]
-
-
-### Functions
-
-# find indices corresponding to geneString in the list geneNames
[…]

### 2020-09-25 ScoreGenesParallel.R

@@ -1,121 +1,0 @@
-# script for calculating scores on individual genes (27383 total genes)
-# using full-scale results from Stan complex model (model3), including PTprob (run 3)
-# input taskID (1-28) corresponding to 1000 gene chunks starting at site (N-1)*1000+1
-# incorporating parallel processing to run through all genes faster
-# just need "StanCfullResults3.Rdata" and "GeneInfo" in working directory
-
-library(doParallel)
-
-### Initialize
-
-print(paste("Cores: ", detectCores()))
-
-args <- commandArgs(trailingOnly = TRUE)
-print(args)
-N <- as.numeric(args[1])
-print(paste("Task #:", N))
-out.file <- args[2]
-log.file <- paste0("logTask",N,".txt")
-writeLines(c(""), log.file)
-
-# load data: Stan parameters, geneNames and geneRegions lists
-load("StanCfullResults3.Rdata")
-load("GeneInfo.Rdata")
-
-# make vector of unique genes and regions
-uniqueGenes <- unique(unlist(geneNames))
-regionTypes <- unique(unlist(geneRegions))
-
-# count sites at each geneName
-geneCount <- data.frame(table(unlist(geneNames)))
-
-# make dataframes for median values of each parameter
-dfMedians <- cbind.data.frame(mu_Cfull$p50, betaT_Cfull$p50, sigmaE_Cfull$p50, sigmaP_Cfull$p50, sigmaPT_Cfull$p50, sigmaT_Cfull$p50)
-colnames(dfMedians) <- c("mu","betaT","sigmaE","sigmaP","sigmaPT","sigmaT")
-
-# calculate log ratios: P/T, P/PT, PT/T
-logRatios <- cbind.data.frame(log(dfMedians$sigmaP/dfMedians$sigmaT), log(dfMedians$sigmaP/dfMedians$sigmaPT), log(dfMedians$sigmaPT/dfMedians$sigmaT))
-colnames(logRatios) <- c("logP/T","logP/PT","logPT/T")
-logRatioMeans <- colMeans(logRatios)
-
-# Choose 1000 genes from uniqueGenes by N, and select geneChunk within gene list
-geneIDs <- (1:1000) + (N - 1) * 1000
-if (N > 27) {
-  geneIDs <- geneIDs[1:383]
-}
-numGenes <- length(geneIDs)
-
-print(paste("Making chunk of numGenes =", numGenes))
[…]

### 2020-09-25 StanCParallel.R

@@ -7,10 +7,11 @@ library(rstan)
 library(gtools)
 
 rstan_options(auto_write = FALSE)
-print(paste("Cores: ", detectCores()))
+numCPUs = as.integer(Sys.getenv('SLURM_CPUS_PER_TASK'))
+print(paste("Cores: ", numCPUs))
 
 # load data
-load("myFA.Rdata")
+load("myFA_bulkonly.Rdata")
 
 args <- commandArgs(trailingOnly = TRUE)
 print(args)
@@ -23,7 +24,7 @@ writeLines(c(""), log.file)
 # Choose 10000 sites by N, and select chunk within FullAnnotation, then remove FA
 siteInds <- (1:10000) + (N - 1) * 10000
 if (N > 86) {
-  siteInds <- siteInds[1:6836]
+  siteInds <- siteInds[1:6091]
 }
 nsites <- length(siteInds)
 
@@ -31,9 +32,6 @@ print(paste("Making chunk of nsites =", nsites))
 chunk <- FullAnnotation[siteInds,]
 rm(FullAnnotation)
 gc()
-
-# must add to prevent crash
-checkFile <- "model3.rds"
 
 
 ### FXNS ###
@@ -45,13 +43,12 @@ site <- function (ind) {
   temp <- chunk[ind,]
   
   # here we use all patient samples, excluding glands
-  indices <- c(9:13, 15:39, 46, 47, 57, 58, 72:75)
+  indices <-  c(10:57) # if using bulk only data file
   patientLabel <- substr(colnames(temp[indices]), 1, 1)
   patientLabel[10:12] <- "K*"
   sideLabel <- substr(colnames(temp[indices]), 2, 2)
   tissueLabel <- sideLabel
-  tissueLabel[tissueLabel %in% c("A", "B")] <-
-    "T"   #replace A and B with T
+  tissueLabel[!(sideLabel %in% c("N"))] <- "T"   #replace A/B/D/M with T
   tumorIndicator <- 1 * (tissueLabel == "T")
   
   work <-
@@ -66,8 +63,8 @@ site <- function (ind) {
   return(work)
 }
 
-# Function to build stan model for each site
-stanfit3 <- function (dataset) {
+# Using Model with TCGA priors (previously model 3)
+stanfit_model <- function (dataset) {
   stanDat <- list(
[…]

### 2020-09-25 StanCParallel2.R

@@ -1,263 +1,0 @@
-# script for extracting variance from Stan complex model (model3)
-# input taskID corresponding to site chunks, here set as 1
-# testing on first 100 sites
-# Now incorporating parallel processing with parallel package
-
-library(doParallel)
-library(rstan)
-library(gtools)
-library(pryr)
-
-rstan_options(auto_write = FALSE)
-#options(mc.cores = detectCores())
-print(paste("Cores: ", detectCores()))
-
-# Kevin's working directory
-setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
-# load data
-load("myFA.Rdata")
-
-N <- 1
-#args <- commandArgs(trailingOnly = TRUE)
-#print(args)
-#N <- as.numeric(args[1])
-print(paste("Task #: ", N))
-#out.file <- args[2]
-log.file <- paste("logTask",N,".txt",sep="")
-writeLines(c(""), log.file)
-
-# Choose 1000  sites by N, and select chunk within FullAnnotation
-# siteInds <- (1:1000) + (N - 1) * 1000
-# if (N > 866) {
-#   siteInds <- siteInds[1:836]
-# }
-# nsites <- length(siteInds)
-
-siteInds <- (1:3) + (N - 1) * 3
-nsites <- length(siteInds)
-
-print(paste("Making chunk of nsites =", nsites))
-chunk <- FullAnnotation[siteInds,]
-rm(FullAnnotation)
-gc()
-
-# must add to prevent crash, by checking for presence of rogue decoy Stan (.rds) file
-checkFile <- "model3.rds"
-
-
-### FXNS ###
-# Function used to read in data from each site
-# changed to now use a "chunk" instead of FullAnnotation, for speed and memory
-# requires index 1-1000 instead of siteIndex
-site <- function (ind) {
-  # extract CpG site xx to start
-  temp <- chunk[ind,]
-  
-  # here we use all patient samples, excluding glands
-  indices <- c(9:13, 15:39, 46, 47, 57, 58, 72:75)
-  patientLabel <- substr(colnames(temp[indices]), 1, 1)
-  patientLabel[10:12] <- "K*"
-  sideLabel <- substr(colnames(temp[indices]), 2, 2)
-  tissueLabel <- sideLabel
[…]

### 2020-09-25 StanCVarsScript1.R

@@ -1,137 +1,0 @@
-# script for extracting variance from Stan model on random 5K sites
-# part 1
-
-
-library(rstan)
-library(gtools)
-
-# load data
-load("myFA.Rdata")
-
-args <- commandArgs(trailingOnly = TRUE)
-N <- as.numeric(args[1])
-out.file <- paste("Results_", args[1], ".Rdata", sep="")
-
-# Function used to read in data from each site
-
-site <- function (site_no) {
-  # extract CpG site xx to start
-  temp <- FullAnnotation[site_no,]
-  
-  # here we use all patient samples, excluding glands
-  indices <- c(9:13, 15:39,46,47,57,58,72:75)
-  patientLabel <- substr(colnames(temp[indices]),1,1)
-  patientLabel[10:12] <- "K*"
-  sideLabel <- substr(colnames(temp[indices]),2,2)
-  tissueLabel <- sideLabel
-  tissueLabel[tissueLabel %in% c("A","B")] <- "T"   #replace A and B with T
-  tumorIndicator <- 1*(tissueLabel=="T")
-  
-  work <- data.frame(beta=logit(t(temp[indices])), patient = patientLabel, tissue=tissueLabel, side=sideLabel, tInd=tumorIndicator)
-  
-  return(work)
-}
-
-
-
-# Function to build stan model for each site
-
-stanfit3 <- function (dataset) {
-  
-  stanDat <- list(pID = as.integer(factor(dataset$patient)),
-                  tInd = dataset$tInd,
-                  N = nrow(dataset),
-                  P = nlevels(dataset$patient),
-                  y = dataset[,1])
-  
-  
-  # Using Model 3: add intra-tumoral variances
-  stanFit3 <- stan(file="model3.stan", data=stanDat, control = list(adapt_delta = 0.999))
-  
-  return(stanFit3=stanFit3)
-}
-
-
-
-# Function to get mode
-
-getmode <- function(v) {
-  uniqv <- unique(v)
-  uniqv[which.max(tabulate(match(v, uniqv)))]
-}
-
-
-# Choose 1k sites by N
-
-siteInds <- ((N*1000-999):(N*1000))
-if (N>866) { siteInds <- ((N*1000-999):866836) }
-nsites <- length(siteInds)
-
-
-
[…]

### 2020-09-25 estimateTCGApriors.R

@@ -9,8 +9,8 @@ library(ggplot2)
 
 ### INITIALIZE ###
 
-log.file <- "logTCGA.txt"
-writeLines(c(""), log.file)
+#log.file <- "logTCGA.txt"
+#writeLines(c(""), log.file)
 
 # Kevin's working directory
 setwd("/Users/kevinmurgas/Documents/Data+ project/TCGA data")
@@ -105,12 +105,90 @@ fitplot <- function(data, line) {
   return(p1)
 }
 
+calcSigmaAll <- function (siteData, plotFlag=FALSE) {
+  siteBeta <- siteData$beta
+  
+  # sigmaP: take mean of normal samples
+  mu <- mean(siteBeta[!tumorLabel])
+  b <- siteBeta[!tumorLabel]
+  sigmaP <- var(b)
+  
+  # sigmaPT: take mean of patient's tumor samples and normalize by patient normal
+  normBeta <- c()
+  for(pat in unique(patLabel)){
+    if(length(intersect(which(pat==patLabel),which(!tumorLabel)))){ # only if pat has normal
+      normIndex <- intersect(which(pat==patLabel),which(!tumorLabel)) # normal sample index
+      tumIndex <- intersect(which(pat==patLabel),which(!!tumorLabel)) # tumor sample indices
+      if(length(tumIndex)>1){ tum<-mean(siteBeta[tumIndex]) }else{ tum<-siteBeta[tumIndex] }
+      normBeta <- c(normBeta, tum-siteBeta[normIndex])
+    }
+  }
+  betaT <- mean(normBeta)
+  c <- normBeta
+  sigmaPT <- var(c)
+  
+  # sigmaT: all samples var method, normalize all tumor samples by patient tumor mean
+  normBetaT <- c()
+  for(pat in unique(patLabel)){
+    if(length(intersect(which(pat==patLabel),which(!!tumorLabel)))>1){ # only if pat has multiple tumor
+      tumIndex <- intersect(which(pat==patLabel),which(!!tumorLabel)) # tumor sample indices
+      mean <- mean(siteBeta[tumIndex])
+      normBetaT <- c(normBetaT, siteBeta[tumIndex] - mean)
+    }
+  }
+  d <- normBetaT
+  sigmaT <- var(d)
+  
[…]

### 2020-09-25 genewise.R

@@ -1,0 +1,315 @@
+##### Gene-wise analyses
+# Kevin Murgas
+# Looking at CpG sites within each individual gene
+# Using full-scale (850K site) Stan model results
+
+# Some analyses below (will) include:
+# Gene Average of Sigma Parameters (PCA, clustering)
+# Distance-based analysis (sigmas vs CpG distance in gene)
+# F vs R strand methylation correlation
+# Differential methylation validation with mu+betaT
+
+# Kevin's working directory
+setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
+
+# load annotation, Stan results
+load("myFA_bulkonly.Rdata")
+load("FullResultsTCGA_relaxgamma3_postsum3.Rdata")
+
+# load GeneInfo (gene names and regions for each site)
+load("GeneInfo.Rdata")
+
+### if GeneInfo not made yet ###
+
+# make siteGenes, a list of genes at each site
+fsplit <- function(str) unique(unlist(strsplit(str,split=";")))
+siteGenes <- lapply(FullAnnotation$UCSC_RefGene_Name, fsplit) # takes ~38 sec
+
+# make siteRegions, a label for regulatory region at each site
+# use 5'>3' priority: TSS1500>TSS200>5UTR>1stExon>Body>ExonBnd>3UTR>blank
+fsplitReg <- function(str) {
+    worklist <- unique(unlist(strsplit(str, split = ";")))
+    worklist[worklist == "3'UTR"] <- "3UTR"
+    worklist[worklist == "5'UTR"] <- "5UTR"
+    worklist[worklist == "5URT"] <- "5UTR"
+    #if (length(worklist)==0) { worklist <- "blank"}
+    return(unique(worklist))
+  }
+siteRegions <- lapply(FullAnnotation$UCSC_RefGene_Group, fsplitReg)
+
+# pre-index all genes
+geneInd <- function (geneStr) {
+  strx = paste0("(?<=^|;)", geneStr, "(?=$|;)") # match string - look for start/end of string or semicolon on either side of match
+  matchinds = grep(strx,FullAnnotation$UCSC_RefGene_Name,perl=TRUE); # check UCSC name
+  return(matchinds)
+}
+ptm<-proc.time()
[…]

### 2020-09-25 genewise2.R

@@ -1,0 +1,469 @@
+##### Gene-wise analysis
+# UPDATED
+# Kevin Murgas
+
+library(tidyverse)
+select <- dplyr::select
+rename <- dplyr::rename
+filter <- dplyr::filter
+
+setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
+library(here)
+
+# load annotated data, Stan results
+load(here("myFA_bulkonly.Rdata"))
+load(here("FullResultsTCGA_relaxgamma3_postsum3.Rdata"))
+
+
+med_vals <-
+  tibble(
+    Name = FullAnnotation[, 1],
+    mu = mu_full$p50,
+    betaT = betaT_full$p50,
+    sigmaE = sigmaE_full$p50,
+    sigmaP = sigmaP_full$p50,
+    sigmaPT = sigmaPT_full$p50,
+    sigmaT = sigmaT_full$p50,
+    prob1 = prob1_full,
+    prob2 = 1 - prob2_full
+  ) %>% mutate(
+    logP = log(sigmaP),
+    logPT = log(sigmaPT),
+    logT = log(sigmaT)
+  )
+
+tibble(FullAnnotation[,c(1,2,3,5,6,8)]) %>% rename(Name = IlmnID) %>%
+  left_join(by="Name", med_vals)
+
+# first take gene averages of every parameter (and # sites)
+
+
+
+
+#### test code for plotting overlaid histograms ####
+
+library(tidyverse)
+setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
+librray(here)
+
+prepped_data <- read_csv(here("data-stan_rg3ps2", "prepped_data.csv"))
+genemask = !is.na(prepped_data$UCSC_RefGene_Name)
+
+fBG <- function(x, a, m1, s1, m2, s2) a*dnorm(x,m1,s1) + (1-a)*dnorm(x,m2,s2) # bimodal gaussian
+
+post_prior_histogram <- function(postvals,mask,Rhat,priorfun,bw,xl,titlestr) {
+  # stacked histogram colored by gene-associated sites
+  # and prior curve overlaid
+  plot_df <- data.frame(val=postvals, gene_assoc=mask)
+  p1 <- ggplot(plot_df, aes(x=val)) +
+    geom_histogram(aes(y=..count../length(mask),fill=gene_assoc), binwidth=bw) +
+    stat_function(fun=function(x) bw*priorfun(x),geom="line",color="black") +
[…]

### 2020-09-25 individualGenes.R

@@ -16,21 +16,22 @@ setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
 setwd("D:/DataPlus2017/Data")
 
 # load annotation, Stan results, and gene scores
-load("myFA.Rdata")
-load("StanCfullResults.Rdata")
-load("StanCfullResults2.Rdata")
-load("StanCfullResults3.Rdata")
-load("ScoresFull.Rdata")
+load("myFA_bulkonly.Rdata")
+#load("StanCfullResults.Rdata")
+#load("StanCfullResults2.Rdata")
+#load("StanCfullResults3.Rdata")
+#load("ScoresFull.Rdata")
+load("FullResultsTCGA.Rdata")
 
 
 ### INITIALIZE
 
 # # create list of unique gene names and gene regions at each site
-# geneNames <- list(866836)
-# geneRegions <- list(866836)
-# for (i in 1:866836) {
-#   if (!(i %% 86684)) {
-#     print(paste(i/8668.4,"% complete"))
+# geneNames <- list(866091)
+# geneRegions <- list(866091)
+# for (i in 1:866091) {
+#   if (!(i %% 86609)) {
+#     print(paste(i/8660.9,"% complete"))
 #   }
 #   geneNames[[i]] <- unique(unlist(strsplit(FullAnnotation$UCSC_RefGene_Name[i],split=";")))
 #   worklist <- unique(unlist(strsplit(FullAnnotation$UCSC_RefGene_Group[i],split=";")))
@@ -44,8 +45,10 @@ load("ScoresFull.Rdata")
 load("GeneInfo.Rdata")
 
 # make vector of unique genes and regions
-uniqueGenes <- unique(unlist(geneNames))
-regionTypes <- unique(unlist(geneRegions))
+uniqueGenes <- unique(unlist(siteGenes))
+regionTypes <- unique(unlist(siteRegions))
+#uniqueGenes <- unique(unlist(geneNames))
+#regionTypes <- unique(unlist(geneRegions))
 
 # make dataframes for median and mean values of each parameter
 dfMedians <- cbind.data.frame(mu_Cfull$p50, betaT_Cfull$p50, sigmaE_Cfull$p50, sigmaP_Cfull$p50, sigmaPT_Cfull$p50, sigmaT_Cfull$p50)

### 2020-09-25 loadSCRuns.R

@@ -10,24 +10,27 @@ library(reshape2)
 ### LOADING FOR STAN MODEL RESULTS
 
 # Set working directory to normal data wd then folder ResultsC
-setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data/NewResultsC")
+setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data/ResultsTCGA_relaxgamma3")
 
-betaT_Cfull3 <-
+mu_full <-
   data.frame(
-    mean = numeric(866836),
-    p2.5 = numeric(866836),
-    p25 = numeric(866836),
-    p50 = numeric(866836),
-    p75 = numeric(866836),
-    p97.5 = numeric(866836)
+    mean = numeric(866091),
+    sem = numeric(866091),
+    p50 = numeric(866091),
+    n_eff = numeric(866091),
+    Rhat = numeric(866091)
   )
 
-mu_Cfull3 <- betaT_Cfull3
-sigmaE_Cfull3 <- betaT_Cfull3
-sigmaP_Cfull3 <- betaT_Cfull3
-sigmaPT_Cfull3 <- betaT_Cfull3
-sigmaT_Cfull3 <- betaT_Cfull3
-PTprob3 <- numeric(866836)
+betaT_full <- mu_full
+sigmaP_full <- mu_full
+sigmaPT_full <- mu_full
+sigmaT_full <- mu_full
+sigmaE_full <- mu_full
+lp_full <- mu_full
+
+CpGscore_full <- data.frame(prob = numeric(866091),
+                         mean = numeric(866091),
+                         median = numeric(866091))
 
 nsites <- 10000
 filesPresent <- (1:87)
@@ -38,19 +41,20 @@ for (i in filesPresent) {
   load(fileNames[fileNames == paste("StanCParResults_",i,".Rdata", sep="")])
   inds <- (1:nsites) + (i - 1) * nsites
   if (i==87) {
-    inds <- inds[1:6836]
+    inds <- inds[1:6091]
   }
   allInds <- append(allInds, inds)
[…]

### 2020-10-01 StanAnalysis_updated.Rmd
differential methylation, clean up

@@
 
-stanFile = "FullResultsTCGA_relaxgamma3_postsum3.Rdata"
+stanFile = "FullResultsTCGA_relaxgamma3.Rdata"
 load(here(stanFile))
@@
 
-<!-- \newpage -->
-<!-- \section{Patient-specific analysis} -->
-<!-- Basic overview: using a specified number of genes (here 100), take top genes (here using score_prob bootstrapping percentiles) to select a "gene signature" of sorts.
-Then for each patient, calculate the mean pair-wise distance within the individual patients' tumor samples.
-Method 1 is these raw values, method 2 normalizes by the mean PWD of all normal samples.
-Method 3 just examines the overall mean PWD for all tumor samples, normalized by mean PWD of all normal.
--->
-
-<!-- ```{r PatientPWD, echo=FALSE, message=FALSE, warning=FALSE, comment=NA, fig.height=9, fig.width=7} -->
-<!-- library(gtools) -->
-
-<!-- # pull out all samples beta values -->
-<!-- all_beta <- FullAnnotation[,10:57] -->
-
-<!-- # extract patient and tissue labels -->
-<!-- patientLabel <- substr(colnames(all_beta), 1, 1) -->
-<!-- patientLabel[10:12] <- "K*" -->
-<!-- tissueLabel <- substr(colnames(all_beta), 2, 2) -->
-<!-- tissueLabel[!(tissueLabel %in% "N")] <- "T"   #replace A/B/D/M with T -->
-
-<!-- pats = patientLabel %>% sort %>% unique -->
-<!-- npat = length(pats) -->
-
-<!-- enrichedGenes <- permed_data %>% filter(score_prob_v3 < quantile(score_prob_v3, 0.05, na.rm = TRUE)) %>% -->
-<!--   select(UCSC_RefGene_Name, score_prob, score_prob_v3, n) %>% -->
-<!--   rename(Gene=UCSC_RefGene_Name, Pctile=score_prob_v3) -->
-
-<!-- #ngene = 70 # using top N genes -->
-<!-- #genes = enrichedGenes %>% top_n(-ngene,Pctile) %>% pull(Gene) -->
-<!-- #genes = genes[1:ngene] -->
-
-<!-- genes = enrichedGenes$Gene -->
-<!-- ngene = length(genes) -->
-
-<!-- pwd <- function(x) { -->
-<!--   mean(dist(x)) -->
-<!-- } -->
-<!-- pwd2 <- function(x) { -->
-<!--   abs(diff(x)) -->
-<!-- } -->
-
-<!-- signature_mat1 = matrix(,ngene,npat) # raw tumor pwd -->
-<!-- signature_mat2 = matrix(,ngene,npat) # tumor pwd normalized by all normal pwd -->
-<!-- for (igene in 1:ngene) { -->
-<!--   # each gene, get gene inds, calculate PWD for the tumor samples -->
-<!--   geneind = grep(genes[igene], FullAnnotation$UCSC_RefGene_Name) -->
-
-<!--   # calculate PWD of each tumor -->
-<!--   tum_pwd <- numeric(npat) -->
-<!--   for (ipat in 1:npat) { -->
-<!--     # each patient, get tumor sample inds -->
-<!--     patind <- which(patientLabel == pats[ipat] & tissueLabel == "T") -->
-<!--     tum_pwd[ipat] <- all_beta[geneind,patind] %>% logit %>% apply(1,pwd2) %>% mean -->
-<!--   } -->
-
-<!--   # calculate PWD of all normal -->
-<!--   norm_pwd <- all_beta[geneind,tissueLabel == "N"] %>% logit %>% apply(1, pwd) %>% mean -->
-
-<!--   signature_mat1[igene,] = tum_pwd -->
-<!--   signature_mat2[igene,] = tum_pwd/norm_pwd -->
-<!-- } -->
-
-<!-- #cat("  \n\\newline  \nMethod 1 - raw intratumor PWD within each patient") -->
-<!-- #signature_mat1 %>% as_tibble %>% `colnames<-`(pats) %>% gather %>% -->
-<!-- #  mutate(gene=rep(genes,npat)) %>% rename(Patient=key,PWD=value) %>% -->
-<!-- #  ggplot + geom_tile(aes(y=gene,x=Patient,fill=PWD)) + ylab(paste0(ngene," genes")) + -->
-<!-- #  theme(axis.text.y = element_blank()) -->
-<!-- #cat("  \n\\newline  \nMethod 2 - intratumor PWD normalized by all normal PWD") -->
-<!-- #signature_mat2 %>% as_tibble %>% `colnames<-`(pats) %>% gather %>% -->
-<!-- #  mutate(gene=rep(genes,npat)) %>% rename(Patient=key,PWD=value) %>% -->
-<!-- #  ggplot + geom_tile(aes(y=gene,x=Patient,fill=PWD)) + ylab(paste0(ngene," genes")) + -->
-<!-- #  theme(axis.text.y = element_blank()) -->
-<!-- #cat("  \n\\newline  \nMethod 3 - all tumor PWD normalized by all normal PWD") -->
-<!-- #signature_mat3 %>% as_tibble %>% mutate(gene=genes) %>% rename(PWD=V1) %>% -->
-<!-- #  ggplot + geom_tile(aes(y=gene,x=1,fill=PWD)) + ylab(paste0(ngene," genes")) + -->
-<!-- #  theme(axis.text.y = element_blank()) -->
-
-<!-- library(pheatmap) -->
-<!-- library(dendsort) -->
-
-<!-- cat("  \n\\newline  \nMethod 1 - raw intratumor PWD within each patient") -->
-<!-- df1 = as.data.frame(signature_mat1) -->
-<!-- rownames(df1) <- genes -->
-<!-- colnames(df1) <- pats -->
-<!-- sort_hclust <- function(...) as.hclust(dendsort(as.dendrogram(...))) -->
-<!-- mat_cluster_cols <- sort_hclust(hclust(dist(t(df1)))) -->
-<!-- mat_cluster_rows <- sort_hclust(hclust(dist(df1))) -->
-<!-- pheatmap( -->
-<!--   mat=df1, -->
-<!--   cluster_cols = mat_cluster_cols, -->
-<!--   cluster_rows = mat_cluster_rows, -->
-<!--   show_rownames = FALSE, -->
-<!--   show_colnames = TRUE, -->
-<!--   main = "Top 1118 Genes: Tumor PWD" -->
-<!-- ) -->
-
-<!-- cat("  \n\\newline  \nMethod 2 - intratumor PWD normalized by all normal PWD") -->
-<!-- df1 = as.data.frame(signature_mat2) -->
-<!-- rownames(df1) <- genes -->
-<!-- colnames(df1) <- pats -->
-<!-- sort_hclust <- function(...) as.hclust(dendsort(as.dendrogram(...))) -->
-<!-- mat_cluster_cols <- sort_hclust(hclust(dist(t(df1)))) -->
-<!-- mat_cluster_rows <- sort_hclust(hclust(dist(df1))) -->
-<!-- pheatmap( -->
-<!--   mat=df1, -->
-<!--   cluster_cols = mat_cluster_cols, -->
-<!--   cluster_rows = mat_cluster_rows, -->
-<!--   show_rownames = FALSE, -->
-<!--   show_colnames = TRUE, -->
-<!--   main = "Top 1118 Genes: Tumor PWD" -->
-<!-- ) -->
-<!-- ``` -->
-
 \newpage
@@
 enrichedPaths <- tibble(Pathway = enrich_sigmaP$Description, ngenes = enrich_sigmaP$Count, padj = enrich_sigmaP$p.adjust) %>% arrange(padj)
+geneList <- enrich_reactome$geneID %>% strsplit(split='/') %>%
+  lapply(function(x) AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, x, 'SYMBOL', 'ENTREZID')) %>%
+  lapply(function(x) paste(x,collapse=";")) %>% unlist
+enrichedPaths$Genes = geneList
 
@@
 enrichedPaths <- tibble(Pathway = enrich_sigmaPT$Description, ngenes = enrich_sigmaPT$Count, padj = enrich_sigmaPT$p.adjust) %>% arrange(padj)
+geneList <- enrich_reactome$geneID %>% strsplit(split='/') %>%
+  lapply(function(x) AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, x, 'SYMBOL', 'ENTREZID')) %>%
+  lapply(function(x) paste(x,collapse=";")) %>% unlist
+enrichedPaths$Genes = geneList
 
@@
 enrichedPaths <- tibble(Pathway = enrich_sigmaT$Description, ngenes = enrich_sigmaT$Count, padj = enrich_sigmaT$p.adjust) %>% arrange(padj)
+geneList <- enrich_reactome$geneID %>% strsplit(split='/') %>%
+  lapply(function(x) AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, x, 'SYMBOL', 'ENTREZID')) %>%
+  lapply(function(x) paste(x,collapse=";")) %>% unlist
+enrichedPaths$Genes = geneList
 
@@
 ```
-
-<!-- \newpage -->
-<!-- \subsection{A4: Score Prob} -->
-<!-- Genes are selected with cutoff >95th percentile (adjusted bootstrap), showing top 50 genes/pathways -->
-<!-- ```{r App_score_prob, echo=FALSE, comment=NA, message=FALSE} -->
-<!-- enrichedGenes <- permed_data %>%  -->
-<!--   select(UCSC_RefGene_Name, score_prob, score_prob_v3, n) %>% -->
-<!--   rename(Gene=UCSC_RefGene_Name, Score_median=score_prob, Pctile=score_prob_v3) %>% -->
-<!--   filter(Pctile > quantile(Pctile, 0.95, na.rm = TRUE)) %>% -->
-<!--   arrange(Pctile) -->
-
-<!-- enrichedPaths <- tibble(Pathway = enrich_score_prob$Description, ngenes = enrich_score_prob$Count, padj = enrich_score_prob$p.adjust) %>% arrange(padj) -->
-
-<!-- ngene <- dim(enrichedGenes)[1] -->
-<!-- npath <- length(enrich_score_prob$Description) -->
-
-<!-- cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath)) -->
-
-<!-- nshow = 50 -->
-<!-- if (ngene > nshow) { -->
-<!--   enrichedGenes <- enrichedGenes[1:nshow,] -->
-<!-- } -->
-<!-- if (npath > nshow) { -->
-<!--   enrichedPaths <- enrichedPaths[1:nshow,] -->
-<!-- } -->
-
-<!-- enrichedGenes %>% pander() -->
-
-<!-- cat("Score_prob top 50 enriched paths:  \n") -->
-<!-- enrichedPaths %>% pander() -->
-<!-- ``` -->
-
-<!-- \newpage -->
-<!-- \subsection{A5: Score Mean} -->
-<!-- Genes are selected with cutoff >95th percentile (adjusted bootstrap), showing top 50 genes/pathways -->
-<!-- ```{r App_score_mean, echo=FALSE, comment=NA, message=FALSE} -->
-<!-- enrichedGenes <- permed_data %>%  -->
-<!--   select(UCSC_RefGene_Name, score_mean, score_mean_v3, n) %>% -->
-<!--   rename(Gene=UCSC_RefGene_Name, Score_median=score_mean, Pctile=score_mean_v3) %>% -->
-<!--   filter(Pctile > quantile(Pctile, 0.95, na.rm = TRUE)) %>% -->
-<!--   arrange(Pctile) -->
-
-<!-- enrichedPaths <- tibble(Pathway = enrich_score_mean$Description, ngenes = enrich_score_mean$Count, padj = enrich_score_mean$p.adjust) %>% arrange(padj) -->
-
-<!-- ngene <- dim(enrichedGenes)[1] -->
-<!-- npath <- length(enrich_score_mean$Description) -->
-
-<!-- cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath)) -->
-
-<!-- nshow = 50 -->
-<!-- if (ngene > nshow) { -->
-<!--   enrichedGenes <- enrichedGenes[1:nshow,] -->
-<!-- } -->
-<!-- if (npath > nshow) { -->
-<!--   enrichedPaths <- enrichedPaths[1:nshow,] -->
-<!-- } -->
-
-<!-- enrichedGenes %>% pander() -->
-
-<!-- cat("Score_mean top 50 enriched paths:  \n") -->
-<!-- enrichedPaths %>% pander() -->
-<!-- ``` -->
 
@@
 ```
-
-\newpage
-\subsection{A7: Score Median2}
-Genes are selected with cutoff >95th percentile (adjusted bootstrap), showing top 50 genes/pathways
-```{r App_score_median2, echo=FALSE, comment=NA, message=FALSE}
-enrichedGenes <- permed_data %>% 
-  select(UCSC_RefGene_Name, score_median2, score_median2_v3, n) %>%
-  rename(Gene=UCSC_RefGene_Name, Score_median2=score_median2, Pctile=score_median2_v3) %>%
-  filter(Pctile > quantile(Pctile, 0.95, na.rm = TRUE)) %>%
-  arrange(Pctile)
-
-enrichedPaths <- tibble(Pathway = enrich_score_median2$Description, ngenes = enrich_score_median2$Count, padj = enrich_score_median2$p.adjust) %>% arrange(padj)
-
-ngene <- dim(enrichedGenes)[1]
-npath <- length(enrich_score_median2$Description)
-
-cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath))
-
-nshow = 50
-if (ngene > nshow) {
-  enrichedGenes <- enrichedGenes[1:nshow,]
-}
-if (npath > nshow) {
-  enrichedPaths <- enrichedPaths[1:nshow,]
-}
-
-enrichedGenes %>% pander()
-
-cat("Score_median2 top 50 enriched paths:  \n")
-enrichedPaths %>% pander()
-```
-
-\newpage
-\subsection{A9: Score Regularized}
-Genes are selected with cutoff >95th percentile (adjusted bootstrap), showing top 50 genes/pathways
-```{r App_score_reg, echo=FALSE, comment=NA, message=FALSE}
-enrichedGenes <- permed_data %>% 
-  select(UCSC_RefGene_Name, score_reg, score_reg_v3, n) %>%
-  rename(Gene=UCSC_RefGene_Name, Score_reg=score_reg, Pctile=score_reg_v3) %>%
-  filter(Pctile > quantile(Pctile, 0.95, na.rm = TRUE)) %>%
-  arrange(Pctile)
-
-enrichedPaths <- tibble(Pathway = enrich_score_reg$Description, ngenes = enrich_score_reg$Count, padj = enrich_score_reg$p.adjust) %>% arrange(padj)
-
-ngene <- dim(enrichedGenes)[1]
-npath <- length(enrich_score_reg$Description)
-
-cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath))
-
-nshow = 50
-if (ngene > nshow) {
-  enrichedGenes <- enrichedGenes[1:nshow,]
-}
-if (npath > nshow) {
-  enrichedPaths <- enrichedPaths[1:nshow,]
-}
-
-enrichedGenes %>% pander()
-
-cat("Score_reg top 50 enriched paths:  \n")
-enrichedPaths %>% pander()
-```

### 2020-10-01 conserved_genes_and_paths.Rmd
differential methylation, clean up

@@
 ```
-
-<!-- \newpage -->
-<!-- \section{Patient-specific analysis} -->
-<!-- Basic overview: using a specified number of genes (here 100), take top genes (here using score_med bootstrapping percentiles) to select a "gene signature" of sorts.
-Then for each patient, calculate the mean pair-wise distance within the individual patients' tumor samples.
-Method 1 is these raw values, method 2 normalizes by the mean PWD of all normal samples.
-Method 3 just examines the overall mean PWD for all tumor samples, normalized by mean PWD of all normal.
--->
-
-<!-- ```{r PatientPWD, echo=FALSE, message=FALSE, warning=FALSE, comment=NA, fig.height=9, fig.width=7} -->
-<!-- library(gtools) -->
-
-<!-- # Load raw methylation data -->
-<!-- load(here("myFA_bulkonly.Rdata")) -->
-
-<!-- # pull out all samples beta values -->
-<!-- all_beta <- FullAnnotation[,10:57] -->
-
-<!-- # extract patient and tissue labels -->
-<!-- patientLabel <- substr(colnames(all_beta), 1, 1) -->
-<!-- patientLabel[10:12] <- "K*" -->
-<!-- tissueLabel <- substr(colnames(all_beta), 2, 2) -->
-<!-- tissueLabel[!(tissueLabel %in% "N")] <- "T"   #replace A/B/D/M with T -->
-
-<!-- pats = patientLabel %>% sort %>% unique -->
-<!-- npat = length(pats) -->
-
-<!-- enrichedGenes <- permed_data %>% filter(score_median_v3 > 0.95) %>% -->
-<!--   select(UCSC_RefGene_Name, score_median, score_median_v3, n) %>% -->
-<!--   rename(Gene=UCSC_RefGene_Name, Pctile=score_median_v3) -->
-
-<!-- #ngene = 70 # using top N genes -->
-<!-- #genes = enrichedGenes %>% top_n(-ngene,Pctile) %>% pull(Gene) -->
-<!-- #genes = genes[1:ngene] -->
-
-<!-- genes = enrichedGenes$Gene -->
-<!-- ngene = length(genes) -->
-
-<!-- pwd <- function(x) { -->
-<!--   mean(dist(x)) -->
-<!-- } -->
-<!-- pwd2 <- function(x) { -->
-<!--   abs(diff(x)) -->
-<!-- } -->
-
-<!-- signature_mat1 = matrix(,ngene,npat) # raw tumor pwd -->
-<!-- for (igene in 1:ngene) { -->
-<!--   # each gene, get gene inds, calculate PWD for the tumor samples -->
-<!--   geneind = grep(genes[igene], FullAnnotation$UCSC_RefGene_Name) -->
-
-<!--   # calculate PWD of each tumor -->
-<!--   tum_pwd <- numeric(npat) -->
-<!--   for (ipat in 1:npat) { -->
-<!--     # each patient, get tumor sample inds -->
-<!--     patind <- which(patientLabel == pats[ipat] & tissueLabel == "T") -->
-<!--     tum_pwd[ipat] <- all_beta[geneind,patind] %>% logit %>% apply(1,pwd2) %>% mean -->
-<!--   } -->
-
-<!--   signature_mat1[igene,] = tum_pwd -->
-<!-- } -->
-
-<!-- library(pheatmap) -->
-<!-- library(dendsort) -->
-
-<!-- cat("  \n\\newline  \nMethod 1 - raw intratumor PWD within each patient") -->
-<!-- df1 = as.data.frame(signature_mat1) -->
-<!-- rownames(df1) <- genes -->
-<!-- colnames(df1) <- pats -->
-<!-- sort_hclust <- function(...) as.hclust(dendsort(as.dendrogram(...))) -->
-<!-- mat_cluster_cols <- sort_hclust(hclust(dist(t(df1)))) -->
-<!-- mat_cluster_rows <- sort_hclust(hclust(dist(df1))) -->
-<!-- pheatmap( -->
-<!--   mat=df1, -->
-<!--   cluster_cols = mat_cluster_cols, -->
-<!--   cluster_rows = mat_cluster_rows, -->
-<!--   show_rownames = FALSE, -->
-<!--   show_colnames = TRUE, -->
-<!--   main = sprintf("Top %i Genes: Tumor PWD", ngene) -->
-<!-- ) -->
-
-<!-- df2 = df1 -->
-<!-- for (ipat in 1:npat) { -->
-<!--   df2[,ipat] = df1[,ipat] > quantile(df1[,ipat],0.5) -->
-<!-- } -->
-<!-- cat("  \n\\newline  \nMethod 1 - Lowest 50% PWD Genes within each patient") -->
-<!-- pheatmap( -->
-<!--   mat=df2, -->
-<!--   cluster_cols = mat_cluster_cols, -->
-<!--   cluster_rows = mat_cluster_rows, -->
-<!--   show_rownames = FALSE, -->
-<!--   show_colnames = TRUE, -->
-<!--   main = sprintf("Top %i Genes: Lowest 50 percentile Tumor PWD per patient", ngene) -->
-<!-- ) -->
-
-<!-- ``` -->

### 2020-10-01 differential_methylation.Rmd
differential methylation, clean up

---
title: "DNAm Hierarchical Model - Fixed Effects Differential Methylation"
author: "Kevin Murgas"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: true
    number_section: true
    toc: true
---

\newpage
\section{Differential Methylation}
Differential methylation is typically defined using a basic linear model over the data points between normal and tumor samples.
$$ M_i = \mu + \Delta M*\text{tInd}(i) $$
Where tInd is a binary indicator for tumor samples.
\newline This linear fit captures the normal mean (intercept) and shift in methylation in tumors (slope). The slope ($\Delta M$) is used to define differentially methylated regions by first z-score normalizing and then determining statistical significance using a permutation based false discovery rate (FDR)
$$ z = \text{mean}(\Delta M) / \text{s.d.}(\Delta M) $$
\newline I am currently a bit unclear on the FDR step. Irizarry et al 2009 reference : Efron B, Tibshirani R, Storey JD, Tusher V. Empirical Bayes analysis of a microarray experiment. Journal of the American statistical association. 2001 Dec 1;96(456):1151-60.
\newline Z-scores can convert to p-values, and Benjamini-Hochberg step-up can account for multiple hypothesis correction. The permutation could be from shuffling the normal/tumor labels on 1000 draws.

\newline The minfi dmpFinder approach uses a linear fit (limma package), and computes a t-statistic for the slope by the following formula:
$$ t = (\Delta M) / (\sigma_{\Delta M} * \sigma_{resid}) $$
Which is then converted to a p-value using the t-distribution with df= # samples - 2. The p-value is then corrected by Storey et al. 2003 Bayesian q-value method, which is similar to Benjamini-Hochberg linear step-up FDR procedure but divides by a factor of an estimated prior.

\section{Model Description: Fixed Effects}
\subsection{Model equation}
Let i=sample, j=patient
$$ y_i = \mu + a_j +  \text{tInd}(i)*(\beta_T + b_j + c_i) $$
$$ a_j \sim N(0,\sigma_P) , b_j \sim N(0,\sigma_{PT}) , c_i \sim N(0,\sigma_T) $$
\subsection{Differential Methylation using betaT}
Effects $\mu$ and $\beta_T$ describe the normal mean methylation and tumor shift in methylation, respectively. Therefore, differentially methylated CpG sites will have significant betaT values. A similar procedure is used as the $\Delta M$ , using a z-score normalization, and permutation based FDR.
\newline This provides a way to identify CpG sites with significant differential methylation. These sites can then be directly compared to C-DMRs of Irizarry 2009 (provided as regions, i.e. base ranges on a given chromosome), or go up to identify genes with either any d.m. CpG sites or a certain fraction. Genes can also be compared to Irizarry C-DMRs (each region has a single gene listed)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
library(gtools)

# set working directory
setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
library(here)

select <- dplyr::select
rename <- dplyr::rename
filter <- dplyr::filter
slice <- dplyr::slice

set.seed(1234)
```

\newpage
\section{Compute DM - raw beta values}
```{r dm_beta, echo=FALSE}
# Load raw methylation data
load(here("myFA_bulkonly.Rdata"))

# extract beta values from data
all_beta <- FullAnnotation[,10:57]
rownames(all_beta)<-FullAnnotation$IlmnID
tissueLabel <- substr(colnames(all_beta), 2, 2)
tissueLabel[!(tissueLabel %in% "N")] <- "T"

# convert to M values, tInd
all_M <- logit(all_beta)
tInd <- tissueLabel=="T"

nsamp = dim(all_beta)[2]
ymean = rowSums(all_M)/nsamp
SStot = rowSums( (all_M - ymean)^2 )

# same procedure as minfi dmpFinder
# use limma to compute linear fits
library(limma)
design <- cbind(muM=1,deltaM=tInd)
fit <- lmFit(all_M, design)
tstat.ord <- fit$coef[,2] / fit$stdev.unscaled[,2] / fit$sigma
pval <- 2*pt(abs(tstat.ord), fit$df.residual, lower.tail=FALSE)
qvalBH <- p.adjust(pval, method="BH")

# qvalue approach used by minfi dmpFinder (from Storey 2003)
library(siggenes)
p0 <- pi0.est(pval[!is.na(pval)])$p0
qvalS <- qvalue.cal(pval, p0)

fit_stats <- cbind(as.data.frame(fit), tstat.ord, pval, qvalBH, p0, qvalS)

## using emprical Bayes stats (not necessary unless <10 samples)
#stats <- as.data.frame(eBayes(fit)) %>%
#  select(-c(stdev.unscaled.muM, stdev.unscaled.deltaM)) %>% #df.residual, df.total 
#  mutate(t.ord.muM=tstat.ord[,1], t.ord.deltaM=tstat.ord[,2]) %>%
#  mutate(qval = p.adjust(pt.ord.deltaM, method="BH"))
## for some reason the qval is off from dmpfinder qval by a constant factor 3.496509

## minfi method
# library(minfi)
# dmpsM <- dmpFinder(as.matrix(all_M), tInd, type="continuous")
# dmpsM_new <- dmpsM[ order(row.names(dmpsM)), ]

library(gt)
data.frame(sumBH = sum(qvalBH<0.05), sumS = sum(qvalS<0.05)) %>%
  rename('# Signif. sites (Benjamini-Hochberg)' = sumBH, '# Signif. sites (Storey)' = sumS)
```

\section{Compute DM - Stan fit betaT}
Use a similar metric, with a t-score based on the median estimate and SEM of posterior. Unclear on how to run significance bootstrapping, but should be find with using the t-test p-value and FDR correction with B-H step up. 
\newline Second approach would be use a z-score based on all of the estimates.
```{r dm_stan, echo=FALSE}
# load Stan fits
stanFile = "FullResultsTCGA_relaxgamma3.Rdata"
load(here(stanFile))

nsamp = dim(all_beta)[2]
sd_all = sd(betaT_full$p50)
betaT_stats <- betaT_full %>% mutate(tstat = p50/(sem*sqrt(n_eff))) %>%
#  mutate(SSreg = sem*n_eff) %>%
#  mutate(SSres = SStot - SSreg) %>%
  mutate(pval = 2*pt(abs(tstat), nsamp-2, lower.tail=FALSE)) %>%
  mutate(qvalBH = p.adjust(pval, method="BH")) %>%
  mutate(zscore = p50/sd_all) %>%
  mutate(pz = 2*pnorm(abs(zscore),lower.tail=FALSE)) %>%
  mutate(qvalBHz = p.adjust(pz, method="BH"))

```

\section{Compare to Irizarry et al 2009}
Irizarry use a similar differential methylation profiling approach in colorectal cancers (different methylation assay), and identify numerous cancer-specific diff. meth. regions (C-DMRs). Compare their list, which reports ranges of bases with single gene labels, to our differential methylation findings.
\newline Doi et al. 2009 (same group as Irizarry 2009), perform a similar analysis to determine C-DMRs in colorectal cancer. These C-DMRs are reported in the "DMR" column of the Illumina EPIC array manifest.
\newline Summarize % of overlap in each case, using specific bases of CpG sites or using gene names.
```{r compare_irizarry, echo=FALSE}
DMR_data = fit_stats$qvalBH <0.05
DMR_beta = betaT_stats$qvalBH < 0.05
CDMR_epic = (FullAnnotation$DMR == "CDMR") #| (FullAnnotation$DMR == "DMR")

DMR_list = cbind(DMR_data, CDMR_epic)
crosstab <- xtabs(~DMR_data + CDMR_epic, DMR_list)
crosstab

DMR_list = cbind(DMR_beta, CDMR_epic)
crosstab <- xtabs(~DMR_beta + CDMR_epic, DMR_list)
crosstab

DMR_list = cbind(DMR_beta, DMR_data)
crosstab <- xtabs(~DMR_beta + DMR_data, DMR_list)
crosstab

chisq.test(crosstab)
fisher.test(crosstab)
OR = (crosstab[1,1]+crosstab[2,2])/(crosstab[1,2]+crosstab[2,1])
phi = (prod(c(crosstab[1,1],crosstab[2,2]))-prod(crosstab[1,2],crosstab[2,1]))/sqrt(prod(rowSums(crosstab))*prod(rowSums(t(crosstab))))
```

### 2020-10-01 prior_sensitivity.Rmd
differential methylation, clean up

---
title: "DNAm Hierarchical Model - Prior Sensitivity Analysis "
author: "Kevin Murgas"
date: "`r Sys.Date()`"
output:
  pdf_document:
    fig_caption: true
    number_section: true
    toc: true
---

\newpage
Prior sensitivity is assessed by looking at Stan fit results of different sets of priors, each with an increasing level of variance relaxation. Current data: 1x (original), 2x, 3x, and 5x.
\newline Priors are assessed by 3 metrics: n_eff and Rhat to assess Stan MCMC convergence, and Kullback-Leibler divergence to assess prior constraints.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
library(gtools)

# set working directory
setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
library(here)

select <- dplyr::select
rename <- dplyr::rename
filter <- dplyr::filter
slice <- dplyr::slice

set.seed(1234)
```

```{r prior_sensitivity, echo=FALSE}

for (i in c(1:4)) {
  stanFile = "FullResultsTCGA_relaxgamma3.Rdata"
  stanFile = switch(i,
                    "FullResultsTCGA_gamma.Rdata",
                    "FullResultsTCGA_relaxgamma2.Rdata",
                    "FullResultsTCGA_relaxgamma3.Rdata",
                    "FullResultsTCGA_relaxgamma5.Rdata")
  print(stanFile)
  #load(here(stanFile))
}

```

### 2020-10-01 genewise2.R

@@ -1,469 +1,0 @@
-##### Gene-wise analysis
-# UPDATED
-# Kevin Murgas
-
-library(tidyverse)
-select <- dplyr::select
-rename <- dplyr::rename
-filter <- dplyr::filter
-
-setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
-library(here)
-
-# load annotated data, Stan results
-load(here("myFA_bulkonly.Rdata"))
-load(here("FullResultsTCGA_relaxgamma3_postsum3.Rdata"))
-
-
-med_vals <-
-  tibble(
-    Name = FullAnnotation[, 1],
-    mu = mu_full$p50,
-    betaT = betaT_full$p50,
-    sigmaE = sigmaE_full$p50,
-    sigmaP = sigmaP_full$p50,
-    sigmaPT = sigmaPT_full$p50,
-    sigmaT = sigmaT_full$p50,
-    prob1 = prob1_full,
-    prob2 = 1 - prob2_full
-  ) %>% mutate(
-    logP = log(sigmaP),
-    logPT = log(sigmaPT),
-    logT = log(sigmaT)
-  )
-
-tibble(FullAnnotation[,c(1,2,3,5,6,8)]) %>% rename(Name = IlmnID) %>%
-  left_join(by="Name", med_vals)
-
-# first take gene averages of every parameter (and # sites)
-
-
-
-
-#### test code for plotting overlaid histograms ####
-
-library(tidyverse)
-setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
-librray(here)
-
-prepped_data <- read_csv(here("data-stan_rg3ps2", "prepped_data.csv"))
-genemask = !is.na(prepped_data$UCSC_RefGene_Name)
-
-fBG <- function(x, a, m1, s1, m2, s2) a*dnorm(x,m1,s1) + (1-a)*dnorm(x,m2,s2) # bimodal gaussian
-
-post_prior_histogram <- function(postvals,mask,Rhat,priorfun,bw,xl,titlestr) {
-  # stacked histogram colored by gene-associated sites
-  # and prior curve overlaid
-  plot_df <- data.frame(val=postvals, gene_assoc=mask)
-  p1 <- ggplot(plot_df, aes(x=val)) +
-    geom_histogram(aes(y=..count../length(mask),fill=gene_assoc), binwidth=bw) +
-    stat_function(fun=function(x) bw*priorfun(x),geom="line",color="black") +
[…]

### 2020-10-02 conserved_genes_and_paths.Rmd
updated prior sensitivity analysis

@@
 
-knit_exit()
-
 ```
@@
   v <- cumsum(runsum)
-  score <- max(c(0,v)) + min(c(0,v))
+  #score <- max(v) # GSEA score
+  score <- max(c(0,v)) + min(c(0,v)) # GSVA score
   GSEA_df$score[iset] <- score
@@
     # for random shuffles just make a random logical mask
-    inds = sample(1:ngene, n_geneset, replace = F)
+    inds = sample(1:ngene, n_geneset, replace = T)
     inmask = logical(ngene)
@@
     v <- cumsum(runsum)
-    score <- max(c(0, v)) + min(c(0, v))
+    #score <- max(v) # GSEA score
+    score <- max(c(0, v)) + min(c(0, v)) # GSVA score
     bootstrap[ishuf] <- score
@@
 names(setlengths) <- setlengths
-cat(sprintf("Shuffling %i set-lengths: 1000x random shuffles (w/o replacement)",length(setlengths)))
+cat(sprintf("Shuffling %i set-lengths: 5000x random shuffles (w/ replacement)",length(setlengths)))
 ptm<-proc.time()
-res <- lapply(setlengths, shuffle_genesets, n_rep = 1000, data = evidence)
+res <- lapply(setlengths, shuffle_genesets, n_rep = 5000, data = evidence)
 proc.time()-ptm

### 2020-10-02 prior_sensitivity.Rmd
updated prior sensitivity analysis

@@
 library(knitr)
-library(gtools)
+#library(gtools)
+library(emdist)
 
@@
 slice <- dplyr::slice
-
-set.seed(1234)
 ```
 
-```{r prior_sensitivity, echo=FALSE}
+```{r prior_sensitivity, echo=FALSE, comment=NA}
+
+n_eff_mat <- data.frame(Params=c("mu","betaT","sigmaP","sigmaPT","sigmaT","sigmaE","lp"))
+Rhat_mat <- n_eff_mat
+KLdiv_mat <- n_eff_mat %>% slice(1:6)
+BCdist_mat <- KLdiv_mat
+EMdist_mat <- KLdiv_mat
 
@@
   print(stanFile)
-  #load(here(stanFile))
+  load(here(stanFile))
+  print("loaded")
+
+  n_eff_temp <-
+  data.frame(
+  mu = mu_full$n_eff,
+  betaT = betaT_full$n_eff,
+  sigmaP = sigmaP_full$n_eff,
+  sigmaPT = sigmaPT_full$n_eff,
+  sigmaT = sigmaT_full$n_eff,
+  sigmaE = sigmaE_full$n_eff,
+  lp = lp_full$n_eff
+  ) %>% gather(param,value,mu:lp) %>% group_by(param) %>% summarize(
+  mean = mean(value),
+  .groups = "drop"
+  ) %>% arrange(match(.$param,c("mu","betaT","sigmaP","sigmaPT","sigmaT","sigmaE","lp")))
+
+  Rhat_temp <-
+  data.frame(
+  mu = mu_full$Rhat,
+  betaT = betaT_full$Rhat,
+  sigmaP = sigmaP_full$Rhat,
+  sigmaPT = sigmaPT_full$Rhat,
+  sigmaT = sigmaT_full$Rhat,
+  sigmaE = sigmaE_full$Rhat,
+  lp = lp_full$Rhat
+  ) %>% gather(param,value,mu:lp) %>% group_by(param) %>% summarize(
+  mean = mean(value),
+  .groups = "drop"
+  ) %>% arrange(match(.$param,c("mu","betaT","sigmaP","sigmaPT","sigmaT","sigmaE","lp")))
+
+  # KL divergence
+  # formula: KL(P||Q) = sum(P(i) * log( P(i)/Q(i) )), where P(i),Q(i) = density
+  KL_temp <- numeric(6)
+  #Bhattacharya distance
+  # formula: BC = sum( sqrt(P(i)*Q(i)) )
+  BC_temp <- numeric(6)
+  # Earth-movers (Wasserstein) distance
+  EM_temp <- numeric(6)
+
+  for (j in c(1:6)) {
+    # choose parameter
+    data <- switch(j,
+                   mu_full$p50,
+                   betaT_full$p50,
+                   sigmaP_full$p50,
+                   sigmaPT_full$p50,
+                   sigmaT_full$p50,
+                   sigmaE_full$p50)
+    breaks <- switch(j,
+                     seq(-5,5,0.1),
+                     seq(-6,6,0.12), #5.5,.11
+                     seq(0,3,0.03), #2.6
+                     seq(0,3,0.03),
+                     seq(0,3,0.03), #1.2
+                     seq(0,3,0.03)) #2.5
+    P <- hist(data, plot=FALSE, breaks=101) # run hist to get density and midpoints
+    x <- P$mids
+    # define prior density by pre-defined priors at P midpoints
+    if (i==1){ # non relaxed (1x)
+      priordens <- switch(j,
+                      0.3557*dnorm(x, -3.0675, 0.75214) + 0.6443*dnorm(x, 1.1290, 1.3920),
+                      dcauchy(x, -0.04051, 0.19330),
+                      dgamma(x, 5.1989, 16.6192),
+                      dgamma(x, 3.5539, 5.5324),
+                      dgamma(x, 6.8383, 25.7560),
+                      dgamma(x,2,2))
+    } else if (i==2) { # 2x relaxed
+      priordens <- switch(j,
+                      0.3557*dnorm(x, -3.0675, 1.0637) + 0.6443*dnorm(x, 1.1290, 1.9685),
+                      dcauchy(x,-0.04051, 0.38660),
+                      dgamma(x, 3.4016, 9.5057),
+                      dgamma(x, 2.5210, 3.2948),
+                      dgamma(x, 4.2574, 14.370),
+                      dgamma(x,2,2))
+    } else if (i==3) { # 3x relaxed
+      priordens <- switch(j,
+                      0.3557*dnorm(x, -3.0675, 1.3027) + 0.6443*dnorm(x, 1.1290, 2.4110),
+                      dcauchy(x,-0.04051, 0.5799),
+                      dgamma(x, 2.7693, 7.0029),
+                      dgamma(x, 2.1457, 2.4819),
+                      dgamma(x, 3.3643, 10.430),
+                      dgamma(x,2,2))
+    } else if (i==4) { # 5x relaxed
+      priordens <- switch(j,
+                      0.3557*dnorm(x, -3.0675, 1.6818) + 0.6443*dnorm(x, 1.1290, 3.1125),
+                      dcauchy(x,-0.04051, 0.96650),
+                      dgamma(x, 2.2298, 4.8674),
+                      dgamma(x, 1.8166, 1.7689),
+                      dgamma(x, 2.6144, 7.1221),
+                      dgamma(x,2,2))
+    }
+    Pi <- P$density[which(P$density>0)]
+    Qi <- priordens[which(P$density>0)]
+    KL_temp[j] = sum(Pi * log(Pi/Qi))
+
+    # BCD
+    BC_temp[j] = -log(sum(sqrt(P$density*priordens)))
+
+    # EMD
+    A = cbind(P$density,P$mids)
+    B = cbind(priordens,P$mids)
+    EM_temp[j] = emd(A,B)
+
+    # code below is for plotting data vs priors
+    #Q <- P
+    #Q$density <- priordens
+    #Q$counts <- Q$density*866091*(Q$breaks[2]-Q$breaks[1])
+    #c1 <- rgb(173,216,230,max = 255, alpha = 80, names = "lt.blue")
+    #c2 <- rgb(255,192,203, max = 255, alpha = 80, names = "lt.pink")
+    #plot(P,col=c1, main=paste("i=",i,", j=",j))
+    #plot(Q,col=c2,add=TRUE)
   }
 
+  varname = switch(i, "1x", "2x", "3x", "5x")
+  n_eff_mat <- n_eff_mat %>% mutate(!!varname := n_eff_temp$mean)
+  Rhat_mat <- Rhat_mat %>% mutate(!!varname := Rhat_temp$mean)
+  KLdiv_mat <- KLdiv_mat %>% mutate(!!varname := KL_temp)
+  BCdist_mat <- BCdist_mat %>% mutate(!!varname := BC_temp)
+  EMdist_mat <- EMdist_mat %>% mutate(!!varname := EM_temp)
+
+}
+
+print("N_eff")
+kable(n_eff_mat, digits=2)
+print("Rhat")
+kable(Rhat_mat, digits=3)
+print("KLdiv")
+kable(KLdiv_mat, digits=3)
+cat("\newpage")
+print("BCdist")
+kable(BCdist_mat, digits=3)
+print("EMdist")
+kable(EMdist_mat, digits=3)
+
 ```

### 2020-10-02 loadSCRuns.R

@@ -10,7 +10,7 @@ library(reshape2)
 ### LOADING FOR STAN MODEL RESULTS
 
 # Set working directory to normal data wd then folder ResultsC
-setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data/ResultsTCGA_relaxgamma3")
+setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data/Stan Results Archive/FinalRuns/ResultsTCGA_gamma")
 
 mu_full <-
   data.frame(
@@ -54,125 +54,5 @@ for (i in filesPresent) {
   CpGscore_full[inds, ] <- CpGscore
 }
 save(mu_full, betaT_full, sigmaP_full, sigmaT_full, sigmaPT_full, sigmaE_full,
-      lp_full, CpGscore_full, file = "FullResultsTCGA_relaxgamma3.Rdata")
-
-
-### LOADING FOR GENE SCORING RESULTS
-# used to be for 3- and 24-space
-# now for 5 score methods on logPTratio only
-
-# Set working directory to normal data wd then folder GeneScores2
-setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data/GeneScores2")
-
-geneScores5 <- as.data.frame(matrix(ncol = 6, nrow = 27383))
-#geneScores27 <- as.data.frame(matrix(ncol = 27, nrow = 27383))
-#colnames(geneScores27) <- paste(rep(c("P/T","P/PT","PT/T"),8),c(rep("all",3),rep(regionTypes[1],3),rep(regionTypes[2],3),rep(regionTypes[3],3),rep(regionTypes[4],3),rep(regionTypes[5],3),rep(regionTypes[6],3),rep(regionTypes[7],3),rep(regionTypes[8],3)))
-#colnames(geneScores3) <- c("logP/T","logP/PT","logPT/T")
-
-nGenes <- 1000
-filesPresent <- (1:28)
-fileNames <- dir(pattern = "ScoreResults_*")
-allInds <- numeric()
-for (i in filesPresent) {
-  print(paste("Chunk:",i))
-  load(fileNames[fileNames == paste0("ScoreResults_",i,".Rdata")])
-  inds <- (1:nGenes) + (i - 1) * nGenes
[…]

### 2020-10-20 50sites files/50sites.Rmd
updated prior analysis for figures

@@
   indices <- c(10:57) # if using bulk only data file
-  patientLabel <- substr(colnames(temp[indices]),1,1)
+  patientLabel <- substr(colnames(temp[indices]), 1, 1)
   patientLabel[10:12] <- "K*"
-  sideLabel <- substr(colnames(temp[indices]),2,2)
+  sideLabel <- substr(colnames(temp[indices]), 2, 2)
   tissueLabel <- sideLabel
   tissueLabel[!(sideLabel %in% c("N"))] <- "T"   #replace A/B/D/M with T
-  tumorIndicator <- 1*(tissueLabel=="T")
+  tumorIndicator <- 1 * (tissueLabel=="T")
 
-  work <- data.frame(beta=logit(t(temp[indices])), patient = patientLabel, tissue=tissueLabel, side=sideLabel, tInd=tumorIndicator)
+  work <- data.frame(
+    beta = logit(t(temp[indices])),
+    patient = patientLabel,
+    tissue = tissueLabel,
+    side = sideLabel,
+    tInd = tumorIndicator
+    )
   colnames(work)[1] <- "beta"
@@
   stanDat <- list(
+    pID = as.integer(factor(dataset$patient)),
+    tInd = dataset$tInd,
     N = nrow(dataset),
     P = nlevels(dataset$patient),
-    pID = as.integer(factor(dataset$patient)),
-    tInd = dataset$tInd,
     y = dataset[, 1]
@@
 # Plot posterior distribution
-postplot <- function(stanFit) {
+postplot_mubetaT <- function(stanFit) {
+  p2 <- mcmc_areas(
+    as.array(stanFit), 
+    pars = c("mu","betaT"),
+    prob = 0.8, # 80% intervals
+    prob_outer = 0.95, 
+    point_est = "median"
+  )
+  return(p2)
+}
+postplot_sigmas <- function(stanFit) {
   p2 <- mcmc_areas(
@@
 # check for model_TCGApriors.rds, must add to prevent crash
-checkFile <- "model_relaxedgamma5.rds"
+checkFile <- "model_relaxedgamma3.rds"
 if (file.exists(checkFile)) {
@@
 )
-emptyFit <- stan(file="model_relaxedgamma5.stan", data = stanDat1, chains = 0)
+emptyFit <- stan(file="model_relaxedgamma3.stan", data = stanDat1, chains = 0)
 
+# inds for extracting data (mean, sem, median, n_eff, Rhat)
+mInd <- c(1, 2, 6, 9, 10)
 ```
@@
 ```{r echo=FALSE,include=TRUE,error=FALSE,message=FALSE,warning=FALSE,results='asis',fig.keep='all',fig.show='asis',fig.height=2.9,fig.width=5}
-# fig.align='center'
+
 ptm <- proc.time()
@@
   p1 <- fitplot(data, stanFit) + ggtitle(paste("Site", i, ": Beta Fits by Patient"))
-  p2 <- postplot(stanFit) + ggtitle("Sigma Posteriors")
+  p15 <- postplot_mubetaT(stanFit) + ggtitle("Mu/BetaT Posteriors")
+  p2 <- postplot_sigmas(stanFit) + ggtitle("Sigma Posteriors")
   p3 <- traceplot(stanFit,pars=c("sigma_p","sigma_pt","sigma_t","sigma_e"),inc_warmup=TRUE,ncol=1,window=c(0,1000)) + theme(axis.title.x=element_blank(),axis.text.x=element_blank(),axis.ticks.x=element_blank(),axis.text.y=element_text(angle=45),legend.position="none")
@@
   cat(paste("  \nSite ",i,": Parameter posterior summary\n",sep=''))
-  # print (or table) mean, median, R_hat, n_eff for each of the 6 params
-  datasumm <-summary(stanFit)$summary[1:6,c(1,6,9,10)]
-  print(kable(datasumm))
+
+  # take summary values for first 6 params (mu, betaT, sigmas x4) and last param (lp__)
+  fitSumm <- summary(stanFit)$summary[c(1:6,97), mInd]
+
+  # compute posterior log-ratio: log(sigmaP/sigmaT), take integral>0, mean, median
+  posterior <- as.matrix(stanFit,pars=c("sigma_p","sigma_t"))
+  logPTratio <- log(posterior[,1]/posterior[,2])
+  CpGscore <- c(mean(logPTratio > 0), mean(logPTratio), median(logPTratio))
+  fitSumm <- rbind(fitSumm,CpGscore)
+
+  # print kable of summary values
+  print(kable(fitSumm))
   cat("  \n")
-  n_eff_sum <- n_eff_sum + datasumm[,3]
+
+  n_eff_sum <- n_eff_sum + fitSumm[,3]
 }

### 2020-10-20 StanAnalysis_updated.Rmd
updated prior analysis for figures

@@
 load(here(stanFile))
+
+prepdatafolder = "data_final_rg3"
 ```
@@
   plot_df <- data.frame(val=postvals, gene_assoc=mask)
-  p1 <- ggplot(plot_df, aes(x=val)) +
-    geom_histogram(aes(y=..count../length(mask),fill=gene_assoc), binwidth=bw) +
-    stat_function(fun=function(x) bw*priorfun(x),geom="line",color="black") +
+  p1 <- ggplot(plot_df, aes(x=val,fill=gene_assoc)) +
+    #geom_histogram(aes(y=..count../length(mask)), binwidth=bw, position="dodge") +
+    #stat_function(fun=function(x) bw*priorfun(x),geom="line",color="black") +
+    geom_histogram(aes(y=..density..), binwidth=bw, position="dodge") +
+    stat_function(fun=function(x) priorfun(x),geom="line",color="black") +
     ggtitle(titlestr) + xlim(xl) + theme_classic() + theme(legend.position = "none") + xlab(titlestr) + ylab("Freq.")
@@
 # plot mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT with priors overlaid
-pM <- post_prior_histogram(mu_full$p50, genemask, function(x) 1.5*fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), 0.1, c(-5,5), "Mu")
-pB <- post_prior_histogram(betaT_full$p50, genemask, function(x) 1.7*dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "BetaT")
-pE <- post_prior_histogram(sigmaE_full$p50, genemask, function(x) 4*dgamma(x,2,2), 0.05, c(0,2), "SigmaE")
-pP <- post_prior_histogram(sigmaP_full$p50, genemask, function(x) 2*dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
-pPT <- post_prior_histogram(sigmaPT_full$p50, genemask, function(x) 1.8*dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
-pT <- post_prior_histogram(sigmaT_full$p50, genemask, function(x) 1.2*dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT")
+pM <- post_prior_histogram(mu_full$p50, genemask, function(x) 1.3*fBG(x,0.3557,-3.0675, 1.3027, 1.1290, 2.4110), 0.25, c(-5,5), "Mu")
+pB <- post_prior_histogram(betaT_full$p50, genemask, function(x) 1.5*dcauchy(x,-0.04051, 0.5799), 0.25, c(-5,5), "BetaT")
+pE <- post_prior_histogram(sigmaE_full$p50, genemask, function(x) 3.3*dgamma(x,2,2), 0.05, c(0,2), "SigmaE")
+pP <- post_prior_histogram(sigmaP_full$p50, genemask, function(x) 1.52*dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
+pPT <- post_prior_histogram(sigmaPT_full$p50, genemask, function(x) 1.6*dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
+pT <- post_prior_histogram(sigmaT_full$p50, genemask, function(x) 0.97*dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT")
 # log sigmas
-pLP <- post_prior_histogram(log(sigmaP_full$p50), genemask, function(x) 0.52*dgamma(exp(x), 2.7693, 7.0029), 0.1, c(-5,5), "log(SigmaP)")
-pLPT <- post_prior_histogram(log(sigmaPT_full$p50), genemask, function(x) 0.7*dgamma(exp(x), 2.1457, 2.4819), 0.1, c(-5,5), "log(SigmaPT)")
-pLT <- post_prior_histogram(log(sigmaT_full$p50), genemask, function(x) 0.4*dgamma(exp(x), 3.3643, 10.430), 0.1, c(-5,5), "log(SigmaT)") + theme(legend.position = c(0.8, 0.6))
+pLP <- post_prior_histogram(log(sigmaP_full$p50), genemask, function(x) 0.34*dgamma(exp(x), 2.7693, 7.0029), 0.25, c(-5,5), "log(SigmaP)")
+pLPT <- post_prior_histogram(log(sigmaPT_full$p50), genemask, function(x) 0.42*dgamma(exp(x), 2.1457, 2.4819), 0.25, c(-5,5), "log(SigmaPT)")
+pLT <- post_prior_histogram(log(sigmaT_full$p50), genemask, function(x) 0.22*dgamma(exp(x), 3.3643, 10.430), 0.25, c(-5,5), "log(SigmaT)") + theme(legend.position = c(0.8, 0.6))
 
@@
 print(p_all)
+# ggsave("posteriors_geneassoc.svg",p_all)
 rm(pM,pB,pE,pP,pPT,pT,pLP,pLPT,pLT)
@@
   p1 <- ggplot(plot_df, aes(x=val)) +
-    geom_histogram(aes(y=..count../length(mask),fill=gene_assoc), binwidth=bw) +
-    ggtitle(titlestr) + xlim(xl) + theme_classic() + theme(legend.position = "none") + xlab(titlestr) + ylab("Freq.")
+    #geom_histogram(aes(y=..count../length(mask),fill=gene_assoc), binwidth=bw, position="dodge") +
+    geom_histogram(aes(y=..density..,fill=gene_assoc), binwidth=bw, position="dodge") +
+    ggtitle(titlestr) + xlim(xl) + theme_classic() + theme(legend.position = "none") + xlab(titlestr) + ylab("Density")
+  p1
 }
@@
 # plot the two conservations score prob sums
-pP1 <- post_histogram(prob1_full$median,genemask,0.05,c(-3,3),expression("Median log("~sigma[P]/sigma[T]~")"))
+pP1 <- post_histogram(CpGscore_full$median,genemask,0.05,c(-3,3),expression("Median log("~sigma[P]/sigma[T]~")")) + scale_x_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3))
 prob1_proxy = log(sigmaP_full$p50/sigmaT_full$p50)
@@
 med_vals <- tibble(Name=FullAnnotation[,1], mu=mu_full$p50, betaT=betaT_full$p50, sigmaE=sigmaE_full$p50, sigmaP=sigmaP_full$p50, sigmaPT=sigmaPT_full$p50, sigmaT=sigmaT_full$p50, Rhat=lp_full$Rhat) %>% mutate(logP=log(sigmaP), logPT=log(sigmaPT), logT=log(sigmaT)) %>%
-  mutate(L2norm = sqrt(sigmaP^2 + sigmaT^2)) %>%
-  mutate(score_median=prob1_full$median, score_median2=log(sigmaP/sigmaT), score_reg=log(sigmaP/sigmaT * L2norm))
-# score_prob=prob1_full$prob, score_mean=prob1_full$mean,
+  mutate(score_median=CpGscore_full$median)
 
@@
   left_join(by="Name", med_vals) %>% filter(Rhat<Rhat_cutoff)
-write_csv(prepped_data, here("data-stan_rg3ps3", "prepped_data.csv"))
+write_csv(prepped_data, here(prepdatafolder, "prepped_data.csv"))
 cat(proc.time()-ptm)
@@
   distinct()
-write_csv(gened_data, here("data-stan_rg3ps3", "gened_data.csv"))
+write_csv(gened_data, here(prepdatafolder, "gened_data.csv"))
 cat(proc.time()-ptm)
@@
   filter(!is.na(UCSC_RefGene_Name)) %>%
-  gene_summarize(UCSC_RefGene_Name, c(mu:sigmaT,score_median:score_reg)) %>% filter(n>=n_cutoff)
+  gene_summarize(UCSC_RefGene_Name, c(mu:sigmaT,score_median)) %>% filter(n>=n_cutoff)
 cat(proc.time()-ptm)
@@
   boot_naive(gened_data, score_median, UCSC_RefGene_Name, 1000) %>%
-  boot_naive(gened_data, score_median2, UCSC_RefGene_Name, 1000) %>%
-  boot_naive(gened_data, score_reg, UCSC_RefGene_Name, 1000) %>%
   boot_adj(gened_data, sigmaP, UCSC_RefGene_Name, 1000) %>%
@@
   boot_adj(gened_data, sigmaT, UCSC_RefGene_Name, 1000) %>%
-  boot_adj(gened_data, score_median, UCSC_RefGene_Name, 1000) %>%
-  boot_adj(gened_data, score_median2, UCSC_RefGene_Name, 1000) %>%
-  boot_adj(gened_data, score_reg, UCSC_RefGene_Name, 1000)
-write_csv(permed_data, here("data-stan_rg3ps3", "permed_data.csv"))
+  boot_adj(gened_data, score_median, UCSC_RefGene_Name, 1000)
+write_csv(permed_data, here(prepdatafolder, "permed_data.csv"))
 cat(proc.time()-ptm)
-#  boot_naive(gened_data, score_prob, UCSC_RefGene_Name, 1000) %>%
-#  boot_naive(gened_data, score_mean, UCSC_RefGene_Name, 1000) %>%
 
-# look at mean methylation conservation in promoters
-cat("Calculate mean methylation conservation in promoters")
-ptm<-proc.time()
-promoted_data <- prepped_data %>%
-  mutate(gene = str_extract(UCSC_RefGene_Name, "^[^;]+")) %>%
-  filter(!is.na(gene)) %>%
-  left_join(read_csv(here("data-raw","MethylationEPIC_v-1-0_B4.csv"), skip = 7) %>%  
-              select(Name, Regulatory_Feature_Group), 
-            by = "Name") %>%
-  filter(Regulatory_Feature_Group == "Promoter_Associated") %>%
-  group_by(gene) %>%
-  summarise(sigmaP_pro = mean(sigmaP), sigmaPT_pro = mean(sigmaPT), sigmaT_pro = mean(sigmaT), 
-            score_median_pro = mean(score_median), score_median2_pro = mean(score_median2), score_reg_pro = mean(score_reg))
-write_csv(promoted_data, here("data-stan_rg3ps3", "promoted_data.csv"))
-cat(proc.time()-ptm)
-cat("Done with data preparation (BOTH)")
+cat("Done with data preparation")
 ```
@@
 cat("Param:prep=FALSE -> load saved data (BOTH)...")
-gened_data <- read_csv(here("data-stan_rg3ps3", "gened_data.csv"), col_types = list("CHR" = col_character()))
-prepped_data <- read_csv(here("data-stan_rg3ps3", "prepped_data.csv"))
-permed_data <- read_csv(here("data-stan_rg3ps3", "permed_data.csv"))
-promoted_data <- read_csv(here("data-stan_rg3ps3", "promoted_data.csv"))
+gened_data <- read_csv(here(prepdatafolder, "gened_data.csv"), col_types = list("CHR" = col_character()))
+prepped_data <- read_csv(here(prepdatafolder, "prepped_data.csv"))
+permed_data <- read_csv(here(prepdatafolder, "permed_data.csv"))
+promoted_data <- read_csv(here(prepdatafolder, "promoted_data.csv"))
 cat("done")
@@
   group_by(`Gene Associated` = !is.na(UCSC_RefGene_Name)) %>%
-  summarise_at(vars(sigmaP:sigmaT,score_median:score_reg), mean, na.rm = TRUE) %>%
+  summarise_at(vars(sigmaP:sigmaT,score_median), mean, na.rm = TRUE) %>%
   mutate(`Gene Associated` = factor(`Gene Associated`, c(TRUE, FALSE), c("Yes", "No"))) %>%
@@
   gt() %>%
-  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, score_median, score_median2, score_reg), decimals = 3) %>%
+  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, score_median), decimals = 3) %>%
   tab_spanner(label = "Hierarchical Variance", columns = vars(sigmaP, sigmaPT, sigmaT)) %>%
-  tab_spanner(label = "Conservation Scores", columns = vars(score_median, score_median2, score_reg))
-#score_prob, score_mean,
+  tab_spanner(label = "Conservation Score", columns = vars(score_median))
 
@@
   group_by(`Island relation` = Relation_to_UCSC_CpG_Island) %>%
-  summarise_at(vars(sigmaP:sigmaT,score_median:score_reg), mean, na.rm = TRUE) %>%
+  summarise_at(vars(sigmaP:sigmaT,score_median), mean, na.rm = TRUE) %>%
   dplyr::rename(ir = `Island relation`) %>%
@@
   gt() %>%
-  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, score_median, score_median2, score_reg), decimals = 3) %>%
+  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, score_median), decimals = 3) %>%
   tab_spanner(label = "Hierarchical Variance", columns = vars(sigmaP, sigmaPT, sigmaT)) %>%
-  tab_spanner(label = "Conservation Scores", columns = vars(score_median, score_median2, score_reg))
+  tab_spanner(label = "Conservation Score", columns = vars(score_median))
 
@@
   group_by(`5'UTR` = str_detect(UCSC_RefGene_Group, "5'UTR") & !is.na(UCSC_RefGene_Group)) %>%
-  summarise_at(vars(sigmaP:sigmaT,score_median:score_reg), mean, na.rm = TRUE) %>%
+  summarise_at(vars(sigmaP:sigmaT,score_median), mean, na.rm = TRUE) %>%
   mutate(`5'UTR` = factor(`5'UTR`, c(TRUE, FALSE), c("Yes", "No"))) %>%
@@
   group_by(`TSS1500` = str_detect(UCSC_RefGene_Group, "TSS1500") & !is.na(UCSC_RefGene_Group)) %>%
-  summarise_at(vars(sigmaP:sigmaT,score_median:score_reg), mean, na.rm = TRUE) %>%
+  summarise_at(vars(sigmaP:sigmaT,score_median), mean, na.rm = TRUE) %>%
   mutate(`TSS1500` = factor(`TSS1500`, c(TRUE, FALSE), c("Yes", "No"))) %>%
@@
   group_by(`TSS200` = str_detect(UCSC_RefGene_Group, "TSS200") & !is.na(UCSC_RefGene_Group)) %>%
-  summarise_at(vars(sigmaP:sigmaT,score_median:score_reg), mean, na.rm = TRUE) %>%
+  summarise_at(vars(sigmaP:sigmaT,score_median), mean, na.rm = TRUE) %>%
   mutate(`TSS200` = factor(`TSS200`, c(TRUE, FALSE), c("Yes", "No"))) %>%
@@
   gt() %>%
-  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, score_median, score_median2, score_reg), decimals = 3) %>%
+  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, score_median), decimals = 3) %>%
   tab_spanner(label = "Hierarchical Variance", columns = vars(sigmaP, sigmaPT, sigmaT)) %>%
-  tab_spanner(label = "Conservation Scores", columns = vars(score_median, score_median2, score_reg))
+  tab_spanner(label = "Conservation Scores", columns = vars(score_median))
+
+# plot CpG island relation score distributions
+prepped_data %>% select(Name, Relation_to_UCSC_CpG_Island, score_median) %>%
+  dplyr::rename(ir = Relation_to_UCSC_CpG_Island, score=score_median) %>%
+  mutate(ir = case_when(ir=="" ~ "Sea",
+                        str_detect(ir, "^S") ~ str_replace(ir, "^S_", ""),
+                        str_detect(ir, "^N") ~ str_replace(ir, "^N_", ""),
+                        TRUE ~ ir)) %>%
+  mutate(ir = factor(ir, levels = c("Island", "Shore", "Shelf", "Sea"))) %>%
+  #ggplot() + geom_density(aes(x=score,y=..scaled..,color=ir)) + theme_classic() + scale_x_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3)) + ylab("Normalized Density")
+  ggplot() + geom_violin(aes(x=ir,y=score,fill=ir),draw_quantiles = 0.5) + theme_classic() + scale_y_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3))
+# ggsave("CpGscore_islands.svg")
+
+# plot functional region score distributions
+temp <- prepped_data %>% select(Name, UCSC_RefGene_Group, score_median) %>%
+  separate_rows(UCSC_RefGene_Group, sep=";") %>%
+  distinct() %>%
+  dplyr::rename(fr = UCSC_RefGene_Group, score=score_median) %>%
+  mutate(fr = case_when(fr=="" ~ "NA",
+                        str_detect(fr, "5URT") ~ "5'UTR",
+                        str_detect(fr, "3UTR") ~ "3'UTR",
+                        TRUE ~ fr)) %>%
+  distinct()
+keepregions = c("TSS200","5'UTR","Body","3'UTR")
+# temp %>%
+#   filter(fr %in% keepregions) %>%
+#   #mutate(fr = factor(fr, levels = c("Island", "Shore", "Shelf", "Sea"))) %>%
+#   ggplot() + geom_density(aes(x=score,color=fr)) + theme_classic() + scale_x_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3)) + ylab("Normalized Density")
+temp %>%
+  mutate(fr = factor(fr, levels = keepregions)) %>%
+  filter(fr %in% keepregions) %>%
+  ggplot() + geom_violin(aes(x=fr,y=score,fill=fr),draw_quantiles = 0.5) + theme_classic() + scale_y_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3))
+# ggsave("CpGscore_regions_violin.svg")
+rm(temp)
 ```

### 2020-10-20 differential_methylation.Rmd
updated prior analysis for figures

@@
 # convert to M values, tInd
-all_M <- logit(all_beta)
+all_M <- logit(all_beta) # base e
+#all_M <- log2(all_beta/(1-all_beta)) # base 2
 tInd <- tissueLabel=="T"
@@
   mutate(qvalBHz = p.adjust(pz, method="BH"))
-
 ```
@@
 ```{r compare_irizarry, echo=FALSE}
-DMR_data = fit_stats$qvalBH <0.05
-DMR_beta = betaT_stats$qvalBH < 0.05
+DMR_data = (fit_stats$qvalBH < 0.05) & (abs(fit_stats$coefficients.deltaM) > 1.4)
+DMR_beta = (betaT_stats$qvalBH < 0.05) & (abs(betaT_stats$p50) > 1.4)
 CDMR_epic = (FullAnnotation$DMR == "CDMR") #| (FullAnnotation$DMR == "DMR")
 
-DMR_list = cbind(DMR_data, CDMR_epic)
-crosstab <- xtabs(~DMR_data + CDMR_epic, DMR_list)
-crosstab
+cor(fit_stats$tstat.ord,betaT_stats$tstat)
+cor(fit_stats$coefficients.deltaM,betaT_stats$p50)
 
-DMR_list = cbind(DMR_beta, CDMR_epic)
-crosstab <- xtabs(~DMR_beta + CDMR_epic, DMR_list)
-crosstab
+# DMR_list = cbind(DMR_data, CDMR_epic)
+# crosstab <- xtabs(~DMR_data + CDMR_epic, DMR_list)
+# crosstab
+
+# DMR_list = cbind(DMR_beta, CDMR_epic)
+# crosstab <- xtabs(~DMR_beta + CDMR_epic, DMR_list)
+# crosstab
 
@@
 phi = (prod(c(crosstab[1,1],crosstab[2,2]))-prod(crosstab[1,2],crosstab[2,1]))/sqrt(prod(rowSums(crosstab))*prod(rowSums(t(crosstab))))
+
+# a few correlations between linear and betaT methods
+cor(fit_stats$tstat.ord,betaT_stats$tstat)
+cor(fit_stats$coefficients.deltaM,betaT_stats$p50)
+
 ```

### 2020-10-20 prior_sensitivity.Rmd
updated prior analysis for figures

@@
 library(emdist)
+library(ggpubr)
 
@@
 slice <- dplyr::slice
+```
+
+```{r prior_dataAndFits, echo=FALSE, comment=NA}
+# load TCGA-based 450K array estimates (see estimateTCGApriors.R for data+fits)
+load("/Users/kevinmurgas/Documents/Data+ project/TCGA data/estimates_allpats.Rdata")
+
+# remove all zeros, and take square root of variance (stored variable) to get sigma/stdev
+sigmaP_ests <- sqrt(sigmaP_ests[which(!!sigmaP_ests)])
+sigmaPT_ests <- sqrt(sigmaPT_ests[which(!!sigmaPT_ests)])
+sigmaT_ests <- sqrt(sigmaT_ests[which(!!sigmaT_ests)])
+mu_ests <- mu_ests[which(!!mu_ests)]
+betaT_ests <- betaT_ests[which(!!betaT_ests)]
+
+gg_color_hue <- function(n) {
+  hues = seq(15, 375, length = n + 1)
+  hcl(h = hues, l = 65, c = 100)[1:n]
+}
+cols = gg_color_hue(4)
+
+# ggplot histograms, with fit overlays
+# mu
+# 1x: 0.35570*normal(-3.0675,0.75214) + 0.64430*normal(1.1290, 1.3920)
+# 2x: 0.35570*normal(-3.0675, 1.0637) + 0.64430*normal(1.1290, 1.9685)
+# 3x: 0.35570*normal(-3.0675, 1.3027) + 0.64430*normal(1.1290, 2.4110)
+# 5x: 0.35570*normal(-3.0675, 1.6818) + 0.64430*normal(1.1290, 3.1125)
+f <- function(x, a, m1, s1, m2, s2) a*dnorm(x,m1,s1) + (1-a)*dnorm(x,m2,s2)
+p1 <- ggplot(data.frame(p=mu_ests), aes(x=p)) + geom_histogram(binwidth=0.1,aes(y=..density..),fill="lightgray") + theme_classic() +
+  #labs(x=expression(mu)) +
+  xlim(-6,6) +
+  stat_function(fun=function(x) f(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197),geom="line",color=cols[1]) +
+  stat_function(fun=function(x) f(x,0.3557,-3.0675, 1.0637,1.1290, 1.9685),geom="line",color=cols[2]) +
+  stat_function(fun=function(x) f(x,0.3557,-3.0675, 1.3027,1.1290, 2.4110),geom="line",color=cols[3]) +
+  stat_function(fun=function(x) f(x,0.3557,-3.0675, 1.6818,1.1290, 3.1125),geom="line",color=cols[4])
+# betaT
+# 1x: Cauchy(-0.04051, 0.19330)
+# 2x: Cauchy(-0.04051, 0.38660)
+# 3x: Cauchy(-0.04051, 0.57990)
+# 5x: Cauchy(-0.04051, 0.96650)
+p2 <- ggplot(data.frame(p=betaT_ests), aes(x=p)) + geom_histogram(binwidth=0.1,aes(y=..density..),fill="lightgray") + theme_classic() +
+  #labs(x=expression(nu)) +
+  xlim(-5,5) +
+  stat_function(fun=function(x) dcauchy(x,-0.04051, 0.19330),geom="line",color=cols[1]) +
+  stat_function(fun=function(x) dcauchy(x,-0.04051, 0.38660),geom="line",color=cols[2]) +
+  stat_function(fun=function(x) dcauchy(x,-0.04051, 0.57990),geom="line",color=cols[3]) +
+  stat_function(fun=function(x) dcauchy(x,-0.04051, 0.96650),geom="line",color=cols[4])
+# sigmaE: all gamma(2,2)
+p3 <- ggplot() + stat_function(fun=function(x) dgamma(x,2,2),geom="line",color="black") + xlim(0,2) + theme_classic() + labs(x=expression(sigma[E]))
+# sigmaP
+# 1x: gamma(5.1989, 16.6192)
+# 2x: gamma(3.4016, 9.5057)
+# 3x: gamma(2.7693, 7.0029)
+# 5x: gamma(2.2298, 4.8674)
+p4 <- ggplot(data.frame(p=sigmaP_ests), aes(x=p)) + geom_histogram(binwidth=0.02,aes(y=..density..),fill="lightgray") + theme_classic() +
+  #labs(x=expression(sigma[P])) +
+  xlim(0,2) +
+  stat_function(fun=function(x) dgamma(x,5.1989, 16.6192),geom="line",color=cols[1]) +
+  stat_function(fun=function(x) dgamma(x,3.4016, 9.5057),geom="line",color=cols[2]) +
+  stat_function(fun=function(x) dgamma(x,2.7693, 7.0029),geom="line",color=cols[3]) +
+  stat_function(fun=function(x) dgamma(x,2.2298, 4.8674),geom="line",color=cols[4])
+# sigmaPT
+# 1x: gamma(3.5539, 5.5324)
+# 2x: gamma(2.5210, 3.2948)
+# 3x: gamma(2.1457, 2.4819)
+# 5x: gamma(1.8166, 1.7689)
+p5 <- ggplot(data.frame(p=sigmaPT_ests), aes(x=p)) + geom_histogram(binwidth=0.02,aes(y=..density..),fill="lightgray") + theme_classic() +
+  #labs(x=expression(sigma[PT])) +
+  xlim(0,2) +
+  stat_function(fun=function(x) dgamma(x,3.5539, 5.5324),geom="line",color=cols[1]) +
+  stat_function(fun=function(x) dgamma(x,2.5210, 3.2948),geom="line",color=cols[2]) +
+  stat_function(fun=function(x) dgamma(x,2.1457, 2.4819),geom="line",color=cols[3]) +
+  stat_function(fun=function(x) dgamma(x,1.8166, 1.7689),geom="line",color=cols[4])
+# sigmaT
+# 1x: gamma(6.8383, 25.7560)
+# 2x: gamma(4.2574, 14.370)
+# 3x: gamma(3.3643, 10.430)
+# 5x: gamma(2.6144, 7.1221)
+p6 <- ggplot(data.frame(p=sigmaT_ests), aes(x=p)) + geom_histogram(binwidth=0.02,aes(y=..density..),fill="lightgray") + theme_classic() +
+  xlim(0,2) +
+  #labs(x=expression(sigma[T])) +
+  stat_function(fun=function(x) dgamma(x,6.8383, 25.7560),geom="line",color=cols[1]) +
+  stat_function(fun=function(x) dgamma(x,4.2574, 14.370),geom="line",color=cols[2]) +
+  stat_function(fun=function(x) dgamma(x,3.3643, 10.430),geom="line",color=cols[3]) +
+  stat_function(fun=function(x) dgamma(x,2.6144, 7.1221),geom="line",color=cols[4])
+
+p_all <- ggarrange(p1,p2,p3,p4,p5,p6)
+# ggsave("prior_fits_all.svg",p_all,height=3,width=5.53,dpi=300)
 ```
@@
     # BCD
-    BC_temp[j] = -log(sum(sqrt(P$density*priordens)))
+    dx <- x[2]-x[1]
+    BC_temp[j] = -log(sum(sqrt(P$density*priordens)*dx))
 
@@
 kable(KLdiv_mat, digits=3)
-cat("\newpage")
+cat("\n")
 print("BCdist")
@@
 ```
+
+\newpage
+Plots
+```{r plots, echo=FALSE, comment=NA}
+# goal here is to just get the distributions of parameters for each prior set
+# store histogram density, and KDE
+# plot histogram with side-by-side bar plots
+# plot KDE's as line plots
+
+mu_allprior <- data.frame(x1=numeric(866091),
+                          x2=numeric(866091),
+                          x3=numeric(866091),
+                          x5=numeric(866091))
+betaT_allprior <- mu_allprior
+sigmaP_allprior <- mu_allprior
+sigmaPT_allprior <- mu_allprior
+sigmaT_allprior <- mu_allprior
+sigmaE_allprior <- mu_allprior
+
+for (i in c(1:4)) {
+  stanFile = "FullResultsTCGA_relaxgamma3.Rdata"
+  stanFile = switch(i,
+                    "FullResultsTCGA_gamma.Rdata",
+                    "FullResultsTCGA_relaxgamma2.Rdata",
+                    "FullResultsTCGA_relaxgamma3.Rdata",
+                    "FullResultsTCGA_relaxgamma5.Rdata")
+  print(stanFile)
+  load(here(stanFile))
+  print("loaded")
+
+  mu_allprior[,i] <- mu_full$p50
+  betaT_allprior[,i] <- betaT_full$p50
+  sigmaP_allprior[,i] <- sigmaP_full$p50
+  sigmaPT_allprior[,i] <- sigmaPT_full$p50
+  sigmaT_allprior[,i] <- sigmaT_full$p50
+  sigmaE_allprior[,i] <- sigmaE_full$p50
+}
+
+cat("Posterior estimates for varying prior relaxation")
+pM <- mu_allprior %>% gather("relax","val",1:4) %>% ggplot(aes(x=val,y=..density..,color=relax)) + geom_density() + xlim(-6,6) + ggtitle("Mu") + theme_classic() + theme(legend.position = "none")
+pB <- betaT_allprior %>% gather("relax","val",1:4) %>% ggplot(aes(x=val,y=..density..,color=relax)) + geom_density() + xlim(-5,5) + ggtitle("BetaT") + theme_classic() + theme(legend.position = "none")
+pP <- sigmaP_allprior %>% gather("relax","val",1:4) %>% ggplot(aes(x=val,y=..density..,color=relax)) + geom_density() + xlim(0,2) + ggtitle("sigmaP") + theme_classic() + theme(legend.position = "none")
+pPT <- sigmaPT_allprior %>% gather("relax","val",1:4) %>% ggplot(aes(x=val,y=..density..,color=relax)) + geom_density() + xlim(0,2) + ggtitle("sigmaPT") + theme_classic() + theme(legend.position = "none")
+pT <- sigmaT_allprior %>% gather("relax","val",1:4) %>% ggplot(aes(x=val,y=..density..,color=relax)) + geom_density() + xlim(0,2) + ggtitle("sigmaT") + theme_classic() + theme(legend.position = "none")
+pE <- sigmaE_allprior %>% gather("relax","val",1:4) %>% ggplot(aes(x=val,y=..density..,color=relax)) + geom_density() + xlim(0,2) + ggtitle("sigmaE") + theme_classic() + theme(legend.position = "none")
+pLP <- sigmaP_allprior %>% gather("relax","val",1:4) %>% mutate(val=log(val)) %>% ggplot(aes(x=val,y=..density..,color=relax)) + geom_density() + xlim(-4,1) + ggtitle("log(sigmaP)") + theme_classic() + theme(legend.position = "none")
+pLPT <- sigmaPT_allprior %>% gather("relax","val",1:4) %>% mutate(val=log(val)) %>% ggplot(aes(x=val,y=..density..,color=relax)) + geom_density() + xlim(-4,1) + ggtitle("log(sigmaPT)") + theme_classic() + theme(legend.position = "none")
+pLT <- sigmaT_allprior %>% gather("relax","val",1:4) %>% mutate(val=log(val)) %>% ggplot(aes(x=val,y=..density..,color=relax)) + geom_density() + xlim(-4,1) + ggtitle("log(sigmaT)") + theme_classic() + theme(legend.position = "none")
+
+p_all<- ggarrange(pM, pB, pE, pP, pPT, pT, pLP, pLPT, pLT, nrow=3, ncol=3)
+print(p_all)
+# ggsave("prior_sens_kdes.svg",p1)
+```

### 2020-10-20 estimateTCGApriors.R

@@ -307,7 +307,7 @@ betaT_ests <- betaT_ests[which(!!betaT_ests)]
 
 # examine histograms of estimates
 hist(sigmaP_ests,breaks=seq(0,ceiling(max(sigmaP_ests)),0.02),xlim=c(0,2),main="sigmaP, all patients")
-hist(sigmaPT_ests,breaks=seq(-3.2,ceiling(max(sigmaPT_ests)),0.02),xlim=c(-1.5,3),main="sigmaPT, all patients")
+hist(sigmaPT_ests,breaks=seq(-3.2,ceiling(max(sigmaPT_ests)),0.02),xlim=c(0,2),main="sigmaPT, all patients")
 hist(sigmaT_ests,breaks=seq(0,ceiling(max(sigmaT_ests)),0.02),xlim=c(0,2),main="sigmaT, all patients")
 hist(mu_ests,breaks=100,xlim=c(-6,6),main="mu, all patients")
 hist(betaT_ests,breaks=100,xlim=c(-5,5),main="betaT, all patients")

### 2020-11-24 Paper_Figures.Rmd
Edits to figures

@@
 library(tidyverse)
+library(ggpubr)
+
 library(gt)
 library(rlang)
-library(ggpubr)
 library(knitr)
+library(gtools)
 
@@
 
-stanFile = "FullResultsTCGA_relaxgamma3_postsum3.Rdata"
+stanFile = "FullResultsTCGA_relaxgamma3.Rdata"
 load(here(stanFile))
 
-prepdir <- "data-stan_rg3ps3"
+prepdatafolder = "data_final_rg3"
 cat(sprintf("Load prepped data dir: %s (needs to be pre-processed)...",prepdir))
-gened_data <- read_csv(here(prepdir, "gened_data.csv"), col_types = list("CHR" = col_character()))
-prepped_data <- read_csv(here(prepdir, "prepped_data.csv"))
-permed_data <- read_csv(here(prepdir, "permed_data.csv"))
-
-```
-
-\newpage
-\section{Figure 1: Model Overview, Priors, Example Site Fit}
+gened_data <- read_csv(here(prepdatafolder, "gened_data.csv"), col_types = list("CHR" = col_character()))
+prepped_data <- read_csv(here(prepdatafolder, "prepped_data.csv"))
+permed_data <- read_csv(here(prepdatafolder, "permed_data.csv"))
+
+# # plot beta/M values
+# all_beta = FullAnnotation[,-c(1,2,3,4,5,6,7,8,9)]
+# mean_beta = rowMeans(all_beta)
+# all_M = logit(all_beta)
+# mean_M = rowMeans(all_M)
+# p_beta <- data.frame(value=mean_beta) %>% gather %>% ggplot() + geom_histogram(aes(x=value),binwidth=0.01,fill="lightgray",color="grey50") + theme_classic() + xlim(c(0,1)) + theme(aspect.ratio = 1) + xlab("Beta value")
+# p_M <- data.frame(value=mean_M) %>% gather %>% ggplot() + geom_histogram(aes(x=value),binwidth=0.1,fill="lightgray",color="grey50") + theme_classic() + xlim(c(-5,5)) + theme(aspect.ratio = 1) + xlab("M value")
+# p_BM <- ggarrange(p_beta, p_M, nrow = 1, ncol = 2)
+# # ggsave("betaMdists.svg",p_BM)
+```
+
+\newpage
+\section{Model Overview, Priors, Example Site Fit}
+
 \subsection{Data}
@@
 By tissue: 6 normal samples, 42 paired bulk tumor samples.
+
 \subsection{Model equation}
-Let i=sample, j=patient
-$$ y_i = \mu + a_j +  \text{tInd}(i)*(\beta_T + b_j + c_i) $$
-$$ a_j \sim N(0,\sigma_P) , b_j \sim N(0,\sigma_{PT}) , c_i \sim N(0,\sigma_T) $$
+Let i=patient, j=sample (0=normal, 1,2=tumor sample 1,2)
+$$ y_{ij} = \mu + \alpha_i +  (1-\delta_{j0})*(\nu + \beta_i + \gamma_{ij}) + \epsilon_{ij} $$
+$$ \alpha_i \sim N(0,\sigma_P) , \beta_i \sim N(0,\sigma_{PT}) , \gamma_{ij} \sim N(0,\sigma_T) $$
 \newline Model is fit at each individual CpG site using RStan's MCMC sampling algorithm. \newline
@@
 \subsection{Prior Probabilities}
-Priors for each parameter (mu, betaT, sigmaP/PT/T) were defined using TCGA 450k array data from colorectal tumor patients with paired multiple tumor and/or normal+tumor samples.
+Priors for each parameter (mu, nu, sigmaP/PT/T) were defined using TCGA 450k array data from colorectal tumor patients with paired multiple tumor and/or normal+tumor samples.
 \newline Empirical values of mean and standard deviations were assesed for all 450k CpG sites.
@@
 \newline
-SigmaE (error variance) prior was defined using a non-informative gamma distribution
-\newline Additionally: Prior relaxation was performed for the Cauchy and gamma distribtions by algebraically increasing the variance of the distributions while maintaining the same mode (peak)
+SigmaE (error variance) prior was defined using a non-informative gamma distribution: gamma(2,2).
+\newline Additionally: Prior relaxation was performed by algebraically increasing the variance of the distributions (or each Gaussian component in the case of mu) while maintaining the same mode (peak).
 
@@
 pM <- post_prior_histogram(mu_ests, function(x) fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), function(x) fBG(x,0.3557,-3.0675, sqrt(3)*0.75214,1.1290, sqrt(3)*1.39197), 0.1, c(-5,5), "Mu")
-pB <- post_prior_histogram(betaT_ests, function(x) dcauchy(x,-0.04051, 0.19330), function(x) dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "BetaT")
+pB <- post_prior_histogram(betaT_ests, function(x) dcauchy(x,-0.04051, 0.19330), function(x) dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "Nu")
 pP <- post_prior_histogram(sigmaP_ests, function(x) dgamma(x, 5.1989, 16.6192), function(x) dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
@@
 lower variation in methlyation within tumor (sigmaT) relative to the variation within normal healthy tissue (sigmaP).
-Therefore we choose the log-ratio of sigmaP/sigmaT to represent a conservation score, which is positive when tumor variation is lower than normal variation.
-We take the median of the posterior distribution for this parameter.
-
-$$\text{score}_\text{med} = \text{Median} (\text{log}\frac{\sigma_{P}}{\sigma_T}) $$
-
-\newpage
-\subsection{Example site fit}
-![Fit](/Users/kevinmurgas/Documents/Data+ project/Paper Figures/example_fit.png)
+Therefore we choose the log-ratio (base 2) of sigmaP/sigmaT to represent a relative conservation score, which is positive when tumor variation is lower than normal variation.
+We take the median of the posterior distribution constructed from the two sigmaP,T posteriors for this score.
+
+$$\text{score}_\text{med} = \text{Median} (\text{log}_2\frac{\sigma_{P}}{\sigma_T}) $$
+
+\newpage
+\subsection{Figure 2A: Example site fit, 2B: Posteriors}
+![Fit](/Users/kevinmurgas/Documents/Data+ project/Manuscript/example_fit.png)
 \newline \newline
-![Posteriors](/Users/kevinmurgas/Documents/Data+ project/Paper Figures/example_posteriors.png)
+![Posteriors](/Users/kevinmurgas/Documents/Data+ project/Manuscript/example_posteriors.png)
 \newpage
@@
 
+# overlaid transparent histograms
+post_histogram <- function(postvals,mask,bw,titlestr) {
+  # stacked histogram colored by gene-associated sites
+  # and prior curve overlaid
+  plot_df <- data.frame(val=postvals, gene_assoc=mask)
+  gene_assoc_df <- plot_df %>% filter(gene_assoc)
+  not_gene_assoc_df <- plot_df %>% filter(!gene_assoc)
+  p1 <- ggplot(plot_df, aes(x=val)) +
+    geom_histogram(data=not_gene_assoc_df, aes(y=..density..,fill="#00BFC4",alpha=0.4), binwidth=bw) +
+    geom_histogram(data=gene_assoc_df, aes(y=..density..,fill="#F8766D",alpha=0.4), binwidth=bw) +
+    ggtitle(titlestr) + xlab(titlestr) + ylab("Density") + theme_classic() 
+  p1
+}
+
 # plot the two conservations score prob sums
-pP1 <- post_histogram(prob1_full$median,genemask,0.05,c(-3,3),expression("Median log("~sigma[P]/sigma[T]~")"))
-#prob1_proxy = log(sigmaP_full$p50/sigmaT_full$p50)
-#pP2 <- post_histogram(prob1_proxy,genemask,0.05,c(-3,3),expression(" log( median" ~sigma[P]/~"median"~sigma[T]~")"))
-p_scores <- pP1
+genemask = !(FullAnnotation$UCSC_RefGene_Name=="")
+p_scores <- post_histogram(CpGscore_full$median,genemask,0.05,"CpG Conservation score") +
+  theme(aspect.ratio=1/4,legend.position = "none") +
+  scale_x_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3))
 print(p_scores)
+# ggsave("CpGscores_overlay.svg")
+
+# plot gene scores and adjusted bootstrap distribution
+p1 <- permed_data %>% ggplot(aes(x=score_median)) + geom_histogram(aes(y=..count..), binwidth=0.05, fill="lightgray",color = "grey50") + xlim(c(-1,2.5)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
+p2 <- permed_data %>% ggplot(aes(x=score_median_v3)) + geom_histogram(aes(y=..count..), binwidth=0.02, fill="lightgray", color="grey50") + xlim(c(0,1)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
+p_genes <- ggarrange(p1,p2,nrows=2,ncol=1)
+# ggsave("gene_scores_bootstrap.svg")
 ```
@@
 
+# stats for cpg island relations
+# summarize mean of each group, do kruskal wallis and post hoc Tukey
+my_data = prepped_data %>% select(score_median, Relation_to_UCSC_CpG_Island) %>%
+  group_by(ir = Relation_to_UCSC_CpG_Island) %>%
+  mutate(ir = case_when(is.na(ir) ~ "Sea",
+                        str_detect(ir, "^S") ~ str_replace(ir, "^S_", ""),
+                        str_detect(ir, "^N") ~ str_replace(ir, "^N_", ""),
+                        TRUE ~ ir))
+my_data %>%
+  summarise_at(vars(score_median), mean, na.rm = TRUE)
+kruskal.test(score_median ~ ir, data = my_data)
+#install.packages("MultNonParam")
+library(MultNonParam)
+tukey.kruskal.test(my_data$score_median, my_data$ir, alpha = 0.001)
+
+# stats for functional regions
 summarize_conservation <- function(data,str) {
@@
           {prepped_data %>% summarize_conservation("Body") %>% mutate(region="Body")},
-          {prepped_data %>% summarize_conservation("3'UTR|5UTR") %>% mutate(region="3'UTR")}) %>%
+          {prepped_data %>% summarize_conservation("3'UTR|3UTR") %>% mutate(region="3'UTR")}) %>%
   select(c(region,cons_in,cons_out,nsites,pval)) %>%
@@
   fmt_number(columns = vars(cons_in,cons_out,pval), decimals = 3)
+
+# new stats
+FRdata = data.frame(CpGscore = log2(prepped_data$sigmaP / prepped_data$sigmaT) ) %>%
+  mutate(FR = pull(prepped_data,UCSC_RefGene_Group)) %>%
+  mutate(FR=if_else(is.na(FR),"NA",FR)) %>%
+  mutate(iTSS200 = str_detect(FR,"TSS200"),
+         i5UTR = str_detect(FR,"5'UTR|5UTR|5URT"),
+         iBody = str_detect(FR,"Body"),
+         i3UTR = str_detect(FR,"3'UTR|3UTR"),
+         iTSS1500 = str_detect(FR,"TSS1500"),
+         i1stEx = str_detect(FR,"1stExon"),
+         iExBnd = str_detect(FR,"ExonBnd"))
+fr.lm = lm(formula = CpGscore ~ iTSS200 + i5UTR + iBody + i3UTR + iTSS1500 + i1stEx + iExBnd, data = FRdata)
+#fr.lm = lm(formula = CpGscore ~ iTSS200 + i5UTR + iBody + i3UTR, data = FRdata)
+summary(fr.lm)
+
+library(rcompanion)
+fr.srh = scheirerRayHare(formula = CpGscore ~ iTSS200 + i5UTR + iBody + i3UTR + iTSS1500 + i1stEx + iExBnd, data = FRdata)
 ```
@@
 
-plotGeneScore <- function(geneStr) {
+plotGeneScore <- function(geneStr,norm) {
   CpGscores <- gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^",geneStr,"$"))) %>% pull(score_median)
 
-  ggplot(prepped_data) + geom_histogram(aes(x=score_median, y=..count../nrow(prepped_data)), binwidth=0.05) + xlim(-3,3) + ggtitle(sprintf("%s: CpG scores (%i sites)",geneStr, length(CpGscores))) + xlab("CpG conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") + geom_vline(xintercept=CpGscores, color="red", alpha=0.5, size=0.2) + geom_vline(xintercept=mean(CpGscores), color="blue", alpha=0.8, size=2)
+  ggplot(prepped_data) + geom_histogram(aes(x=score_median, y=..count../nrow(prepped_data)), binwidth=0.05,fill="lightgray") +
+    xlim(-3,3) + ggtitle(sprintf("%s: CpG scores (%i sites)",geneStr, length(CpGscores))) +
+    xlab("CpG conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") +
+    #geom_vline(xintercept=CpGscores, color="red", alpha=0.5, size=0.2) +
+    geom_histogram(data=data.frame(score=CpGscores),aes(x=score,y=norm*..count../length(CpGscores)),binwidth=0.05, fill="blue",alpha=0.5) +
+    geom_vline(xintercept=mean(CpGscores), color="blue",linetype="dashed", size=1)
 }
@@
 cat("\n")
-plotGeneScore("TP53")
+#plotGeneScore("TP53",0.5)
+plotGeneScore("TUBA1A",0.285)
+#ggsave("genescore_TUBA1A.svg")
 cat("\n")
-plotGeneScore("TTN")
+plotGeneScore("TTN",0.37)
+#ggsave("genescore_TTN.svg")
 cat("\n")
-plotGeneScore("HLA-A")
+plotGeneScore("HLA-A",0.26)
+#ggsave("genescore_HLA-A.svg")
+
+# stats for each gene
+# do one sample wilcoxon rank sum
+gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^TUBA1A$"))) %>% pull(score_median) %>% wilcox.test(mu = 0, alternative = "two.sided")
+gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^TTN$"))) %>% pull(score_median) %>% wilcox.test(mu = 0, alternative = "two.sided")
+gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^HLA-A$"))) %>% pull(score_median) %>% wilcox.test(mu = 0, alternative = "two.sided")
 ```
@@
 
-cancergenes <- read_csv("/Users/kevinmurgas/Documents/Data+ project/Cancer gene lists/SangerCensus_COSMIC_CRC.csv")
-genescores <- gene_summary %>% filter(UCSC_RefGene_Name %in% (cancergenes %>% pull(`Gene Symbol`))) %>% pull(score_median)
+cancergenes1 <- read_csv("/Users/kevinmurgas/Documents/Data+ project/Cancer gene lists/SangerCensus_COSMIC_CRC.csv")
+genescores <- gene_summary %>% filter(UCSC_RefGene_Name %in% (cancergenes1 %>% pull(`Gene Symbol`))) %>% pull(score_median)
 ggplot(gene_summary) + geom_histogram(aes(x=score_median, y=..count../nrow(gene_summary)), binwidth=0.05) + xlim(-3,3) + ggtitle(sprintf("Gene Scores of %i CRC Genes", length(genescores))) + xlab("Gene conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") + geom_vline(xintercept=genescores, color="blue", alpha=0.5, size=0.2)
@@
 # source 2
-cancergenes <- read_csv("/Users/kevinmurgas/Documents/Data+ project/Cancer gene lists/CRCgenes_atlasgeneticsoncology.csv")
-genescores <- gene_summary %>% filter(UCSC_RefGene_Name %in% (cancergenes %>% pull(CancerGenes))) %>% pull(score_median)
+cancergenes2 <- read_csv("/Users/kevinmurgas/Documents/Data+ project/Cancer gene lists/CRCgenes_atlasgeneticsoncology.csv")
+genescores <- gene_summary %>% filter(UCSC_RefGene_Name %in% (cancergenes2 %>% pull(CancerGenes))) %>% pull(score_median)
 ggplot(gene_summary) + geom_histogram(aes(x=score_median, y=..count../nrow(gene_summary)), binwidth=0.05) + xlim(-3,3) + ggtitle(sprintf("Gene Scores of %i CRC Genes", length(genescores))) + xlab("Gene conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") + geom_vline(xintercept=genescores, color="blue", alpha=0.1, size=0.2)
+
+# compute % overlap of database genes and significant genes (By adj bootstrap)
+n1 = sum(permed_data %>% filter(score_median_v3>0.95) %>% pull(UCSC_RefGene_Name) %in% (cancergenes1 %>% pull(`Gene Symbol`)))
+n2 = sum(permed_data %>% filter(score_median_v3>0.95) %>% pull(UCSC_RefGene_Name) %in% (cancergenes2 %>% pull(CancerGenes)))
+
+# plot violins of: all genes, COSMIC, AGO
+perm_temp <- permed_data %>% select(c(UCSC_RefGene_Name, score_median,score_median_v3,n))
+temp <- rbind(perm_temp %>% mutate(group="all"),
+              perm_temp %>% filter(UCSC_RefGene_Name %in% (cancergenes1 %>% pull(`Gene Symbol`))) %>% mutate(group="COSMIC"),
+              perm_temp %>% filter(UCSC_RefGene_Name %in% (cancergenes2 %>% pull(CancerGenes))) %>% mutate(group="AGO")
+              )
+temp %>% mutate(group = factor(group, levels = c("COSMIC", "AGO", "all"))) %>%
+  ggplot() + geom_violin(aes(y=score_median,x=group,fill=group), draw_quantiles = 0.5) + scale_y_continuous(limits=c(-1,2),breaks = c(-1,0,1,2)) + theme_classic() + xlab("Database") + ylab("Gene Conservation Score")
+# ggsave("database_violins.svg")
+
+# try with histogram
+plotDBScore <- function(geneStr,norm) {
+  CpGscores <- gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^",geneStr,"$"))) %>% pull(score_median)
+
+  ggplot(prepped_data) + geom_histogram(aes(x=score_median, y=..count../nrow(prepped_data)), binwidth=0.05,fill="lightgray") +
+    xlim(-3,3) + ggtitle(sprintf("%s: CpG scores (%i sites)",geneStr, length(CpGscores))) +
+    xlab("CpG conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") +
+    #geom_vline(xintercept=CpGscores, color="red", alpha=0.5, size=0.2) +
+    geom_histogram(data=data.frame(score=CpGscores),aes(x=score,y=norm*..count../length(CpGscores)),binwidth=0.05, fill="blue",alpha=0.5) +
+    geom_vline(xintercept=mean(CpGscores), color="blue",linetype="dashed", size=1)
+}
+
 ```
@@
   pull(UCSC_RefGene_Name) %>% tidy_gsea()
-#enrich_score_prob <- permed_data %>%
-#  filter(score_prob_v3 > quantile(score_prob_v3, 0.95, na.rm = TRUE)) %>%
-#  pull(UCSC_RefGene_Name) %>% tidy_gsea()
-#enrich_score_mean <- permed_data %>%
-#  filter(score_mean_v3 > quantile(score_mean_v3, 0.95, na.rm = TRUE)) %>%
-#  pull(UCSC_RefGene_Name) %>% tidy_gsea()
 enrich_score_median <- permed_data %>%
@@
   filter(score_median_v3 > 0.95) %>%
-  pull(UCSC_RefGene_Name) %>% tidy_gsea()
-enrich_score_median2 <- permed_data %>%
-  #filter(score_median2_v3 > quantile(score_median2_v3, 0.95, na.rm = TRUE)) %>%
-  filter(score_median2_v3 > 0.95) %>%
-  pull(UCSC_RefGene_Name) %>% tidy_gsea()
-enrich_score_reg <- permed_data %>%
-  #filter(score_reg_v3 > quantile(score_reg_v3, 0.95, na.rm = TRUE)) %>%
-  filter(score_reg_v3 > 0.95) %>%
   pull(UCSC_RefGene_Name) %>% tidy_gsea()
@@
   filter(!is.na(UCSC_RefGene_Name)) %>%
-  gene_summarize(UCSC_RefGene_Name, c(mu:sigmaT,score_median:score_reg))
+  gene_summarize(UCSC_RefGene_Name, c(mu:sigmaT,score_median))
 cat(proc.time()-ptm)
@@
 pP1 <- post_histogram(gene_summary$score_median,0.05,c(-3,3),"Score_median")
-pP2 <- post_histogram(gene_summary$score_median2,0.05,c(-3,3),"Score_median2")
-pP3 <- post_histogram(gene_summary$score_reg,0.05,c(-3,3),"Score_reg")
-
-#pP1 <- post_histogram(gene_summary$score_prob,0.01,c(0,1),"Score_prob")
-#pP2 <- post_histogram(gene_summary$score_mean,0.05,c(-3,3),"Score_mean")
-
 # arrange each parameter as a subplot
-p_all <- ggarrange(pN,pM,pB,pP,pPT,pT,pP1,pP2,pP3,hjust=0,nrow=3,ncol=3)
+p_all <- ggarrange(pN,pM,pB,pP,pPT,pT,pP1,hjust=0,nrow=3,ncol=3)
 print(p_all)
-rm(pN,pM,pB,pP,pPT,pT,pP1,pP2,pP3)
+rm(pN,pM,pB,pP,pPT,pT,pP1)
 
@@
 
-#pP1 <- ggplot(gene_summary) + geom_point(aes(x=n,y=score_prob), alpha=0.01) + ggtitle("Score_prob vs N-sites")
-#pP2 <- ggplot(gene_summary) + geom_point(aes(x=n,y=score_mean), alpha=0.01) + ggtitle("Score_mean vs N-sites")
 pP1 <- ggplot(gene_summary) + geom_point(aes(x=n,y=score_median), alpha=0.01) + ggtitle("Score_median vs N-sites")
-pP2 <- ggplot(gene_summary) + geom_point(aes(x=n,y=score_median2), alpha=0.01) + ggtitle("Score_median2 vs N-sites")
-pP3 <- ggplot(gene_summary) + geom_point(aes(x=n,y=score_reg), alpha=0.01) + ggtitle("Score_reg vs N-sites")
 
 # arrange each parameter as a subplot
-p_all <- ggarrange(pM,pB,pE,pP,pPT,pT,pP1,pP2,pP3,hjust=0,nrow=3,ncol=3)
+p_all <- ggarrange(pM,pB,pE,pP,pPT,pT,pP1,hjust=0,nrow=3,ncol=3)
 print(p_all)
-rm(pM,pB,pE,pP,pPT,pT,pP1,pP2,pP3)
+rm(pM,pB,pE,pP,pPT,pT,pP1)
 ```
@@
   mutate(method = factor(method, c("1", "3"), c("Naive bootstrap", "Adjusted bootstrap")),
-         type = factor(type, c("score_median","score_median2","score_reg"), c("Score_median","Score_median2","Score_reg"))) %>%
+         type = factor(type, c("score_median"), c("Score_median"))) %>%
   drop_na() %>%
@@
 cat("Number of enrichment pathways (probs):")
-c(score_median = length(enrich_score_median$Description),
-  score_median2 = length(enrich_score_median2$Description),
-  score_reg = length(enrich_score_reg$Description)) %>%
+c(score_median = length(enrich_score_median$Description)) %>%
   sort()
@@
 list(
-  score_median = enrich_score_median$Description,
-  score_median2 = enrich_score_median2$Description,
-  score_reg = enrich_score_reg$Description
+  score_median = enrich_score_median$Description
 ) %>%
   UpSetR::fromList() %>%
-  UpSetR::upset(sets.bar.color = RColorBrewer::brewer.pal(3, "Set2")[c( 2, 1)])
+  UpSetR::upset(sets.bar.color = RColorBrewer::brewer.pal(3, "Set2")[c( 1)])
 grid::grid.text("Figure 3B: Frequency of overlap between pathways that are called \nas most conserved for each conservation score.", x=0.65, y=0.05, gp=grid::gpar(fontsize=10))
@@
 
-# score_median2
-cat("score_median2")
-ggg <- enrich_score_median2 %>%
-  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
-  labs(caption = "Figure 4E: Relationship between pathways that are called as conserved based on score_median2.", title = NULL)
-ggg$layers <- ggg$layers[1:2]
-ggg + coord_flip() + 
-  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
-  scale_color_gradientn(colors = "black") +
-  guides(color = "none", size = "none")
-
-# score_reg
-cat("score_reg")
-#ggg <- enrich_score_reg %>%
-#  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
-#  labs(caption = "Figure 4G: Relationship between pathways that are called as conserved based on score_reg.", title = NULL)
-#ggg$layers <- ggg$layers[1:2]
-#ggg + coord_flip() +
-#  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
-#  scale_color_gradientn(colors = "black") +
-#  guides(color = "none", size = "none")
-```
-
-\subsection{Figure 5: Gene Conservation vs Gene Expression}
-```{r Fig5, echo=FALSE, message=FALSE}
-# Figure 5A
-figure5_data <- permed_data %>% 
-  left_join(by = c("UCSC_RefGene_Name" = "gene"), promoted_data) %>%
-  left_join(expression_atlas %>% select(`Gene Name`, colon),
-            by = c("UCSC_RefGene_Name" = "Gene Name"))
-
-bind_rows(
-  figure5_data %>%
-    select(sigmaP_v3, sigmaP_pro, colon) %>%
-    pivot_longer(-colon) %>%
-    drop_na() %>%
-    rename(expression = colon),
-  figure5_data %>%
-    select(sigmaPT_v3, sigmaPT_pro, colon) %>%
-    pivot_longer(-colon) %>%
-    drop_na() %>%
-    rename(expression = colon),
-  figure5_data %>%
-    select(sigmaT_v3, sigmaT_pro, colon) %>%
-    pivot_longer(-colon) %>%
-    drop_na() %>%
-    rename(expression = colon)
-) %>%
-  group_by(name) %>%
-  mutate(value = factor(11-as.numeric(cut_number(value, 10)))) %>%
-  separate(name, c("type", "method"), sep = "_", extra = "merge") %>%
-  mutate(type = factor(type, c("sigmaP","sigmaPT","sigmaT"), c("SigmaP-Colon","SigmaPT-Colon","SigmaT-colon")),
-         method = factor(method, c("v3", "pro"), c("Boot-strapped values", "Promoter region"))) %>%
-  ggplot(aes(value, expression, fill = type)) +
-  geom_boxplot() +
-  scale_y_log10() +
-  guides(color = "none") +
-  facet_grid(method ~ type) +
-  labs(x = "quantile of conservation by gene", y = "Expression", title = NULL,
-       caption = "Figure 5A: Relationship between hierarchical variance (Sigmas) and gene expression.") +
-  theme_minimal() +
-  scale_fill_brewer(palette = "Set2") +
-  theme(legend.position = "none")
-
-# Figure 5B
-bind_rows(
-  figure5_data %>%
-    select(score_median_v3, score_median_pro, colon) %>%
-    rename(scoremedian_v3 = score_median_v3, scoremedian_pro = score_median_pro) %>%
-    pivot_longer(-colon) %>%
-    drop_na() %>%
-    rename(expression = colon),
-  figure5_data %>%
-    select(score_median2_v3, score_median2_pro, colon) %>%
-    rename(scoremedian2_v3 = score_median2_v3, scoremedian2_pro = score_median2_pro) %>%
-    pivot_longer(-colon) %>%
-    drop_na() %>%
-    rename(expression = colon),
-  figure5_data %>%
-    select(score_reg_v3, score_reg_pro, colon) %>%
-    rename(scorereg_v3 = score_reg_v3, scorereg_pro = score_reg_pro) %>%
-    pivot_longer(-colon) %>%
-    drop_na() %>%
-    rename(expression = colon)
-) %>%
-  group_by(name) %>%
-  mutate(value = factor(as.numeric(cut_number(value, 10)))) %>%
-  separate(name, c("type", "method"), sep = "_", extra = "merge") %>%
-  mutate(type = factor(type, c("scoremedian","scoremedian2","scorereg"), c("ScoreMedian-colon","ScoreMedian2-colon","Scorereg-colon")),
-         method = factor(method, c("v3", "pro"), c("Boot-strapped values", "Promoter region"))) %>%
-  ggplot(aes(value, expression, fill = type)) +
-  geom_boxplot() +
-  scale_y_log10() +
-  guides(color = "none") +
-  facet_grid(method ~ type) +
-  labs(x = "quantile of conservation by gene", y = "Expression", title = NULL,
-       caption = "Figure 5B: Relationship between conservation scores and gene expression.") +
-  theme_minimal() +
-  scale_fill_brewer(palette = "Set2") +
-  theme(legend.position = "none")
 ```
@@
 
-<!-- \newpage -->
-<!-- \subsection{A4: Score Prob} -->
-<!-- Genes are selected with cutoff >95th percentile (adjusted bootstrap), showing top 50 genes/pathways -->
-<!-- ```{r App_score_prob, echo=FALSE, comment=NA, message=FALSE} -->
-<!-- enrichedGenes <- permed_data %>%  -->
-<!--   select(UCSC_RefGene_Name, score_prob, score_prob_v3, n) %>% -->
-<!--   rename(Gene=UCSC_RefGene_Name, Score_median=score_prob, Pctile=score_prob_v3) %>% -->
-<!--   filter(Pctile > quantile(Pctile, 0.95, na.rm = TRUE)) %>% -->
-<!--   arrange(Pctile) -->
-
-<!-- enrichedPaths <- tibble(Pathway = enrich_score_prob$Description, ngenes = enrich_score_prob$Count, padj = enrich_score_prob$p.adjust) %>% arrange(padj) -->
-
-<!-- ngene <- dim(enrichedGenes)[1] -->
-<!-- npath <- length(enrich_score_prob$Description) -->
-
-<!-- cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath)) -->
-
-<!-- nshow = 50 -->
-<!-- if (ngene > nshow) { -->
-<!--   enrichedGenes <- enrichedGenes[1:nshow,] -->
-<!-- } -->
-<!-- if (npath > nshow) { -->
-<!--   enrichedPaths <- enrichedPaths[1:nshow,] -->
-<!-- } -->
-
-<!-- enrichedGenes %>% pander() -->
-
-<!-- cat("Score_prob top 50 enriched paths:  \n") -->
-<!-- enrichedPaths %>% pander() -->
-<!-- ``` -->
-
-<!-- \newpage -->
-<!-- \subsection{A5: Score Mean} -->
-<!-- Genes are selected with cutoff >95th percentile (adjusted bootstrap), showing top 50 genes/pathways -->
-<!-- ```{r App_score_mean, echo=FALSE, comment=NA, message=FALSE} -->
-<!-- enrichedGenes <- permed_data %>%  -->
-<!--   select(UCSC_RefGene_Name, score_mean, score_mean_v3, n) %>% -->
-<!--   rename(Gene=UCSC_RefGene_Name, Score_median=score_mean, Pctile=score_mean_v3) %>% -->
-<!--   filter(Pctile > quantile(Pctile, 0.95, na.rm = TRUE)) %>% -->
-<!--   arrange(Pctile) -->
-
-<!-- enrichedPaths <- tibble(Pathway = enrich_score_mean$Description, ngenes = enrich_score_mean$Count, padj = enrich_score_mean$p.adjust) %>% arrange(padj) -->
-
-<!-- ngene <- dim(enrichedGenes)[1] -->
-<!-- npath <- length(enrich_score_mean$Description) -->
-
-<!-- cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath)) -->
-
-<!-- nshow = 50 -->
-<!-- if (ngene > nshow) { -->
-<!--   enrichedGenes <- enrichedGenes[1:nshow,] -->
-<!-- } -->
-<!-- if (npath > nshow) { -->
-<!--   enrichedPaths <- enrichedPaths[1:nshow,] -->
-<!-- } -->
-
-<!-- enrichedGenes %>% pander() -->
-
-<!-- cat("Score_mean top 50 enriched paths:  \n") -->
-<!-- enrichedPaths %>% pander() -->
-<!-- ``` -->
-
-\newpage
-\subsection{A6: Score Median}
+\newpage
+\subsection{A4: Score Median}
 Genes are selected with cutoff >95th percentile (adjusted bootstrap), showing top 50 genes/pathways

### 2020-11-24 StanAnalysis_updated.Rmd
Edits to figures

@@
   plot_df <- data.frame(val=postvals, gene_assoc=mask)
-  p1 <- ggplot(plot_df, aes(x=val,fill=gene_assoc)) +
+  p1 <- ggplot(plot_df, aes(x=val)) + #,fill=gene_assoc
     #geom_histogram(aes(y=..count../length(mask)), binwidth=bw, position="dodge") +
     #stat_function(fun=function(x) bw*priorfun(x),geom="line",color="black") +
-    geom_histogram(aes(y=..density..), binwidth=bw, position="dodge") +
-    stat_function(fun=function(x) priorfun(x),geom="line",color="black") +
+    geom_histogram(aes(y=..density..), binwidth=bw, position="dodge", fill="lightgray") +
+    stat_function(fun=function(x) priorfun(x), geom="line", color="black") +
     ggtitle(titlestr) + xlim(xl) + theme_classic() + theme(legend.position = "none") + xlab(titlestr) + ylab("Freq.")
@@
 # plot mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT with priors overlaid
-pM <- post_prior_histogram(mu_full$p50, genemask, function(x) 1.3*fBG(x,0.3557,-3.0675, 1.3027, 1.1290, 2.4110), 0.25, c(-5,5), "Mu")
-pB <- post_prior_histogram(betaT_full$p50, genemask, function(x) 1.5*dcauchy(x,-0.04051, 0.5799), 0.25, c(-5,5), "BetaT")
-pE <- post_prior_histogram(sigmaE_full$p50, genemask, function(x) 3.3*dgamma(x,2,2), 0.05, c(0,2), "SigmaE")
-pP <- post_prior_histogram(sigmaP_full$p50, genemask, function(x) 1.52*dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
-pPT <- post_prior_histogram(sigmaPT_full$p50, genemask, function(x) 1.6*dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
-pT <- post_prior_histogram(sigmaT_full$p50, genemask, function(x) 0.97*dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT")
+pM <- post_prior_histogram(mu_full$p50, genemask, function(x) 1.3*fBG(x,0.3557,-3.0675, 1.3027, 1.1290, 2.4110), 0.25, c(-5,5), "Mu") + theme(aspect.ratio = 1)
+pB <- post_prior_histogram(betaT_full$p50, genemask, function(x) 1.5*dcauchy(x,-0.04051, 0.5799), 0.25, c(-5,5), "BetaT") + theme(aspect.ratio = 1)
+pE <- post_prior_histogram(sigmaE_full$p50, genemask, function(x) 3.3*dgamma(x,2,2), 0.05, c(0,2), "SigmaE") + theme(aspect.ratio = 1)
+pP <- post_prior_histogram(sigmaP_full$p50, genemask, function(x) 1.52*dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP") + theme(aspect.ratio = 1)
+pPT <- post_prior_histogram(sigmaPT_full$p50, genemask, function(x) 1.6*dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT") + theme(aspect.ratio = 1)
+pT <- post_prior_histogram(sigmaT_full$p50, genemask, function(x) 0.97*dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT") + theme(aspect.ratio = 1)
 # log sigmas
-pLP <- post_prior_histogram(log(sigmaP_full$p50), genemask, function(x) 0.34*dgamma(exp(x), 2.7693, 7.0029), 0.25, c(-5,5), "log(SigmaP)")
-pLPT <- post_prior_histogram(log(sigmaPT_full$p50), genemask, function(x) 0.42*dgamma(exp(x), 2.1457, 2.4819), 0.25, c(-5,5), "log(SigmaPT)")
-pLT <- post_prior_histogram(log(sigmaT_full$p50), genemask, function(x) 0.22*dgamma(exp(x), 3.3643, 10.430), 0.25, c(-5,5), "log(SigmaT)") + theme(legend.position = c(0.8, 0.6))
+#pLP <- post_prior_histogram(log(sigmaP_full$p50), genemask, function(x) 0.34*dgamma(exp(x), 2.7693, 7.0029), 0.25, c(-5,5), "log(SigmaP)")
+#pLPT <- post_prior_histogram(log(sigmaPT_full$p50), genemask, function(x) 0.42*dgamma(exp(x), 2.1457, 2.4819), 0.25, c(-5,5), "log(SigmaPT)")
+#pLT <- post_prior_histogram(log(sigmaT_full$p50), genemask, function(x) 0.22*dgamma(exp(x), 3.3643, 10.430), 0.25, c(-5,5), "log(SigmaT)") + theme(legend.position = c(0.8, 0.6))
 
 # arrange each parameter as a subplot # ,labels=c("M","B","E","P","PT","T","lP","lPT","lT")
-p_all <- ggarrange(pM,pB,pE,pP,pPT,pT,pLP,pLPT,pLT,hjust=0,nrow=3,ncol=3)
+p_all <- ggarrange(pM,pB,pE,pP,pPT,pT,hjust=0,nrow=2,ncol=3) #,pLP,pLPT,pLT
 print(p_all)
 # ggsave("posteriors_geneassoc.svg",p_all)
-rm(pM,pB,pE,pP,pPT,pT,pLP,pLPT,pLT)
+rm(pM,pB,pE,pP,pPT,pT)
 ```
@@
 # plot the two conservations score prob sums
-pP1 <- post_histogram(CpGscore_full$median,genemask,0.05,c(-3,3),expression("Median log("~sigma[P]/sigma[T]~")")) + scale_x_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3))
-prob1_proxy = log(sigmaP_full$p50/sigmaT_full$p50)
-pP2 <- post_histogram(prob1_proxy,genemask,0.05,c(-3,3),expression(" log( median" ~sigma[P]/~"median"~sigma[T]~")"))
-# regularized version
-L2norm = sqrt(sigmaP_full$p50^2 + sigmaT_full$p50^2)
-prob1_reg = log(sigmaP_full$p50/sigmaT_full$p50) - 1/L2norm
-pP3 <- post_histogram(prob1_reg,genemask,0.1,c(-10,5),"Regularized")
-p_probs <- ggarrange(pP1,pP2,pP3,nrow=1,ncol=3)
-print(p_probs)
+pP1 <- post_histogram(CpGscore_full$median,genemask,0.1,c(-3,3),expression("Median log("~sigma[P]/sigma[T]~")")) + scale_x_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3))
+# ggsave("CpGscore_geneassoc.svg")
+# prob1_proxy = log(sigmaP_full$p50/sigmaT_full$p50)
+# pP2 <- post_histogram(prob1_proxy,genemask,0.05,c(-3,3),expression(" log( median" ~sigma[P]/~"median"~sigma[T]~")"))
+# # regularized version
+# L2norm = sqrt(sigmaP_full$p50^2 + sigmaT_full$p50^2)
+# prob1_reg = log(sigmaP_full$p50/sigmaT_full$p50) - 1/L2norm
+# pP3 <- post_histogram(prob1_reg,genemask,0.1,c(-10,5),"Regularized")
+# p_probs <- ggarrange(pP1,pP2,pP3,nrow=1,ncol=3)
+# print(p_probs)
 ```
@@
 permed_data <- read_csv(here(prepdatafolder, "permed_data.csv"))
-promoted_data <- read_csv(here(prepdatafolder, "promoted_data.csv"))
 cat("done")
@@
   dplyr::rename(ir = Relation_to_UCSC_CpG_Island, score=score_median) %>%
-  mutate(ir = case_when(ir=="" ~ "Sea",
+  mutate(ir = case_when(is.na(ir) ~ "Sea",
                         str_detect(ir, "^S") ~ str_replace(ir, "^S_", ""),
@@
                         TRUE ~ ir)) %>%
-  mutate(ir = factor(ir, levels = c("Island", "Shore", "Shelf", "Sea"))) %>%
-  #ggplot() + geom_density(aes(x=score,y=..scaled..,color=ir)) + theme_classic() + scale_x_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3)) + ylab("Normalized Density")
-  ggplot() + geom_violin(aes(x=ir,y=score,fill=ir),draw_quantiles = 0.5) + theme_classic() + scale_y_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3))
-# ggsave("CpGscore_islands.svg")
+  #mutate(ir = factor(ir, levels = c("Island", "Shore", "Shelf", "Sea"))) %>%
+  #ggplot() + geom_density(aes(x=score,y=..scaled..,color=ir)) + theme_classic() + scale_x_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3)) + ylab("Normalized Density") + theme(aspect.ratio=1/4)
+  mutate(ir = factor(ir, levels = c("Sea", "Shelf", "Shore", "Island"))) %>%
+  ggplot() + geom_violin(aes(y=ir,x=score,fill=ir),draw_quantiles = 0.5) + theme_classic() + scale_x_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3)) + theme(aspect.ratio=1/4,legend.position = "none")
+# ggsave("CpGscore_islands_violin.svg")
 
@@
 #   filter(fr %in% keepregions) %>%
-#   #mutate(fr = factor(fr, levels = c("Island", "Shore", "Shelf", "Sea"))) %>%
 #   ggplot() + geom_density(aes(x=score,color=fr)) + theme_classic() + scale_x_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3)) + ylab("Normalized Density")
+keepregions = c("3'UTR","Body","5'UTR","TSS200")
 temp %>%
@@
   filter(fr %in% keepregions) %>%
-  ggplot() + geom_violin(aes(x=fr,y=score,fill=fr),draw_quantiles = 0.5) + theme_classic() + scale_y_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3))
+  ggplot() + geom_violin(aes(y=fr,x=score,fill=fr), draw_quantiles = 0.5) + theme_classic() + scale_x_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3)) + theme(aspect.ratio=1/4,legend.position = "none")
 # ggsave("CpGscore_regions_violin.svg")
@@
 pP1 <- post_histogram(gene_summary$score_median,0.05,c(-3,3),"Score_median")
-pP2 <- post_histogram(gene_summary$score_median2,0.05,c(-3,3),"Score_median2")
-pP3 <- post_histogram(gene_summary$score_reg,0.05,c(-3,3),"Score_reg")
-
-#pP1 <- post_histogram(gene_summary$score_prob,0.01,c(0,1),"Score_prob")
-#pP2 <- post_histogram(gene_summary$score_mean,0.05,c(-3,3),"Score_mean")
 
 # arrange each parameter as a subplot
-p_all <- ggarrange(pN,pM,pB,pP,pPT,pT,pP1,pP2,pP3,hjust=0,nrow=3,ncol=3)
+p_all <- ggarrange(pN,pM,pB,pP,pPT,pT,pP1,hjust=0,nrow=3,ncol=3)
 print(p_all)
-rm(pN,pM,pB,pP,pPT,pT,pP1,pP2,pP3)
+rm(pN,pM,pB,pP,pPT,pT,pP1)
 
@@
   filter(!is.na(UCSC_RefGene_Name)) %>%
-  select(CHR, UCSC_RefGene_Name, cpgdist, c(sigmaP:sigmaT,score_median:score_reg)) %>%
+  select(CHR, UCSC_RefGene_Name, cpgdist, c(sigmaP:sigmaT,score_median)) %>%
   mutate(location_bin = cut(cpgdist, c(seq(-1500, 20000, by = 100)), include.lowest = TRUE)) %>%
   group_by(location_bin) %>%
-  summarise(mean_sigmaP = mean(sigmaP), mean_sigmaPT = mean(sigmaPT), mean_sigmaT = mean(sigmaT),  mean_score_median = mean(score_median), mean_score_median2 = mean(score_median2), n_obs = n(), .groups='drop') %>%
+  summarise(mean_sigmaP = mean(sigmaP), mean_sigmaPT = mean(sigmaPT), mean_sigmaT = mean(sigmaT),  mean_score_median = mean(score_median), n_obs = n(), .groups='drop') %>%
   mutate(location_bin = as.numeric(str_extract(location_bin, "(?<=,)(.*)(?=])")))
@@
   drop_na() %>%
-  pivot_longer(mean_score_median:mean_score_median2) %>% 
-  mutate(name = factor(name, c("mean_score_median","mean_score_median2"), c("Score_median","Score_median2"))) %>%
+  pivot_longer(mean_score_median) %>% 
+  #mutate(name = factor(name, c("mean_score_median"), c("Score_median"))) %>%
   sample_frac() %>%
@@
        y = "Average Conservation Score", 
-       caption = "Figure 1: Average conservation scores of single CpGs as a function of position relative to gene.") +
+       caption = "Figure 1: Average conservation scores of single CpGs as a function of position relative to gene."
+       ) +
   scale_color_brewer(palette = "Set2") +
@@
 cat("\n")
+# ggsave("CpGdist.svg")
 

### 2020-11-24 conserved_genes_and_paths.Rmd
Edits to figures

@@
 library(gt)
-library(dplyr)
 # set working directory
@@
 
-permed_data <- read_csv(here("data-stan_rg3ps3", "permed_data.csv"))
+permed_data <- read_csv(here("data_final_rg3", "permed_data.csv"))
 #enrichedGenes <- read_csv(here("prob1genes.csv"))
@@
 
-# Gene Set Enrichment
-tidy_gsea <- function(genes) {
+# Gene Set Enrichment with Reactome
+tidy_reactome <- function(genes) {
   AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, genes, 'ENTREZID', 'SYMBOL') %>%
     ReactomePA::enrichPathway()
+}
+# GSEA with KEGG
+tidy_kegg <- function(genes) {
+  AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, genes, 'ENTREZID', 'SYMBOL') %>%
+    clusterProfiler::enrichKEGG()
 }
@@
 enrich_reactome <-enrichedGenes %>%
-  pull(Gene) %>% tidy_gsea()
+  pull(Gene) %>% tidy_reactome()
+proc.time()-ptm
+cat("Begin KEGG GSEA...")
+ptm<-proc.time()
+enrich_kegg <-enrichedGenes %>%
+  pull(Gene) %>% tidy_kegg()
 proc.time()-ptm

### 2021-01-05 Paper_Figures.Rmd
Removed extraneous code, minor code edits

@@
 prepdatafolder = "data_final_rg3"
-cat(sprintf("Load prepped data dir: %s (needs to be pre-processed)...",prepdir))
+cat(sprintf("Load prepped data dir: %s (needs to be pre-processed)...",prepdatafolder))
 gened_data <- read_csv(here(prepdatafolder, "gened_data.csv"), col_types = list("CHR" = col_character()))
@@
 
-# # plot beta/M values
-# all_beta = FullAnnotation[,-c(1,2,3,4,5,6,7,8,9)]
-# mean_beta = rowMeans(all_beta)
-# all_M = logit(all_beta)
-# mean_M = rowMeans(all_M)
-# p_beta <- data.frame(value=mean_beta) %>% gather %>% ggplot() + geom_histogram(aes(x=value),binwidth=0.01,fill="lightgray",color="grey50") + theme_classic() + xlim(c(0,1)) + theme(aspect.ratio = 1) + xlab("Beta value")
-# p_M <- data.frame(value=mean_M) %>% gather %>% ggplot() + geom_histogram(aes(x=value),binwidth=0.1,fill="lightgray",color="grey50") + theme_classic() + xlim(c(-5,5)) + theme(aspect.ratio = 1) + xlab("M value")
-# p_BM <- ggarrange(p_beta, p_M, nrow = 1, ncol = 2)
-# # ggsave("betaMdists.svg",p_BM)
 ```
@@
 
-\subsection{Prior Probabilities}
+\subsection{Prior Probabilities (Fig S1)}
 Priors for each parameter (mu, nu, sigmaP/PT/T) were defined using TCGA 450k array data from colorectal tumor patients with paired multiple tumor and/or normal+tumor samples.
@@
 
-# plot mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT with priors overlaid
+# plot TCGA estimates for mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT; with prior fits overlaid
 pM <- post_prior_histogram(mu_ests, function(x) fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), function(x) fBG(x,0.3557,-3.0675, sqrt(3)*0.75214,1.1290, sqrt(3)*1.39197), 0.1, c(-5,5), "Mu")
@@
 \subsection{Figure 2A: Example site fit, 2B: Posteriors}
-![Fit](/Users/kevinmurgas/Documents/Data+ project/Manuscript/example_fit.png)
+![Fit](/Users/kevinmurgas/Documents/Data+ project/Manuscript/Paper Figures (screenshots)/example_fit.png)
 \newline \newline
-![Posteriors](/Users/kevinmurgas/Documents/Data+ project/Manuscript/example_posteriors.png)
+![Posteriors](/Users/kevinmurgas/Documents/Data+ project/Manuscript/Paper Figures (screenshots)/example_posteriors.png)
+
 \newpage
-\section{Figure 2: Model Fit Results}
-\subsection{Parameter Distributions}
+\section{Model Fit Results}
+\subsection{Fig 3C: Parameter Distributions}
 Histograms are median values of posterior distribution
@@
 
-# plot mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT with priors overlaid
+# plot hierarchical model posterior medians for mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT; with priors overlaid
 pM <- post_prior_histogram(mu_full$p50, genemask, function(x) 1.5*fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), 0.1, c(-5,5), "Mu")
@@
 rm(pM,pB,pE,pP,pPT,pT,pLP,pLPT,pLT)
-```
-
-\subsection{Conservation score}
-```{r Fig1B, echo=FALSE, fig.height=2, fig.width=6.5, fig.fullwidth=TRUE, warning=FALSE}
-
-post_histogram <- function(postvals,mask,bw,xl,titlestr) {
-  # stacked histogram colored by gene-associated sites
-  # and prior curve overlaid
-  plot_df <- data.frame(val=postvals, gene_assoc=mask)
-  p1 <- ggplot(plot_df, aes(x=val)) +
-    geom_histogram(aes(y=..count../length(mask),fill=gene_assoc), binwidth=bw) +
-    ggtitle(titlestr) + xlim(xl) + theme_classic() + theme(legend.position = "none") + xlab(titlestr) + ylab("Freq.")
-}
-
-# overlaid transparent histograms
-post_histogram <- function(postvals,mask,bw,titlestr) {
-  # stacked histogram colored by gene-associated sites
-  # and prior curve overlaid
-  plot_df <- data.frame(val=postvals, gene_assoc=mask)
-  gene_assoc_df <- plot_df %>% filter(gene_assoc)
-  not_gene_assoc_df <- plot_df %>% filter(!gene_assoc)
-  p1 <- ggplot(plot_df, aes(x=val)) +
-    geom_histogram(data=not_gene_assoc_df, aes(y=..density..,fill="#00BFC4",alpha=0.4), binwidth=bw) +
-    geom_histogram(data=gene_assoc_df, aes(y=..density..,fill="#F8766D",alpha=0.4), binwidth=bw) +
-    ggtitle(titlestr) + xlab(titlestr) + ylab("Density") + theme_classic() 
-  p1
-}
-
-# plot the two conservations score prob sums
-genemask = !(FullAnnotation$UCSC_RefGene_Name=="")
-p_scores <- post_histogram(CpGscore_full$median,genemask,0.05,"CpG Conservation score") +
-  theme(aspect.ratio=1/4,legend.position = "none") +
-  scale_x_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3))
-print(p_scores)
-# ggsave("CpGscores_overlay.svg")
-
-# plot gene scores and adjusted bootstrap distribution
-p1 <- permed_data %>% ggplot(aes(x=score_median)) + geom_histogram(aes(y=..count..), binwidth=0.05, fill="lightgray",color = "grey50") + xlim(c(-1,2.5)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
-p2 <- permed_data %>% ggplot(aes(x=score_median_v3)) + geom_histogram(aes(y=..count..), binwidth=0.02, fill="lightgray", color="grey50") + xlim(c(0,1)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
-p_genes <- ggarrange(p1,p2,nrows=2,ncol=1)
-# ggsave("gene_scores_bootstrap.svg")
 ```
@@
 
+\subsection{Fig 3A: Conservation score}
+```{r Fig3A, echo=FALSE, fig.height=2, fig.width=6.5, fig.fullwidth=TRUE, warning=FALSE}
+
+# staggered (dodge) histograms
+post_histogram <- function(postvals,mask,bw,titlestr) {
+  # stacked histogram colored by gene-associated sites
+  plot_df <- data.frame(val=postvals, gene_assoc=mask)
+  p1 <- ggplot(plot_df, aes(x=val)) +
+    geom_histogram(aes(y=..density..,fill=gene_assoc), binwidth=bw, position="dodge") +
+    ggtitle(titlestr) + theme_classic() + theme(legend.position = "none") + xlab(titlestr) + ylab("Freq.")
+}
+
+# overlaid transparent histograms
+post_histogram <- function(postvals,mask,bw,titlestr) {
+  # transparent histogram colored by gene-associated sites
+  plot_df <- data.frame(val=postvals, gene_assoc=mask)
+  gene_assoc_df <- plot_df %>% filter(gene_assoc)
+  not_gene_assoc_df <- plot_df %>% filter(!gene_assoc)
+  p1 <- ggplot(plot_df, aes(x=val)) +
+    geom_histogram(data=not_gene_assoc_df, aes(y=..density..,fill="#00BFC4",alpha=0.4), binwidth=bw) +
+    geom_histogram(data=gene_assoc_df, aes(y=..density..,fill="#F8766D",alpha=0.4), binwidth=bw) +
+    ggtitle(titlestr) + xlab(titlestr) + ylab("Density") + theme_classic() 
+  p1
+}
+
+# plot the two conservations score prob sums (using /log(2) to convert log_e to log_2)
+genemask = !(FullAnnotation$UCSC_RefGene_Name=="")
+p_scores <- post_histogram(CpGscore_full$median/log(2),genemask,0.05,"CpG Conservation score") +
+  theme(aspect.ratio=1/4,legend.position = "none") +
+  scale_x_continuous(limits=c(-4,4),breaks=c(-4,-2,0,2,4)) + 
+  scale_y_continuous(breaks=c(0,0.1,0.2,0.3,0.4,0.5))
+print(p_scores)
+# ggsave("CpGscores_overlay.svg")
+
+# Fig 4B: plot gene scores and adjusted bootstrap distribution
+p1 <- permed_data %>% ggplot(aes(x=score_median/log(2))) + geom_histogram(aes(y=..count..), binwidth=0.1, fill="lightgray",color = "grey50") + xlim(c(-1.5,3.5)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
+p2 <- permed_data %>% ggplot(aes(x=score_median_v3)) + geom_histogram(aes(y=..count..), binwidth=0.02, fill="lightgray", color="grey50") + xlim(c(0,1)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
+p_genes <- ggarrange(p1,p2,nrows=2,ncol=1)
+# ggsave("gene_scores_bootstrap.svg")
+```
+
 \newpage
@@
 
-\subsection{Regulatory regions}
-Table comparing different groups of CpG sites:
-```{r Table1, echo=FALSE, warning=FALSE}
-
+\subsection{CpG Island Relation (Fig 3B)}
+Table comparing different groups of CpG sites based on CpG Island relationship: Island, Shore, Shelf, Sea
+```{r CpGIslandRelation, echo=FALSE, warning=FALSE}
 #, include=TRUE, results='asis', fig.keep='all', fig.show='asis', fig.height=3, fig.width=5
 
+# for a table summary of sigmas and conservation score
 prepped_data %>%
   group_by(`Island relation` = Relation_to_UCSC_CpG_Island) %>%
+  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
   summarise_at(vars(sigmaP:sigmaT,score_median), mean, na.rm = TRUE) %>%
@@
 # stats for cpg island relations
-# summarize mean of each group, do kruskal wallis and post hoc Tukey
 my_data = prepped_data %>% select(score_median, Relation_to_UCSC_CpG_Island) %>%
+  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
   group_by(ir = Relation_to_UCSC_CpG_Island) %>%
@@
                         TRUE ~ ir))
-my_data %>%
-  summarise_at(vars(score_median), mean, na.rm = TRUE)
+# summarize mean of each group and convert to fold change
+my_data %>% summarise_at(vars(score_median), mean, na.rm = TRUE) %>% mutate(fc = 2^score_median)
+# run Kruskal Wallis ANOVA and post-hoc Tukey
 kruskal.test(score_median ~ ir, data = my_data)
@@
 
+# Fig 3B: plot CpG island relation score distributions
+prepped_data %>% select(Name, Relation_to_UCSC_CpG_Island, score_median) %>%
+  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
+  dplyr::rename(ir = Relation_to_UCSC_CpG_Island, score=score_median) %>%
+  mutate(ir = case_when(is.na(ir) ~ "Sea",
+                        str_detect(ir, "^S") ~ str_replace(ir, "^S_", ""),
+                        str_detect(ir, "^N") ~ str_replace(ir, "^N_", ""),
+                        TRUE ~ ir)) %>%
+  mutate(ir = factor(ir, levels = c("Island", "Shore", "Shelf", "Sea"))) %>%
+  #mutate(ir = factor(ir, levels = c("Sea", "Shelf", "Shore", "Island"))) %>%
+  ggplot() + geom_violin(aes(x=ir,y=score,fill=ir),draw_quantiles = 0.5) + theme_classic() + scale_y_continuous(limits=c(-5,5),breaks=c(-4,-2,0,2,4)) + theme(aspect.ratio=1/4,legend.position = "none")
+# ggsave("CpGscore_islands_violin.svg")
+```
+
+\subsection{Functional Regions (Fig 3C)}
+Table comparing different groups of CpG sites based on functional regions: 5'UTR, TSS1500, TSS200, 1stExon, Body, ExonBnd, 3'UTR
+```{r FunctionalRegions, echo=FALSE, warning=FALSE}
 # stats for functional regions
@@
          nsites=sum(indsin),
+         nout=sum(indsout),
          pval=wilcox.test(data$score_median[indsin],data$score_median[indsout])$p.value)
@@
 
-bind_rows({prepped_data %>% summarize_conservation("any") %>% mutate(region="Gene-Associated")},
+temp <- bind_rows({prepped_data %>% summarize_conservation("any") %>% mutate(region="Gene-Associated")},
           {prepped_data %>% summarize_conservation("TSS1500") %>% mutate(region="TSS1500")},
@@
           {prepped_data %>% summarize_conservation("1stExon") %>% mutate(region="1stExon")},
+          {prepped_data %>% summarize_conservation("ExonBnd") %>% mutate(region="ExonBnd")},
           {prepped_data %>% summarize_conservation("Body") %>% mutate(region="Body")},
           {prepped_data %>% summarize_conservation("3'UTR|3UTR") %>% mutate(region="3'UTR")}) %>%
-  select(c(region,cons_in,cons_out,nsites,pval)) %>%
-  gt() %>%
-  fmt_number(columns = vars(cons_in,cons_out,pval), decimals = 3)
+  select(c(region,cons_in,cons_out,nsites,nout,pval)) %>%
+  mutate(cons_in = cons_in/log(2), cons_out = cons_out/log(2)) %>% # change of base e to base 2
+  mutate(fold_change = 2^cons_in, fold_diff = 2^(cons_in - cons_out)) # calculating fold difference (2^ (log2 difference))
+temp %>% gt() %>% fmt_number(columns = vars(cons_in,cons_out,pval), decimals = 3)
 
-# new stats
+# multiple regression stats (using log2 score)
 FRdata = data.frame(CpGscore = log2(prepped_data$sigmaP / prepped_data$sigmaT) ) %>%
@@
          iExBnd = str_detect(FR,"ExonBnd"))
+
+# lm model
 fr.lm = lm(formula = CpGscore ~ iTSS200 + i5UTR + iBody + i3UTR + iTSS1500 + i1stEx + iExBnd, data = FRdata)
-#fr.lm = lm(formula = CpGscore ~ iTSS200 + i5UTR + iBody + i3UTR, data = FRdata)
 summary(fr.lm)
 
-library(rcompanion)
-fr.srh = scheirerRayHare(formula = CpGscore ~ iTSS200 + i5UTR + iBody + i3UTR + iTSS1500 + i1stEx + iExBnd, data = FRdata)
+# Fig 3C: plot functional region score distributions
+temp <- prepped_data %>% select(Name, UCSC_RefGene_Group, score_median) %>%
+  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
+  separate_rows(UCSC_RefGene_Group, sep=";") %>%
+  distinct() %>%
+  dplyr::rename(fr = UCSC_RefGene_Group, score=score_median) %>%
+  mutate(fr = case_when(fr=="" ~ "NA",
+                        str_detect(fr, "5URT") ~ "5'UTR",
+                        str_detect(fr, "3UTR") ~ "3'UTR",
+                        TRUE ~ fr)) %>%
+  distinct()
+#keepregions = c("TSS1500","Body", "3'UTR","1stExon","TSS200","5'UTR","ExonBnd")
+keepregions = c("TSS200","1stExon", "5'UTR","TSS1500","Body","3'UTR","ExonBnd")
+temp %>%
+  mutate(fr = factor(fr, levels = keepregions)) %>%
+  filter(fr %in% keepregions) %>%
+  ggplot() + geom_violin(aes(x=fr,y=score,fill=fr), draw_quantiles = 0.5) + theme_classic() + scale_y_continuous(limits=c(-5,5),breaks=c(-4,-2,0,2,4,6)) + theme(aspect.ratio=1/4,legend.position = "none") + geom_hline(yintercept = 0, color="black", alpha=0.2)
+# ggsave("CpGscore_regions_violin.svg")
+#rm(temp)
+
 ```
@@
 \newpage
-\subsection{Figure 1: CpG Distance Analysis}
+\subsection{Figure 3D: CpG Distance Analysis}
 Corrected for gene direction, based on Ensembl gene database \newline
-```{r Fig1, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=3, comment=NA}
+```{r Fig3D, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=3, comment=NA}
 # load ensembl db
@@
 distanced_data <- gened_data %>%
+  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
   filter(!is.na(UCSC_RefGene_Name)) %>%
@@
   filter(!is.na(UCSC_RefGene_Name)) %>%
-  select(CHR, UCSC_RefGene_Name, cpgdist, c(sigmaP:sigmaT,score_median:score_reg)) %>%
+  select(CHR, UCSC_RefGene_Name, cpgdist, c(score_median)) %>%
   mutate(location_bin = cut(cpgdist, c(seq(-1500, 20000, by = 100)), include.lowest = TRUE)) %>%
   group_by(location_bin) %>%
-  summarise(mean_sigmaP = mean(sigmaP), mean_sigmaPT = mean(sigmaPT), mean_sigmaT = mean(sigmaT),  mean_score_median = mean(score_median), mean_score_median2 = mean(score_median2), n_obs = n(), .groups='drop') %>%
+  summarise(mean_score_median = mean(score_median), n_obs = n(), .groups='drop') %>%
   mutate(location_bin = as.numeric(str_extract(location_bin, "(?<=,)(.*)(?=])")))
 
-distance_length %>%
-  drop_na() %>%
-  pivot_longer(mean_sigmaP:mean_sigmaT) %>% 
-  mutate(name = factor(name, c("mean_sigmaP","mean_sigmaPT","mean_sigmaT"), c("SigmaP","SigmaPT","SigmaT"))) %>%
-  sample_frac() %>%
-  ggplot(aes(location_bin, value, color = name, fill = name)) +
-  geom_point(key_glyph = draw_key_rect) +
-  ggforce::facet_zoom(location_bin < 5000) +
-  theme_light() +
-  theme(legend.position = "bottom") +
-  labs(title = NULL,
-       fill = NULL,
-       color = NULL,
-       x = "Distance from gene start site (binned to 100 bp)",
-       y = "Average Sigma Value", 
-       caption = "Figure 1: Average Sigma values of single CpGs as a function of position relative to gene.") +
-  scale_color_brewer(palette = "Set2") +
-  scale_fill_brewer(palette = "Set2")
-cat("\n")
-
-# again for conservation scores
+# plot for conservation scores
 distance_length %>%
@@
        y = "Average Conservation Score", 
-       caption = "Figure 1: Average conservation scores of single CpGs as a function of position relative to gene.") +
+       caption = "Figure 3D: Average conservation scores of single CpGs as a function of position relative to gene.") +
   scale_color_brewer(palette = "Set2") +
@@
 cat("\n")
-
-# and look at number of CpG in each bin
-distance_length %>%
-  drop_na() %>%
-  pivot_longer(n_obs) %>%
-  mutate(name = factor(name, c("n_obs"), c("n_obs"))) %>%
-  sample_frac() %>%
-  ggplot(aes(location_bin, value, color = name, fill = name)) +
-  geom_point(key_glyph = draw_key_rect) +
-  ggforce::facet_zoom(location_bin < 5000) +
-  theme_light() +
-  theme(legend.position = "bottom") +
-  labs(title = NULL,
-       fill = NULL,
-       color = NULL,
-       x = "Distance from gene start site (binned to 100 bp)",
-       y = "Number of CpGs", 
-       caption = "Figure 1: Frequency of single CpGs as a function of position relative to gene start site.") +
-  scale_color_brewer(palette = "Set2") +
-  scale_fill_brewer(palette = "Set2")
+# ggsave("CpGdist.svg")
 ```
@@
 Genes are scored by average of all CpG sites within gene
-\subsection{Example Genes: APC, TP53, TTN, HLA-A, GAPDH}
+\subsection{Fig 4A: Example Genes: TUBA1A, TTN, HLA-A}
 ```{r examplegenes, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=2, comment=NA}
@@
   CpGscores <- gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^",geneStr,"$"))) %>% pull(score_median)
-
-  ggplot(prepped_data) + geom_histogram(aes(x=score_median, y=..count../nrow(prepped_data)), binwidth=0.05,fill="lightgray") +
-    xlim(-3,3) + ggtitle(sprintf("%s: CpG scores (%i sites)",geneStr, length(CpGscores))) +
+  # convert log_e to log_2
+  ggplot(prepped_data %>% mutate(score_median = score_median/log(2))) + geom_histogram(aes(x=score_median, y=..count../nrow(prepped_data)), binwidth=0.05,fill="lightgray") + ggtitle(sprintf("%s: CpG scores (%i sites)",geneStr, length(CpGscores))) +
+    scale_x_continuous(limits=c(-5,5), breaks=c(-4,-2,0,2,4)) +
+    scale_y_continuous(limits=c(0,0.03),breaks=c(0,0.01,0.02,0.03)) +
     xlab("CpG conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") +
@@
 cat("\n")
-#plotGeneScore("TP53",0.5)
-plotGeneScore("TUBA1A",0.285)
+# Fig 4A: Example genes CpG site distributions
+plotGeneScore("TUBA1A",0.2)
 #ggsave("genescore_TUBA1A.svg")
 cat("\n")
-plotGeneScore("TTN",0.37)
+plotGeneScore("TTN",0.26)
 #ggsave("genescore_TTN.svg")
 cat("\n")
-plotGeneScore("HLA-A",0.26)
+plotGeneScore("HLA-A",0.18)
 #ggsave("genescore_HLA-A.svg")
@@
 # do one sample wilcoxon rank sum
+# TUBA1A
+gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^TUBA1A$"))) %>%
+  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
+  summarize(n = n(), mean=mean(score_median)) %>% mutate(fc = 2^mean)
 gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^TUBA1A$"))) %>% pull(score_median) %>% wilcox.test(mu = 0, alternative = "two.sided")
+# TTN
+gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^TTN$"))) %>%
+  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
+  summarize(n = n(), mean=mean(score_median)) %>% mutate(fc = 2^mean)
 gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^TTN$"))) %>% pull(score_median) %>% wilcox.test(mu = 0, alternative = "two.sided")
+# HLA-A
+gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^HLA-A$"))) %>%
+  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
+  summarize(n = n(), mean=mean(score_median)) %>% mutate(fc = 2^mean)
 gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^HLA-A$"))) %>% pull(score_median) %>% wilcox.test(mu = 0, alternative = "two.sided")
@@
 
-\subsection{COSMIC Cancer Genes}
+\subsection{CRC Cancer Genes Databases}
 ```{r cancergenes, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=2, comment=NA}
-# make gene-wise averages
-gene_summarize <- function(data, id, value, fun = mean, ...) {
-  grouped_data <- data %>% group_by({{id}})
 
-  bind_cols(
-    summarise_at(grouped_data, dplyr::vars({{value}}), fun, ...),
-    count(grouped_data) %>% dplyr::ungroup() %>% dplyr::select(.data$n)
-  )
-}
-
-gene_summary <- gened_data %>%
-  filter(!is.na(UCSC_RefGene_Name)) %>%
-  gene_summarize(UCSC_RefGene_Name, c(score_median))
-
+# CRC gene database 1: COSMIC
 cancergenes1 <- read_csv("/Users/kevinmurgas/Documents/Data+ project/Cancer gene lists/SangerCensus_COSMIC_CRC.csv")
-genescores <- gene_summary %>% filter(UCSC_RefGene_Name %in% (cancergenes1 %>% pull(`Gene Symbol`))) %>% pull(score_median)
-ggplot(gene_summary) + geom_histogram(aes(x=score_median, y=..count../nrow(gene_summary)), binwidth=0.05) + xlim(-3,3) + ggtitle(sprintf("Gene Scores of %i CRC Genes", length(genescores))) + xlab("Gene conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") + geom_vline(xintercept=genescores, color="blue", alpha=0.5, size=0.2)
-cat("\n")
-
-# source 2
+# CRC gene database 2: AGO
 cancergenes2 <- read_csv("/Users/kevinmurgas/Documents/Data+ project/Cancer gene lists/CRCgenes_atlasgeneticsoncology.csv")
-genescores <- gene_summary %>% filter(UCSC_RefGene_Name %in% (cancergenes2 %>% pull(CancerGenes))) %>% pull(score_median)
-ggplot(gene_summary) + geom_histogram(aes(x=score_median, y=..count../nrow(gene_summary)), binwidth=0.05) + xlim(-3,3) + ggtitle(sprintf("Gene Scores of %i CRC Genes", length(genescores))) + xlab("Gene conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") + geom_vline(xintercept=genescores, color="blue", alpha=0.1, size=0.2)
 
 # compute % overlap of database genes and significant genes (By adj bootstrap)
-n1 = sum(permed_data %>% filter(score_median_v3>0.95) %>% pull(UCSC_RefGene_Name) %in% (cancergenes1 %>% pull(`Gene Symbol`)))
-n2 = sum(permed_data %>% filter(score_median_v3>0.95) %>% pull(UCSC_RefGene_Name) %in% (cancergenes2 %>% pull(CancerGenes)))
+# n1 = sum(permed_data %>% filter(score_median_v3>0.95) %>% pull(UCSC_RefGene_Name) %in% (cancergenes1 %>% pull(`Gene Symbol`)))
+# n2 = sum(permed_data %>% filter(score_median_v3>0.95) %>% pull(UCSC_RefGene_Name) %in% (cancergenes2 %>% pull(CancerGenes)))
 
 # plot violins of: all genes, COSMIC, AGO
-perm_temp <- permed_data %>% select(c(UCSC_RefGene_Name, score_median,score_median_v3,n))
-temp <- rbind(perm_temp %>% mutate(group="all"),
+perm_temp <- permed_data %>% select(c(UCSC_RefGene_Name, score_median,score_median_v3,n)) %>%
+  mutate(score_median2 = score_median/log(2)) # change of base from e to 2
+genes_dbs <- rbind(perm_temp %>% mutate(group="all"),
               perm_temp %>% filter(UCSC_RefGene_Name %in% (cancergenes1 %>% pull(`Gene Symbol`))) %>% mutate(group="COSMIC"),
@@
               )
-temp %>% mutate(group = factor(group, levels = c("COSMIC", "AGO", "all"))) %>%
-  ggplot() + geom_violin(aes(y=score_median,x=group,fill=group), draw_quantiles = 0.5) + scale_y_continuous(limits=c(-1,2),breaks = c(-1,0,1,2)) + theme_classic() + xlab("Database") + ylab("Gene Conservation Score")
+# Fig 4B: plot violins (with log2 base change)
+pDB <- genes_dbs %>% mutate(group = factor(group, levels = c("all", "AGO", "COSMIC"))) %>%
+  ggplot() + geom_violin(aes(x=score_median2,y=group,fill=group), draw_quantiles = 0.5) + scale_x_continuous(limits=c(-2,4),breaks = c(-2,0,2,4)) + theme_classic() + ylab("Database") + xlab("Gene Conservation Score") + geom_vline(xintercept=0, color="black", alpha=0.2) + theme(aspect.ratio=60/156.3,legend.position = "none")
+# add hashmarks for each gene (doesn't look great)
+marks <- genes_dbs %>% mutate(x=score_median2, y=if_else(group=="COSMIC",3,if_else(group=="AGO",2,1))) %>% select(x,y)
+marks2 <- genes_dbs %>% filter(score_median_v3 > 0.95) %>% mutate(x=score_median2, y=if_else(group=="COSMIC",3,if_else(group=="AGO",2,1))) %>% select(x,y)
+pDB + geom_segment(data=marks2, aes(x=x, xend=x, y=y, yend=y+0.2), color="red") + geom_segment(data=marks, aes(x=x, xend=x, y=y, yend=y+0.1), color="black", alpha=0.5)
 # ggsave("database_violins.svg")
 
-# try with histogram
-plotDBScore <- function(geneStr,norm) {
-  CpGscores <- gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^",geneStr,"$"))) %>% pull(score_median)
-
-  ggplot(prepped_data) + geom_histogram(aes(x=score_median, y=..count../nrow(prepped_data)), binwidth=0.05,fill="lightgray") +
-    xlim(-3,3) + ggtitle(sprintf("%s: CpG scores (%i sites)",geneStr, length(CpGscores))) +
-    xlab("CpG conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") +
-    #geom_vline(xintercept=CpGscores, color="red", alpha=0.5, size=0.2) +
-    geom_histogram(data=data.frame(score=CpGscores),aes(x=score,y=norm*..count../length(CpGscores)),binwidth=0.05, fill="blue",alpha=0.5) +
-    geom_vline(xintercept=mean(CpGscores), color="blue",linetype="dashed", size=1)
-}
-
-```
-
-\newpage
-\section{Gene Set Conservation Analysis}
-```{r load_gene_sets, include=FALSE}
-# load in the C5 gene sets from MSigDB
-library(GSA)
-gmtfile <- file.path("/Users/kevinmurgas/Documents/Data+ project/EPIC data","c2.cp.v7.1.symbols.gmt")
-genesets <- GSA.read.gmt(gmtfile)
-```
-
-```{r GSCA, echo=FALSE, comment=NA, fig.width=8}
-# make a data.frame for gene sets and scores
-nsets <- length(genesets[[1]])
-GSEA_df <- data.frame(GeneSet=genesets[[2]][1:nsets], score=numeric(nsets),
-                      num_gene = unlist(lapply(genesets[[1]][1:nsets],length)),
-                      num_gene_cut = numeric(nsets))
-
-# have to order all genes first by rank of conservation (conserved genes come first)
-genenames = permed_data$UCSC_RefGene_Name
-ngene = length(genenames)
-
-# reorder by evidence rank
-ind = order(-permed_data$score_median) # use ranked conservation score as evidence
-evidence = seq(ngene,1,-1)
-genenames = genenames[ind]
-
-# plot evidence ranked: permed_data$score_median
-ranks <- permed_data$score_median
-names(ranks) <- permed_data$UCSC_RefGene_Name
-barplot(sort(ranks, decreasing = T))
-
-ptm <- proc.time()
-for (iset in 1:nsets) {
-  # take only genes that we have in full gene list
-  set <- unlist(genesets[[1]][iset]) %>% intersect(genenames)
-  GSEA_df$num_gene_cut[iset] = length(set)
-
-  # skip if num_gene <10 or >500
-  if (length(set)<10 | length(set)>500) {next}
-
-  # compute running sum (as PDF first)
-  runsum_in = numeric(ngene)
-  runsum_out = numeric(ngene)
-  for (i in 1:ngene) {
-    if (genenames[i] %in% set) {
-      # evidence = permed_data$median, or ranks
-      runsum_in[i] = evidence[i]
-      runsum_out[i] = 0
-    } else {
-      runsum_in[i] = 0
-      runsum_out[i] = 1
-    }
-  }
-  # convert to CDFs
-  runsum_in <- cumsum(runsum_in)/sum(runsum_in)
-  runsum_out <- cumsum(runsum_out)/sum(runsum_out)
-
-  # compute K-S-like statistic
-  v <- runsum_in - runsum_out
-  score <- max(c(0,v)) + min(c(0,v))
-  GSEA_df$score[iset] <- score
-}
-cat(proc.time()-ptm)
-cat(sprintf("Processed %i genes, %i pathways. Top 50 paths:",ngene,nsets))
-
-# insert significance bootstrapping
-
-# make table of top scores
-GSEA_df %>% arrange(-score) %>%
-  select(score,num_gene,GeneSet) %>%
-  slice(1:50) %>%
-  mutate(GeneSet = gsub('(.{1,90})(\\s|$)', '\\1\n', GeneSet)) %>%
-  gt()
+# stats
+wilcox.test(x = genes_dbs$score_median2[genes_dbs$group=="COSMIC"], y = genes_dbs$score_median2[genes_dbs$group=="all"], alternative = "greater")
+wilcox.test(x = genes_dbs$score_median2[genes_dbs$group=="AGO"], y = genes_dbs$score_median2[genes_dbs$group=="all"], alternative = "greater")
 ```
@@
 
-``` {r enrich, include=FALSE}
-# expression data
-# downloaded file: E-MTAB-2836-query-results.tpms.tsv
-# source: https://www.ebi.ac.uk/gxa/experiments/E-MTAB-2836/Results
-expression_atlas <- read_tsv(here("data-raw", "E-MTAB-2836-query-results.tpms.tsv"), skip = 4, 
-                             col_types = cols(
-                               .default = col_double(),
-                               `Gene ID` = col_character(),
-                               `Gene Name` = col_character()
-                             ))
+<!-- \newpage -->
+<!-- \section{Gene Set Conservation Analysis} -->
+<!-- Based on Gene Set Enrichment Analysis (GSEA) -->
+<!-- ```{r load_gene_sets, include=FALSE} -->
+<!-- # load in the C5 gene sets from MSigDB -->
+<!-- library(GSA) -->
+<!-- gmtfile <- file.path("/Users/kevinmurgas/Documents/Data+ project/EPIC data","c2.cp.v7.1.symbols.gmt") -->
+<!-- genesets <- GSA.read.gmt(gmtfile) -->
+<!-- ``` -->
 
-# Gene Set Enrichment
-tidy_gsea <- function(genes) {
-  AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, genes, 'ENTREZID', 'SYMBOL') %>%
-    ReactomePA::enrichPathway()
-}
+<!-- ```{r GSCA, echo=FALSE, comment=NA, fig.width=8} -->
+<!-- # make a data.frame for gene sets and scores -->
+<!-- nsets <- length(genesets[[1]]) -->
+<!-- GSEA_df <- data.frame(GeneSet=genesets[[2]][1:nsets], score=numeric(nsets), -->
+<!--                       num_gene = unlist(lapply(genesets[[1]][1:nsets],length)), -->
+<!--                       num_gene_cut = numeric(nsets)) -->
 
-enrich_sigmaP <- permed_data %>%
-  #filter(sigmaP_v3 < quantile(sigmaP_v3, 0.05, na.rm = TRUE)) %>%
-  filter(sigmaP_v3 < 0.05) %>%
-  pull(UCSC_RefGene_Name) %>% tidy_gsea()
-enrich_sigmaPT <- permed_data %>%
-  #filter(sigmaPT_v3 < quantile(sigmaPT_v3, 0.05, na.rm = TRUE)) %>%
-  filter(sigmaPT_v3 < 0.05) %>%
-  pull(UCSC_RefGene_Name) %>% tidy_gsea()
-enrich_sigmaT <- permed_data %>%
-  #filter(sigmaT_v3 < quantile(sigmaT_v3, 0.05, na.rm = TRUE)) %>%
-  filter(sigmaT_v3 < 0.05) %>%
-  pull(UCSC_RefGene_Name) %>% tidy_gsea()
-enrich_score_median <- permed_data %>%
-  #filter(score_median_v3 > quantile(score_median_v3, 0.95, na.rm = TRUE)) %>%
-  filter(score_median_v3 > 0.95) %>%
-  pull(UCSC_RefGene_Name) %>% tidy_gsea()
-```
+<!-- # have to order all genes first by rank of conservation (conserved genes come first) -->
+<!-- genenames = permed_data$UCSC_RefGene_Name -->
+<!-- ngene = length(genenames) -->
+
+<!-- # reorder by evidence rank -->
+<!-- ind = order(-permed_data$score_median) # use ranked conservation score as evidence -->
+<!-- evidence = seq(ngene,1,-1) -->
+<!-- genenames = genenames[ind] -->
+
+<!-- # plot evidence ranked: permed_data$score_median -->
+<!-- ranks <- permed_data$score_median -->
+<!-- names(ranks) <- permed_data$UCSC_RefGene_Name -->
+<!-- barplot(sort(ranks, decreasing = T)) -->
+
+<!-- ptm <- proc.time() -->
+<!-- for (iset in 1:nsets) { -->
+<!--   # take only genes that we have in full gene list -->
+<!--   set <- unlist(genesets[[1]][iset]) %>% intersect(genenames) -->
+<!--   GSEA_df$num_gene_cut[iset] = length(set) -->
+
+<!--   # skip if num_gene <10 or >500 -->
+<!--   if (length(set)<10 | length(set)>500) {next} -->
+
+<!--   # compute running sum (as PDF first) -->
+<!--   runsum_in = numeric(ngene) -->
+<!--   runsum_out = numeric(ngene) -->
+<!--   for (i in 1:ngene) { -->
+<!--     if (genenames[i] %in% set) { -->
+<!--       # evidence = permed_data$median, or ranks -->
+<!--       runsum_in[i] = evidence[i] -->
+<!--       runsum_out[i] = 0 -->
+<!--     } else { -->
+<!--       runsum_in[i] = 0 -->
+<!--       runsum_out[i] = 1 -->
+<!--     } -->
+<!--   } -->
+<!--   # convert to CDFs -->
+<!--   runsum_in <- cumsum(runsum_in)/sum(runsum_in) -->
+<!--   runsum_out <- cumsum(runsum_out)/sum(runsum_out) -->
+
+<!--   # compute K-S-like statistic -->
+<!--   v <- runsum_in - runsum_out -->
+<!--   score <- max(c(0,v)) + min(c(0,v)) -->
+<!--   GSEA_df$score[iset] <- score -->
+<!-- } -->
+<!-- cat(proc.time()-ptm) -->
+<!-- cat(sprintf("Processed %i genes, %i pathways. Top 50 paths:",ngene,nsets)) -->
+
+<!-- # insert significance bootstrapping -->
+
+<!-- # make table of top scores -->
+<!-- GSEA_df %>% arrange(-score) %>% -->
+<!--   select(score,num_gene,GeneSet) %>% -->
+<!--   slice(1:50) %>% -->
+<!--   mutate(GeneSet = gsub('(.{1,90})(\\s|$)', '\\1\n', GeneSet)) %>% -->
+<!--   gt() -->
+<!-- ``` -->
 
@@
 ```
-
-\subsection{Figure 2: Bootstrapping}
-```{r Fig2, echo=FALSE, warning=FALSE, fig.height=4, comment=NA}
-# Figure 2
-permed_data %>%
-  pivot_longer(dplyr::matches("1|3")) %>%
-  separate(name, into = c("type", "method"), sep = "_v") %>%
-  mutate(method = factor(method, c("1", "3"), c("Naive bootstrap", "Adjusted bootstrap")),
-         type = factor(type, c("sigmaP","sigmaPT","sigmaT"), c("SigmaP","SigmaPT","SigmaT"))) %>%
-  drop_na() %>%
-  ggplot(aes(value, fill = type)) +
-  geom_histogram(bins = 50, color = "grey30") +
-  facet_grid(method~type, scales = "free") +
-  theme_minimal() +
-  labs(x = "p-value",
-       y = "Number of genes",
-       title = NULL,
-       caption = "Figure 2A: Distribution of boot-strapping p-values for genes as a function of sigma level and bootstrap method.")  +
-  theme(legend.position = "none") +
-  scale_color_brewer(palette = "Set2") +
-  scale_fill_brewer(palette = "Set2")
-
-cat("***")
-permed_data %>%
-  pivot_longer(dplyr::matches("1|3")) %>%
-  separate(name, into = c("type", "method"), sep = "_v") %>%
-  mutate(method = factor(method, c("1", "3"), c("Naive bootstrap", "Adjusted bootstrap")),
-         type = factor(type, c("score_median"), c("Score_median"))) %>%
-  drop_na() %>%
-  ggplot(aes(value, fill = type)) +
-  geom_histogram(bins = 50, color = "grey30") +
-  facet_grid(method~type, scales = "free") +
-  theme_minimal() +
-  labs(x = "p-value",
-       y = "Number of genes",
-       title = NULL,
-       caption = "Figure 2B: Distribution of boot-strapping p-values for genes as a function of conservation score and bootstrap method.")  +
-  theme(legend.position = "none") +
-  scale_color_brewer(palette = "Set2") +
-  scale_fill_brewer(palette = "Set2")
-```
-
-\newpage
-\subsection{Figure 3: Reactome Pathway Enrichment Analysis}
-Enrichment defined from the group of genes <5th percentile of each sigma, or >95th percentile of conservation score.
-```{r Fig3, echo=FALSE, fig.height=3.5, comment=NA}
-cat("Number of enrichment pathways (sigmas):")
-c(sigmaP = length(enrich_sigmaP$Description),
-  sigmaPT = length(enrich_sigmaPT$Description),
-  sigmaT = length(enrich_sigmaT$Description)) %>%
-  sort()
-
-list(
-  sigmaP = enrich_sigmaP$Description,
-  sigmaPT = enrich_sigmaPT$Description,
-  sigmaT = enrich_sigmaT$Description
-) %>%
-  UpSetR::fromList() %>%
-  UpSetR::upset(sets.bar.color = RColorBrewer::brewer.pal(3, "Set2")[c(3, 2, 1)])
-grid::grid.text("Figure 3A: Frequency of overlap between pathways that are called \nas most conserved for each hierarchical variance.", x=0.65, y=0.05, gp=grid::gpar(fontsize=10))
-
-cat("Number of enrichment pathways (probs):")
-c(score_median = length(enrich_score_median$Description)) %>%
-  sort()
-
-list(
-  score_median = enrich_score_median$Description
-) %>%
-  UpSetR::fromList() %>%
-  UpSetR::upset(sets.bar.color = RColorBrewer::brewer.pal(3, "Set2")[c( 1)])
-grid::grid.text("Figure 3B: Frequency of overlap between pathways that are called \nas most conserved for each conservation score.", x=0.65, y=0.05, gp=grid::gpar(fontsize=10))
-```
-
-\newpage
-\subsection{Figure 4: Pathway Enrichment Network}
-```{r Fig4, echo=FALSE, message=FALSE, fig.height=2.5}
-# Figure 4
-cat("sigmaP:")
-ggg <- enrich_sigmaP %>%
-  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
-  labs(caption = "Figure 4A: Relationship between pathways that are called as conserved based on sigmaP.", title = NULL)
-
-ggg$layers <- ggg$layers[1:2]
-
-ggg + coord_flip() + 
-  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
-  scale_color_gradientn(colors = "black") +
-  guides(color = "none", size = "none")
-
-#ggsave(filename = here("figures", "figure-4.png"), width = 7, dpi = 320, height = 4.326)
-
-# sigmaPT
-cat("sigmaPT:")
-ggg <- enrich_sigmaPT %>%
-  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
-  labs(caption = "Figure 4B: Relationship between pathways that are called as conserved based on sigmaPT.", title = NULL)
-
-ggg$layers <- ggg$layers[1:2]
-ggg + coord_flip() + 
-  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
-  scale_color_gradientn(colors = "black") +
-  guides(color = "none", size = "none")
-
-# sigmaT
-cat("sigmaT:")
-ggg <- enrich_sigmaT %>%
-  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
-  labs(caption = "Figure 4C: Relationship between pathways that are called as conserved based on sigmaT.", title = NULL)
-
-ggg$layers <- ggg$layers[1:2]
-ggg + coord_flip() + 
-  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
-  scale_color_gradientn(colors = "black") +
-  guides(color = "none", size = "none")
-```
-
-\newpage
-```{r Fig4DEF, echo=FALSE, message=FALSE, fig.height=2.5}
-# score_median
-cat("score_median")
-ggg <- enrich_score_median %>%
-  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
-  labs(caption = "Figure 4D: Relationship between pathways that are called as conserved based on score_median.", title = NULL)
-ggg$layers <- ggg$layers[1:2]
-ggg + coord_flip() + 
-  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
-  scale_color_gradientn(colors = "black") +
-  guides(color = "none", size = "none")
-
-```
-
-<!-- \newpage -->
-<!-- \section{Patient-specific analysis} -->
-<!-- Basic overview: using a specified number of genes (here 100), take top genes (here using score_prob bootstrapping percentiles) to select a "gene signature" of sorts.
-Then for each patient, calculate the mean pair-wise distance within the individual patients' tumor samples.
-Method 1 is these raw values, method 2 normalizes by the mean PWD of all normal samples.
-Method 3 just examines the overall mean PWD for all tumor samples, normalized by mean PWD of all normal.
--->
-
-<!-- ```{r PatientPWD, echo=FALSE, message=FALSE, warning=FALSE, comment=NA, fig.height=9, fig.width=7} -->
-<!-- library(gtools) -->
-
-<!-- # pull out all samples beta values -->
-<!-- all_beta <- FullAnnotation[,10:57] -->
-
-<!-- # extract patient and tissue labels -->
-<!-- patientLabel <- substr(colnames(all_beta), 1, 1) -->
-<!-- patientLabel[10:12] <- "K*" -->
-<!-- tissueLabel <- substr(colnames(all_beta), 2, 2) -->
-<!-- tissueLabel[!(tissueLabel %in% "N")] <- "T"   #replace A/B/D/M with T -->
-
-<!-- pats = patientLabel %>% sort %>% unique -->
-<!-- npat = length(pats) -->
-
-<!-- enrichedGenes <- permed_data %>% filter(score_prob_v3 < quantile(score_prob_v3, 0.05, na.rm = TRUE)) %>% -->
-<!--   select(UCSC_RefGene_Name, score_prob, score_prob_v3, n) %>% -->
-<!--   rename(Gene=UCSC_RefGene_Name, Pctile=score_prob_v3) -->
-
-<!-- #ngene = 70 # using top N genes -->
-<!-- #genes = enrichedGenes %>% top_n(-ngene,Pctile) %>% pull(Gene) -->
-<!-- #genes = genes[1:ngene] -->
-
-<!-- genes = enrichedGenes$Gene -->
-<!-- ngene = length(genes) -->
-
-<!-- pwd <- function(x) { -->
-<!--   mean(dist(x)) -->
-<!-- } -->
-<!-- pwd2 <- function(x) { -->
-<!--   abs(diff(x)) -->
-<!-- } -->
-
-<!-- signature_mat1 = matrix(,ngene,npat) # raw tumor pwd -->
-<!-- signature_mat2 = matrix(,ngene,npat) # tumor pwd normalized by all normal pwd -->
-<!-- for (igene in 1:ngene) { -->
-<!--   # each gene, get gene inds, calculate PWD for the tumor samples -->
-<!--   geneind = grep(genes[igene], FullAnnotation$UCSC_RefGene_Name) -->
-
-<!--   # calculate PWD of each tumor -->
-<!--   tum_pwd <- numeric(npat) -->
-<!--   for (ipat in 1:npat) { -->
-<!--     # each patient, get tumor sample inds -->
-<!--     patind <- which(patientLabel == pats[ipat] & tissueLabel == "T") -->
-<!--     tum_pwd[ipat] <- all_beta[geneind,patind] %>% logit %>% apply(1,pwd2) %>% mean -->
-<!--   } -->
-
-<!--   # calculate PWD of all normal -->
-<!--   norm_pwd <- all_beta[geneind,tissueLabel == "N"] %>% logit %>% apply(1, pwd) %>% mean -->
-
-<!--   signature_mat1[igene,] = tum_pwd -->
-<!--   signature_mat2[igene,] = tum_pwd/norm_pwd -->
-<!-- } -->
-
-<!-- #cat("  \n\\newline  \nMethod 1 - raw intratumor PWD within each patient") -->
-<!-- #signature_mat1 %>% as_tibble %>% `colnames<-`(pats) %>% gather %>% -->
-<!-- #  mutate(gene=rep(genes,npat)) %>% rename(Patient=key,PWD=value) %>% -->
-<!-- #  ggplot + geom_tile(aes(y=gene,x=Patient,fill=PWD)) + ylab(paste0(ngene," genes")) + -->
-<!-- #  theme(axis.text.y = element_blank()) -->
-<!-- #cat("  \n\\newline  \nMethod 2 - intratumor PWD normalized by all normal PWD") -->
-<!-- #signature_mat2 %>% as_tibble %>% `colnames<-`(pats) %>% gather %>% -->
-<!-- #  mutate(gene=rep(genes,npat)) %>% rename(Patient=key,PWD=value) %>% -->
-<!-- #  ggplot + geom_tile(aes(y=gene,x=Patient,fill=PWD)) + ylab(paste0(ngene," genes")) + -->
-<!-- #  theme(axis.text.y = element_blank()) -->
-<!-- #cat("  \n\\newline  \nMethod 3 - all tumor PWD normalized by all normal PWD") -->
-<!-- #signature_mat3 %>% as_tibble %>% mutate(gene=genes) %>% rename(PWD=V1) %>% -->
-<!-- #  ggplot + geom_tile(aes(y=gene,x=1,fill=PWD)) + ylab(paste0(ngene," genes")) + -->
-<!-- #  theme(axis.text.y = element_blank()) -->
-
-<!-- library(pheatmap) -->
-<!-- library(dendsort) -->
-
-<!-- cat("  \n\\newline  \nMethod 1 - raw intratumor PWD within each patient") -->
-<!-- df1 = as.data.frame(signature_mat1) -->
-<!-- rownames(df1) <- genes -->
-<!-- colnames(df1) <- pats -->
-<!-- sort_hclust <- function(...) as.hclust(dendsort(as.dendrogram(...))) -->
-<!-- mat_cluster_cols <- sort_hclust(hclust(dist(t(df1)))) -->
-<!-- mat_cluster_rows <- sort_hclust(hclust(dist(df1))) -->
-<!-- pheatmap( -->
-<!--   mat=df1, -->
-<!--   cluster_cols = mat_cluster_cols, -->
-<!--   cluster_rows = mat_cluster_rows, -->
-<!--   show_rownames = FALSE, -->
-<!--   show_colnames = TRUE, -->
-<!--   main = "Top 1118 Genes: Tumor PWD" -->
-<!-- ) -->
-
-<!-- cat("  \n\\newline  \nMethod 2 - intratumor PWD normalized by all normal PWD") -->
-<!-- df1 = as.data.frame(signature_mat2) -->
-<!-- rownames(df1) <- genes -->
-<!-- colnames(df1) <- pats -->
-<!-- sort_hclust <- function(...) as.hclust(dendsort(as.dendrogram(...))) -->
-<!-- mat_cluster_cols <- sort_hclust(hclust(dist(t(df1)))) -->
-<!-- mat_cluster_rows <- sort_hclust(hclust(dist(df1))) -->
-<!-- pheatmap( -->
-<!--   mat=df1, -->
-<!--   cluster_cols = mat_cluster_cols, -->
-<!--   cluster_rows = mat_cluster_rows, -->
-<!--   show_rownames = FALSE, -->
-<!--   show_colnames = TRUE, -->
-<!--   main = "Top 1118 Genes: Tumor PWD" -->
-<!-- ) -->
-<!-- ``` -->
-
-\newpage
-\section{Appendix: Enrichment Gene/Pathway Lists}
-\subsection{A1: SigmaP}
-Genes are selected with cutoff <5th percentile (adjusted bootstrap), showing top 30 genes/pathways
-```{r App_sigmaP, echo=FALSE, fig.fullwidth=TRUE, comment=NA, message=FALSE}
-enrichedGenes <- permed_data %>%
-  select(UCSC_RefGene_Name, sigmaP, sigmaP_v3, n) %>%
-  rename(Gene=UCSC_RefGene_Name, Pctile=sigmaP_v3) %>%
-  filter(Pctile < quantile(Pctile, 0.05, na.rm = TRUE)) %>%
-  arrange(Pctile)
-
-enrichedPaths <- tibble(Pathway = enrich_sigmaP$Description, ngenes = enrich_sigmaP$Count, padj = enrich_sigmaP$p.adjust) %>% arrange(padj)
-
-ngene <- dim(enrichedGenes)[1]
-npath <- length(enrich_sigmaP$Description)
-
-cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath))
-
-nshow = 30
-if (ngene > nshow) {
-  enrichedGenes <- enrichedGenes[1:nshow,]
-}
-if (npath > nshow) {
-  enrichedPaths <- enrichedPaths[1:nshow,]
-}
-
-library(pander)
-enrichedGenes %>% pander()
-
-cat("SigmaP top 30 enriched paths:  \n")
-enrichedPaths %>% pander()
-```
-
-\newpage
-\subsection{A2: SigmaPT}
-Genes are selected with cutoff <5th percentile (adjusted bootstrap), showing top 30 genes/pathways
-```{r App_sigmaPT, echo=FALSE, comment=NA, message=FALSE}
-enrichedGenes <- permed_data %>%
-  select(UCSC_RefGene_Name, sigmaPT, sigmaPT_v3, n) %>%
-  rename(Gene=UCSC_RefGene_Name, Pctile=sigmaPT_v3) %>%
-  filter(Pctile < quantile(Pctile, 0.05, na.rm = TRUE)) %>%
-  arrange(Pctile)
-
-enrichedPaths <- tibble(Pathway = enrich_sigmaPT$Description, ngenes = enrich_sigmaPT$Count, padj = enrich_sigmaPT$p.adjust) %>% arrange(padj)
-
-ngene <- dim(enrichedGenes)[1]
-npath <- length(enrich_sigmaPT$Description)
-
-cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath))
-
-nshow = 30
-if (ngene > nshow) {
-  enrichedGenes <- enrichedGenes[1:nshow,]
-}
-if (npath > nshow) {
-  enrichedPaths <- enrichedPaths[1:nshow,]
-}
-
-enrichedGenes %>% pander()
-
-cat("\nSigmaPT top 30 enriched paths:  \n")
-enrichedPaths %>% pander()
-```
-
-\newpage
-\subsection{A3: SigmaT}
-Genes are selected with cutoff <5th percentile (adjusted bootstrap), showing top 30 genes/pathways
-```{r App_sigmaT, echo=FALSE, comment=NA, message=FALSE}
-enrichedGenes <- permed_data %>%
-  select(UCSC_RefGene_Name, sigmaT, sigmaT_v3, n) %>%
-  rename(Gene=UCSC_RefGene_Name, Pctile=sigmaT_v3) %>%
-  filter(Pctile < quantile(Pctile, 0.05, na.rm = TRUE)) %>%
-  arrange(Pctile)
-
-enrichedPaths <- tibble(Pathway = enrich_sigmaT$Description, ngenes = enrich_sigmaT$Count, padj = enrich_sigmaT$p.adjust) %>% arrange(padj)
-
-ngene <- dim(enrichedGenes)[1]
-npath <- length(enrich_sigmaT$Description)
-
-cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath))
-
-nshow = 30
-if (ngene > nshow) {
-  enrichedGenes <- enrichedGenes[1:nshow,]
-}
-if (npath > nshow) {
-  enrichedPaths <- enrichedPaths[1:nshow,]
-}
-
-enrichedGenes %>% pander()
-
-cat("SigmaT top 30 enriched paths:  \n")
-enrichedPaths %>% pander()
-```
-
-\newpage
-\subsection{A4: Score Median}
-Genes are selected with cutoff >95th percentile (adjusted bootstrap), showing top 50 genes/pathways
-```{r App_score_median, echo=FALSE, comment=NA, message=FALSE}
-enrichedGenes <- permed_data %>% 
-  select(UCSC_RefGene_Name, score_median, score_median_v3, n) %>%
-  rename(Gene=UCSC_RefGene_Name, Score_median=score_median, Pctile=score_median_v3) %>%
-  filter(Pctile > quantile(Pctile, 0.95, na.rm = TRUE)) %>%
-  arrange(Pctile)
-
-enrichedPaths <- tibble(Pathway = enrich_score_median$Description, ngenes = enrich_score_median$Count, padj = enrich_score_median$p.adjust) %>% arrange(padj)
-
-ngene <- dim(enrichedGenes)[1]
-npath <- length(enrich_score_median$Description)
-
-cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath))
-
-nshow = 50
-if (ngene > nshow) {
-  enrichedGenes <- enrichedGenes[1:nshow,]
-}
-if (npath > nshow) {
-  enrichedPaths <- enrichedPaths[1:nshow,]
-}
-
-enrichedGenes %>% pander()
-
-cat("Score_median top 50 enriched paths:  \n")
-enrichedPaths %>% pander()
-```

### 2021-01-05 GetPosteriorsSingleSites.R

@@ -25,6 +25,7 @@ load("myFA_bulkonly.Rdata")
 # randomly select 5 sites in 1-866091
 #siteInds <- sample(1:866091,5)
 siteInds <- c(525490, 570532, 827025, 287316, 854143)
+siteInds <- 124022
 nsites <- length(siteInds)
 
 
@@ -46,13 +47,13 @@ site <- function (ind) {
   
   work <-
     data.frame(
-      beta = logit(t(temp[indices])),
+      M = logit(t(temp[indices])),
       patient = patientLabel,
       tissue = tissueLabel,
       side = sideLabel,
       tInd = tumorIndicator
     )
-  colnames(work)[1] <- "beta"
+  colnames(work)[1] <- "M"
   return(work)
 }
 
@@ -61,11 +62,11 @@ stanfit_model <- function (dataset) {
   stanDat <- list(
     N = nrow(dataset),
     P = nlevels(dataset$patient),
-    pID = as.integer(factor(dataset$patient)),
+    y = dataset[, 1],
     tInd = dataset$tInd,
-    y = dataset[, 1]
+    pID = as.integer(factor(dataset$patient))
   )
-  stanFit <-
+  stanFit <- 
     stan(
       fit = emptyFit,
       warmup = 200,
@@ -76,7 +77,7 @@ stanfit_model <- function (dataset) {
       refresh = 0,
       seed=123
     )
-  return(stanFit = stanFit)
+  return(stanFit=stanFit)
 }
 
 # Plot raw data and fits
@@ -124,7 +125,7 @@ postplot <- function(stanFit) {
 ### CODE ###
 
 # check for model_TCGApriors.rds, must add to prevent crash
-checkFile <- "model_TCGApriors2.rds"
+checkFile <- "model_TCGApriors.rds"
 if (file.exists(checkFile)) {
   file.remove(checkFile)
   print("caught one!") 
@@ -139,7 +140,7 @@ stanDat1 <- list(
   P = nlevels(data1$patient),
   y = data1[, 1]
 )
-emptyFit <- stan(file="model_TCGApriors2.stan", data = stanDat1, chains = 0)
+emptyFit <- stan(file="model_TCGApriors.stan", data = stanDat1, chains = 0)
 
 # # run in parallel via doParallel
[…]

### 2021-01-05 StanCParallel.R

@@ -53,7 +53,7 @@ site <- function (ind) {
   
   work <-
     data.frame(
-      beta = logit(t(temp[indices])),
+      M = logit(t(temp[indices])),
       patient = patientLabel,
       tissue = tissueLabel,
       side = sideLabel,
@@ -66,11 +66,11 @@ site <- function (ind) {
 # Using Model with TCGA priors (previously model 3)
 stanfit_model <- function (dataset) {
   stanDat <- list(
-    pID = as.integer(factor(dataset$patient)),
-    tInd = dataset$tInd,
     N = nrow(dataset),
     P = nlevels(dataset$patient),
-    y = dataset[, 1]
+    y = dataset[, 1],
+    tInd = dataset$tInd,
+    pID = as.integer(factor(dataset$patient))
   )
   stanFit <- 
     stan(
@@ -104,7 +104,7 @@ mu_C <-
     n_eff = numeric(nsites),
     Rhat = numeric(nsites)
   )
-betaT_C <- mu_C
+nu_C <- mu_C
 sigmaP_C <- mu_C
 sigmaPT_C <- mu_C
 sigmaT_C <- mu_C
@@ -123,7 +123,7 @@ stanDat1 <- list(
   P = nlevels(data1$patient),
   y = data1[, 1]
 )
-emptyFit <- stan(file="model_TCGApriors_nosigmae.stan", data = stanDat1, chains = 0)
+emptyFit <- stan(file="model_TCGApriors.stan", data = stanDat1, chains = 0)
 
 # run in parallel via doParallel
 cl <- makeCluster(numCPUs)
@@ -142,16 +142,16 @@ parData <- foreach(i = iter(1:nsites), .combine = 'comb', .multicombine = TRUE, 
   data <- site(i)
   stanFit <- stanfit_model(data)
   
-  # take summary values for first 6 params (mu, betaT, sigmas x4) and last param (lp__)
-  fitSumm <- summary(stanFit)$summary[c(1:6,97), mInd]
+  # take summary values for first 6 params (mu, nu, sigmas x4) and last param (lp__)
+  fitSumm <- summary(stanFit)$summary[c(1:6,dim(summary(stanFit)$summary)[1]), mInd]
   
-  # compute posterior log-ratio: log(sigmaP/sigmaT), take integral>0, mean, median
[…]

### 2021-01-05 estimateTCGApriors.R

@@ -68,7 +68,7 @@ for(i in dataLabel){
 print(paste("# sites with at least one NA: ",length(NAsites)))
 
 # randomly select 5 sites in 1-485577 (excluding sites with any NA samples)
-#nsites <- 10000
+#nsites <- 5
 #siteInds <- sample(setdiff(1:485577,NAsites),nsites)
 
 # alternatively, use the entire set

### 2021-01-05 jobTCGA.sh

@@ -1,0 +1,16 @@
+#!/bin/bash
+#SBATCH --mem=512
+#SBATCH --array=1-87
+#SBATCH --error=jParErr-%j.err
+#SBATCH --partition=common
+
+export TMPDIR=/dscrhome/kam109/temp
+export PATH=/opt/apps/rhel7/R-3.4.4/bin:$PATH
+echo "working directory = "$SLURM_SUBMIT_DIR
+echo "SLURM_JOBID="$SLURM_JOBID
+echo "SLURM_JOB_NODELIST"=$SLURM_JOB_NODELIST
+echo "SLURM_NNODES"=$SLURM_NNODES
+echo "SLURM_ARRAY_TASK_COUNT="$SLURM_ARRAY_TASK_COUNT
+echo "Slurm Array Task ID="${SLURM_ARRAY_TASK_ID}
+Rscript StanCParallel.R ${SLURM_ARRAY_TASK_ID} ~/ResultsTCGA/StanCParResults_${SLURM_ARRAY_TASK_ID}.Rdata > ~/ParOut/taskOut${SLURM_ARRAY_TASK_ID}.txt
+

### 2021-01-05 loadSCRuns.R

@@ -9,8 +9,8 @@ library(reshape2)
 
 ### LOADING FOR STAN MODEL RESULTS
 
-# Set working directory to normal data wd then folder ResultsC
-setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data/Stan Results Archive/FinalRuns/ResultsTCGA_gamma")
+# Set working directory to normal data wd then folder with Results files from StanCParallel.R on all sites
+setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data/Stan Results Archive/FinalRuns/ResultsTCGA_relaxgamma5")
 
 mu_full <-
   data.frame(
@@ -21,7 +21,7 @@ mu_full <-
     Rhat = numeric(866091)
   )
 
-betaT_full <- mu_full
+nu_full <- mu_full
 sigmaP_full <- mu_full
 sigmaPT_full <- mu_full
 sigmaT_full <- mu_full
@@ -45,7 +45,7 @@ for (i in filesPresent) {
   }
   allInds <- append(allInds, inds)
   mu_full[inds, ] <- mu_C
-  betaT_full[inds, ] <- betaT_C
+  nu_full[inds, ] <- nu_C
   sigmaP_full[inds, ] <- sigmaP_C
   sigmaPT_full[inds, ] <- sigmaPT_C
   sigmaT_full[inds, ] <- sigmaT_C
@@ -53,6 +53,6 @@ for (i in filesPresent) {
   lp_full[inds, ] <- lp_C
   CpGscore_full[inds, ] <- CpGscore
 }
-save(mu_full, betaT_full, sigmaP_full, sigmaT_full, sigmaPT_full, sigmaE_full,
-      lp_full, CpGscore_full, file = "FullResultsTCGA_gamma.Rdata")
+save(mu_full, nu_full, sigmaP_full, sigmaT_full, sigmaPT_full, sigmaE_full,
+      lp_full, CpGscore_full, file = "FullResultsTCGA_relaxgamma5.Rdata")
 

### 2021-01-06 estimateTCGApriors.R
updated estimate TCGA priors code

@@ -47,6 +47,7 @@ print("Total time to load data:")
 print(proc.time()-ptmTot)
 
 save(concatData, file="testdata_allpats.Rdata")
+
 load("testdata_allpats.Rdata") # start here to load concatenated patient data
 
 # create labels
@@ -98,8 +99,8 @@ fitplot <- function(data, line) {
   pats <- as.integer(factor(data$patient))
   npats <- max(pats)
   mu <- line[1]
-  betaT <- line[2]
-  linedf <- data.frame(tInd = c(0,1), est = c(mu, (mu+betaT)))
+  nu <- line[2]
+  linedf <- data.frame(tInd = c(0,1), est = c(mu, (mu+nu)))
   p1 <- p1 + geom_line(data=linedf, aes(x=tInd, y=est))
   
   return(p1)
@@ -110,8 +111,8 @@ calcSigmaAll <- function (siteData, plotFlag=FALSE) {
   
   # sigmaP: take mean of normal samples
   mu <- mean(siteBeta[!tumorLabel])
-  b <- siteBeta[!tumorLabel]
-  sigmaP <- var(b)
+  a <- siteBeta[!tumorLabel]
+  sigmaP <- var(a)
   
   # sigmaPT: take mean of patient's tumor samples and normalize by patient normal
   normBeta <- c()
@@ -123,9 +124,9 @@ calcSigmaAll <- function (siteData, plotFlag=FALSE) {
       normBeta <- c(normBeta, tum-siteBeta[normIndex])
     }
   }
-  betaT <- mean(normBeta)
-  c <- normBeta
-  sigmaPT <- var(c)
+  nu <- mean(normBeta)
+  b <- normBeta
+  sigmaPT <- var(b)
   
   # sigmaT: all samples var method, normalize all tumor samples by patient tumor mean
   normBetaT <- c()
@@ -136,8 +137,8 @@ calcSigmaAll <- function (siteData, plotFlag=FALSE) {
       normBetaT <- c(normBetaT, siteBeta[tumIndex] - mean)
     }
   }
-  d <- normBetaT
-  sigmaT <- var(d)
+  g <- normBetaT
+  sigmaT <- var(g)
   
   if (plotFlag == TRUE){
   ### PLOTS ###
[…]

### 2021-01-06 PlotBetas.R
remove extraneous files

@@ -1,390 +1,0 @@
-### plot beta values between specific samples
-
-### INITIALIZE
-
-library(ggplot2)
-library(minfi)
-library(MASS)
-
-# here set the working directory that points to the data folder
-# e.g. the folder with annotated data saved as "myFA.Rdata"
-# please just comment out the directories already here, so each
-# user can uncomment the setwd for them as we move code around
-
-# Kevin's working directory
-setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
-
-# Yanlin's working directory
-setwd("D:/DataPlus2017/Data")
-
-
-##### LOAD DATA
-
-# load fully annotated data (saved from LoadDataAndQC.R)
-load("myFA.Rdata")
-
-
-### mds plot
-
-# mds plot
-cutData <- FullAnnotation[-(1:8)]
-
-pat <- data.frame(Sample.ID = colnames(cutData))
-
-pat$Sample.ID = colnames(cutData)
-
-aInd = c(1,2,14,15,21,22,29,30,52:55)
-cInd = c(3:13,16:20,23:28,31,38,39,49,50,56:71)
-oInd = c(32:37,40:48)
-
-pat$type[aInd] <- "Adenomas"
-pat$type[cInd] <- "Carcinomas"
-pat$type[oInd] <- "Other"
-
-patLabel <- pat$Sample.ID
-
-cutData2 <- cutData[pat$type=="Adenomas"|pat$type=="Carcinomas"]
-
-mdsPlot(as.matrix(cutData),numPositions = 1000000,sampNames=patLabel,sampGroups = pat$type,
-        main = "PCA Plot Using Noob (with JA)", pch=16,cex=0.5)
-
-
-
-
-##### COMPARE BETA
-
-### tumor vs normal for patients H E K* W C J (have both)
-# average tumor side A and B for tumor data
-
-# H
-betaHNor <- FullAnnotation$HN
-betaHTum <- (FullAnnotation$HA + FullAnnotation$HB)/2
[…]

### 2021-01-06 genewise.R

@@ -1,315 +1,0 @@
-##### Gene-wise analyses
-# Kevin Murgas
-# Looking at CpG sites within each individual gene
-# Using full-scale (850K site) Stan model results
-
-# Some analyses below (will) include:
-# Gene Average of Sigma Parameters (PCA, clustering)
-# Distance-based analysis (sigmas vs CpG distance in gene)
-# F vs R strand methylation correlation
-# Differential methylation validation with mu+betaT
-
-# Kevin's working directory
-setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
-
-# load annotation, Stan results
-load("myFA_bulkonly.Rdata")
-load("FullResultsTCGA_relaxgamma3_postsum3.Rdata")
-
-# load GeneInfo (gene names and regions for each site)
-load("GeneInfo.Rdata")
-
-### if GeneInfo not made yet ###
-
-# make siteGenes, a list of genes at each site
-fsplit <- function(str) unique(unlist(strsplit(str,split=";")))
-siteGenes <- lapply(FullAnnotation$UCSC_RefGene_Name, fsplit) # takes ~38 sec
-
-# make siteRegions, a label for regulatory region at each site
-# use 5'>3' priority: TSS1500>TSS200>5UTR>1stExon>Body>ExonBnd>3UTR>blank
-fsplitReg <- function(str) {
-    worklist <- unique(unlist(strsplit(str, split = ";")))
-    worklist[worklist == "3'UTR"] <- "3UTR"
-    worklist[worklist == "5'UTR"] <- "5UTR"
-    worklist[worklist == "5URT"] <- "5UTR"
-    #if (length(worklist)==0) { worklist <- "blank"}
-    return(unique(worklist))
-  }
-siteRegions <- lapply(FullAnnotation$UCSC_RefGene_Group, fsplitReg)
-
-# pre-index all genes
-geneInd <- function (geneStr) {
-  strx = paste0("(?<=^|;)", geneStr, "(?=$|;)") # match string - look for start/end of string or semicolon on either side of match
-  matchinds = grep(strx,FullAnnotation$UCSC_RefGene_Name,perl=TRUE); # check UCSC name
-  return(matchinds)
-}
-ptm<-proc.time()
[…]

### 2021-01-06 individualGenes.R

@@ -1,299 +1,0 @@
-# mark individual genes on histogram of Stan results
-
-library(lme4)
-library(arm)
-library(rstan)
-library(coda)
-library(gtools)
-library(ggplot2)
-library(bayesplot)
-library(reshape2)
-
-# Kevin's working directory
-setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
-
-# Yanlin's working directory
-setwd("D:/DataPlus2017/Data")
-
-# load annotation, Stan results, and gene scores
-load("myFA_bulkonly.Rdata")
-#load("StanCfullResults.Rdata")
-#load("StanCfullResults2.Rdata")
-#load("StanCfullResults3.Rdata")
-#load("ScoresFull.Rdata")
-load("FullResultsTCGA.Rdata")
-
-
-### INITIALIZE
-
-# # create list of unique gene names and gene regions at each site
-# geneNames <- list(866091)
-# geneRegions <- list(866091)
-# for (i in 1:866091) {
-#   if (!(i %% 86609)) {
-#     print(paste(i/8660.9,"% complete"))
-#   }
-#   geneNames[[i]] <- unique(unlist(strsplit(FullAnnotation$UCSC_RefGene_Name[i],split=";")))
-#   worklist <- unique(unlist(strsplit(FullAnnotation$UCSC_RefGene_Group[i],split=";")))
-#   worklist[worklist=="3'UTR"] <- "3UTR"
-#   worklist[worklist=="5'UTR"] <- "5UTR"
-#   worklist[worklist=="5URT"] <- "5UTR"
-#   if (length(worklist)==0) { worklist <- "blank"}
-#   geneRegions[[i]] <- worklist
-# }
-# save(geneNames, geneRegions, file = "GeneInfo.Rdata")
-load("GeneInfo.Rdata")
-
-# make vector of unique genes and regions
-uniqueGenes <- unique(unlist(siteGenes))
-regionTypes <- unique(unlist(siteRegions))
-#uniqueGenes <- unique(unlist(geneNames))
-#regionTypes <- unique(unlist(geneRegions))
-
-# make dataframes for median and mean values of each parameter
-dfMedians <- cbind.data.frame(mu_Cfull$p50, betaT_Cfull$p50, sigmaE_Cfull$p50, sigmaP_Cfull$p50, sigmaPT_Cfull$p50, sigmaT_Cfull$p50)
[…]

### 2021-01-06 varyChainParams.R

@@ -1,384 +1,0 @@
-# script for extracting variance from Stan complex model (model3)
-# input taskID corresponding to 10000 site chunks starting at site (N-1)*10000+1
-# Now incorporating parallel processing by RStan methods
-
-library(doParallel)
-library(rstan)
-library(gtools)
-library(ggplot2)
-library(bayesplot)
-
-## INITIALIZE
-rstan_options(auto_write = FALSE)
-print(paste("Cores: ", detectCores()))
-
-# Kevin's working directory
-setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data")
-
-# Yanlin's working directory
-#setwd("D:/DataPlus2017/Data")
-
-# load data
-load("myFA.Rdata")
-
-#args <- commandArgs(trailingOnly = TRUE)
-#print(args)
-#N <- as.numeric(args[1])
-N <- 1
-print(paste("Task #: ", N))
-#out.file <- args[2]
-log.file <- paste("logTask",N,".txt",sep="")
-writeLines(c(""), log.file)
-
-# randomly select 5 sites in 1-866836
-siteInds <- sample(1:866836,1)
-nsites <- length(siteInds)
-
-# select chunk within FullAnnotation, then remove FA
-print(paste("Making chunk of nsites =", nsites))
-chunk <- FullAnnotation[siteInds,]
-rm(FullAnnotation)
-gc()
-
-# must add to prevent crash
-checkFile <- "model3.rds"
-
-
-### FXNS ###
-# Function used to read in data from each site
-# changed to now use a "chunk" instead of FullAnnotation, for speed and memory
-# requires index 1-10000 instead of siteIndex
-site <- function (ind) {
-  # extract CpG site xx to start
-  temp <- chunk[ind,]
-  
-  # here we use all patient samples, excluding glands
-  indices <- c(9:13, 15:39, 46, 47, 57, 58, 72:75)
-  patientLabel <- substr(colnames(temp[indices]), 1, 1)
-  patientLabel[10:12] <- "K*"
-  sideLabel <- substr(colnames(temp[indices]), 2, 2)
-  tissueLabel <- sideLabel
-  tissueLabel[tissueLabel %in% c("A", "B")] <-
-    "T"   #replace A and B with T
-  tumorIndicator <- 1 * (tissueLabel == "T")
-  
[…]

### 2021-01-06 Paper_Figures.Rmd
Fixed log2 issue, updated readme

@@
 prepdatafolder = "data_final_rg3"
-cat(sprintf("Load prepped data dir: %s (needs to be pre-processed)...",prepdatafolder))
+cat(sprintf("Load prepped data dir: %s (needs to be pre-processed in StanAnalysis.Rmd)...",prepdatafolder))
 gened_data <- read_csv(here(prepdatafolder, "gened_data.csv"), col_types = list("CHR" = col_character()))
@@
 \newline Empirical values of mean and standard deviations were assesed for all 450k CpG sites.
-\newline The resulting distributions were fit with: bimodal gaussian (mu), Cauchy (betaT), or gamma distributions (sigmas).
+\newline The resulting distributions were fit with: bimodal gaussian (mu), Cauchy (nu), or gamma distributions (sigmas).
 \newline
@@
 ```{r Priors, echo=FALSE, fig.fullwidth=TRUE, warning=FALSE, comment=NA}
-# load TCGA estimates for mu, betaT, sigmaP, sigmaPT, sigmaT
+# load TCGA estimates for mu, nu, sigmaP, sigmaPT, sigmaT
 load(file="/Users/kevinmurgas/Documents/Data+ project/TCGA data/estimates_allpats.Rdata")
@@
 mu_ests <- mu_ests[which(!!mu_ests)]
-betaT_ests <- betaT_ests[which(!!betaT_ests)]
+nu_ests <- nu_ests[which(!!nu_ests)]
 
@@
 
-# plot TCGA estimates for mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT; with prior fits overlaid
+# plot TCGA estimates for mu,nu,sigmaE,sigmaP,sigmaPT,sigmaT; with prior fits overlaid
 pM <- post_prior_histogram(mu_ests, function(x) fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), function(x) fBG(x,0.3557,-3.0675, sqrt(3)*0.75214,1.1290, sqrt(3)*1.39197), 0.1, c(-5,5), "Mu")
-pB <- post_prior_histogram(betaT_ests, function(x) dcauchy(x,-0.04051, 0.19330), function(x) dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "Nu")
+pN <- post_prior_histogram(nu_ests, function(x) dcauchy(x,-0.04051, 0.19330), function(x) dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "Nu")
 pP <- post_prior_histogram(sigmaP_ests, function(x) dgamma(x, 5.1989, 16.6192), function(x) dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
@@
 
-p_all <- ggarrange(pM,pB,pE,pP,pPT,pT,hjust=0,nrow=2,ncol=3)
+p_all <- ggarrange(pM,pN,pE,pP,pPT,pT,hjust=0,nrow=2,ncol=3)
 print(p_all)
-rm(pM,pB,pBlank,pP,pPT,pT,pE)
+rm(pM,pN,pBlank,pP,pPT,pT,pE)
 
@@
 
-# plot hierarchical model posterior medians for mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT; with priors overlaid
+# plot hierarchical model posterior medians for mu,nu,sigmaE,sigmaP,sigmaPT,sigmaT; with priors overlaid
 pM <- post_prior_histogram(mu_full$p50, genemask, function(x) 1.5*fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), 0.1, c(-5,5), "Mu")
-pB <- post_prior_histogram(betaT_full$p50, genemask, function(x) 1.7*dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "BetaT")
+pN <- post_prior_histogram(nu_full$p50, genemask, function(x) 1.7*dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "Nu")
 pE <- post_prior_histogram(sigmaE_full$p50, genemask, function(x) 4*dgamma(x,2,2), 0.05, c(0,2), "SigmaE")
@@
 # arrange each parameter as a subplot # ,labels=c("M","B","E","P","PT","T","lP","lPT","lT")
-p_all <- ggarrange(pM,pB,pE,pP,pPT,pT,pLP,pLPT,pLT,hjust=0,nrow=3,ncol=3)
+p_all <- ggarrange(pM,pN,pE,pP,pPT,pT,pLP,pLPT,pLT,hjust=0,nrow=3,ncol=3)
 print(p_all)
-rm(pM,pB,pE,pP,pPT,pT,pLP,pLPT,pLT)
+rm(pM,pN,pE,pP,pPT,pT,pLP,pLPT,pLT)
 ```
@@
 ```{r neffRhat, echo=FALSE, warning=FALSE}
-# for each parameter (mu, betaT, sigmaP, sigmaPT, sigmaT, sigmaE, lp__)
+# for each parameter (mu, nu, sigmaP, sigmaPT, sigmaT, sigmaE, lp__)
 # reporting: mean n_eff, min n_eff, # n_eff<500, %, # n_eff<100, %
@@
   mu = mu_full$n_eff,
-  betaT = betaT_full$n_eff,
+  nu = nu_full$n_eff,
   sigmaP = sigmaP_full$n_eff,
@@
   .groups = "drop"
-  ) %>% arrange(match(.$param,c("mu","betaT","sigmaP","sigmaPT","sigmaT","sigmaE","lp")))
+  ) %>% arrange(match(.$param,c("mu","nu","sigmaP","sigmaPT","sigmaT","sigmaE","lp")))
 
@@
   mu = mu_full$Rhat,
-  betaT = betaT_full$Rhat,
+  nu = nu_full$Rhat,
   sigmaP = sigmaP_full$Rhat,
@@
 
-# plot the two conservations score prob sums (using /log(2) to convert log_e to log_2)
+# plot the two conservations score prob sums
 genemask = !(FullAnnotation$UCSC_RefGene_Name=="")
-p_scores <- post_histogram(CpGscore_full$median/log(2),genemask,0.05,"CpG Conservation score") +
+p_scores <- post_histogram(CpGscore_full$median,genemask,0.05,"CpG Conservation score") +
   theme(aspect.ratio=1/4,legend.position = "none") +
@@
 # Fig 4B: plot gene scores and adjusted bootstrap distribution
-p1 <- permed_data %>% ggplot(aes(x=score_median/log(2))) + geom_histogram(aes(y=..count..), binwidth=0.1, fill="lightgray",color = "grey50") + xlim(c(-1.5,3.5)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
+p1 <- permed_data %>% ggplot(aes(x=score_median)) + geom_histogram(aes(y=..count..), binwidth=0.1, fill="lightgray",color = "grey50") + xlim(c(-1.5,3.5)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
 p2 <- permed_data %>% ggplot(aes(x=score_median_v3)) + geom_histogram(aes(y=..count..), binwidth=0.02, fill="lightgray", color="grey50") + xlim(c(0,1)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
@@
   group_by(`Island relation` = Relation_to_UCSC_CpG_Island) %>%
-  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
   summarise_at(vars(sigmaP:sigmaT,score_median), mean, na.rm = TRUE) %>%
@@
 my_data = prepped_data %>% select(score_median, Relation_to_UCSC_CpG_Island) %>%
-  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
   group_by(ir = Relation_to_UCSC_CpG_Island) %>%
@@
 prepped_data %>% select(Name, Relation_to_UCSC_CpG_Island, score_median) %>%
-  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
   dplyr::rename(ir = Relation_to_UCSC_CpG_Island, score=score_median) %>%
@@
   select(c(region,cons_in,cons_out,nsites,nout,pval)) %>%
-  mutate(cons_in = cons_in/log(2), cons_out = cons_out/log(2)) %>% # change of base e to base 2
   mutate(fold_change = 2^cons_in, fold_diff = 2^(cons_in - cons_out)) # calculating fold difference (2^ (log2 difference))
@@
 
-# multiple regression stats (using log2 score)
-FRdata = data.frame(CpGscore = log2(prepped_data$sigmaP / prepped_data$sigmaT) ) %>%
+# multiple regression stats
+FRdata = data.frame(CpGscore = prepped_data$score_median ) %>%
   mutate(FR = pull(prepped_data,UCSC_RefGene_Group)) %>%
@@
 temp <- prepped_data %>% select(Name, UCSC_RefGene_Group, score_median) %>%
-  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
   separate_rows(UCSC_RefGene_Group, sep=";") %>%
@@
 distanced_data <- gened_data %>%
-  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
   filter(!is.na(UCSC_RefGene_Name)) %>%
@@
   CpGscores <- gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^",geneStr,"$"))) %>% pull(score_median)
-  # convert log_e to log_2
-  ggplot(prepped_data %>% mutate(score_median = score_median/log(2))) + geom_histogram(aes(x=score_median, y=..count../nrow(prepped_data)), binwidth=0.05,fill="lightgray") + ggtitle(sprintf("%s: CpG scores (%i sites)",geneStr, length(CpGscores))) +
+  ggplot(prepped_data) + geom_histogram(aes(x=score_median, y=..count../nrow(prepped_data)), binwidth=0.05,fill="lightgray") + ggtitle(sprintf("%s: CpG scores (%i sites)",geneStr, length(CpGscores))) +
     scale_x_continuous(limits=c(-5,5), breaks=c(-4,-2,0,2,4)) +
@@
 gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^TUBA1A$"))) %>%
-  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
   summarize(n = n(), mean=mean(score_median)) %>% mutate(fc = 2^mean)
@@
 gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^TTN$"))) %>%
-  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
   summarize(n = n(), mean=mean(score_median)) %>% mutate(fc = 2^mean)
@@
 gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^HLA-A$"))) %>%
-  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
   summarize(n = n(), mean=mean(score_median)) %>% mutate(fc = 2^mean)
@@
 # plot violins of: all genes, COSMIC, AGO
-perm_temp <- permed_data %>% select(c(UCSC_RefGene_Name, score_median,score_median_v3,n)) %>%
-  mutate(score_median2 = score_median/log(2)) # change of base from e to 2
+perm_temp <- permed_data %>% select(c(UCSC_RefGene_Name, score_median,score_median_v3,n))
 genes_dbs <- rbind(perm_temp %>% mutate(group="all"),
@@
               )
-# Fig 4B: plot violins (with log2 base change)
+# Fig 4B: plot violins
 pDB <- genes_dbs %>% mutate(group = factor(group, levels = c("all", "AGO", "COSMIC"))) %>%
-  ggplot() + geom_violin(aes(x=score_median2,y=group,fill=group), draw_quantiles = 0.5) + scale_x_continuous(limits=c(-2,4),breaks = c(-2,0,2,4)) + theme_classic() + ylab("Database") + xlab("Gene Conservation Score") + geom_vline(xintercept=0, color="black", alpha=0.2) + theme(aspect.ratio=60/156.3,legend.position = "none")
+  ggplot() + geom_violin(aes(x=score_median,y=group,fill=group), draw_quantiles = 0.5) + scale_x_continuous(limits=c(-2,4),breaks = c(-2,0,2,4)) + theme_classic() + ylab("Database") + xlab("Gene Conservation Score") + geom_vline(xintercept=0, color="black", alpha=0.2) + theme(aspect.ratio=60/156.3,legend.position = "none")
 # add hashmarks for each gene (doesn't look great)
-marks <- genes_dbs %>% mutate(x=score_median2, y=if_else(group=="COSMIC",3,if_else(group=="AGO",2,1))) %>% select(x,y)
-marks2 <- genes_dbs %>% filter(score_median_v3 > 0.95) %>% mutate(x=score_median2, y=if_else(group=="COSMIC",3,if_else(group=="AGO",2,1))) %>% select(x,y)
+marks <- genes_dbs %>% mutate(x=score_median, y=if_else(group=="COSMIC",3,if_else(group=="AGO",2,1))) %>% select(x,y)
+marks2 <- genes_dbs %>% filter(score_median_v3 > 0.95) %>% mutate(x=score_median, y=if_else(group=="COSMIC",3,if_else(group=="AGO",2,1))) %>% select(x,y)
 pDB + geom_segment(data=marks2, aes(x=x, xend=x, y=y, yend=y+0.2), color="red") + geom_segment(data=marks, aes(x=x, xend=x, y=y, yend=y+0.1), color="black", alpha=0.5)
@@
 # stats
-wilcox.test(x = genes_dbs$score_median2[genes_dbs$group=="COSMIC"], y = genes_dbs$score_median2[genes_dbs$group=="all"], alternative = "greater")
-wilcox.test(x = genes_dbs$score_median2[genes_dbs$group=="AGO"], y = genes_dbs$score_median2[genes_dbs$group=="all"], alternative = "greater")
+wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="COSMIC"], y = genes_dbs$score_median[genes_dbs$group=="all"], alternative = "greater")
+wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="AGO"], y = genes_dbs$score_median[genes_dbs$group=="all"], alternative = "greater")
 ```
@@
 ```
-
-<!-- \newpage -->
-<!-- \section{Gene Set Conservation Analysis} -->
-<!-- Based on Gene Set Enrichment Analysis (GSEA) -->
-<!-- ```{r load_gene_sets, include=FALSE} -->
-<!-- # load in the C5 gene sets from MSigDB -->
-<!-- library(GSA) -->
-<!-- gmtfile <- file.path("/Users/kevinmurgas/Documents/Data+ project/EPIC data","c2.cp.v7.1.symbols.gmt") -->
-<!-- genesets <- GSA.read.gmt(gmtfile) -->
-<!-- ``` -->
-
-<!-- ```{r GSCA, echo=FALSE, comment=NA, fig.width=8} -->
-<!-- # make a data.frame for gene sets and scores -->
-<!-- nsets <- length(genesets[[1]]) -->
-<!-- GSEA_df <- data.frame(GeneSet=genesets[[2]][1:nsets], score=numeric(nsets), -->
-<!--                       num_gene = unlist(lapply(genesets[[1]][1:nsets],length)), -->
-<!--                       num_gene_cut = numeric(nsets)) -->
-
-<!-- # have to order all genes first by rank of conservation (conserved genes come first) -->
-<!-- genenames = permed_data$UCSC_RefGene_Name -->
-<!-- ngene = length(genenames) -->
-
-<!-- # reorder by evidence rank -->
-<!-- ind = order(-permed_data$score_median) # use ranked conservation score as evidence -->
-<!-- evidence = seq(ngene,1,-1) -->
-<!-- genenames = genenames[ind] -->
-
-<!-- # plot evidence ranked: permed_data$score_median -->
-<!-- ranks <- permed_data$score_median -->
-<!-- names(ranks) <- permed_data$UCSC_RefGene_Name -->
-<!-- barplot(sort(ranks, decreasing = T)) -->
-
-<!-- ptm <- proc.time() -->
-<!-- for (iset in 1:nsets) { -->
-<!--   # take only genes that we have in full gene list -->
-<!--   set <- unlist(genesets[[1]][iset]) %>% intersect(genenames) -->
-<!--   GSEA_df$num_gene_cut[iset] = length(set) -->
-
-<!--   # skip if num_gene <10 or >500 -->
-<!--   if (length(set)<10 | length(set)>500) {next} -->
-
-<!--   # compute running sum (as PDF first) -->
-<!--   runsum_in = numeric(ngene) -->
-<!--   runsum_out = numeric(ngene) -->
-<!--   for (i in 1:ngene) { -->
-<!--     if (genenames[i] %in% set) { -->
-<!--       # evidence = permed_data$median, or ranks -->
-<!--       runsum_in[i] = evidence[i] -->
-<!--       runsum_out[i] = 0 -->
-<!--     } else { -->
-<!--       runsum_in[i] = 0 -->
-<!--       runsum_out[i] = 1 -->
-<!--     } -->
-<!--   } -->
-<!--   # convert to CDFs -->
-<!--   runsum_in <- cumsum(runsum_in)/sum(runsum_in) -->
-<!--   runsum_out <- cumsum(runsum_out)/sum(runsum_out) -->
-
-<!--   # compute K-S-like statistic -->
-<!--   v <- runsum_in - runsum_out -->
-<!--   score <- max(c(0,v)) + min(c(0,v)) -->
-<!--   GSEA_df$score[iset] <- score -->
-<!-- } -->
-<!-- cat(proc.time()-ptm) -->
-<!-- cat(sprintf("Processed %i genes, %i pathways. Top 50 paths:",ngene,nsets)) -->
-
-<!-- # insert significance bootstrapping -->
-
-<!-- # make table of top scores -->
-<!-- GSEA_df %>% arrange(-score) %>% -->
-<!--   select(score,num_gene,GeneSet) %>% -->
-<!--   slice(1:50) %>% -->
-<!--   mutate(GeneSet = gsub('(.{1,90})(\\s|$)', '\\1\n', GeneSet)) %>% -->
-<!--   gt() -->
-<!-- ``` -->
 
@@
 
-# plot: n_sites, mu, betaT, sigmaP, sigmaPT, sigmaT
+# plot: n_sites, mu, nu, sigmaP, sigmaPT, sigmaT
 n_cutlo = 5
 n_cuthi = quantile(gene_summary$n, 0.99)
-pN <- post_histogram(gene_summary$n,1,c(0,200),paste0("N sites (max = ",max(gene_summary$n),")")) + geom_vline(xintercept=n_cutlo-0.5,linetype = "dashed",color="red") + annotate(geom="text",x=80,y=0.03,size=2,label=paste0("# genes <",n_cutlo," sites:\n", sum(gene_summary$n<n_cutlo)," / ",nrow(gene_summary))) +
+pNS <- post_histogram(gene_summary$n,1,c(0,200),paste0("N sites (max = ",max(gene_summary$n),")")) + geom_vline(xintercept=n_cutlo-0.5,linetype = "dashed",color="red") + annotate(geom="text",x=80,y=0.03,size=2,label=paste0("# genes <",n_cutlo," sites:\n", sum(gene_summary$n<n_cutlo)," / ",nrow(gene_summary))) +
 geom_vline(xintercept=n_cuthi,linetype = "dashed",color="red") +annotate(geom="text",x=120,y=0.013,size=2,label=sprintf("# genes >%0.1f sites:\n %i / %i (99th pctile)", n_cuthi,sum(gene_summary$n>n_cuthi),nrow(gene_summary)))
@@
 pM <- post_histogram(gene_summary$mu,0.1,c(-5,5),"Mu")
-pB <- post_histogram(gene_summary$betaT,0.1,c(-5,5),"BetaT")
+pN <- post_histogram(gene_summary$nu,0.1,c(-5,5),"Nu")
 pP <- post_histogram(gene_summary$sigmaP,0.05,c(0,2),"SigmaP")
@@
 # arrange each parameter as a subplot
-p_all <- ggarrange(pN,pM,pB,pP,pPT,pT,pP1,hjust=0,nrow=3,ncol=3)
+p_all <- ggarrange(pNS,pM,pN,pP,pPT,pT,pP1,hjust=0,nrow=3,ncol=3)
 print(p_all)
-rm(pN,pM,pB,pP,pPT,pT,pP1)
+rm(pNS,pM,pN,pP,pPT,pT,pP1)
 
@@
 pM <- ggplot(gene_summary) + geom_point(aes(x=n,y=mu), alpha=0.01) + ggtitle("Mu vs N-sites")
-pB <- ggplot(gene_summary) + geom_point(aes(x=n,y=betaT), alpha=0.01) + ggtitle("BetaT vs N-sites")
+pN <- ggplot(gene_summary) + geom_point(aes(x=n,y=nu), alpha=0.01) + ggtitle("Nu vs N-sites")
 pE <- ggplot(gene_summary) + geom_point(aes(x=n,y=sigmaE), alpha=0.01) + ggtitle("SigmaE vs N-sites")
@@
 # arrange each parameter as a subplot
-p_all <- ggarrange(pM,pB,pE,pP,pPT,pT,pP1,hjust=0,nrow=3,ncol=3)
+p_all <- ggarrange(pM,pN,pE,pP,pPT,pT,pP1,hjust=0,nrow=3,ncol=3)
 print(p_all)
-rm(pM,pB,pE,pP,pPT,pT,pP1)
+rm(pM,pN,pE,pP,pPT,pT,pP1)
 ```

### 2021-01-06 prior_sensitivity.Rmd
Fixed log2 issue, updated readme

--- 2020-10-01 prior_sensitivity.Rmd
@@
 library(knitr)
-library(gtools)
+#library(gtools)
+library(emdist)
+library(ggpubr)
 
@@
 slice <- dplyr::slice
-
-set.seed(1234)
 ```
 
-```{r prior_sensitivity, echo=FALSE}
+```{r prior_dataAndFits, echo=FALSE, comment=NA}
+# load TCGA-based 450K array estimates (see estimateTCGApriors.R for data+fits)
+load("/Users/kevinmurgas/Documents/Data+ project/TCGA data/estimates_allpats.Rdata")
+
+# remove all zeros, and take square root of variance (stored variable) to get sigma/stdev
+sigmaP_ests <- sqrt(sigmaP_ests[which(!!sigmaP_ests)])
+sigmaPT_ests <- sqrt(sigmaPT_ests[which(!!sigmaPT_ests)])
+sigmaT_ests <- sqrt(sigmaT_ests[which(!!sigmaT_ests)])
+mu_ests <- mu_ests[which(!!mu_ests)]
+nu_ests <- nu_ests[which(!!nu_ests)]
+
+gg_color_hue <- function(n) {
+  hues = seq(15, 375, length = n + 1)
+  hcl(h = hues, l = 65, c = 100)[1:n]
+}
+cols = gg_color_hue(4)
+
+# ggplot histograms, with fit overlays
+# mu
+# 1x: 0.35570*normal(-3.0675,0.75214) + 0.64430*normal(1.1290, 1.3920)
+# 2x: 0.35570*normal(-3.0675, 1.0637) + 0.64430*normal(1.1290, 1.9685)
+# 3x: 0.35570*normal(-3.0675, 1.3027) + 0.64430*normal(1.1290, 2.4110)
+# 5x: 0.35570*normal(-3.0675, 1.6818) + 0.64430*normal(1.1290, 3.1125)
+f <- function(x, a, m1, s1, m2, s2) a*dnorm(x,m1,s1) + (1-a)*dnorm(x,m2,s2)
+p1 <- ggplot(data.frame(p=mu_ests), aes(x=p)) + geom_histogram(binwidth=0.1,aes(y=..density..),fill="lightgray") + theme_classic() +
+  #labs(x=expression(mu)) +
+  xlim(-6,6) +
+  stat_function(fun=function(x) f(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197),geom="line",color=cols[1]) +
+  stat_function(fun=function(x) f(x,0.3557,-3.0675, 1.0637,1.1290, 1.9685),geom="line",color=cols[2]) +
+  stat_function(fun=function(x) f(x,0.3557,-3.0675, 1.3027,1.1290, 2.4110),geom="line",color=cols[3]) +
+  stat_function(fun=function(x) f(x,0.3557,-3.0675, 1.6818,1.1290, 3.1125),geom="line",color=cols[4])
+# nu
+# 1x: Cauchy(-0.04051, 0.19330)
+# 2x: Cauchy(-0.04051, 0.38660)
+# 3x: Cauchy(-0.04051, 0.57990)
+# 5x: Cauchy(-0.04051, 0.96650)
+p2 <- ggplot(data.frame(p=nu_ests), aes(x=p)) + geom_histogram(binwidth=0.1,aes(y=..density..),fill="lightgray") + theme_classic() +
+  #labs(x=expression(nu)) +
+  xlim(-5,5) +
+  stat_function(fun=function(x) dcauchy(x,-0.04051, 0.19330),geom="line",color=cols[1]) +
+  stat_function(fun=function(x) dcauchy(x,-0.04051, 0.38660),geom="line",color=cols[2]) +
+  stat_function(fun=function(x) dcauchy(x,-0.04051, 0.57990),geom="line",color=cols[3]) +
+  stat_function(fun=function(x) dcauchy(x,-0.04051, 0.96650),geom="line",color=cols[4])
+# sigmaE: all gamma(2,2)
+p3 <- ggplot() + stat_function(fun=function(x) dgamma(x,2,2),geom="line",color="black") + xlim(0,2) + theme_classic() + labs(x=expression(sigma[E]))
+# sigmaP
+# 1x: gamma(5.1989, 16.6192)
+# 2x: gamma(3.4016, 9.5057)
+# 3x: gamma(2.7693, 7.0029)
+# 5x: gamma(2.2298, 4.8674)
+p4 <- ggplot(data.frame(p=sigmaP_ests), aes(x=p)) + geom_histogram(binwidth=0.02,aes(y=..density..),fill="lightgray") + theme_classic() +
+  #labs(x=expression(sigma[P])) +
+  xlim(0,2) +
+  stat_function(fun=function(x) dgamma(x,5.1989, 16.6192),geom="line",color=cols[1]) +
+  stat_function(fun=function(x) dgamma(x,3.4016, 9.5057),geom="line",color=cols[2]) +
+  stat_function(fun=function(x) dgamma(x,2.7693, 7.0029),geom="line",color=cols[3]) +
+  stat_function(fun=function(x) dgamma(x,2.2298, 4.8674),geom="line",color=cols[4])
+# sigmaPT
+# 1x: gamma(3.5539, 5.5324)
+# 2x: gamma(2.5210, 3.2948)
+# 3x: gamma(2.1457, 2.4819)
+# 5x: gamma(1.8166, 1.7689)
+p5 <- ggplot(data.frame(p=sigmaPT_ests), aes(x=p)) + geom_histogram(binwidth=0.02,aes(y=..density..),fill="lightgray") + theme_classic() +
+  #labs(x=expression(sigma[PT])) +
+  xlim(0,2) +
+  stat_function(fun=function(x) dgamma(x,3.5539, 5.5324),geom="line",color=cols[1]) +
+  stat_function(fun=function(x) dgamma(x,2.5210, 3.2948),geom="line",color=cols[2]) +
+  stat_function(fun=function(x) dgamma(x,2.1457, 2.4819),geom="line",color=cols[3]) +
+  stat_function(fun=function(x) dgamma(x,1.8166, 1.7689),geom="line",color=cols[4])
+# sigmaT
+# 1x: gamma(6.8383, 25.7560)
+# 2x: gamma(4.2574, 14.370)
+# 3x: gamma(3.3643, 10.430)
+# 5x: gamma(2.6144, 7.1221)
+p6 <- ggplot(data.frame(p=sigmaT_ests), aes(x=p)) + geom_histogram(binwidth=0.02,aes(y=..density..),fill="lightgray") + theme_classic() +
+  xlim(0,2) +
+  #labs(x=expression(sigma[T])) +
+  stat_function(fun=function(x) dgamma(x,6.8383, 25.7560),geom="line",color=cols[1]) +
+  stat_function(fun=function(x) dgamma(x,4.2574, 14.370),geom="line",color=cols[2]) +
+  stat_function(fun=function(x) dgamma(x,3.3643, 10.430),geom="line",color=cols[3]) +
+  stat_function(fun=function(x) dgamma(x,2.6144, 7.1221),geom="line",color=cols[4])
+
+p_all <- ggarrange(p1,p2,p3,p4,p5,p6)
+# ggsave("prior_fits_all.svg",p_all,height=3,width=5.53,dpi=300)
+```
+
+```{r prior_sensitivity, echo=FALSE, comment=NA}
+
+n_eff_mat <- data.frame(Params=c("mu","nu","sigmaP","sigmaPT","sigmaT","sigmaE","lp"))
+Rhat_mat <- n_eff_mat
+KLdiv_mat <- n_eff_mat %>% slice(1:6)
+BCdist_mat <- KLdiv_mat
+EMdist_mat <- KLdiv_mat
 
@@
   print(stanFile)
-  #load(here(stanFile))
+  load(here(stanFile))
+  print("loaded")
+
+  n_eff_temp <-
+  data.frame(
+  mu = mu_full$n_eff,
+  nu = nu_full$n_eff,
+  sigmaP = sigmaP_full$n_eff,
+  sigmaPT = sigmaPT_full$n_eff,
+  sigmaT = sigmaT_full$n_eff,
+  sigmaE = sigmaE_full$n_eff,
+  lp = lp_full$n_eff
+  ) %>% gather(param,value,mu:lp) %>% group_by(param) %>% summarize(
+  mean = mean(value),
+  .groups = "drop"
+  ) %>% arrange(match(.$param,c("mu","nu","sigmaP","sigmaPT","sigmaT","sigmaE","lp")))
+
+  Rhat_temp <-
+  data.frame(
+  mu = mu_full$Rhat,
+  nu = nu_full$Rhat,
+  sigmaP = sigmaP_full$Rhat,
+  sigmaPT = sigmaPT_full$Rhat,
+  sigmaT = sigmaT_full$Rhat,
+  sigmaE = sigmaE_full$Rhat,
+  lp = lp_full$Rhat
+  ) %>% gather(param,value,mu:lp) %>% group_by(param) %>% summarize(
+  mean = mean(value),
+  .groups = "drop"
+  ) %>% arrange(match(.$param,c("mu","nu","sigmaP","sigmaPT","sigmaT","sigmaE","lp")))
+
+  # KL divergence
+  # formula: KL(P||Q) = sum(P(i) * log( P(i)/Q(i) )), where P(i),Q(i) = density
+  KL_temp <- numeric(6)
+  #Bhattacharya distance
+  # formula: BC = sum( sqrt(P(i)*Q(i)) )
+  BC_temp <- numeric(6)
+  # Earth-movers (Wasserstein) distance
+  EM_temp <- numeric(6)
+
+  for (j in c(1:6)) {
+    # choose parameter
+    data <- switch(j,
+                   mu_full$p50,
+                   nu_full$p50,
+                   sigmaP_full$p50,
+                   sigmaPT_full$p50,
+                   sigmaT_full$p50,
+                   sigmaE_full$p50)
+    breaks <- switch(j,
+                     seq(-5,5,0.1),
+                     seq(-6,6,0.12), #5.5,.11
+                     seq(0,3,0.03), #2.6
+                     seq(0,3,0.03),
+                     seq(0,3,0.03), #1.2
+                     seq(0,3,0.03)) #2.5
+    P <- hist(data, plot=FALSE, breaks=101) # run hist to get density and midpoints
+    x <- P$mids
+    # define prior density by pre-defined priors at P midpoints
+    if (i==1){ # non relaxed (1x)
+      priordens <- switch(j,
+                      0.3557*dnorm(x, -3.0675, 0.75214) + 0.6443*dnorm(x, 1.1290, 1.3920),
+                      dcauchy(x, -0.04051, 0.19330),
+                      dgamma(x, 5.1989, 16.6192),
+                      dgamma(x, 3.5539, 5.5324),
+                      dgamma(x, 6.8383, 25.7560),
+                      dgamma(x,2,2))
+    } else if (i==2) { # 2x relaxed
+      priordens <- switch(j,
+                      0.3557*dnorm(x, -3.0675, 1.0637) + 0.6443*dnorm(x, 1.1290, 1.9685),
+                      dcauchy(x,-0.04051, 0.38660),
+                      dgamma(x, 3.4016, 9.5057),
+                      dgamma(x, 2.5210, 3.2948),
+                      dgamma(x, 4.2574, 14.370),
+                      dgamma(x,2,2))
+    } else if (i==3) { # 3x relaxed
+      priordens <- switch(j,
+                      0.3557*dnorm(x, -3.0675, 1.3027) + 0.6443*dnorm(x, 1.1290, 2.4110),
+                      dcauchy(x,-0.04051, 0.5799),
+                      dgamma(x, 2.7693, 7.0029),
+                      dgamma(x, 2.1457, 2.4819),
+                      dgamma(x, 3.3643, 10.430),
+                      dgamma(x,2,2))
+    } else if (i==4) { # 5x relaxed
+      priordens <- switch(j,
+                      0.3557*dnorm(x, -3.0675, 1.6818) + 0.6443*dnorm(x, 1.1290, 3.1125),
+                      dcauchy(x,-0.04051, 0.96650),
+                      dgamma(x, 2.2298, 4.8674),
+                      dgamma(x, 1.8166, 1.7689),
+                      dgamma(x, 2.6144, 7.1221),
+                      dgamma(x,2,2))
+    }
+    Pi <- P$density[which(P$density>0)]
+    Qi <- priordens[which(P$density>0)]
+    KL_temp[j] = sum(Pi * log(Pi/Qi))
+
+    # BCD
+    dx <- x[2]-x[1]
+    BC_temp[j] = -log(sum(sqrt(P$density*priordens)*dx))
+
+    # EMD
+    A = cbind(P$density,P$mids)
+    B = cbind(priordens,P$mids)
+    EM_temp[j] = emd(A,B)
+
+    # code below is for plotting data vs priors
+    #Q <- P
+    #Q$density <- priordens
+    #Q$counts <- Q$density*866091*(Q$breaks[2]-Q$breaks[1])
+    #c1 <- rgb(173,216,230,max = 255, alpha = 80, names = "lt.blue")
+    #c2 <- rgb(255,192,203, max = 255, alpha = 80, names = "lt.pink")
+    #plot(P,col=c1, main=paste("i=",i,", j=",j))
+    #plot(Q,col=c2,add=TRUE)
   }
 
+  varname = switch(i, "1x", "2x", "3x", "5x")
+  n_eff_mat <- n_eff_mat %>% mutate(!!varname := n_eff_temp$mean)
+  Rhat_mat <- Rhat_mat %>% mutate(!!varname := Rhat_temp$mean)
+  KLdiv_mat <- KLdiv_mat %>% mutate(!!varname := KL_temp)
+  BCdist_mat <- BCdist_mat %>% mutate(!!varname := BC_temp)
+  EMdist_mat <- EMdist_mat %>% mutate(!!varname := EM_temp)
+
+}
+
+print("N_eff")
+kable(n_eff_mat, digits=2)
+print("Rhat")
+kable(Rhat_mat, digits=3)
+print("KLdiv")
+kable(KLdiv_mat, digits=3)
+cat("\n")
+print("BCdist")
+kable(BCdist_mat, digits=3)
+print("EMdist")
+kable(EMdist_mat, digits=3)
+
 ```
+
+\newpage
+Plots
+```{r plots, echo=FALSE, comment=NA}
+# goal here is to just get the distributions of parameters for each prior set
+# store histogram density, and KDE
+# plot histogram with side-by-side bar plots
+# plot KDE's as line plots
+
+mu_allprior <- data.frame(x1=numeric(866091),
+                          x2=numeric(866091),
+                          x3=numeric(866091),
+                          x5=numeric(866091))
+nu_allprior <- mu_allprior
+sigmaP_allprior <- mu_allprior
+sigmaPT_allprior <- mu_allprior
+sigmaT_allprior <- mu_allprior
+sigmaE_allprior <- mu_allprior
+
+for (i in c(1:4)) {
+  stanFile = "FullResultsTCGA_relaxgamma3.Rdata"
+  stanFile = switch(i,
+                    "FullResultsTCGA_gamma.Rdata",
+                    "FullResultsTCGA_relaxgamma2.Rdata",
+                    "FullResultsTCGA_relaxgamma3.Rdata",
+                    "FullResultsTCGA_relaxgamma5.Rdata")
+  print(stanFile)
+  load(here(stanFile))
+  print("loaded")
+
+  mu_allprior[,i] <- mu_full$p50
+  nu_allprior[,i] <- nu_full$p50
+  sigmaP_allprior[,i] <- sigmaP_full$p50
+  sigmaPT_allprior[,i] <- sigmaPT_full$p50
+  sigmaT_allprior[,i] <- sigmaT_full$p50
+  sigmaE_allprior[,i] <- sigmaE_full$p50
+}
+
+cat("Posterior estimates for varying prior relaxation")
+pM <- mu_allprior %>% gather("relax","val",1:4) %>% ggplot(aes(x=val,y=..density..,color=relax)) + geom_density() + xlim(-6,6) + ggtitle("Mu") + theme_classic() + theme(legend.position = "none")
+pN <- nu_allprior %>% gather("relax","val",1:4) %>% ggplot(aes(x=val,y=..density..,color=relax)) + geom_density() + xlim(-5,5) + ggtitle("Nu") + theme_classic() + theme(legend.position = "none")
+pP <- sigmaP_allprior %>% gather("relax","val",1:4) %>% ggplot(aes(x=val,y=..density..,color=relax)) + geom_density() + xlim(0,2) + ggtitle("sigmaP") + theme_classic() + theme(legend.position = "none")
+pPT <- sigmaPT_allprior %>% gather("relax","val",1:4) %>% ggplot(aes(x=val,y=..density..,color=relax)) + geom_density() + xlim(0,2) + ggtitle("sigmaPT") + theme_classic() + theme(legend.position = "none")
+pT <- sigmaT_allprior %>% gather("relax","val",1:4) %>% ggplot(aes(x=val,y=..density..,color=relax)) + geom_density() + xlim(0,2) + ggtitle("sigmaT") + theme_classic() + theme(legend.position = "none")
+pE <- sigmaE_allprior %>% gather("relax","val",1:4) %>% ggplot(aes(x=val,y=..density..,color=relax)) + geom_density() + xlim(0,2) + ggtitle("sigmaE") + theme_classic() + theme(legend.position = "none")
+pLP <- sigmaP_allprior %>% gather("relax","val",1:4) %>% mutate(val=log(val)) %>% ggplot(aes(x=val,y=..density..,color=relax)) + geom_density() + xlim(-4,1) + ggtitle("log(sigmaP)") + theme_classic() + theme(legend.position = "none")
+pLPT <- sigmaPT_allprior %>% gather("relax","val",1:4) %>% mutate(val=log(val)) %>% ggplot(aes(x=val,y=..density..,color=relax)) + geom_density() + xlim(-4,1) + ggtitle("log(sigmaPT)") + theme_classic() + theme(legend.position = "none")
+pLT <- sigmaT_allprior %>% gather("relax","val",1:4) %>% mutate(val=log(val)) %>% ggplot(aes(x=val,y=..density..,color=relax)) + geom_density() + xlim(-4,1) + ggtitle("log(sigmaT)") + theme_classic() + theme(legend.position = "none")
+
+p_all<- ggarrange(pM, pN, pE, pP, pPT, pT, pLP, pLPT, pLT, nrow=3, ncol=3)
+print(p_all)
+# ggsave("prior_sens_kdes.svg",p1)
+```

### 2021-01-06 loadSCRuns.R
Fixed log2 issue

@@ -10,7 +10,7 @@ library(reshape2)
 ### LOADING FOR STAN MODEL RESULTS
 
 # Set working directory to normal data wd then folder with Results files from StanCParallel.R on all sites
-setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data/Stan Results Archive/FinalRuns/ResultsTCGA_relaxgamma5")
+setwd("/Users/kevinmurgas/Documents/Data+ project/EPIC data/Stan Results Archive/FinalRuns/ResultsTCGA_gamma")
 
 mu_full <-
   data.frame(
@@ -54,5 +54,5 @@ for (i in filesPresent) {
   CpGscore_full[inds, ] <- CpGscore
 }
 save(mu_full, nu_full, sigmaP_full, sigmaT_full, sigmaPT_full, sigmaE_full,
-      lp_full, CpGscore_full, file = "FullResultsTCGA_relaxgamma5.Rdata")
+      lp_full, CpGscore_full, file = "FullResultsTCGA_gamma.Rdata")
 

### 2021-01-06 StanAnalysis.Rmd
Edit to StanAnalysis

--- 2021-01-05 StanAnalysis_updated.Rmd
@@
 \subsection{Prior Probabilities}
-Priors for each parameter (mu, betaT, sigmaP/PT/T) were defined using TCGA 450k array data from colorectal tumor patients with paired multiple tumor and/or normal+tumor samples.
+Priors for each parameter (mu, nu, sigmaP/PT/T) were defined using TCGA 450k array data from colorectal tumor patients with paired multiple tumor and/or normal+tumor samples.
 \newline Empirical values of mean and standard deviations were assesed for all 450k CpG sites.
-\newline The resulting distributions were fit with: bimodal gaussian (mu), Cauchy (betaT), or gamma distributions (sigmas).
+\newline The resulting distributions were fit with: bimodal gaussian (mu), Cauchy (nu), or gamma distributions (sigmas).
 \newline
@@
 lower variation in methlyation within tumor (sigmaT) relative to the variation within normal healthy tissue (sigmaP).
-Therefore we choose the log-ratio of sigmaP/sigmaT to represent a conservation score, which is positive when tumor variation is lower than normal variation.
-We take the median of the posterior distribution for this parameter. Additionally, we calculate a regularization term based on the inverse of the L2 norm of sigmaP and sigmaT.
+Therefore we choose the log2-ratio of sigmaP/sigmaT to represent a conservation score, which is positive when tumor variation is lower than normal variation.
+We take the median of the posterior distribution for this parameter.
 
-<!--  $$ \text{score}_\text{prob} = \Pr (\text{log}\frac{\sigma_{P}}{\sigma_T} > 0) = \int \text{log}\frac{\sigma_{P}}{\sigma_T} > 0 ~ dp $$
-$$ \text{score}_\text{mean} = \text{Mean} (\text{log}\frac{\sigma_{P}}{\sigma_T}) ,    ~ -->
-$$\text{score}_\text{med} = \text{Median} (\text{log}\frac{\sigma_{P}}{\sigma_T}) $$
+$$\text{score}_\text{med} = \text{Median} (\text{log}_2\frac{\sigma_{P}}{\sigma_T}) $$
 
@@
 
-# plot mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT with priors overlaid
+# plot mu,nu,sigmaE,sigmaP,sigmaPT,sigmaT with priors overlaid
 pM <- post_prior_histogram(mu_full$p50, genemask, function(x) 1.3*fBG(x,0.3557,-3.0675, 1.3027, 1.1290, 2.4110), 0.25, c(-5,5), "Mu") + theme(aspect.ratio = 1)
-pB <- post_prior_histogram(betaT_full$p50, genemask, function(x) 1.5*dcauchy(x,-0.04051, 0.5799), 0.25, c(-5,5), "BetaT") + theme(aspect.ratio = 1)
+pN <- post_prior_histogram(nu_full$p50, genemask, function(x) 1.5*dcauchy(x,-0.04051, 0.5799), 0.25, c(-5,5), "Nu") + theme(aspect.ratio = 1)
 pE <- post_prior_histogram(sigmaE_full$p50, genemask, function(x) 3.3*dgamma(x,2,2), 0.05, c(0,2), "SigmaE") + theme(aspect.ratio = 1)
@@
 
-# arrange each parameter as a subplot # ,labels=c("M","B","E","P","PT","T","lP","lPT","lT")
-p_all <- ggarrange(pM,pB,pE,pP,pPT,pT,hjust=0,nrow=2,ncol=3) #,pLP,pLPT,pLT
+# arrange each parameter as a subplot # ,labels=c("M","N","E","P","PT","T","lP","lPT","lT")
+p_all <- ggarrange(pM,pN,pE,pP,pPT,pT,hjust=0,nrow=2,ncol=3) #,pLP,pLPT,pLT
 print(p_all)
 # ggsave("posteriors_geneassoc.svg",p_all)
-rm(pM,pB,pE,pP,pPT,pT)
+rm(pM,pN,pE,pP,pPT,pT)
 ```
@@
 # ggsave("CpGscore_geneassoc.svg")
-# prob1_proxy = log(sigmaP_full$p50/sigmaT_full$p50)
-# pP2 <- post_histogram(prob1_proxy,genemask,0.05,c(-3,3),expression(" log( median" ~sigma[P]/~"median"~sigma[T]~")"))
-# # regularized version
-# L2norm = sqrt(sigmaP_full$p50^2 + sigmaT_full$p50^2)
-# prob1_reg = log(sigmaP_full$p50/sigmaT_full$p50) - 1/L2norm
-# pP3 <- post_histogram(prob1_reg,genemask,0.1,c(-10,5),"Regularized")
-# p_probs <- ggarrange(pP1,pP2,pP3,nrow=1,ncol=3)
-# print(p_probs)
 ```
@@
 
-med_vals <- tibble(Name=FullAnnotation[,1], mu=mu_full$p50, betaT=betaT_full$p50, sigmaE=sigmaE_full$p50, sigmaP=sigmaP_full$p50, sigmaPT=sigmaPT_full$p50, sigmaT=sigmaT_full$p50, Rhat=lp_full$Rhat) %>% mutate(logP=log(sigmaP), logPT=log(sigmaPT), logT=log(sigmaT)) %>%
+med_vals <- tibble(Name=FullAnnotation[,1], mu=mu_full$p50, nu=nu_full$p50, sigmaE=sigmaE_full$p50, sigmaP=sigmaP_full$p50, sigmaPT=sigmaPT_full$p50, sigmaT=sigmaT_full$p50, Rhat=lp_full$Rhat) %>% mutate(logP=log(sigmaP), logPT=log(sigmaPT), logT=log(sigmaT)) %>%
   mutate(score_median=CpGscore_full$median)
@@
   filter(n < quantile(n, 0.99)) %>%
-  boot_naive(gened_data, sigmaP, UCSC_RefGene_Name, 1000) %>%
-  boot_naive(gened_data, sigmaPT, UCSC_RefGene_Name, 1000) %>%
-  boot_naive(gened_data, sigmaT, UCSC_RefGene_Name, 1000) %>%
   boot_naive(gened_data, score_median, UCSC_RefGene_Name, 1000) %>%
-  boot_adj(gened_data, sigmaP, UCSC_RefGene_Name, 1000) %>%
-  boot_adj(gened_data, sigmaPT, UCSC_RefGene_Name, 1000) %>%
-  boot_adj(gened_data, sigmaT, UCSC_RefGene_Name, 1000) %>%
   boot_adj(gened_data, score_median, UCSC_RefGene_Name, 1000)
@@
 cat("done")
-```
-
-``` {r enrich, include=FALSE}
-# expression data
-# downloaded file: E-MTAB-2836-query-results.tpms.tsv
-# source: https://www.ebi.ac.uk/gxa/experiments/E-MTAB-2836/Results
-expression_atlas <- read_tsv(here("data-raw", "E-MTAB-2836-query-results.tpms.tsv"), skip = 4, 
-                             col_types = cols(
-                               .default = col_double(),
-                               `Gene ID` = col_character(),
-                               `Gene Name` = col_character()
-                             ))
-
-# Gene Set Enrichment
-tidy_gsea <- function(genes) {
-  AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, genes, 'ENTREZID', 'SYMBOL') %>%
-    ReactomePA::enrichPathway()
-}
-
-enrich_sigmaP <- permed_data %>%
-  #filter(sigmaP_v3 < quantile(sigmaP_v3, 0.05, na.rm = TRUE)) %>%
-  filter(sigmaP_v3 < 0.05) %>%
-  pull(UCSC_RefGene_Name) %>% tidy_gsea()
-enrich_sigmaPT <- permed_data %>%
-  #filter(sigmaPT_v3 < quantile(sigmaPT_v3, 0.05, na.rm = TRUE)) %>%
-  filter(sigmaPT_v3 < 0.05) %>%
-  pull(UCSC_RefGene_Name) %>% tidy_gsea()
-enrich_sigmaT <- permed_data %>%
-  #filter(sigmaT_v3 < quantile(sigmaT_v3, 0.05, na.rm = TRUE)) %>%
-  filter(sigmaT_v3 < 0.05) %>%
-  pull(UCSC_RefGene_Name) %>% tidy_gsea()
-enrich_score_median <- permed_data %>%
-  #filter(score_median_v3 > quantile(score_median_v3, 0.95, na.rm = TRUE)) %>%
-  filter(score_median_v3 > 0.95) %>%
-  pull(UCSC_RefGene_Name) %>% tidy_gsea()
 ```
@@
 pM <- ggplot(gene_summary) + geom_point(aes(x=n,y=mu), alpha=0.01) + ggtitle("Mu vs N-sites")
-pB <- ggplot(gene_summary) + geom_point(aes(x=n,y=betaT), alpha=0.01) + ggtitle("BetaT vs N-sites")
+pB <- ggplot(gene_summary) + geom_point(aes(x=n,y=nu), alpha=0.01) + ggtitle("Nu vs N-sites")
 pE <- ggplot(gene_summary) + geom_point(aes(x=n,y=sigmaE), alpha=0.01) + ggtitle("SigmaE vs N-sites")
@@
 ```
-
-\newpage
-\subsection{Figure 3: Reactome Pathway Enrichment Analysis}
-Enrichment defined from the group of genes <5th percentile of each sigma, or >95th percentile of conservation score.
-```{r Fig3, echo=FALSE, fig.height=3.5, comment=NA}
-cat("Number of enrichment pathways (sigmas):")
-c(sigmaP = length(enrich_sigmaP$Description),
-  sigmaPT = length(enrich_sigmaPT$Description),
-  sigmaT = length(enrich_sigmaT$Description)) %>%
-  sort()
-
-list(
-  sigmaP = enrich_sigmaP$Description,
-  sigmaPT = enrich_sigmaPT$Description,
-  sigmaT = enrich_sigmaT$Description
-) %>%
-  UpSetR::fromList() %>%
-  UpSetR::upset(sets.bar.color = RColorBrewer::brewer.pal(3, "Set2")[c(3, 2, 1)])
-grid::grid.text("Figure 3A: Frequency of overlap between pathways that are called \nas most conserved for each hierarchical variance.", x=0.65, y=0.05, gp=grid::gpar(fontsize=10))
-
-cat("Number of enrichment pathways (probs):")
-c(score_median = length(enrich_score_median$Description),
-  score_median2 = length(enrich_score_median2$Description),
-  score_reg = length(enrich_score_reg$Description)) %>%
-  sort()
-
-list(
-  score_median = enrich_score_median$Description,
-  score_median2 = enrich_score_median2$Description,
-  score_reg = enrich_score_reg$Description
-) %>%
-  UpSetR::fromList() %>%
-  UpSetR::upset(sets.bar.color = RColorBrewer::brewer.pal(3, "Set2")[c( 2, 1)])
-grid::grid.text("Figure 3B: Frequency of overlap between pathways that are called \nas most conserved for each conservation score.", x=0.65, y=0.05, gp=grid::gpar(fontsize=10))
-```
-
-\newpage
-\subsection{Figure 4: Pathway Enrichment Network}
-```{r Fig4, echo=FALSE, message=FALSE, fig.height=2.5}
-# Figure 4
-cat("sigmaP:")
-ggg <- enrich_sigmaP %>%
-  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
-  labs(caption = "Figure 4A: Relationship between pathways that are called as conserved based on sigmaP.", title = NULL)
-
-ggg$layers <- ggg$layers[1:2]
-
-ggg + coord_flip() + 
-  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
-  scale_color_gradientn(colors = "black") +
-  guides(color = "none", size = "none")
-
-#ggsave(filename = here("figures", "figure-4.png"), width = 7, dpi = 320, height = 4.326)
-
-# sigmaPT
-cat("sigmaPT:")
-ggg <- enrich_sigmaPT %>%
-  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
-  labs(caption = "Figure 4B: Relationship between pathways that are called as conserved based on sigmaPT.", title = NULL)
-
-ggg$layers <- ggg$layers[1:2]
-ggg + coord_flip() + 
-  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
-  scale_color_gradientn(colors = "black") +
-  guides(color = "none", size = "none")
-
-# sigmaT
-cat("sigmaT:")
-ggg <- enrich_sigmaT %>%
-  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
-  labs(caption = "Figure 4C: Relationship between pathways that are called as conserved based on sigmaT.", title = NULL)
-
-ggg$layers <- ggg$layers[1:2]
-ggg + coord_flip() + 
-  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
-  scale_color_gradientn(colors = "black") +
-  guides(color = "none", size = "none")
-```
-
-\newpage
-```{r Fig4DEF, echo=FALSE, message=FALSE, fig.height=2.5}
-# score_median
-cat("score_median")
-ggg <- enrich_score_median %>%
-  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
-  labs(caption = "Figure 4D: Relationship between pathways that are called as conserved based on score_median.", title = NULL)
-ggg$layers <- ggg$layers[1:2]
-ggg + coord_flip() + 
-  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
-  scale_color_gradientn(colors = "black") +
-  guides(color = "none", size = "none")
-
-# score_median2
-cat("score_median2")
-ggg <- enrich_score_median2 %>%
-  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
-  labs(caption = "Figure 4E: Relationship between pathways that are called as conserved based on score_median2.", title = NULL)
-ggg$layers <- ggg$layers[1:2]
-ggg + coord_flip() + 
-  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
-  scale_color_gradientn(colors = "black") +
-  guides(color = "none", size = "none")
-
-# score_reg
-cat("score_reg")
-#ggg <- enrich_score_reg %>%
-#  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
-#  labs(caption = "Figure 4G: Relationship between pathways that are called as conserved based on score_reg.", title = NULL)
-#ggg$layers <- ggg$layers[1:2]
-#ggg + coord_flip() +
-#  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
-#  scale_color_gradientn(colors = "black") +
-#  guides(color = "none", size = "none")
-```
-
-\subsection{Figure 5: Gene Conservation vs Gene Expression}
-```{r Fig5, echo=FALSE, message=FALSE}
-# Figure 5A
-figure5_data <- permed_data %>% 
-  left_join(by = c("UCSC_RefGene_Name" = "gene"), promoted_data) %>%
-  left_join(expression_atlas %>% select(`Gene Name`, colon),
-            by = c("UCSC_RefGene_Name" = "Gene Name"))
-
-bind_rows(
-  figure5_data %>%
-    select(sigmaP_v3, sigmaP_pro, colon) %>%
-    pivot_longer(-colon) %>%
-    drop_na() %>%
-    rename(expression = colon),
-  figure5_data %>%
-    select(sigmaPT_v3, sigmaPT_pro, colon) %>%
-    pivot_longer(-colon) %>%
-    drop_na() %>%
-    rename(expression = colon),
-  figure5_data %>%
-    select(sigmaT_v3, sigmaT_pro, colon) %>%
-    pivot_longer(-colon) %>%
-    drop_na() %>%
-    rename(expression = colon)
-) %>%
-  group_by(name) %>%
-  mutate(value = factor(11-as.numeric(cut_number(value, 10)))) %>%
-  separate(name, c("type", "method"), sep = "_", extra = "merge") %>%
-  mutate(type = factor(type, c("sigmaP","sigmaPT","sigmaT"), c("SigmaP-Colon","SigmaPT-Colon","SigmaT-colon")),
-         method = factor(method, c("v3", "pro"), c("Boot-strapped values", "Promoter region"))) %>%
-  ggplot(aes(value, expression, fill = type)) +
-  geom_boxplot() +
-  scale_y_log10() +
-  guides(color = "none") +
-  facet_grid(method ~ type) +
-  labs(x = "quantile of conservation by gene", y = "Expression", title = NULL,
-       caption = "Figure 5A: Relationship between hierarchical variance (Sigmas) and gene expression.") +
-  theme_minimal() +
-  scale_fill_brewer(palette = "Set2") +
-  theme(legend.position = "none")
-
-# Figure 5B
-bind_rows(
-  figure5_data %>%
-    select(score_median_v3, score_median_pro, colon) %>%
-    rename(scoremedian_v3 = score_median_v3, scoremedian_pro = score_median_pro) %>%
-    pivot_longer(-colon) %>%
-    drop_na() %>%
-    rename(expression = colon),
-  figure5_data %>%
-    select(score_median2_v3, score_median2_pro, colon) %>%
-    rename(scoremedian2_v3 = score_median2_v3, scoremedian2_pro = score_median2_pro) %>%
-    pivot_longer(-colon) %>%
-    drop_na() %>%
-    rename(expression = colon),
-  figure5_data %>%
-    select(score_reg_v3, score_reg_pro, colon) %>%
-    rename(scorereg_v3 = score_reg_v3, scorereg_pro = score_reg_pro) %>%
-    pivot_longer(-colon) %>%
-    drop_na() %>%
-    rename(expression = colon)
-) %>%
-  group_by(name) %>%
-  mutate(value = factor(as.numeric(cut_number(value, 10)))) %>%
-  separate(name, c("type", "method"), sep = "_", extra = "merge") %>%
-  mutate(type = factor(type, c("scoremedian","scoremedian2","scorereg"), c("ScoreMedian-colon","ScoreMedian2-colon","Scorereg-colon")),
-         method = factor(method, c("v3", "pro"), c("Boot-strapped values", "Promoter region"))) %>%
-  ggplot(aes(value, expression, fill = type)) +
-  geom_boxplot() +
-  scale_y_log10() +
-  guides(color = "none") +
-  facet_grid(method ~ type) +
-  labs(x = "quantile of conservation by gene", y = "Expression", title = NULL,
-       caption = "Figure 5B: Relationship between conservation scores and gene expression.") +
-  theme_minimal() +
-  scale_fill_brewer(palette = "Set2") +
-  theme(legend.position = "none")
-```
-
-\newpage
-\section{Appendix: Enrichment Gene/Pathway Lists}
-\subsection{A1: SigmaP}
-Genes are selected with cutoff <5th percentile (adjusted bootstrap), showing top 30 genes/pathways
-```{r App_sigmaP, echo=FALSE, fig.fullwidth=TRUE, comment=NA, message=FALSE}
-enrichedGenes <- permed_data %>%
-  select(UCSC_RefGene_Name, sigmaP, sigmaP_v3, n) %>%
-  rename(Gene=UCSC_RefGene_Name, Pctile=sigmaP_v3) %>%
-  filter(Pctile < quantile(Pctile, 0.05, na.rm = TRUE)) %>%
-  arrange(Pctile)
-
-enrichedPaths <- tibble(Pathway = enrich_sigmaP$Description, ngenes = enrich_sigmaP$Count, padj = enrich_sigmaP$p.adjust) %>% arrange(padj)
-geneList <- enrich_reactome$geneID %>% strsplit(split='/') %>%
-  lapply(function(x) AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, x, 'SYMBOL', 'ENTREZID')) %>%
-  lapply(function(x) paste(x,collapse=";")) %>% unlist
-enrichedPaths$Genes = geneList
-
-ngene <- dim(enrichedGenes)[1]
-npath <- length(enrich_sigmaP$Description)
-
-cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath))
-
-nshow = 30
-if (ngene > nshow) {
-  enrichedGenes <- enrichedGenes[1:nshow,]
-}
-if (npath > nshow) {
-  enrichedPaths <- enrichedPaths[1:nshow,]
-}
-
-library(pander)
-enrichedGenes %>% pander()
-
-cat("SigmaP top 30 enriched paths:  \n")
-enrichedPaths %>% pander()
-```
-
-\newpage
-\subsection{A2: SigmaPT}
-Genes are selected with cutoff <5th percentile (adjusted bootstrap), showing top 30 genes/pathways
-```{r App_sigmaPT, echo=FALSE, comment=NA, message=FALSE}
-enrichedGenes <- permed_data %>%
-  select(UCSC_RefGene_Name, sigmaPT, sigmaPT_v3, n) %>%
-  rename(Gene=UCSC_RefGene_Name, Pctile=sigmaPT_v3) %>%
-  filter(Pctile < quantile(Pctile, 0.05, na.rm = TRUE)) %>%
-  arrange(Pctile)
-
-enrichedPaths <- tibble(Pathway = enrich_sigmaPT$Description, ngenes = enrich_sigmaPT$Count, padj = enrich_sigmaPT$p.adjust) %>% arrange(padj)
-geneList <- enrich_reactome$geneID %>% strsplit(split='/') %>%
-  lapply(function(x) AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, x, 'SYMBOL', 'ENTREZID')) %>%
-  lapply(function(x) paste(x,collapse=";")) %>% unlist
-enrichedPaths$Genes = geneList
-
-ngene <- dim(enrichedGenes)[1]
-npath <- length(enrich_sigmaPT$Description)
-
-cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath))
-
-nshow = 30
-if (ngene > nshow) {
-  enrichedGenes <- enrichedGenes[1:nshow,]
-}
-if (npath > nshow) {
-  enrichedPaths <- enrichedPaths[1:nshow,]
-}
-
-enrichedGenes %>% pander()
-
-cat("\nSigmaPT top 30 enriched paths:  \n")
-enrichedPaths %>% pander()
-```
-
-\newpage
-\subsection{A3: SigmaT}
-Genes are selected with cutoff <5th percentile (adjusted bootstrap), showing top 30 genes/pathways
-```{r App_sigmaT, echo=FALSE, comment=NA, message=FALSE}
-enrichedGenes <- permed_data %>%
-  select(UCSC_RefGene_Name, sigmaT, sigmaT_v3, n) %>%
-  rename(Gene=UCSC_RefGene_Name, Pctile=sigmaT_v3) %>%
-  filter(Pctile < quantile(Pctile, 0.05, na.rm = TRUE)) %>%
-  arrange(Pctile)
-
-enrichedPaths <- tibble(Pathway = enrich_sigmaT$Description, ngenes = enrich_sigmaT$Count, padj = enrich_sigmaT$p.adjust) %>% arrange(padj)
-geneList <- enrich_reactome$geneID %>% strsplit(split='/') %>%
-  lapply(function(x) AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, x, 'SYMBOL', 'ENTREZID')) %>%
-  lapply(function(x) paste(x,collapse=";")) %>% unlist
-enrichedPaths$Genes = geneList
-
-ngene <- dim(enrichedGenes)[1]
-npath <- length(enrich_sigmaT$Description)
-
-cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath))
-
-nshow = 30
-if (ngene > nshow) {
-  enrichedGenes <- enrichedGenes[1:nshow,]
-}
-if (npath > nshow) {
-  enrichedPaths <- enrichedPaths[1:nshow,]
-}
-
-enrichedGenes %>% pander()
-
-cat("SigmaT top 30 enriched paths:  \n")
-enrichedPaths %>% pander()
-```
-
-\newpage
-\subsection{A6: Score Median}
-Genes are selected with cutoff >95th percentile (adjusted bootstrap), showing top 50 genes/pathways
-```{r App_score_median, echo=FALSE, comment=NA, message=FALSE}
-enrichedGenes <- permed_data %>% 
-  select(UCSC_RefGene_Name, score_median, score_median_v3, n) %>%
-  rename(Gene=UCSC_RefGene_Name, Score_median=score_median, Pctile=score_median_v3) %>%
-  filter(Pctile > quantile(Pctile, 0.95, na.rm = TRUE)) %>%
-  arrange(Pctile)
-
-enrichedPaths <- tibble(Pathway = enrich_score_median$Description, ngenes = enrich_score_median$Count, padj = enrich_score_median$p.adjust) %>% arrange(padj)
-
-ngene <- dim(enrichedGenes)[1]
-npath <- length(enrich_score_median$Description)
-
-cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath))
-
-nshow = 50
-if (ngene > nshow) {
-  enrichedGenes <- enrichedGenes[1:nshow,]
-}
-if (npath > nshow) {
-  enrichedPaths <- enrichedPaths[1:nshow,]
-}
-
-enrichedGenes %>% pander()
-
-cat("Score_median top 50 enriched paths:  \n")
-enrichedPaths %>% pander()
-```

### 2021-01-06 GetPosteriorsSingleSites.R
minor edits

@@ -92,7 +92,7 @@ fitplot <- function(data,stanFit) {
   npats <- max(pats)
   sumTemp <- summary(stanFit)$summary
   mu <- sumTemp[1,1] # mean estimate
-  betaT <- sumTemp[2,1] # mean
+  nu <- sumTemp[2,1] # mean
   sampcoefs <- data.frame(pat = pats, b = sumTemp[6+pats], c = sumTemp[6+npats+pats], d = sumTemp[6+2*npats+c(1:nsamp)])
   patcoefs <- data.frame(pat = 1:npats, b = sumTemp[6+1:npats], c = sumTemp[6+npats+1:npats])
   rm(sumTemp)
@@ -103,7 +103,7 @@ fitplot <- function(data,stanFit) {
   patdf <- data.frame(pat = rep(levels(data$patient),2), tInd = c(rep(0,npats),rep(1,npats)), est = c(patcoefs$b+mu, patcoefs$b+patcoefs$c+mu+betaT))
   p1 <- p1 + geom_line(data=patdf, aes(x=tInd, y=est, colour=pat))
   # add overall mean estimates as black dots
-  meandf <- data.frame(tInd = c(0,1), est = c(mu,mu+betaT))
+  meandf <- data.frame(tInd = c(0,1), est = c(mu,mu+nu))
   p1 <- p1 + geom_point(data=meandf, aes(x=tInd, y=est), size=3, alpha=0.7) + geom_line(data=meandf, aes(x=tInd, y=est), size=2, alpha=0.5)
   
   return(p1)
@@ -143,7 +143,7 @@ stanDat1 <- list(
 emptyFit <- stan(file="model_TCGApriors.stan", data = stanDat1, chains = 0)
 
 # # run in parallel via doParallel
-# # want to collect data of random effect coefficients (bcd), fixed effects (betaT,mu), and sigmas
+# # want to collect data of random effect coefficients (abg), fixed effects (nu,mu), and sigmas
 # cl <- makeCluster(detectCores()-3)
 # registerDoParallel(cl)
 # print(paste("Cores registered:",getDoParWorkers()))
@@ -155,29 +155,33 @@ emptyFit <- stan(file="model_TCGApriors.stan", data = stanDat1, chains = 0)
 #   print(paste("site:", i,"/",nsites))
 #   sink()
 # 
[…]

### 2021-01-06 LoadDataAndQC.R

@@ -501,7 +501,7 @@ map <- mapToGenome(raw_1639)
 gr <- granges(map, use.names=TRUE)
 anno <- getAnnotation(map)
 
-# DMP finder (can compare to mu/betaT later)
+# DMP finder (can compare to mu/nu later)
 temp <- as.matrix(FullAnnotation[,10:57])
 rownames(temp) <- FullAnnotation$IlmnID
 patientLabel <- substr(colnames(temp),1,1)

### 2021-01-06 loadSCRuns.R

@@ -1,5 +1,5 @@
 # load StanCRun data
-# produce full size data frames for mu,betaT,sigmas (P,PT,T,E)
+# produce full size data frames for mu,nu,sigmas (P,PT,T,E)
 # read in individual save files containing 10000 site runs
 # store 10000 site chunks in corresponding positions in full data frame
 # repeat for each completed run

### 2021-01-29 Paper_Figures.Rmd
Updates

@@
 load(here(stanFile))
+CpGscore_full = CpGscore_full %>% mutate(mean = mean*2, median=median*2)
 
@@
 permed_data <- read_csv(here(prepdatafolder, "permed_data.csv"))
+
+gened_data <- gened_data %>% mutate(score_median = score_median*2)
+prepped_data <- prepped_data %>% mutate(score_median = score_median*2)
+permed_data <- permed_data %>% mutate(score_median = score_median*2)
 
@@
 fBG <- function(x, a, m1, s1, m2, s2) a*dnorm(x,m1,s1) + (1-a)*dnorm(x,m2,s2) # bimodal gaussian
-post_prior_histogram <- function(postvals,mask,priorfun,bw,xl,titlestr) {
-  # stacked histogram colored by gene-associated sites
-  # and prior curve overlaid
-  plot_df <- data.frame(val=postvals, gene_assoc=mask)
+post_prior_histogram <- function(postvals,priorfun,bw,xl,titlestr) {
+  # histogram and prior curve overlaid
+  plot_df <- data.frame(val=postvals)
   p1 <- ggplot(plot_df, aes(x=val)) +
-    geom_histogram(aes(y=..count../length(mask),fill=gene_assoc), binwidth=bw) +
+    geom_histogram(aes(y=..density..), binwidth=bw) +
     stat_function(fun=function(x) bw*priorfun(x),geom="line",color="black") +
@@
 # plot hierarchical model posterior medians for mu,nu,sigmaE,sigmaP,sigmaPT,sigmaT; with priors overlaid
-pM <- post_prior_histogram(mu_full$p50, genemask, function(x) 1.5*fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), 0.1, c(-5,5), "Mu")
-pN <- post_prior_histogram(nu_full$p50, genemask, function(x) 1.7*dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "Nu")
-pE <- post_prior_histogram(sigmaE_full$p50, genemask, function(x) 4*dgamma(x,2,2), 0.05, c(0,2), "SigmaE")
-pP <- post_prior_histogram(sigmaP_full$p50, genemask, function(x) 2*dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
-pPT <- post_prior_histogram(sigmaPT_full$p50, genemask, function(x) 1.8*dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
-pT <- post_prior_histogram(sigmaT_full$p50, genemask, function(x) 1.2*dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT")
+pM <- post_prior_histogram(mu_full$p50, function(x) 10*fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), 0.1, c(-5,5), "Mu")
+pN <- post_prior_histogram(nu_full$p50, function(x) 15*dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "Nu")
+#pE <- post_prior_histogram(sigmaE_full$p50, function(x) 4*dgamma(x,2,2), 0.05, c(0,2), "SigmaE")
+#pP <- post_prior_histogram(sigmaP_full$p50^2, function(x) 2*dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
+#pPT <- post_prior_histogram(sigmaPT_full$p50^2, function(x) 1.8*dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
+#pT <- post_prior_histogram(sigmaT_full$p50^2, function(x) 1.2*dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT")
+pE <- post_prior_histogram(sigmaE_full$p50^2, function(x) 100*dgamma(sqrt(x),2,2), 0.05, c(0,2), "SigmaE^2")
+pP <- post_prior_histogram(sigmaP_full$p50^2, function(x) 250*dgamma(sqrt(x), 2.7693, 7.0029), 0.01, c(0,1), "SigmaP^2")
+pPT <- post_prior_histogram(sigmaPT_full$p50^2, function(x) 40*dgamma(sqrt(x), 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT^2")
+pT <- post_prior_histogram(sigmaT_full$p50^2, function(x) 220*dgamma(sqrt(x), 3.3643, 10.430), 0.01, c(0,1), "SigmaT^2")
 # log sigmas
-pLP <- post_prior_histogram(log(sigmaP_full$p50), genemask, function(x) 0.52*dgamma(exp(x), 2.7693, 7.0029), 0.1, c(-5,5), "log(SigmaP)")
-pLPT <- post_prior_histogram(log(sigmaPT_full$p50), genemask, function(x) 0.7*dgamma(exp(x), 2.1457, 2.4819), 0.1, c(-5,5), "log(SigmaPT)")
-pLT <- post_prior_histogram(log(sigmaT_full$p50), genemask, function(x) 0.4*dgamma(exp(x), 3.3643, 10.430), 0.1, c(-5,5), "log(SigmaT)") + theme(legend.position = c(0.8, 0.6), legend.text=element_text(size=6), legend.title=element_text(size=6))
+#pLP <- post_prior_histogram(log(sigmaP_full$p50), genemask, function(x) 0.52*dgamma(exp(x), 2.7693, 7.0029), 0.1, c(-5,5), "log(SigmaP)")
+#pLPT <- post_prior_histogram(log(sigmaPT_full$p50), genemask, function(x) 0.7*dgamma(exp(x), 2.1457, 2.4819), 0.1, c(-5,5), "log(SigmaPT)")
+#pLT <- post_prior_histogram(log(sigmaT_full$p50), genemask, function(x) 0.4*dgamma(exp(x), 3.3643, 10.430), 0.1, c(-5,5), "log(SigmaT)") + theme(legend.position = c(0.8, 0.6), legend.text=element_text(size=6), legend.title=element_text(size=6))
 
 # arrange each parameter as a subplot # ,labels=c("M","B","E","P","PT","T","lP","lPT","lT")
-p_all <- ggarrange(pM,pN,pE,pP,pPT,pT,pLP,pLPT,pLT,hjust=0,nrow=3,ncol=3)
+p_all <- ggarrange(pM,pN,pE,pP,pPT,pT,hjust=0,nrow=2,ncol=3)
 print(p_all)
-rm(pM,pN,pE,pP,pPT,pT,pLP,pLPT,pLT)
+rm(pM,pN,pE,pP,pPT,pT)
 ```
@@
 genemask = !(FullAnnotation$UCSC_RefGene_Name=="")
-p_scores <- post_histogram(CpGscore_full$median,genemask,0.05,"CpG Conservation score") +
+p_scores <- post_histogram(CpGscore_full$median,genemask,0.2,"CpG Conservation score") +
   theme(aspect.ratio=1/4,legend.position = "none") +
-  scale_x_continuous(limits=c(-4,4),breaks=c(-4,-2,0,2,4)) + 
+  scale_x_continuous(limits=c(-8,8),breaks=c(-8,-4,0,4,8)) + 
   scale_y_continuous(breaks=c(0,0.1,0.2,0.3,0.4,0.5))
@@
 # Fig 4B: plot gene scores and adjusted bootstrap distribution
-p1 <- permed_data %>% ggplot(aes(x=score_median)) + geom_histogram(aes(y=..count..), binwidth=0.1, fill="lightgray",color = "grey50") + xlim(c(-1.5,3.5)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
+p1 <- permed_data %>% ggplot(aes(x=score_median)) + geom_histogram(aes(y=..count..), binwidth=0.1, fill="lightgray",color = "grey50") + xlim(c(-3,7)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
 p2 <- permed_data %>% ggplot(aes(x=score_median_v3)) + geom_histogram(aes(y=..count..), binwidth=0.02, fill="lightgray", color="grey50") + xlim(c(0,1)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
@@
   select(CHR, UCSC_RefGene_Name, cpgdist, c(score_median)) %>%
-  mutate(location_bin = cut(cpgdist, c(seq(-1500, 20000, by = 100)), include.lowest = TRUE)) %>%
+  mutate(location_bin = cut(cpgdist, c(seq(-1500, 9000, by = 100)), include.lowest = TRUE)) %>%
   group_by(location_bin) %>%
@@
 # plot for conservation scores
-distance_length %>%
-  drop_na() %>%
-  pivot_longer(mean_score_median) %>% 
-  mutate(name = factor(name, c("mean_score_median"), c("Score_median"))) %>%
-  sample_frac() %>%
-  ggplot(aes(location_bin, value, color = name, fill = name)) +
-  geom_point(key_glyph = draw_key_rect) +
-  ggforce::facet_zoom(location_bin < 5000) +
-  theme_light() +
-  theme(legend.position = "bottom") +
-  labs(title = NULL,
-       fill = NULL,
-       color = NULL,
-       x = "Distance from gene start site (binned to 100 bp)",
-       y = "Average Conservation Score", 
-       caption = "Figure 3D: Average conservation scores of single CpGs as a function of position relative to gene.") +
-  scale_color_brewer(palette = "Set2") +
-  scale_fill_brewer(palette = "Set2")
+distance_length %>% drop_na() %>%
+  ggplot(aes(x = location_bin, y = mean_score_median)) +
+  geom_point(color="black") +
+  theme_bw() + theme(legend.position = "none", aspect.ratio = 3/8) +
+  xlab("Distance from gene start site (binned to 100 bp)") + ylab("Average Conservation Score") +
+  scale_x_continuous(breaks = seq(-1500,10000,1500))
 cat("\n")

### 2021-01-29 GetPosteriorsSingleSites.R

@@ -25,7 +25,7 @@ load("myFA_bulkonly.Rdata")
 # randomly select 5 sites in 1-866091
 #siteInds <- sample(1:866091,5)
 siteInds <- c(525490, 570532, 827025, 287316, 854143)
-siteInds <- 124022
+#siteInds <- 124022
 nsites <- length(siteInds)
 
 
@@ -82,9 +82,9 @@ stanfit_model <- function (dataset) {
 
 # Plot raw data and fits
 fitplot <- function(data,stanFit) {
-  # plot beta and fits at this site for all patients, using mean estimates for b and c coefficients
+  # plot M and fits at this site for all patients, using mean estimates for b and c coefficients
   # first plot data points
-  p1 <- ggplot() + geom_point(data = data, aes(x=tInd, y=beta, colour=patient))
+  p1 <- ggplot() + geom_point(data = data, aes(x=tInd, y=M, colour=patient))
   
   # extract coefficients from stan fit
   pats <- as.integer(factor(data$patient))
@@ -97,10 +97,10 @@ fitplot <- function(data,stanFit) {
   patcoefs <- data.frame(pat = 1:npats, b = sumTemp[6+1:npats], c = sumTemp[6+npats+1:npats])
   rm(sumTemp)
   # add individual sample estimates
-  sampdf <- data.frame(pat = rep(factor(data$patient),2), patg = rep(factor(1:nsamp),2), tInd = c(rep(0,nsamp),rep(1,nsamp)), est = c(sampcoefs$b+mu, (sampcoefs$b+sampcoefs$c++sampcoefs$d+mu+betaT)))
+  sampdf <- data.frame(pat = rep(factor(data$patient),2), patg = rep(factor(1:nsamp),2), tInd = c(rep(0,nsamp),rep(1,nsamp)), est = c(sampcoefs$b+mu, (sampcoefs$b+sampcoefs$c++sampcoefs$d+mu+nu)))
   p1 <- p1 + geom_line(data=sampdf, aes(x=tInd, y=est, colour=pat, group=patg), linetype="dashed", alpha=0.5)
   # add patient mean estimates
-  patdf <- data.frame(pat = rep(levels(data$patient),2), tInd = c(rep(0,npats),rep(1,npats)), est = c(patcoefs$b+mu, patcoefs$b+patcoefs$c+mu+betaT))
[…]

### 2021-01-29 estimateTCGApriors.R

@@ -343,9 +343,11 @@ text(0.7,5,expression(Gamma*"("*alpha*"=6.8383,"~beta*"=25.7560)"),col="blue")
 
 # calculate relaxed priors (for gammas)
 # increase variance by 2 but keep same mode
-a1 = 6.8383; b1 = 25.7560
+a1 = 5.1989; b1 = 16.6192 # sigmaP
+a1 = 3.5539; b1 = 5.5324 # sigmaPT
+a1 = 6.8383; b1 = 25.7560 # sigmaT
 f1 <- function(a2, a1, b1) (a2-1)*b1/(a1-1) # mode
-f2 <- function(a2, a1, b1) sqrt(a2*b1^2/(3*a1)) # 3x variance
+f2 <- function(a2, a1, b1) sqrt(a2*b1^2/(10*a1)) # 3x variance
 a2range = c(0:1000000)*0.00001
 out1 <- f1(a2range,a1,b1) - f2(a2range,a1,b1)
 a2 = a2range[which(abs(out1)==min(abs(out1)))]

### 2021-02-19 Paper_Figures.Rmd
added boxplots

@@
 print(p_scores)
+
+df = data.frame(med=CpGscore_full$median, mask=genemask)
+p_scores <- ggplot(data=df) + geom_boxplot(aes(x=med, y = genemask, fill = genemask), outlier.shape = NA) +
+  scale_x_continuous(limits=c(-8,8),breaks=c(-8,-4,0,4,8)) + xlab("CpG score") + ylab("Gene-associated") + theme_classic() + theme(aspect.ratio=1/4,legend.position = "none")
+print(p_scores)
 # ggsave("CpGscores_overlay.svg")
@@
 p2 <- permed_data %>% ggplot(aes(x=score_median_v3)) + geom_histogram(aes(y=..count..), binwidth=0.02, fill="lightgray", color="grey50") + xlim(c(0,1)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
-p_genes <- ggarrange(p1,p2,nrows=2,ncol=1)
+p_genes <- ggarrange(p1,p2,nrows=1,ncol=2)
 # ggsave("gene_scores_bootstrap.svg")
@@
                         TRUE ~ ir)) %>%
-  mutate(ir = factor(ir, levels = c("Island", "Shore", "Shelf", "Sea"))) %>%
-  #mutate(ir = factor(ir, levels = c("Sea", "Shelf", "Shore", "Island"))) %>%
-  ggplot() + geom_violin(aes(x=ir,y=score,fill=ir),draw_quantiles = 0.5) + theme_classic() + scale_y_continuous(limits=c(-5,5),breaks=c(-4,-2,0,2,4)) + theme(aspect.ratio=1/4,legend.position = "none")
+  #mutate(ir = factor(ir, levels = c("Island", "Shore", "Shelf", "Sea"))) %>%
+  #ggplot() + geom_violin(aes(x=ir,y=score,fill=ir),draw_quantiles = 0.5) + theme_classic() + 
+  mutate(ir = factor(ir, levels = c("Sea", "Shelf", "Shore", "Island"))) %>%
+  ggplot() + geom_boxplot(aes(y=ir,x=score, fill=ir), outlier.shape = NA) + theme_classic() +
+  scale_x_continuous(limits=c(-5,5),breaks=c(-4,-2,0,2,4)) + theme(aspect.ratio=1/4,legend.position = "none")
 # ggsave("CpGscore_islands_violin.svg")
+ggsave("CpGscore_islands_boxplot.svg")
 ```
@@
   distinct()
-#keepregions = c("TSS1500","Body", "3'UTR","1stExon","TSS200","5'UTR","ExonBnd")
-keepregions = c("TSS200","1stExon", "5'UTR","TSS1500","Body","3'UTR","ExonBnd")
+#regions = c("TSS1500","Body", "3'UTR","1stExon","TSS200","5'UTR","ExonBnd")
+regions = rev(c("TSS200","1stExon", "5'UTR","TSS1500","Body","3'UTR","ExonBnd"))
 temp %>%
-  mutate(fr = factor(fr, levels = keepregions)) %>%
-  filter(fr %in% keepregions) %>%
-  ggplot() + geom_violin(aes(x=fr,y=score,fill=fr), draw_quantiles = 0.5) + theme_classic() + scale_y_continuous(limits=c(-5,5),breaks=c(-4,-2,0,2,4,6)) + theme(aspect.ratio=1/4,legend.position = "none") + geom_hline(yintercept = 0, color="black", alpha=0.2)
+  mutate(fr = factor(fr, levels = regions)) %>%
+  filter(fr %in% regions) %>%
+  #ggplot() + geom_violin(aes(x=fr,y=score,fill=fr), draw_quantiles = 0.5) + theme_classic() +
+  ggplot() + geom_boxplot(aes(y=fr,x=score, fill=fr), outlier.shape = NA) + theme_classic() +
+  scale_x_continuous(limits=c(-5,5),breaks=c(-4,-2,0,2,4)) + theme(aspect.ratio=1/4,legend.position = "none") + geom_vline(xintercept = 0, color="black", alpha=0.2)
 # ggsave("CpGscore_regions_violin.svg")
-#rm(temp)
+ggsave("CpGscore_regions_boxplot.svg")
+#rm(temp)3
 
@@
               )
-# Fig 4B: plot violins
+# Fig 4C: plot distribution
 pDB <- genes_dbs %>% mutate(group = factor(group, levels = c("all", "AGO", "COSMIC"))) %>%
-  ggplot() + geom_violin(aes(x=score_median,y=group,fill=group), draw_quantiles = 0.5) + scale_x_continuous(limits=c(-2,4),breaks = c(-2,0,2,4)) + theme_classic() + ylab("Database") + xlab("Gene Conservation Score") + geom_vline(xintercept=0, color="black", alpha=0.2) + theme(aspect.ratio=60/156.3,legend.position = "none")
-# add hashmarks for each gene (doesn't look great)
-marks <- genes_dbs %>% mutate(x=score_median, y=if_else(group=="COSMIC",3,if_else(group=="AGO",2,1))) %>% select(x,y)
-marks2 <- genes_dbs %>% filter(score_median_v3 > 0.95) %>% mutate(x=score_median, y=if_else(group=="COSMIC",3,if_else(group=="AGO",2,1))) %>% select(x,y)
-pDB + geom_segment(data=marks2, aes(x=x, xend=x, y=y, yend=y+0.2), color="red") + geom_segment(data=marks, aes(x=x, xend=x, y=y, yend=y+0.1), color="black", alpha=0.5)
+  #ggplot() + geom_violin(aes(x=score_median,y=group,fill=group), draw_quantiles = 0.5) +
+  ggplot() + geom_boxplot(aes(x=score_median,y=group,fill=group), outlier.shape = NA) +
+  scale_x_continuous(limits=c(-2,4),breaks = c(-2,0,2,4)) + theme_classic() + ylab("Database") + xlab("Gene Conservation Score") + geom_vline(xintercept=0, color="black", alpha=0.2) + theme(aspect.ratio=60/156.3,legend.position = "none")
+# # add hashmarks for each gene (doesn't look great)
+# marks <- genes_dbs %>% mutate(x=score_median, y=if_else(group=="COSMIC",3,if_else(group=="AGO",2,1))) %>% select(x,y)
+# marks2 <- genes_dbs %>% filter(score_median_v3 > 0.95) %>% mutate(x=score_median, y=if_else(group=="COSMIC",3,if_else(group=="AGO",2,1))) %>% select(x,y)
+# pDB + geom_segment(data=marks2, aes(x=x, xend=x, y=y, yend=y+0.2), color="red") + geom_segment(data=marks, aes(x=x, xend=x, y=y, yend=y+0.1), color="black", alpha=0.5)
+pDB
 # ggsave("database_violins.svg")
+ggsave("database_boxplot.svg")
 

### 2021-02-19 prior_sensitivity.Rmd
added boxplots

@@
 ```
+
+```{r posterior_sensitivity, echo=FALSE, comment=NA}
+
+n_eff_mat <- data.frame(Params=c("mu","nu","sigmaP","sigmaPT","sigmaT","sigmaE","lp"))
+Rhat_mat <- n_eff_mat
+KLdiv_mat <- n_eff_mat %>% slice(1:6)
+BCdist_mat <- KLdiv_mat
+EMdist_mat <- KLdiv_mat
+
+stanFile = "FullResultsTCGA_relaxgamma3.Rdata"
+load(here(stanFile))
+
+mu_ref <- mu_full
+nu_ref <- nu_full
+sigmaP_ref <- sigmaP_full
+sigmaPT_ref <- sigmaPT_full
+sigmaT_ref <- sigmaT_full
+sigmaE_ref <- sigmaE_full
+
+for (i in c(1:4)) {
+  stanFile = switch(i,
+                    "FullResultsTCGA_gamma.Rdata",
+                    "FullResultsTCGA_relaxgamma2.Rdata",
+                    "FullResultsTCGA_relaxgamma3.Rdata",
+                    "FullResultsTCGA_relaxgamma5.Rdata")
+  print(stanFile)
+  load(here(stanFile))
+  print("loaded")
+
+  n_eff_temp <-
+  data.frame(
+  mu = mu_full$n_eff,
+  nu = nu_full$n_eff,
+  sigmaP = sigmaP_full$n_eff,
+  sigmaPT = sigmaPT_full$n_eff,
+  sigmaT = sigmaT_full$n_eff,
+  sigmaE = sigmaE_full$n_eff,
+  lp = lp_full$n_eff
+  ) %>% gather(param,value,mu:lp) %>% group_by(param) %>% summarize(
+  mean = mean(value),
+  .groups = "drop"
+  ) %>% arrange(match(.$param,c("mu","nu","sigmaP","sigmaPT","sigmaT","sigmaE","lp")))
+
+  Rhat_temp <-
+  data.frame(
+  mu = mu_full$Rhat,
+  nu = nu_full$Rhat,
+  sigmaP = sigmaP_full$Rhat,
+  sigmaPT = sigmaPT_full$Rhat,
+  sigmaT = sigmaT_full$Rhat,
+  sigmaE = sigmaE_full$Rhat,
+  lp = lp_full$Rhat
+  ) %>% gather(param,value,mu:lp) %>% group_by(param) %>% summarize(
+  mean = mean(value),
+  .groups = "drop"
+  ) %>% arrange(match(.$param,c("mu","nu","sigmaP","sigmaPT","sigmaT","sigmaE","lp")))
+
+  # KL divergence
+  # formula: KL(P||Q) = sum(P(i) * log( P(i)/Q(i) )), where P(i),Q(i) = density
+  KL_temp <- numeric(6)
+  #Bhattacharya distance
+  # formula: BC = sum( sqrt(P(i)*Q(i)) )
+  BC_temp <- numeric(6)
+  # Earth-movers (Wasserstein) distance
+  EM_temp <- numeric(6)
+
+  for (j in c(1:6)) {
+    # choose parameter (first get reference dist (3x), then prior relaxed dist)
+    breaks <- switch(j,
+                     seq(-5,5,0.1),
+                     seq(-6,6,0.12), #5.5,.11
+                     seq(0,3,0.03), #2.6
+                     seq(0,3,0.03),
+                     seq(0,3,0.03), #1.2
+                     seq(0,3,0.03)) #2.5
+
+    refdata <- switch(j,
+                   mu_ref$p50,
+                   nu_ref$p50,
+                   sigmaP_ref$p50,
+                   sigmaPT_ref$p50,
+                   sigmaT_ref$p50,
+                   sigmaE_ref$p50)
+    Q <- hist(refdata, plot=FALSE, breaks=breaks) # run hist to get density and midpoints
+    x <- Q$mids
+
+    data <- switch(j,
+                   mu_full$p50,
+                   nu_full$p50,
+                   sigmaP_full$p50,
+                   sigmaPT_full$p50,
+                   sigmaT_full$p50,
+                   sigmaE_full$p50)
+    P <- hist(data, plot=FALSE, breaks=breaks) # run hist to get density and midpoints
+
+    Pi <- P$density[which(P$density>0 & Q$density>0)]
+    Qi <- Q$density[which(P$density>0 & Q$density>0)]
+    KL_temp[j] = sum(Pi * log(Pi/Qi))
+
+    # BCD
+    dx <- x[2]-x[1]
+    BC_temp[j] = -log(sum(sqrt(P$density*Q$density)*dx))
+
+    # EMD
+    A = cbind(P$density,P$mids)
+    B = cbind(Q$density,Q$mids)
+    EM_temp[j] = emd(A,B)
+
+    # code below is for plotting data vs priors
+    #Q <- P
+    #Q$density <- priordens
+    #Q$counts <- Q$density*866091*(Q$breaks[2]-Q$breaks[1])
+    #c1 <- rgb(173,216,230,max = 255, alpha = 80, names = "lt.blue")
+    #c2 <- rgb(255,192,203, max = 255, alpha = 80, names = "lt.pink")
+    #plot(P,col=c1, main=paste("i=",i,", j=",j))
+    #plot(Q,col=c2,add=TRUE)
+  }
+
+  varname = switch(i, "1x", "2x", "3x", "5x")
+  n_eff_mat <- n_eff_mat %>% mutate(!!varname := n_eff_temp$mean)
+  Rhat_mat <- Rhat_mat %>% mutate(!!varname := Rhat_temp$mean)
+  KLdiv_mat <- KLdiv_mat %>% mutate(!!varname := KL_temp)
+  BCdist_mat <- BCdist_mat %>% mutate(!!varname := BC_temp)
+  EMdist_mat <- EMdist_mat %>% mutate(!!varname := EM_temp)
+
+}
+
+print("N_eff")
+kable(n_eff_mat, digits=2)
+print("Rhat")
+kable(Rhat_mat, digits=3)
+print("KLdiv");kable(KLdiv_mat, digits=3);print("BCdist");kable(BCdist_mat, digits=3);print("EMdist"); kable(EMdist_mat, digits=3)
+
+```

### 2021-07-07 Paper_Figures.Rmd
updates to figures and code

@@
 load(here(stanFile))
-CpGscore_full = CpGscore_full %>% mutate(mean = mean*2, median=median*2)
 
@@
 permed_data <- read_csv(here(prepdatafolder, "permed_data.csv"))
-
-gened_data <- gened_data %>% mutate(score_median = score_median*2)
-prepped_data <- prepped_data %>% mutate(score_median = score_median*2)
-permed_data <- permed_data %>% mutate(score_median = score_median*2)
-
 ```
@@
 
-\subsection{CpG Conservation score}
-We want to understand which sites of DNA methylation are fundamentally conserved within tumors, i.e.
-lower variation in methlyation within tumor (sigmaT) relative to the variation within normal healthy tissue (sigmaP).
-Therefore we choose the log-ratio (base 2) of sigmaP/sigmaT to represent a relative conservation score, which is positive when tumor variation is lower than normal variation.
-We take the median of the posterior distribution constructed from the two sigmaP,T posteriors for this score.
-
-$$\text{score}_\text{med} = \text{Median} (\text{log}_2\frac{\sigma_{P}}{\sigma_T}) $$
-
 \newpage
-\subsection{Figure 2A: Example site fit, 2B: Posteriors}
+\subsection{Figure 2A: Example site fit, 2B: Example site posteriors}
 ![Fit](/Users/kevinmurgas/Documents/Data+ project/Manuscript/Paper Figures (screenshots)/example_fit.png)
@@
 ![Posteriors](/Users/kevinmurgas/Documents/Data+ project/Manuscript/Paper Figures (screenshots)/example_posteriors.png)
-
-\newpage
-\section{Model Fit Results}
-\subsection{Fig 3C: Parameter Distributions}
-Histograms are median values of posterior distribution
-```{r Params_site, echo=FALSE, fig.fullwidth=TRUE, warning=FALSE, comment=NA}
-genemask = !(FullAnnotation$UCSC_RefGene_Name=="")
-cat(sprintf("N sites: total=%i, gene-associated=%i, non-gene=%i  \n",length(genemask),sum(genemask),length(genemask)-sum(genemask)))
-
-fBG <- function(x, a, m1, s1, m2, s2) a*dnorm(x,m1,s1) + (1-a)*dnorm(x,m2,s2) # bimodal gaussian
-post_prior_histogram <- function(postvals,priorfun,bw,xl,titlestr) {
-  # histogram and prior curve overlaid
-  plot_df <- data.frame(val=postvals)
-  p1 <- ggplot(plot_df, aes(x=val)) +
-    geom_histogram(aes(y=..density..), binwidth=bw) +
-    stat_function(fun=function(x) bw*priorfun(x),geom="line",color="black") +
-    ggtitle(titlestr) + xlim(xl) + theme_classic() + theme(legend.position = "none") + xlab(titlestr) + ylab("Freq.")
-  p1
-}
-
-# plot hierarchical model posterior medians for mu,nu,sigmaE,sigmaP,sigmaPT,sigmaT; with priors overlaid
-pM <- post_prior_histogram(mu_full$p50, function(x) 10*fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), 0.1, c(-5,5), "Mu")
-pN <- post_prior_histogram(nu_full$p50, function(x) 15*dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "Nu")
-#pE <- post_prior_histogram(sigmaE_full$p50, function(x) 4*dgamma(x,2,2), 0.05, c(0,2), "SigmaE")
-#pP <- post_prior_histogram(sigmaP_full$p50^2, function(x) 2*dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
-#pPT <- post_prior_histogram(sigmaPT_full$p50^2, function(x) 1.8*dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
-#pT <- post_prior_histogram(sigmaT_full$p50^2, function(x) 1.2*dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT")
-pE <- post_prior_histogram(sigmaE_full$p50^2, function(x) 100*dgamma(sqrt(x),2,2), 0.05, c(0,2), "SigmaE^2")
-pP <- post_prior_histogram(sigmaP_full$p50^2, function(x) 250*dgamma(sqrt(x), 2.7693, 7.0029), 0.01, c(0,1), "SigmaP^2")
-pPT <- post_prior_histogram(sigmaPT_full$p50^2, function(x) 40*dgamma(sqrt(x), 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT^2")
-pT <- post_prior_histogram(sigmaT_full$p50^2, function(x) 220*dgamma(sqrt(x), 3.3643, 10.430), 0.01, c(0,1), "SigmaT^2")
-# log sigmas
-#pLP <- post_prior_histogram(log(sigmaP_full$p50), genemask, function(x) 0.52*dgamma(exp(x), 2.7693, 7.0029), 0.1, c(-5,5), "log(SigmaP)")
-#pLPT <- post_prior_histogram(log(sigmaPT_full$p50), genemask, function(x) 0.7*dgamma(exp(x), 2.1457, 2.4819), 0.1, c(-5,5), "log(SigmaPT)")
-#pLT <- post_prior_histogram(log(sigmaT_full$p50), genemask, function(x) 0.4*dgamma(exp(x), 3.3643, 10.430), 0.1, c(-5,5), "log(SigmaT)") + theme(legend.position = c(0.8, 0.6), legend.text=element_text(size=6), legend.title=element_text(size=6))
-
-# arrange each parameter as a subplot # ,labels=c("M","B","E","P","PT","T","lP","lPT","lT")
-p_all <- ggarrange(pM,pN,pE,pP,pPT,pT,hjust=0,nrow=2,ncol=3)
-print(p_all)
-rm(pM,pN,pE,pP,pPT,pT)
-```
 
@@
 cat("  \n")
-
-# indices to filter out sites Rhat>1,1
-QCinds <- lp_full$Rhat > 1.1
 ```
 
-For the remainder of these results, a filter is applied to remove all sites with Rhat > 1.1 in lp (log-posterior likelihood of entire site fit).
-Another filter is later applied to remove all genes with <5 CpG sites.
+\subsection{Filter CpG sites}
+A filter is applied to remove all sites with Rhat > 1.1 in lp (log-posterior likelihood of entire site fit), and CpG sites mapping to sex chromosomes X and Y are removed as well to eliminate sex-specific bias.
+
+Later, a gene filter is applied to remove all genes with <5 CpG sites.
+
+```{r filter_sites, echo=FALSE, warning=FALSE}
+# indices to filter out sites Rhat>1,1
+QCinds <- which(lp_full$Rhat > 1.1)
+# indices to filter out X and Y chromosome
+XYinds <- which(FullAnnotation$CHR %in% c("X","Y"))
+
+# remove the corresponding rows from the results dataframes
+FullAnnotation <- FullAnnotation[-union(QCinds,XYinds),]
+mu_full <- mu_full[-union(QCinds,XYinds),]
+nu_full <- nu_full[-union(QCinds,XYinds),]
+sigmaP_full <- sigmaP_full[-union(QCinds,XYinds),]
+sigmaPT_full <- sigmaPT_full[-union(QCinds,XYinds),]
+sigmaT_full <- sigmaT_full[-union(QCinds,XYinds),]
+sigmaE_full <- sigmaE_full[-union(QCinds,XYinds),]
+lp_full <- lp_full[-union(QCinds,XYinds),]
+CpGscore_full <- CpGscore_full[-union(QCinds,XYinds),]
+```
+
+\subsection{Figure 2C: Posterior medians over all CpG sites}
+```{r posteriors, echo=FALSE, fig.fullwidth=TRUE, warning=FALSE, comment=NA}
+# for each parameter, plot the distribution and overlay the fits
+fBG <- function(x, a, m1, s1, m2, s2) a*dnorm(x,m1,s1) + (1-a)*dnorm(x,m2,s2) # bimodal gaussian
+post_prior_histogram <- function(postvals,priorfun,bw,xl,titlestr) {
+  # histogram with prior curve overlaid
+  plot_df <- data.frame(val=postvals)
+  p1 <- ggplot(plot_df, aes(x=val)) +
+    geom_histogram(aes(y=..density..), binwidth=bw, position="dodge", fill="lightgray") +
+    stat_function(fun=function(x) priorfun(x), geom="line", color="black") +
+    ggtitle(titlestr) + xlim(xl) + theme_classic() + theme(legend.position = "none") + xlab(titlestr) + ylab("Freq.")
+  p1
+}
+
+# plot mu,nu,sigmaE,sigmaP,sigmaPT,sigmaT with priors overlaid
+pM <- post_prior_histogram(mu_full$p50, function(x) 1.3*fBG(x,0.3557,-3.0675, 1.3027, 1.1290, 2.4110), 0.25, c(-5,5), "Mu") + theme(aspect.ratio = 1)
+pN <- post_prior_histogram(nu_full$p50, function(x) 1.5*dcauchy(x,-0.04051, 0.5799), 0.25, c(-5,5), "Nu") + theme(aspect.ratio = 1)
+pE <- post_prior_histogram(sigmaE_full$p50, function(x) 3.3*dgamma(x,2,2), 0.05, c(0,2), "SigmaE") + theme(aspect.ratio = 1)
+pP <- post_prior_histogram(sigmaP_full$p50, function(x) 1.52*dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP") + theme(aspect.ratio = 1)
+pPT <- post_prior_histogram(sigmaPT_full$p50, function(x) 1.6*dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT") + theme(aspect.ratio = 1)
+pT <- post_prior_histogram(sigmaT_full$p50, function(x) 0.97*dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT") + theme(aspect.ratio = 1)
+
+p_all <- ggarrange(pM,pN,pE,pP,pPT,pT,hjust=0,nrow=2,ncol=3)
+print(p_all)
+rm(pM,pN,pE,pP,pPT,pT)
+
+ggsave(filename = "posteriors.svg", plot=p_all)
+```
+
+\newpage
+\section{Model Fit Results}
+\subsection{Fig 3C: Parameter Distributions}
+Histograms are median values of posterior distribution
+```{r Params_site, echo=FALSE, fig.fullwidth=TRUE, warning=FALSE, comment=NA}
+genemask = !(FullAnnotation$UCSC_RefGene_Name=="")
+cat(sprintf("N sites: total=%i, gene-associated=%i, non-gene=%i  \n",length(genemask),sum(genemask),length(genemask)-sum(genemask)))
+
+fBG <- function(x, a, m1, s1, m2, s2) a*dnorm(x,m1,s1) + (1-a)*dnorm(x,m2,s2) # bimodal gaussian
+post_prior_histogram <- function(postvals,priorfun,bw,xl,titlestr) {
+  # histogram and prior curve overlaid
+  plot_df <- data.frame(val=postvals)
+  p1 <- ggplot(plot_df, aes(x=val)) +
+    geom_histogram(aes(y=..density..), binwidth=bw) +
+    stat_function(fun=function(x) bw*priorfun(x),geom="line",color="black") +
+    ggtitle(titlestr) + xlim(xl) + theme_classic() + theme(legend.position = "none") + xlab(titlestr) + ylab("Freq.")
+  p1
+}
+
+# plot hierarchical model posterior medians for mu,nu,sigmaE,sigmaP,sigmaPT,sigmaT; with priors overlaid
+pM <- post_prior_histogram(mu_full$p50, function(x) 10*fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), 0.1, c(-5,5), "Mu")
+pN <- post_prior_histogram(nu_full$p50, function(x) 15*dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "Nu")
+#pE <- post_prior_histogram(sigmaE_full$p50, function(x) 4*dgamma(x,2,2), 0.05, c(0,2), "SigmaE")
+#pP <- post_prior_histogram(sigmaP_full$p50^2, function(x) 2*dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
+#pPT <- post_prior_histogram(sigmaPT_full$p50^2, function(x) 1.8*dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
+#pT <- post_prior_histogram(sigmaT_full$p50^2, function(x) 1.2*dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT")
+pE <- post_prior_histogram(sigmaE_full$p50^2, function(x) 100*dgamma(sqrt(x),2,2), 0.05, c(0,2), "SigmaE^2")
+pP <- post_prior_histogram(sigmaP_full$p50^2, function(x) 250*dgamma(sqrt(x), 2.7693, 7.0029), 0.01, c(0,1), "SigmaP^2")
+pPT <- post_prior_histogram(sigmaPT_full$p50^2, function(x) 40*dgamma(sqrt(x), 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT^2")
+pT <- post_prior_histogram(sigmaT_full$p50^2, function(x) 220*dgamma(sqrt(x), 3.3643, 10.430), 0.01, c(0,1), "SigmaT^2")
+
+# arrange each parameter as a subplot # ,labels=c("M","B","E","P","PT","T","lP","lPT","lT")
+p_all <- ggarrange(pM,pN,pE,pP,pPT,pT,hjust=0,nrow=2,ncol=3)
+print(p_all)
+rm(pM,pN,pE,pP,pPT,pT)
+```
+
+\subsection{CpG Conservation score}
+We want to understand which sites of DNA methylation are fundamentally conserved within tumors, i.e.
+lower variation in methlyation within tumor (sigmaT) relative to the variation within normal healthy tissue (sigmaP).
+Therefore we choose the log-ratio (base 2) of sigmaP/sigmaT to represent a relative conservation score, which is positive when tumor variation is lower than normal variation.
+We take the median of the posterior distribution constructed from the two sigmaP,T posteriors for this score.
+
+$$\text{score}_\text{med} = \text{Median} (\text{log}_2\frac{\sigma_P^2}{\sigma_T^2}) $$
 
@@
 # overlaid transparent histograms
-post_histogram <- function(postvals,mask,bw,titlestr) {
-  # transparent histogram colored by gene-associated sites
-  plot_df <- data.frame(val=postvals, gene_assoc=mask)
-  gene_assoc_df <- plot_df %>% filter(gene_assoc)
-  not_gene_assoc_df <- plot_df %>% filter(!gene_assoc)
-  p1 <- ggplot(plot_df, aes(x=val)) +
-    geom_histogram(data=not_gene_assoc_df, aes(y=..density..,fill="#00BFC4",alpha=0.4), binwidth=bw) +
-    geom_histogram(data=gene_assoc_df, aes(y=..density..,fill="#F8766D",alpha=0.4), binwidth=bw) +
-    ggtitle(titlestr) + xlab(titlestr) + ylab("Density") + theme_classic() 
-  p1
-}
+# post_histogram <- function(postvals,mask,bw,titlestr) {
+#   # transparent histogram colored by gene-associated sites
+#   plot_df <- data.frame(val=postvals, gene_assoc=mask)
+#   gene_assoc_df <- plot_df %>% filter(gene_assoc)
+#   not_gene_assoc_df <- plot_df %>% filter(!gene_assoc)
+#   p1 <- ggplot(plot_df, aes(x=val)) +
+#     geom_histogram(data=not_gene_assoc_df, aes(y=..density..,fill="#00BFC4",alpha=0.4), binwidth=bw) +
+#     geom_histogram(data=gene_assoc_df, aes(y=..density..,fill="#F8766D",alpha=0.4), binwidth=bw) +
+#     ggtitle(titlestr) + xlab(titlestr) + ylab("Density") + theme_classic() 
+#   p1
+# }
 
-# plot the two conservations score prob sums
+# plot conservation score
 genemask = !(FullAnnotation$UCSC_RefGene_Name=="")
@@
 print(p_scores)
+ggsave("CpGscores_dodge.svg")
 
+# same data but as boxplot
 df = data.frame(med=CpGscore_full$median, mask=genemask)
@@
 print(p_scores)
-# ggsave("CpGscores_overlay.svg")
+
+# statistics: two-sample wilcoxon rank-sum
+wilcox.test(x=CpGscore_full$median[genemask], y=CpGscore_full$median[!genemask], alternative = "greater")
 
 # Fig 4B: plot gene scores and adjusted bootstrap distribution
-p1 <- permed_data %>% ggplot(aes(x=score_median)) + geom_histogram(aes(y=..count..), binwidth=0.1, fill="lightgray",color = "grey50") + xlim(c(-3,7)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
+p1 <- permed_data %>% ggplot(aes(x=score_median)) + geom_histogram(aes(y=..count..), binwidth=0.2, fill="lightgray",color = "grey50") + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1) + scale_x_continuous(breaks = c(-2,0,2,4,6), limits = c(-3,7)) + scale_y_continuous(breaks = c(0,1000,2000,3000), limits = c(0,3000))
 p2 <- permed_data %>% ggplot(aes(x=score_median_v3)) + geom_histogram(aes(y=..count..), binwidth=0.02, fill="lightgray", color="grey50") + xlim(c(0,1)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
 p_genes <- ggarrange(p1,p2,nrows=1,ncol=2)
-# ggsave("gene_scores_bootstrap.svg")
+ggsave("gene_scores_median.svg", plot=p1)
+ggsave("gene_scores_bootstrap.svg", plot=p2)
 ```
@@
 ggsave("CpGscore_regions_boxplot.svg")
-#rm(temp)3
+#rm(temp)
 
@@
 cat("\n")
-# ggsave("CpGdist.svg")
+ggsave("CpGdist.svg")
 ```
@@
 #ggsave("genescore_HLA-A.svg")
+cat("\n")
+plotGeneScore("APC",0.12) + theme(aspect.ratio = 1)
+#ggsave("genescore_APC.svg")
+cat("\n")
+plotGeneScore("KRAS",0.18) + theme(aspect.ratio = 1)
+#ggsave("genescore_KRAS.svg")
 
@@
 cancergenes2 <- read_csv("/Users/kevinmurgas/Documents/Data+ project/Cancer gene lists/CRCgenes_atlasgeneticsoncology.csv")
+essentialgenes <- read_delim("/Users/kevinmurgas/Documents/Data+ project/Manuscript/Bioinformatics Revision/core-essential-genes-sym_HGNCID.txt", delim = "\t", col_names = F)
 
@@
 # plot violins of: all genes, COSMIC, AGO
-perm_temp <- permed_data %>% select(c(UCSC_RefGene_Name, score_median,score_median_v3,n))
-genes_dbs <- rbind(perm_temp %>% mutate(group="all"),
+perm_temp <- permed_data %>% select(c(UCSC_RefGene_Name, score_median,score_median_v3,n)) %>% mutate(score_median = 2*score_median)
+genes_dbs <- rbind(perm_temp %>% mutate(group="all genes"),
               perm_temp %>% filter(UCSC_RefGene_Name %in% (cancergenes1 %>% pull(`Gene Symbol`))) %>% mutate(group="COSMIC"),
-              perm_temp %>% filter(UCSC_RefGene_Name %in% (cancergenes2 %>% pull(CancerGenes))) %>% mutate(group="AGO")
+              perm_temp %>% filter(UCSC_RefGene_Name %in% (cancergenes2 %>% pull(CancerGenes))) %>% mutate(group="AGO"),
+              perm_temp %>% filter(UCSC_RefGene_Name %in% (essentialgenes %>% pull(X1))) %>% mutate(group="DEPMAP")
               )
 # Fig 4C: plot distribution
-pDB <- genes_dbs %>% mutate(group = factor(group, levels = c("all", "AGO", "COSMIC"))) %>%
+pDB <- genes_dbs %>% mutate(group = factor(group, levels = c("all genes", "AGO", "COSMIC", "DEPMAP"))) %>%
   #ggplot() + geom_violin(aes(x=score_median,y=group,fill=group), draw_quantiles = 0.5) +
   ggplot() + geom_boxplot(aes(x=score_median,y=group,fill=group), outlier.shape = NA) +
-  scale_x_continuous(limits=c(-2,4),breaks = c(-2,0,2,4)) + theme_classic() + ylab("Database") + xlab("Gene Conservation Score") + geom_vline(xintercept=0, color="black", alpha=0.2) + theme(aspect.ratio=60/156.3,legend.position = "none")
+  scale_x_continuous(limits=c(-5,7), breaks = c(-4, -2,0, 2, 4, 6)) + theme_classic() + ylab("Database") + xlab("Gene Conservation Score") + geom_vline(xintercept=0, color="black", alpha=0.2) + theme(aspect.ratio=60/156.3,legend.position = "none")
 # # add hashmarks for each gene (doesn't look great)
@@
 # ggsave("database_violins.svg")
-ggsave("database_boxplot.svg")
+ggsave("database_boxplot_depmap.svg")
+
+genes_dbs %>% group_by(group) %>% summarize(mean = mean(score_median))
 
 # stats
-wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="COSMIC"], y = genes_dbs$score_median[genes_dbs$group=="all"], alternative = "greater")
-wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="AGO"], y = genes_dbs$score_median[genes_dbs$group=="all"], alternative = "greater")
+wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="COSMIC"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
+wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="AGO"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
+wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="DEPMAP"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
+```
+
+\newpage
+\subsection{Oncogenes vs Tumor suppressors}
+Based on list of top 20 genes mutated in CRC, manually labelled as oncogene or tumor-suppressor:
+
+| Oncogene | Tumor Suppressor |
+|----------|------------------|
+| KRAS     | TP53             |
+| MYC      | APC              |
+| BRAF     | KLF4             |
+| EGFR     | FBXW7            |
+| CDK8     | LRP1B            |
+| PIK3CA   | PTEN             |
+| NRAS     | FAT4             |
+| GNAS     | ARID1A           |
+| SRC      | AMER1            |
+| CTNNB1   | SMAD4            |
+
+```{r Onco_vs_TSG, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=2, comment=NA, fig.fullwidth=TRUE, dev = "png"}
+
+plotGeneScore <- function(geneStr,norm) {
+  cat("\n")
+  CpGscores <- gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^",geneStr,"$"))) %>% pull(score_median)
+  ggplot(prepped_data) + geom_histogram(aes(x=score_median, y=..count../nrow(prepped_data)), binwidth=0.05,fill="lightgray") + ggtitle(sprintf("%s: CpG scores (%i sites)",geneStr, length(CpGscores))) +
+    scale_x_continuous(limits=c(-5,5), breaks=c(-4,-2,0,2,4)) +
+    scale_y_continuous(limits=c(0,0.03),breaks=c(0,0.01,0.02,0.03)) +
+    xlab("CpG conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") +
+    #geom_vline(xintercept=CpGscores, color="red", alpha=0.5, size=0.2) +
+    geom_histogram(data=data.frame(score=CpGscores),aes(x=score,y=norm*..count../length(CpGscores)),binwidth=0.05, fill="blue",alpha=0.5) +
+    geom_vline(xintercept=mean(CpGscores), color="blue",linetype="dashed", size=1)
+}
+
+cat("\nOncogenes:")
+plotGeneScore("KRAS",0.25)
+plotGeneScore("MYC",0.21)
+plotGeneScore("BRAF",0.5)
+plotGeneScore("EGFR",0.5)
+plotGeneScore("CDK8",0.5)
+plotGeneScore("PIK3CA",0.5)
+plotGeneScore("NRAS",0.5)
+plotGeneScore("GNAS",0.5)
+plotGeneScore("SRC",0.5)
+plotGeneScore("CTNNB1",0.5)
+
+cat("\nTumor Suppressors:")
+plotGeneScore("TP53",0.25)
+plotGeneScore("APC",0.21)
+plotGeneScore("KLF4",0.5)
+plotGeneScore("FBXW7",0.5)
+plotGeneScore("LRP1B",0.5)
+plotGeneScore("PTEN",0.5)
+plotGeneScore("FAT4",0.5)
+plotGeneScore("ARID1A",0.5)
+plotGeneScore("AMER1",0.5)
+plotGeneScore("SMAD4",0.5)
+
 ```

### 2021-07-07 StanAnalysis.Rmd
updates to figures and code

@@
 
-\subsection{Conservation scores}
+\subsection{Conservation score}
 ```{r Fig1B, echo=FALSE, fig.height=2, fig.width=6.5, fig.fullwidth=TRUE, warning=FALSE}
@@
 
-# plot the two conservations score prob sums
+# plot conservation score
 pP1 <- post_histogram(CpGscore_full$median,genemask,0.1,c(-3,3),expression("Median log("~sigma[P]/sigma[T]~")")) + scale_x_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3))
@@
 cat("  \n")
-
-# indices to filter out sites Rhat>1,1
-QCinds <- lp_full$Rhat > 1.1
 ```
 
-For the remainder of these results, a filter is applied to remove all sites with Rhat > 1.1 in lp (log-posterior likelihood of entire site fit).
-Another filter is later applied to remove all genes with <5 CpG sites.
+\subsection{Filter CpG sites}
+A filter is applied to remove all sites with Rhat > 1.1 in lp (log-posterior likelihood of entire site fit), and CpG sites mapping to sex chromosomes X and Y are removed as well to eliminate sex-specific bias.
+
+Later, a gene filter is applied to remove all genes with <5 CpG sites.
+
+```{r filter_sites, echo=FALSE, warning=FALSE}
+# indices to filter out sites Rhat>1,1
+QCinds <- which(lp_full$Rhat > 1.1)
+# indices to filter out X and Y chromosome
+XYinds <- which(FullAnnotation$CHR %in% c("X","Y"))
+
+# remove the corresponding rows from the results dataframes
+FullAnnotation <- FullAnnotation[-union(QCinds,XYinds),]
+mu_full <- mu_full[-union(QCinds,XYinds),]
+nu_full <- nu_full[-union(QCinds,XYinds),]
+sigmaP_full <- sigmaP_full[-union(QCinds,XYinds),]
+sigmaPT_full <- sigmaPT_full[-union(QCinds,XYinds),]
+sigmaT_full <- sigmaT_full[-union(QCinds,XYinds),]
+sigmaE_full <- sigmaE_full[-union(QCinds,XYinds),]
+lp_full <- lp_full[-union(QCinds,XYinds),]
+CpGscore_full <- CpGscore_full[-union(QCinds,XYinds),]
+```
 
@@
 ```
-
-\newpage
-\subsection{Figure 1: CpG Distance Analysis}
-Corrected for gene direction, based on Ensembl gene database
-```{r Fig1, echo=FALSE, warning=FALSE, fig.width=7, fig.height=2.5, comment=NA}
-# load ensembl db
-library(EnsDb.Hsapiens.v75) # ensembl annotation with gene position (hg19 version)
-edb <- EnsDb.Hsapiens.v75
-geneKeys <- keys(edb, keytype="GENENAME") %>% intersect(unique(gened_data$UCSC_RefGene_Name))
-edbData<- AnnotationDbi::select(edb, keys=geneKeys, keytype="GENENAME",
-                                columns=c("GENEBIOTYPE", "GENESEQSTART", "GENESEQEND", "SEQNAME", "SEQSTRAND")) %>%
-  filter(SEQNAME %in% c(as.character(1:22),"X","Y")) %>%
-  mutate(GENESTART = GENESEQSTART*(SEQSTRAND==1) + GENESEQEND*(SEQSTRAND==-1)) %>%
-  select(-c(GENESEQSTART,GENESEQEND))
-
-# calculate CpG distance based on Ensembl gene start position and direction
-distanced_data <- gened_data %>%
-  filter(!is.na(UCSC_RefGene_Name)) %>%
-  filter(UCSC_RefGene_Name %in% geneKeys) %>%
-  mutate(EDBCHR=edbData$SEQNAME[match(UCSC_RefGene_Name,edbData$GENENAME)],
-         EDBSTART=edbData$GENESTART[match(UCSC_RefGene_Name,edbData$GENENAME)],
-         EDBDIR=edbData$SEQSTRAND[match(UCSC_RefGene_Name,edbData$GENENAME)],
-         EDBTYPE=edbData$GENEBIOTYPE[match(UCSC_RefGene_Name,edbData$GENENAME)]
-         ) %>%
-  filter(CHR==EDBCHR) %>%
-  mutate(cpgdist = if_else(EDBDIR==1, MAPINFO-EDBSTART, EDBSTART-MAPINFO))
-
-distance_length <- distanced_data %>%
-  filter(!is.na(UCSC_RefGene_Name)) %>%
-  select(CHR, UCSC_RefGene_Name, cpgdist, c(sigmaP:sigmaT,score_median)) %>%
-  mutate(location_bin = cut(cpgdist, c(seq(-1500, 20000, by = 100)), include.lowest = TRUE)) %>%
-  group_by(location_bin) %>%
-  summarise(mean_sigmaP = mean(sigmaP), mean_sigmaPT = mean(sigmaPT), mean_sigmaT = mean(sigmaT),  mean_score_median = mean(score_median), n_obs = n(), .groups='drop') %>%
-  mutate(location_bin = as.numeric(str_extract(location_bin, "(?<=,)(.*)(?=])")))
-
-distance_length %>%
-  drop_na() %>%
-  pivot_longer(mean_sigmaP:mean_sigmaT) %>% 
-  mutate(name = factor(name, c("mean_sigmaP","mean_sigmaPT","mean_sigmaT"), c("SigmaP","SigmaPT","SigmaT"))) %>%
-  sample_frac() %>%
-  ggplot(aes(location_bin, value, color = name, fill = name)) +
-  geom_point(key_glyph = draw_key_rect) +
-  ggforce::facet_zoom(location_bin < 5000) +
-  theme_light() +
-  theme(legend.position = "bottom") +
-  labs(title = NULL,
-       fill = NULL,
-       color = NULL,
-       x = "Distance from gene start site (binned to 100 bp)",
-       y = "Average Sigma Value", 
-       caption = "Figure 1: Average Sigma values of single CpGs as a function of position relative to gene.") +
-  scale_color_brewer(palette = "Set2") +
-  scale_fill_brewer(palette = "Set2")
-cat("\n")
-
-# again for conservation scores
-distance_length %>%
-  drop_na() %>%
-  pivot_longer(mean_score_median) %>% 
-  #mutate(name = factor(name, c("mean_score_median"), c("Score_median"))) %>%
-  sample_frac() %>%
-  ggplot(aes(location_bin, value, color = name, fill = name)) +
-  geom_point(key_glyph = draw_key_rect) +
-  ggforce::facet_zoom(location_bin < 5000) +
-  theme_light() +
-  theme(legend.position = "bottom") +
-  labs(title = NULL,
-       fill = NULL,
-       color = NULL,
-       x = "Distance from gene start site (binned to 100 bp)",
-       y = "Average Conservation Score", 
-       caption = "Figure 1: Average conservation scores of single CpGs as a function of position relative to gene."
-       ) +
-  scale_color_brewer(palette = "Set2") +
-  scale_fill_brewer(palette = "Set2")
-cat("\n")
-# ggsave("CpGdist.svg")
-
-# and look at number of CpG in each bin
-distance_length %>%
-  drop_na() %>%
-  pivot_longer(n_obs) %>%
-  mutate(name = factor(name, c("n_obs"), c("n_obs"))) %>%
-  sample_frac() %>%
-  ggplot(aes(location_bin, value, color = name, fill = name)) +
-  geom_point(key_glyph = draw_key_rect) +
-  ggforce::facet_zoom(location_bin < 5000) +
-  theme_light() +
-  theme(legend.position = "bottom") +
-  labs(title = NULL,
-       fill = NULL,
-       color = NULL,
-       x = "Distance from gene start site (binned to 100 bp)",
-       y = "Number of CpGs", 
-       caption = "Figure 1: Frequency of single CpGs as a function of position relative to gene start site.") +
-  scale_color_brewer(palette = "Set2") +
-  scale_fill_brewer(palette = "Set2")
-```
-
-\subsection{Figure 2: Bootstrapping}
-```{r Fig2, echo=FALSE, warning=FALSE, fig.height=4, comment=NA}
-# Figure 2
-permed_data %>%
-  pivot_longer(dplyr::matches("1|3")) %>%
-  separate(name, into = c("type", "method"), sep = "_v") %>%
-  mutate(method = factor(method, c("1", "3"), c("Naive bootstrap", "Adjusted bootstrap")),
-         type = factor(type, c("sigmaP","sigmaPT","sigmaT"), c("SigmaP","SigmaPT","SigmaT"))) %>%
-  drop_na() %>%
-  ggplot(aes(value, fill = type)) +
-  geom_histogram(bins = 50, color = "grey30") +
-  facet_grid(method~type, scales = "free") +
-  theme_minimal() +
-  labs(x = "p-value",
-       y = "Number of genes",
-       title = NULL,
-       caption = "Figure 2A: Distribution of boot-strapping p-values for genes as a function of sigma level and bootstrap method.")  +
-  theme(legend.position = "none") +
-  scale_color_brewer(palette = "Set2") +
-  scale_fill_brewer(palette = "Set2")
-
-cat("***")
-permed_data %>%
-  pivot_longer(dplyr::matches("1|3")) %>%
-  separate(name, into = c("type", "method"), sep = "_v") %>%
-  mutate(method = factor(method, c("1", "3"), c("Naive bootstrap", "Adjusted bootstrap")),
-         type = factor(type, c("score_median"), c("Score_median"))) %>%
-  drop_na() %>%
-  ggplot(aes(value, fill = type)) +
-  geom_histogram(bins = 50, color = "grey30") +
-  facet_grid(method~type, scales = "free") +
-  theme_minimal() +
-  labs(x = "p-value",
-       y = "Number of genes",
-       title = NULL,
-       caption = "Figure 2B: Distribution of boot-strapping p-values for genes as a function of conservation score and bootstrap method.")  +
-  theme(legend.position = "none") +
-  scale_color_brewer(palette = "Set2") +
-  scale_fill_brewer(palette = "Set2")
-```

### 2021-07-07 PrepareDataFromGEO.R

@@ -31,7 +31,7 @@ selectedCols <- c("character", rep("NULL", 10), "character", "integer", "NULL", 
 EPICchar <- read.csv("MethylationEPIC_v-1-0_B4.csv", skip=7) %>%
   select(IlmnID, CHR, MAPINFO, Strand, UCSC_RefGene_Name, UCSC_RefGene_Group, UCSC_CpG_Islands_Name, Relation_to_UCSC_CpG_Island, DMR)
 
-# left join EPICchar to SampleData
+# inner join EPICchar to SampleData
 FullAnnotation = inner_join(EPICchar, SampleData, by="IlmnID") %>% arrange(IlmnID)
 
 # save output file as "myFA_bulkonly.Rdata"

### 2021-07-07 StanCParallel.R

@@ -147,7 +147,7 @@ parData <- foreach(i = iter(1:nsites), .combine = 'comb', .multicombine = TRUE, 
   
   # compute posterior log-ratio: log2(sigmaP/sigmaT), take integral>0, mean, median
   posterior <- as.matrix(stanFit,pars=c("sigma_p","sigma_t"))
-  logPTratio <- log2(posterior[,1]/posterior[,2])
+  logPTratio <- log2(posterior[,1]/posterior[,2])*2 # multiply log-ratio by 2 to get sigmaP^2/sigmaT^2
   CpGscore <- c(mean(logPTratio > 0), mean(logPTratio), median(logPTratio))
   fitSumm <- rbind(fitSumm,CpGscore)
   

### 2021-07-07 Paper_Figures.Rmd
changes to paper figures script

@@
 ggsave("gene_scores_bootstrap.svg", plot=p2)
+
+# save significant bootstrapped genes as csv
+sig_genes = permed_data %>% filter(score_median_v3>=0.95) %>% pull(UCSC_RefGene_Name)
+write_csv(data.frame(Gene = sig_genes), file = "sig_cons_genes.csv")
+# for Reactome pathway analysis at https://reactome.org/PathwayBrowser
 ```

### 2021-07-29 Paper_Figures.Rmd
updated analysis to exclude SNPs, new gene sets for neuron marker and cardiac progenitor genes

@@
 
+# use MethylToSNP package to detect likely SNPs
+# from paper: LaBarre et al. 2019, Epigenetics & Chromatin)
+# methyltosnp_results file should exist from running StanAnalysis.Rmd
+snps <- read.csv(here("methyltosnp_results.csv"))
+rownames(snps) <- snps[,1]
+snps<-snps[,-1]
+SNPinds <- which(FullAnnotation$IlmnID %in% rownames(snps))
+
+remove_inds = union(QCinds, union(XYinds, SNPinds))
+
 # remove the corresponding rows from the results dataframes
-FullAnnotation <- FullAnnotation[-union(QCinds,XYinds),]
-mu_full <- mu_full[-union(QCinds,XYinds),]
-nu_full <- nu_full[-union(QCinds,XYinds),]
-sigmaP_full <- sigmaP_full[-union(QCinds,XYinds),]
-sigmaPT_full <- sigmaPT_full[-union(QCinds,XYinds),]
-sigmaT_full <- sigmaT_full[-union(QCinds,XYinds),]
-sigmaE_full <- sigmaE_full[-union(QCinds,XYinds),]
-lp_full <- lp_full[-union(QCinds,XYinds),]
-CpGscore_full <- CpGscore_full[-union(QCinds,XYinds),]
+FullAnnotation <- FullAnnotation[-remove_inds,]
+mu_full <- mu_full[-remove_inds,]
+nu_full <- nu_full[-remove_inds,]
+sigmaP_full <- sigmaP_full[-remove_inds,]
+sigmaPT_full <- sigmaPT_full[-remove_inds,]
+sigmaT_full <- sigmaT_full[-remove_inds,]
+sigmaE_full <- sigmaE_full[-remove_inds,]
+lp_full <- lp_full[-remove_inds,]
+CpGscore_full <- CpGscore_full[-remove_inds,]
 ```
@@
 
+# additional non-relevant gene sets from MSigDB (LEIN_NEURON, CARDIAC_PROGENITOR)
+# MSigDB: C2: LEIN_NEURON_MARKERS
+neuron <- read_delim("/Users/kevinmurgas/Documents/Data+ project/geneset_neuron.gmt", delim="\t") %>% colnames()
+# MSigDB: C2: WP_CARDIAC_PROGENITOR_DIFFERENTIATION
+cardprog <- read_delim("/Users/kevinmurgas/Documents/Data+ project/geneset_cardprog.gmt", delim="\t") %>% colnames()
+
 # compute % overlap of database genes and significant genes (By adj bootstrap)
@@
               perm_temp %>% filter(UCSC_RefGene_Name %in% (cancergenes2 %>% pull(CancerGenes))) %>% mutate(group="AGO"),
-              perm_temp %>% filter(UCSC_RefGene_Name %in% (essentialgenes %>% pull(X1))) %>% mutate(group="DEPMAP")
+              perm_temp %>% filter(UCSC_RefGene_Name %in% (essentialgenes %>% pull(X1))) %>% mutate(group="DEPMAP"),
+              perm_temp %>% filter(UCSC_RefGene_Name %in% neuron) %>% mutate(group="neuron"),
+              perm_temp %>% filter(UCSC_RefGene_Name %in% cardprog) %>% mutate(group="cardprog")
               )
 # Fig 4C: plot distribution
-pDB <- genes_dbs %>% mutate(group = factor(group, levels = c("all genes", "AGO", "COSMIC", "DEPMAP"))) %>%
+pDB <- genes_dbs %>% mutate(group = factor(group, levels = c("all genes", "AGO", "COSMIC", "DEPMAP", "neuron", "cardprog"))) %>%
   #ggplot() + geom_violin(aes(x=score_median,y=group,fill=group), draw_quantiles = 0.5) +
@@
 wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="DEPMAP"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
-```
-
-\newpage
-\subsection{Oncogenes vs Tumor suppressors}
-Based on list of top 20 genes mutated in CRC, manually labelled as oncogene or tumor-suppressor:
-
-| Oncogene | Tumor Suppressor |
-|----------|------------------|
-| KRAS     | TP53             |
-| MYC      | APC              |
-| BRAF     | KLF4             |
-| EGFR     | FBXW7            |
-| CDK8     | LRP1B            |
-| PIK3CA   | PTEN             |
-| NRAS     | FAT4             |
-| GNAS     | ARID1A           |
-| SRC      | AMER1            |
-| CTNNB1   | SMAD4            |
-
-```{r Onco_vs_TSG, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=2, comment=NA, fig.fullwidth=TRUE, dev = "png"}
-
-plotGeneScore <- function(geneStr,norm) {
-  cat("\n")
-  CpGscores <- gened_data %>% filter(str_detect(UCSC_RefGene_Name,paste0("^",geneStr,"$"))) %>% pull(score_median)
-  ggplot(prepped_data) + geom_histogram(aes(x=score_median, y=..count../nrow(prepped_data)), binwidth=0.05,fill="lightgray") + ggtitle(sprintf("%s: CpG scores (%i sites)",geneStr, length(CpGscores))) +
-    scale_x_continuous(limits=c(-5,5), breaks=c(-4,-2,0,2,4)) +
-    scale_y_continuous(limits=c(0,0.03),breaks=c(0,0.01,0.02,0.03)) +
-    xlab("CpG conservation score") + ylab("Freq.") + theme_classic() + theme(legend.position = "none") +
-    #geom_vline(xintercept=CpGscores, color="red", alpha=0.5, size=0.2) +
-    geom_histogram(data=data.frame(score=CpGscores),aes(x=score,y=norm*..count../length(CpGscores)),binwidth=0.05, fill="blue",alpha=0.5) +
-    geom_vline(xintercept=mean(CpGscores), color="blue",linetype="dashed", size=1)
-}
-
-cat("\nOncogenes:")
-plotGeneScore("KRAS",0.25)
-plotGeneScore("MYC",0.21)
-plotGeneScore("BRAF",0.5)
-plotGeneScore("EGFR",0.5)
-plotGeneScore("CDK8",0.5)
-plotGeneScore("PIK3CA",0.5)
-plotGeneScore("NRAS",0.5)
-plotGeneScore("GNAS",0.5)
-plotGeneScore("SRC",0.5)
-plotGeneScore("CTNNB1",0.5)
-
-cat("\nTumor Suppressors:")
-plotGeneScore("TP53",0.25)
-plotGeneScore("APC",0.21)
-plotGeneScore("KLF4",0.5)
-plotGeneScore("FBXW7",0.5)
-plotGeneScore("LRP1B",0.5)
-plotGeneScore("PTEN",0.5)
-plotGeneScore("FAT4",0.5)
-plotGeneScore("ARID1A",0.5)
-plotGeneScore("AMER1",0.5)
-plotGeneScore("SMAD4",0.5)
-
+wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="neuron"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
+wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="cardprog"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
 ```

### 2021-07-29 StanAnalysis.Rmd
updated analysis to exclude SNPs, new gene sets for neuron marker and cardiac progenitor genes

@@
 \subsection{Filter CpG sites}
-A filter is applied to remove all sites with Rhat > 1.1 in lp (log-posterior likelihood of entire site fit), and CpG sites mapping to sex chromosomes X and Y are removed as well to eliminate sex-specific bias.
+Three filters are applied to remove CpG sites from subsequent analysis:
+1. Sites with poor model convergence: Rhat > 1.1 in lp (log-posterior likelihood of entire site fit),
+2. Sites mapping to sex chromosomes X and Y,
+3. Sites with likely SNPs (determined via MethylToSNP algorithm from LaBarre 2019, *Epigenetics & Chromatin*)
+
+8740, 19632, and 377 sites were identified in each respective filter, for total union of 28557 CpG sites to remove (out of 866091 total CpG sites)
 
@@
 
+# detect likely SNPs with MethylToSNP package
+# from paper: LaBarre et al. 2019, Epigenetics & Chromatin)
+if ( file.exists(here("methyltosnp_results.csv")) ) {
+  # if file already exists then load it
+  snps <- read.csv(here("methyltosnp_results.csv"))
+  rownames(snps) <- snps[,1]
+  snps<-snps[,-1]
+} else {
+  # otherwise, run the MethylToSNP algorithm and save
+  library(MethylToSNP)
+  library(minfiDataEPIC)
+  data(SNPs.147CommonSingle)
+  data_beta = FullAnnotation[,10:57]
+  rownames(data_beta) <- FullAnnotation$IlmnID
+  snps <- MethylToSNP(data_beta, SNP=SNPs.147CommonSingle) %>% filter(confidence>0.5) # use suggested cutoff of 0.5
+  write.csv(snps, file = here("methyltosnp_results.csv"))
+}
+SNPinds <- which(FullAnnotation$IlmnID %in% rownames(snps))
+
+remove_inds = union(QCinds, union(XYinds, SNPinds))
+
 # remove the corresponding rows from the results dataframes
-FullAnnotation <- FullAnnotation[-union(QCinds,XYinds),]
-mu_full <- mu_full[-union(QCinds,XYinds),]
-nu_full <- nu_full[-union(QCinds,XYinds),]
-sigmaP_full <- sigmaP_full[-union(QCinds,XYinds),]
-sigmaPT_full <- sigmaPT_full[-union(QCinds,XYinds),]
-sigmaT_full <- sigmaT_full[-union(QCinds,XYinds),]
-sigmaE_full <- sigmaE_full[-union(QCinds,XYinds),]
-lp_full <- lp_full[-union(QCinds,XYinds),]
-CpGscore_full <- CpGscore_full[-union(QCinds,XYinds),]
+FullAnnotation <- FullAnnotation[-remove_inds,]
+mu_full <- mu_full[-remove_inds,]
+nu_full <- nu_full[-remove_inds,]
+sigmaP_full <- sigmaP_full[-remove_inds,]
+sigmaPT_full <- sigmaPT_full[-remove_inds,]
+sigmaT_full <- sigmaT_full[-remove_inds,]
+sigmaE_full <- sigmaE_full[-remove_inds,]
+lp_full <- lp_full[-remove_inds,]
+CpGscore_full <- CpGscore_full[-remove_inds,]
 ```

### 2021-07-30 Paper_Figures.Rmd
colored boxplots in fig 4C

@@
 
-\subsection{CRC Cancer Genes Databases}
+\subsection{Fig 4B: Gene-average Conservation Score}
+Examine the distribution of gene-average conservation scores, as well as the distribution of adjusted-boostrap p-values.
+```{r Gene_Score, echo=FALSE, fig.fullwidth=TRUE, warning=FALSE, comment=NA}
+
+permed_data %>% ggplot() + geom_histogram(aes(x=score_median, y=..density..), binwidth=0.2, color="gray",fill="lightgray") + theme_classic()
+
+permed_data %>% ggplot() + geom_histogram(aes(x=score_median_v3, y=..count..), binwidth=0.02, color="gray",fill="lightgray") + theme_classic() + geom_vline(aes(xintercept = 0.95), color="red", linetype="dashed")
+```
+
+\subsection{Fig 4C: CRC Cancer Genes Databases}
+Including sets of genes implicated in CRC, based on common mutations from COSMIC database and Atlas of Genetics in Oncology (AGO), as well as a cancer essential gene list based on experimental CRISPRi knockout library.
+We also compared two gene sets from MSigDB which should be irrelevant to cancer, a set of cardiac progenitor differentiation genes (MSigDB: C2: WP_CARDIAC_PROGENITOR_DIFFERENTIATION) and a set of neuron marker genes (MSigDB: C2: LEIN_NEURON_MARKERS).
 ```{r cancergenes, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=2, comment=NA}
@@
               perm_temp %>% filter(UCSC_RefGene_Name %in% neuron) %>% mutate(group="neuron"),
-              perm_temp %>% filter(UCSC_RefGene_Name %in% cardprog) %>% mutate(group="cardprog")
+              perm_temp %>% filter(UCSC_RefGene_Name %in% cardprog) %>% mutate(group="cardiac")
               )
 # Fig 4C: plot distribution
-pDB <- genes_dbs %>% mutate(group = factor(group, levels = c("all genes", "AGO", "COSMIC", "DEPMAP", "neuron", "cardprog"))) %>%
+pDB <- genes_dbs %>% mutate(group = factor(group, levels = c("all genes", "neuron", "cardiac", "AGO", "COSMIC", "DEPMAP"))) %>%
   #ggplot() + geom_violin(aes(x=score_median,y=group,fill=group), draw_quantiles = 0.5) +
   ggplot() + geom_boxplot(aes(x=score_median,y=group,fill=group), outlier.shape = NA) +
-  scale_x_continuous(limits=c(-5,7), breaks = c(-4, -2,0, 2, 4, 6)) + theme_classic() + ylab("Database") + xlab("Gene Conservation Score") + geom_vline(xintercept=0, color="black", alpha=0.2) + theme(aspect.ratio=60/156.3,legend.position = "none")
+  scale_x_continuous(limits=c(-5,7), breaks = c(-4, -2,0, 2, 4, 6)) + theme_classic() + ylab("Gene set") + xlab("Gene Conservation Score") + geom_vline(xintercept=0, color="black", alpha=0.2) + theme(aspect.ratio=60/156.3,legend.position = "none") + scale_fill_manual(values=c("gray","red", "red", "deepskyblue", "deepskyblue", "deepskyblue"))
 # # add hashmarks for each gene (doesn't look great)
@@
 ```
-
-\newpage
-\subsection{Gene-average Parameter Distributions}
-```{r Params_gene, echo=FALSE, fig.fullwidth=TRUE, warning=FALSE, comment=NA}
-# make gene-wise averages
-gene_summarize <- function(data, id, value, fun = mean, ...) {
-  grouped_data <- data %>% group_by({{id}})
-
-  bind_cols(
-    summarise_at(grouped_data, dplyr::vars({{value}}), fun, ...),
-    count(grouped_data) %>% dplyr::ungroup() %>% dplyr::select(.data$n)
-  )
-}
-
-cat("Summarize by gene, taking mean value of all sites in gene")
-ptm<-proc.time()
-gene_summary <- gened_data %>%
-  filter(!is.na(UCSC_RefGene_Name)) %>%
-  gene_summarize(UCSC_RefGene_Name, c(mu:sigmaT,score_median))
-cat(proc.time()-ptm)
-
-post_histogram <- function(postvals,bw,xl,titlestr) {
-  # stacked histogram colored by gene averages
-  # and prior curve overlaid
-  plot_df <- data.frame(val=postvals)
-  p1 <- ggplot(plot_df, aes(x=val)) +
-    geom_histogram(aes(y=..density..), binwidth=bw) +
-    ggtitle(titlestr) + xlim(xl) + theme_classic() + xlab(titlestr) + ylab("Freq.")
-  p1
-}
-
-# plot: n_sites, mu, nu, sigmaP, sigmaPT, sigmaT
-n_cutlo = 5
-n_cuthi = quantile(gene_summary$n, 0.99)
-pNS <- post_histogram(gene_summary$n,1,c(0,200),paste0("N sites (max = ",max(gene_summary$n),")")) + geom_vline(xintercept=n_cutlo-0.5,linetype = "dashed",color="red") + annotate(geom="text",x=80,y=0.03,size=2,label=paste0("# genes <",n_cutlo," sites:\n", sum(gene_summary$n<n_cutlo)," / ",nrow(gene_summary))) +
-geom_vline(xintercept=n_cuthi,linetype = "dashed",color="red") +annotate(geom="text",x=120,y=0.013,size=2,label=sprintf("# genes >%0.1f sites:\n %i / %i (99th pctile)", n_cuthi,sum(gene_summary$n>n_cuthi),nrow(gene_summary)))
-gene_summary <- gene_summary %>% filter(n>=n_cutlo) %>% filter(n<n_cuthi)
-pM <- post_histogram(gene_summary$mu,0.1,c(-5,5),"Mu")
-pN <- post_histogram(gene_summary$nu,0.1,c(-5,5),"Nu")
-pP <- post_histogram(gene_summary$sigmaP,0.05,c(0,2),"SigmaP")
-pPT <- post_histogram(gene_summary$sigmaPT,0.05,c(0,2),"SigmaPT")
-pT <- post_histogram(gene_summary$sigmaT,0.05,c(0,2),"SigmaT")
-pP1 <- post_histogram(gene_summary$score_median,0.05,c(-3,3),"Score_median")
-# arrange each parameter as a subplot
-p_all <- ggarrange(pNS,pM,pN,pP,pPT,pT,pP1,hjust=0,nrow=3,ncol=3)
-print(p_all)
-rm(pNS,pM,pN,pP,pPT,pT,pP1)
-
-```
-
-\newpage
-\subsection{Gene-average parameters vs N sites}
-```{r Params_vsN, echo=FALSE, fig.fullwidth=TRUE, warning=FALSE, comment=NA, dev = "png"}
-pM <- ggplot(gene_summary) + geom_point(aes(x=n,y=mu), alpha=0.01) + ggtitle("Mu vs N-sites")
-pN <- ggplot(gene_summary) + geom_point(aes(x=n,y=nu), alpha=0.01) + ggtitle("Nu vs N-sites")
-pE <- ggplot(gene_summary) + geom_point(aes(x=n,y=sigmaE), alpha=0.01) + ggtitle("SigmaE vs N-sites")
-pP <- ggplot(gene_summary) + geom_point(aes(x=n,y=sigmaP), alpha=0.01) + ggtitle("SigmaP vs N-sites")
-pPT <- ggplot(gene_summary) + geom_point(aes(x=n,y=sigmaPT), alpha=0.01) + ggtitle("SigmaPT vs N-sites")
-pT <- ggplot(gene_summary) + geom_point(aes(x=n,y=sigmaT), alpha=0.01) + ggtitle("SigmaT vs N-sites")
-
-pP1 <- ggplot(gene_summary) + geom_point(aes(x=n,y=score_median), alpha=0.01) + ggtitle("Score_median vs N-sites")
-
-# arrange each parameter as a subplot
-p_all <- ggarrange(pM,pN,pE,pP,pPT,pT,pP1,hjust=0,nrow=3,ncol=3)
-print(p_all)
-rm(pM,pN,pE,pP,pPT,pT,pP1)
-```

### 2021-07-30 Paper_Figures.Rmd
cardiac gene set typo fix

@@
 wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="DEPMAP"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
+wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="cardiac"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
 wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="neuron"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
-wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="cardprog"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
 ```

### 2021-07-30 Paper_Figures.Rmd
fixed direction of one-sided rank sum test in negative control gene sets

@@
 wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="DEPMAP"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
-wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="cardiac"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
-wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="neuron"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "greater")
+wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="cardiac"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "less")
+wilcox.test(x = genes_dbs$score_median[genes_dbs$group=="neuron"], y = genes_dbs$score_median[genes_dbs$group=="all genes"], alternative = "less")
 ```
